### 2017-03-03 mers-structure.tex
Start manuscript.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\usepackage{grffile}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% reference figures across documents
\usepackage{xr}
\externaldocument{mers-structure_supp}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
\usepackage{natbib}
%\renewcommand\citepleft{(}
%\renewcommand\citepright{)}
%\renewcommand\citepform[1]{\textsl{#1}}

\usepackage{amsmath}

% Remove brackets from numbering in list of References
%\renewcommand\refname{\large References}
%\makeatletter
%\renewcommand{\@biblabel}[1]{\quad#1.}
%\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

% comments
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
\def\tbc#1{\textcolor{purple}{[#1]}}

% symbols
\newcommand{\chiSq}{\chi^{2}_{df}}
\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
\newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
\newcommand{\dspr}{d_\mathrm{SPR}}

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf MERS structure}

\author[1]{Gytis Dudas}
\author[2]{Luiz Max Carvalho}
\author[1]{Allison Black}
\author[1]{Trevor Bedford}
\author[2,4,5]{Andrew Rambaut}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
\affil[5]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus causing significant mortality and morbidity largely in the Arabian Peninsula.
The epidemiology of the virus is relatively poorly understood with most MERS patients being older males with comorbidities, although large nosocomial have occurred, notably in Jeddah, Riyadh and South Korea.
Our understanding of MERS-CoV epidemiology is further complicated by some patients not recalling contact with camels, the accepted reservoir for MERS-CoV, or any other livestock, suggesting significant cryptic MERS-CoV circulation in communities.
Seroepidemiology has been employed extensively during the time since the discovery of MERS-CoV to understand the exposure and therefore risk among camels and humans alike, but various sequencing efforts have not been fully leveraged against this zoonotic virus.
Here we use existing MERS-CoV sequencing data to estimate the number of times the virus has been introduced into humans and use the distribution of sequence clusters to estimate the reproductive number for the virus from sequence data alone.
We confirm that MERS-CoV has relatively poor capacity to spread in humans and arrive at an estimate of at least 65 zoonotic introductions of the virus into humans.

\end{abstract}

\pagebreak

\section*{Introduction}
Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
The virus, first discovered in 2012 \cite{WHO_2016}, has gone on to cause more than 1800 infections with at least 600 associated deaths.
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities.
Whilst contact with camels is often reported by patients, many do not recall contact with any livestock, suggesting a significant but unobserved community contribution to the outbreak.
Studies into MERS-CoV epidemiology often rely on serology to identify factors associated with MERS CoV exposure in potential risk groups or hosts.
The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios.
Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios, which are a fundamental part of quantifying risk.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir, nosocomial outbreak investigations and routine surveillance.

It is accepted that human MERS-CoV infections are a result of multiple introductions of the virus into humans, but no study has attempted to go beyond educated guesses about this number.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts: humans and camels.
By explicitly modelling the evolution of MERS-CoV between these two distinct host populations we show that human MERS infections are the result of frequent and unidirectional spillover events from camels into humans.
Inter-outbreak evolution of the virus occurs exclusively in camels, highlighting the need for genomic surveillance of MERS-CoV in the reservoir as well.
We also use the sequence clusters recovered from our population structure analyses to estimate the reproductive number for the virus (R$_{0}$), which is in agreement with previous findings that MERS-CoV is, on average, poor at spreading in human populations.
Our analyses also capture seasonal variation in zoonotic transmission patterns, with indications that particularly large outbreaks are the result of zoonotic transmissions that take place at the beginning of each new year.

%Here we use 307 largely complete MERS-CoV genomes, 91 of which are from camel viruses, to explicitly model the population structure of the virus in humans and camels in order to arrive at an estimate of the number of times the virus has been introduced into human populations.
%We then explore the resulting sequence clusters that can be thought of as replicates of the transmission process, albeit with different settings, as the data include large nosocomial outbreaks.

%\begin{figure}[h]
% \centering		
%	\includegraphics[width=0.45\textwidth]{figures/}
%	\caption{\textbf{}
%	
%	}
%	\label{}
%\end{figure}

\section*{Results}

\subsection*{}

\section*{Discussion}

\subsection*{}

\newpage

\section*{Methods}
\subsection*{Sequence data}
All MERS-CoV sequences were downloaded from GenBank.
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
Sequences were annotated with available collection dates and hosts, designated as camel or human.
For partitioned analyses the alignment was split into two fragments, the first containing all nucleotides up to nucleotide 21000, and the rest of the genome being assigned to the second fragment.
Despite their divergence both Severe Acute Respiratory Syndrome (SARS) and MERS betacoronaviruses appear to recombine across position 21000, which may indicate the existence of a recombinogenic secondary RNA structure that remains to be described in its vicinity. %% "Work by Zhang et al. (66) suggested that the presence of a stable hairpin between two direct repeats increases the rate of homologous recombination in MLV. Their data led to the conclusion that stable secondary structures decrease processive RT synthesis and, further, that a functional NC can increase RT processivity. However, in this case, there were no corresponding in vitro data (such as primer extension assays) to correlate with the in vivo observations. Here, we have demonstrated for the first time a direct relationship between pausing in vitro and viral recombination in vivo."

\subsection*{Data availability}

\section*{Acknowledgements}

\bibliographystyle{mbe.bst}
\bibliography{mers-structure}

\end{document}

### 2017-03-04 mers-structure.tex
Add methods, results.

@@
 
-\subsection*{}
+\subsection*{Cross species transmission of MERS-CoV}
+It is generally accepted that MERS-CoV is a continuously re-emerging human pathogen, however the exact number of cross-species transmission events remains unknown.
+We confirm this by identifying the camel reservoir as the host population where most of MERS-CoV evolution takes place (see figure \ref{mcc}).
+The human side of MERS outbreaks, on the other hand, are caused by repetitive zoonotic transmissions from camels, which we estimated by using an explicit model of population structure.
+We infer that across 216 MERS-CoV sequences collected from humans, we detect a median of 67 individual camel to human cross-species transmissions (95\% highest posterior density: 60--72), with little support for back-transmission scenarios (median 3, 95\% HPDs: 0--13).
+There is also apparent heterogeneity in resulting human cluster sizes, which is not unexpected.
+Large clusters of MERS are apparent and mostly correspond to known large transmission chains, often in a nosocomial setting, such as the outbreak in South Korea and Riyadh.
+We also note a general predisposition for large MERS-CoV clusters to have crossed the species barrier into humans early in each year.
+
+\begin{figure}[h]
+ \centering		
+	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
+	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV sequences from humans and camels.}
+	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered using the structured coalescent.
+	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
+	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
+	}
+	\label{mcc}
+\end{figure}
+
+\subsection*{Estimated epidemiological parameters}
+Removing the confounding effects of shared ancestry for MERS-CoV sequences has allowed us to treat every viral introduction into humans as an independent replicate of the transmission process in humans.
 
@@
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
+Protein coding sequences were extracted and concatenated.
 Sequences were annotated with available collection dates and hosts, designated as camel or human.
@@
 
+\subsection*{Structured coalescent analyses}
+MultiTypeTree module was used in BEAUti v2.4.3 to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
+Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ nucleotide substitution models specified for codon positions 1+2 and 3.
+A relaxed molecular clock with branch rates drawn from a lognormal distribution was used to infer the evolutionary rate from date calibrated tips.
+Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
+An identical set up was used for analysing alignments split into two fragments, with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+Eight independent whole-genome MCMC analyses were run, for 200 million states each subsampling every 20000 states.
+For two-fragment MCMC analyses ten independent runs were set up.
+Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin.
+
+\subsection*{Epidemiological analyses}
+
+\begin{equation}
+r_{j} = \frac{\Gamma(kj+j-1)}{\Gamma(kj)\Gamma(j+1)} \frac{(\frac{R_{0}}{k})^{j-1}}{(1+\frac{R_{0}}{k})^{kj+j-1}}
+\end{equation}
+
 \subsection*{Data availability}

### 2017-03-06 mers-structure.tex
Add text to results and discussion.

@@
 
-\subsection*{Cross species transmission of MERS-CoV}
+\subsection*{Recombination occurs in the reservoir}
+Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
+Since detectable and problematic recombination in RNA viruses requires co-infection with distinct viral genotypes humans were ruled out as a suitable recombination environment \textit{a priori}.
+Here we confirm this assumption by showing that evolution of MERS-CoV within human clusters is clonal.
+This is not surprising, since zoonotic transmission of MERS-CoV into humans is relatively rare, of limited diversity at the point of introduction and viral diversity in humans is ultimately transient.
+All of these factors act to make co-infection of humans with MERS-CoV exceedingly rare.
+It is worth highlighting that MERS-CoV almost certainly recombines during human infection, but the limited diversity of virus populations in individual humans renders it invisible and of little consequence.
+
+Although we do not detect recombination in human clusters, we see evidence for introduction of recombinant viruses into humans which evolve clonally after the transfer.
+This is not unexpected either - there are strong indications that the Korean MERS outbreak was caused by a recombinant virus crossing the species barrier.
+The frequency of MERS-CoV recombination in the reservoir implies that co-infection of camels is quite common, and by extension that the virus is likely to be very prevalent in this host.
+
+\subsection*{Effects of recombination are minimal}
+Most phylogenetic approaches model sequence evolution as a strictly clonal process and when recombination is present traditional phylogenetic approaches can fail to recover correct topologies, branch lengths or demographic history.
+All coronavirus genera are known to be recombinogenic, and MERS-CoV is no exception, but whereas recombination is obligate for some eukaryotes, detectable RNA virus recombination can only occur during co-infection with two or more distinct viral genotypes.
+Previous studies have inferred prevalent recombination in MERS-CoV and found direct evidence for it.
+When not directly looking for recombination, however, most studies carried out their objectives without regard for potential effects of recombination.
+Our independent analyses of the alignment split into two fragments reveals that whilst some effects of recombination are certainly detectable, they do not seem to be systematically biasing phylogenetic inference.
+%So whereas we find some evidence of recombinant viruses crossing the species barrier into humans
+
+\subsection*{Numerous cross-species transmissions of MERS-CoV}
 It is generally accepted that MERS-CoV is a continuously re-emerging human pathogen, however the exact number of cross-species transmission events remains unknown.
@@
 
-\subsection*{Estimated epidemiological parameters}
-Removing the confounding effects of shared ancestry for MERS-CoV sequences has allowed us to treat every viral introduction into humans as an independent replicate of the transmission process in humans.
+\subsection*{MERS epidemiology}
+Although sequences in an outbreak necessarily represent a subset of all the cases, their clustering and resulting cluster sizes can be highly informative of underlying epidemiological processes that are otherwise inaccesible or unfeasible to study.
+In our case we have used sequence data to disentangle the evolution of MERS-CoV into two discrete host compartments.
+By removing the confounding effects of shared ancestry for MERS-CoV sequences we can treat every viral introduction into humans as an independent replicate of the transmission process in humans.
 
@@
 
-\subsection*{}
+\subsection*{MERS-CoV in camels}
+Our observation of recombinant MERS coronaviruses being transmitted to humans is not cause for concern by itself, however, the mere presence of detectable recombination implies that co-infection of camels with MERS-CoV is common.
+By extension, frequent co-infection implies a high prevalence of the virus in camels, which is the main unsettling conclusion.
+There is ample evidence for this from seroepidemiology, which has shown that XX\% of camels are infected by age XX.
+Our results show that zoonotic transmission of MERS-CoV into humans is the main drivers for the continuation of the outbreak in the Arabian Peninsula. 
+Prevalence of MERS-CoV in camels could easily be translated into force of infection and given the role of zoonotic transmission as the main driver for the outbreak only reiterates the importance of controlling the MERS coronavirus in the reservoir if the human side of the outbreak is to be brought under control.
 

### 2017-04-11 mers-structure.tex
Update text.
The core is probably there. Mostly missing bit now is seasonality +
human side with R0 estimates. Tempted to do a camel skygrid from camel
sequences + single sequence per human cluster too.

--- 2017-03-03 mers-structure.tex
@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \LARGE \bf MERS structure}
+\title{\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology}
 
@@
 \author[1]{Trevor Bedford}
-\author[2,4,5]{Andrew Rambaut}
+\author[2,3,4]{Andrew Rambaut}
 
@@
 \affil[2]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
-\affil[5]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}
+\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+\affil[4]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}
 
@@
 
-Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus causing significant mortality and morbidity largely in the Arabian Peninsula.
-The epidemiology of the virus is relatively poorly understood with most MERS patients being older males with comorbidities, although large nosocomial have occurred, notably in Jeddah, Riyadh and South Korea.
-Our understanding of MERS-CoV epidemiology is further complicated by some patients not recalling contact with camels, the accepted reservoir for MERS-CoV, or any other livestock, suggesting significant cryptic MERS-CoV circulation in communities.
-Seroepidemiology has been employed extensively during the time since the discovery of MERS-CoV to understand the exposure and therefore risk among camels and humans alike, but various sequencing efforts have not been fully leveraged against this zoonotic virus.
-Here we use existing MERS-CoV sequencing data to estimate the number of times the virus has been introduced into humans and use the distribution of sequence clusters to estimate the reproductive number for the virus from sequence data alone.
-We confirm that MERS-CoV has relatively poor capacity to spread in humans and arrive at an estimate of at least 65 zoonotic introductions of the virus into humans.
-
+Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
+The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
+Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, sequence data have not been utilised to any appreciable extent to understand transmission patterns within the entire outbreak.
+Here we use existing MERS-CoV sequencing data to explore the phylodynamics of the virus in two of its major known hosts, humans and camels, as well as the interface between the two during cross species transmission.
+We show that all of long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
+Our study identifies MERS-CoV spillovers into humans as largely self-limiting, suggesting that curbing human exposure to camels or reducing the prevalence of the virus in camels are the most efficient ways of preempting future MERS outbreaks.
+Joining a number of recent genomic epidemiology publications, our study highlights the utility of sequence data in understanding the transmission patterns of pathogens which are largely inaccessible to more traditional approaches.
 \end{abstract}
@@
 Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
-The virus, first discovered in 2012 \cite{WHO_2016}, has gone on to cause more than 1800 infections with at least 600 associated deaths.
-Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities.
-Whilst contact with camels is often reported by patients, many do not recall contact with any livestock, suggesting a significant but unobserved community contribution to the outbreak.
-Studies into MERS-CoV epidemiology often rely on serology to identify factors associated with MERS CoV exposure in potential risk groups or hosts.
-The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios.
-Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios, which are a fundamental part of quantifying risk.
-Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir, nosocomial outbreak investigations and routine surveillance.
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 1800 infections with at least 600 associated deaths \citep{who_2016_MERS}.
+Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
+Whilst contact with camels is often reported, many patients do not recall contact with any livestock, suggesting a significant but unobserved community contribution to the outbreak \citep{group_state_2013}.
+Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
+Indeed, such data have been used to show high prevalence of past viral exposure in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and more frequent exposure in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015} compared to the general population.
+Traditional epidemiological approaches have also been used to look at distributions of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.
 
-It is accepted that human MERS-CoV infections are a result of multiple introductions of the virus into humans, but no study has attempted to go beyond educated guesses about this number.
-Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts: humans and camels.
-By explicitly modelling the evolution of MERS-CoV between these two distinct host populations we show that human MERS infections are the result of frequent and unidirectional spillover events from camels into humans.
-Inter-outbreak evolution of the virus occurs exclusively in camels, highlighting the need for genomic surveillance of MERS-CoV in the reservoir as well.
-We also use the sequence clusters recovered from our population structure analyses to estimate the reproductive number for the virus (R$_{0}$), which is in agreement with previous findings that MERS-CoV is, on average, poor at spreading in human populations.
-Our analyses also capture seasonal variation in zoonotic transmission patterns, with indications that particularly large outbreaks are the result of zoonotic transmissions that take place at the beginning of each new year.
+Although traditional approaches yield important clues about exposure patterns and potential for larger outbreaks, the origins of a pathogen remain opaque to traditional approaches.
+The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios \citep{arias_rapid_2016}.
+Modern sequence data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2016}.
+Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
+Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
+A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether the introductions were of a single origin and homogenous or a result of increased cases among incoming patients \citep{fagbo_molecular_2015}.
 
-%Here we use 307 largely complete MERS-CoV genomes, 91 of which are from camel viruses, to explicitly model the population structure of the virus in humans and camels in order to arrive at an estimate of the number of times the virus has been introduced into human populations.
-%We then explore the resulting sequence clusters that can be thought of as replicates of the transmission process, albeit with different settings, as the data include large nosocomial outbreaks.
-
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013}.
+Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or relied on methods agnostic to underlying signals of population structure \citep{zhang_evolutionary_2016}.
+Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels, as well as the zoonotic spillover dynamics at the interface between the two, by using an explicit model of population structure \citep{vaughan_efficient_2014}.
+We show that viral evolution of MERS-CoV occurs almost exclusively in camels, where virus is also most likely to recombine.
+Our results suggest that spillover events into humans appear to be considerably more common than previously thought.
+Once human MERS-CoV infections are established, however, the virus has relatively poor person-to-person transmission and outbreaks are short-lived.
 %\begin{figure}[h]
@@
 
-\subsection*{}
+\subsection*{MERS-CoV is predominantly a camel virus}
+The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies the camel reservoir as the host population where most of MERS-CoV evolution takes place (see figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal to the virus.
+Across XXX MERS-CoV sequences collected from humans, we detect a median of XX individual camel to human cross-species transmissions (95\% highest posterior density: XX--XX).
+Whilst we identify a median of X (95\% HPDs: X--XX) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (figure \ref{mcc}).
+This suggests that these migrations are induced by the migration rate prior, rather than inferred from data.
+The remainder of human-to-camel transmissions are likely derived from phylogenetic proximity of human sequences to the ``backbone'' of the phylogenetic tree.
+It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows the model to consider humans as a potential alternative host, given the proximity of sequences from humans to the ``backbone'' of the tree.
+The first sequences of MERS-CoV from camels were collected in November 2013.
+
+The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission.
+If MERS-CoV had attained any appreciable degree of community transmission we would expect to observe some continuity between human sequence clusters.
+Yet we observe dense and terminal human clusters of MERS-CoV of no subsequent relevance to the evolution of the virus interspersed within the relatively large diversity of MERS-CoV lineages in camels.
+In addition, these observations are derived from sequence data heavily skewed towards human cases and in all likelihood biased towards the most severe outbreaks documented.
+We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases, and probably, on average, milder pathology that do not get reported, let alone sequenced.
+
+% We also detect apparent seasonality in times of zoonotic transfer of viruses.
+% Whilst zoonotic transmissions occur throughout the entire year, particularly large sequence clusters begin as viral introductions into humans earlier in the year.
+% %We identify
+% There is also apparent heterogeneity in resulting human cluster sizes, which is not unexpected.
+% Large clusters of MERS are apparent and mostly correspond to known large transmission chains, often in a nosocomial setting, such as the outbreak in South Korea and Riyadh.
+% At the same time, the long tail of the sequence cluster size distribution is an indicator of transmission chains terminating soon after introduction, many of which could represent genuine singleton infections.
+
+\subsection{MERS-CoV is poorly suited for human transmission}
+Whilst the structured coalescent model clearly shows humans as a terminal host for MERS-CoV, we want to quantify this further.
+
+This explains why MERS-CoV has not caused a detectable general epidemic in the Arabian Peninsula, nor a world-wide pandemic.
+
+%\begin{figure}[h]
+% \centering
+%	\includegraphics[width=0.65\textwidth]{figures/}
+%	\caption{\textbf{}
+%
+%	}
+%	\label{}
+%\end{figure}
+
+\begin{figure}[h]
+ \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
+	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV sequences from humans and camels.}
+	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered using the structured coalescent.
+	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
+  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
+	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
+	}
+	\label{mcc}
+\end{figure}
+
+\subsection*{Recombination shapes MERS-CoV diversity}
+Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
+In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of data corresponding to inferred MERS-CoV clades.
+Both methods used rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
+One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, which gives the appearance of repeated mutations taking place in the recipient lineage, somewhere else in the tree.
+PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment.
+Another hallmark of recombination is spatial clustering of alleles in the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
+3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences.
+
+Both tests can give spurious results in cases of extreme rate heterogeneity \citep{dudas_mers-cov_2016}, but both are not known to fail simultaneously.
+PHI and 3Seq methods consistently identify most of the apparent `backbone' of the phylogeny as encompassing sequences with evidence of recombination (figure \ref{mcc}).
+Neither method can identify where in the tree recombination occurred, but each full asterisk in figure \ref{mcc} should be interpreted as the minimum partition of data that still captures recombinant descendants.
+This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
+
+Amongst clusters inferred to be infecting humans we see extremely limited amounts of genetic diversity, accompanied by fewer homoplasies (figure \ref{}).
+Given that recombination requires co-infection with two or more distinct lineages of the virus, the low prevalence of the MERS-CoV in the general human population, and its poor transmissibility, there is little reason to suspect that distinct MERS-CoV lineages are able to generate widespread recombination we observe.
+This is reassuring, since phylogenetic approaches can be used freely in these scenarios, but limiting analysis to sequences strictly from single-introduction clusters comes at a cost of mutational resolution.
+In contrast, the prevalence of MERS-CoV in camels is thought to be high, camels seemingly suffer only mild symptoms from infection and most of MERS-CoV evolutionary history takes place in camels.
+This makes camels the ideal host where distinct lineages of MERS-CoV can frequently mingle and recombine as a consequence.
+
+% Here we confirm this assumption by showing that evolution of MERS-CoV within human clusters is clonal.
+% This is not surprising, since zoonotic transmission of MERS-CoV into humans is relatively rare, of limited diversity at the point of introduction and viral diversity in humans is ultimately transient.
+% All of these factors act to make co-infection of humans with MERS-CoV exceedingly rare.
+% It is worth highlighting that MERS-CoV almost certainly recombines during human infection, but the limited diversity of virus populations in individual humans renders it invisible and of little consequence.
+%
+% Although we do not detect recombination in human clusters, we see evidence for introduction of recombinant viruses into humans which evolve clonally after the transfer (figure \ref{chain}).
+% This is not unexpected either - there are strong indications that the Korean MERS outbreak was caused by a recombinant virus crossing the species barrier.
+% The frequency of MERS-CoV recombination in the reservoir implies that co-infection of camels is quite common, and by extension that the virus is likely to be very prevalent in this host.
+
+\begin{figure}[h]
+ \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_chain.png}
+	\caption{\textbf{Tangled chain of MERS-CoV genome and its parts.}
+	Tangled chain of the full genome (a), all nucleotides up to position 21000 (b) and all nucleotides past position 21000 (c) of MERS-CoV.
+	The same taxa, if available in neighbouring trees, are connected via coloured lines (human sequences in blue, camel sequences in red).
+	Tree branches are coloured by inferred ancestral host state (human in grey, camel in black).
+	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in the camel host.
+	Human clusters in grey change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
+	This is evidence for recombinant viruses generated in the reservoir entering human populations.
+	Coding sequences in the MERS-CoV genome are indicated with arrows at the bottom of the plot with position 21000 indicated by the black line.
+	}
+	\label{}
+\end{figure}
+
+% \subsection*{Effects of recombination are minimal on human clusters}
+% Most phylogenetic approaches model sequence evolution as a strictly clonal process and when recombination is present traditional phylogenetic approaches can fail to recover correct topologies, branch lengths or demographic history.
+% All coronavirus genera are known to be recombinogenic and MERS-CoV is no exception.
+% Whereas recombination is obligate for some eukaryotes, detectable RNA virus recombination can only occur during co-infection with two or more distinct viral genotypes.
+% Previous studies have inferred prevalent recombination in MERS-CoV \citep{dudas_mers-cov_2016} and found direct evidence for it \citep{}.
+% When not directly looking for recombination, however, most studies carried out their objectives without regard for potential effects of recombination.
+% Our independent analyses of the alignment split into two fragments reveals that effects of recombination are certainly detectable, but limited to the parts of the phylogeny evolving in camels.
+%So whereas we find some evidence of recombinant viruses crossing the species barrier into humans
+
+% \subsection*{MERS epidemiology}
+
+% Although sequences in an outbreak necessarily represent a subset of all the cases, their clustering and resulting cluster sizes can be highly informative of underlying epidemiological processes that are otherwise inaccesible or unfeasible to study.
+% In our case we have used sequence data to disentangle the evolution of MERS-CoV into two discrete host compartments.
+% By removing the confounding effects of shared ancestry for MERS-CoV sequences we can treat every viral introduction into humans as an independent replicate of the transmission process in humans.
 
@@
 
-\subsection*{}
+\subsection{Population structure and inference}
+Phylogeny labelling approaches depend on discrete states being assigned to taxa, followed by ancestral state reconstruction equivalent to that applied to nucleotide sites.
+These methods, when used to infer ancestral migration history across a phylogeny, have thus been called ``mugration'' models.
+When the number of possible discrete states is limited, data are ample and questions are formulated correctly, phylogeny labelling approaches are computationally efficient.
+However, ``mugration'' models ignore potentially relevant information, such as markedly different demographic histories under across distinct populations.
+
+Sequencing of MERS-CoV has been a challenging issue, especially from the reservoir.
+MERS-CoV seroprevalence in camels, as expected of a reservoir, is very high, yet sequencing efforts have overwhelmingly focused on human infections.
+At present MERS-CoV sequences collected from humans outnumber those from camels at a ratio of over 2-to-1, and camel sequences make a late appearance in our data, largely because of the time required to identify camels as the reservoir for the virus.
+Using such data in a ``mugration'' model have led to proposals that MERS-CoV is an epizootic in camels of largely human origins \citep{zhang_evolutionary_2016}, a finding that contradicts other sources of information.
+Similarly, traditional epidemiological approaches have no way of firmly establishing links between human cases outside of known family clusters or hospital outbreaks \citep{cauchemez_unraveling_2016}.
+
+Our study has shown how significantly different demographic regimes can be leveraged against biased sequence data to arrive at conclusions perfectly in line with prior observations.
+It has been apparent for some time that MERS-CoV sequences collected from camels are, on average, relatively distantly related to each other, whereas human outbreaks tended to fall into densely sampled clusters of no apparent relation or consequence to further human outbreaks.
+These hallmarks of distinct demographic modes were easily picked up by the explicit population structure model we employed, and helped us arrive at a comprehensive picture of MERS-CoV epidemiology in agreement with previous empirical studies.
+
+\subsection*{MERS-CoV epidemiology}
+Our analyses firmly support camels as the primary and essential host of human MERS coronaviruses.
+Although sequencing efforts for MERS-CoV have been highly human-biased, we have found that over 80\% of the sampled MERS-CoV evolutionary history has occurred in camels.
+In contrast, sequences recovered from human cases represent MERS-CoV lineages that play no detectable role in the long-term evolution of the virus.
+The difference in apparent host suitability between humans and camels helps explain why unlike some human viruses of zoonotic origins, such as Ebola virus, that have spread internationally, MERS-CoV has remained a problem largely limited to the Arabian Peninsula.
+The sole exception to this has been the MERS outbreak in South Korea.
+
+%Despite the crucial role camels play in MERS-CoV maintenance, we have also identified a relatively small effective population size of the virus in the camel reservoir.
+
+% Our observation of recombinant MERS coronaviruses being transmitted to humans is not cause for concern by itself, however, the mere presence of detectable recombination implies that co-infection of camels with MERS-CoV is common.
+% By extension, frequent co-infection implies a high prevalence of the virus in camels, which is the main unsettling conclusion.
+% There is ample evidence for this from seroepidemiology, which has shown that XX\% of camels are infected by age XX.
+% Our results show that zoonotic transmission of MERS-CoV into humans is the main drivers for the continuation of the outbreak in the Arabian Peninsula.
+% Prevalence of MERS-CoV in camels could easily be translated into force of infection and given the role of zoonotic transmission as the main driver for the outbreak only reiterates the importance of controlling the MERS coronavirus in the reservoir if the human side of the outbreak is to be brought under control.
+
+\subsection*{Issues with recombination}
+Recombination in MERS-CoV is now an accepted feature of the system.
+Prevalent recombination can seriously impede phylogenetic and other sequence analysis techniques.
+We argue that although MERS-CoV sequences contain overwhelming evidence for recombination, we do not expect humans to play a major role in this process.
+Every introduction of the virus into a human population is expected to establish an apparently clonally evolving population.
+However, lineages crossing the species barrier are very likely to have experienced recombination at some point in their past, in some cases occurring right before zoonotic transmission, as was the case in the Korean outbreak of MERS.
+We thus conclude that whilst sequence analyses of MERS-CoV are perfectly legitimate for well-defined human outbreaks, extreme care must be taken when inferring their origins in the context of the reservoir, or even when comparing related outbreaks in humans.
+We also strongly advise against drawing hasty conclusions from MERS-CoV phylogenies recovered from human sequences if they span more than four months of evolution or camel sequences under virtually all circumstances, as there is no guarantee that phylogenies will cover the clonal evolutionary history of the virus in humans.
 
@@
 
+\subsection*{Structured coalescent analyses}
+MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
+Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
+A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
+Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
+An identical set up was used for analysing alignments split into two fragments, with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+Eight independent whole-genome MCMC analyses were run, for 200 million states each subsampling every 20000 states.
+For two-fragment MCMC analyses ten independent runs were set up.
+Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin.
+
+\subsection*{Epidemiological analyses}
+
+\begin{equation}
+r_{j} = \frac{\Gamma(kj+j-1)}{\Gamma(kj)\Gamma(j+1)} \frac{(\frac{R_{0}}{k})^{j-1}}{(1+\frac{R_{0}}{k})^{kj+j-1}}
+\end{equation}
+
 \subsection*{Data availability}

### 2017-05-02 mers-structure.tex
changed to omega notation; added simple model for seasonality

@@
 	}
-	\label{}
+	\label{fig:mers_chain}
 \end{figure}
@@
 \begin{equation}
-r_{j} = \frac{\Gamma(kj+j-1)}{\Gamma(kj)\Gamma(j+1)} \frac{(\frac{R_{0}}{k})^{j-1}}{(1+\frac{R_{0}}{k})^{kj+j-1}}
+p(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}
 \end{equation}
+
+\subsubsection*{Introduction seasonality}
+We computed the times of camel-to-human introductions in the posterior distribution of trees.
+This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
+We model the variable $Y_{ij} = \sum_{k = 1}^M Z_{ijk}$ with the hierarchical model:
+\begin{align}
+ Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
+ logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
+ \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\ 
+  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
+ \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y});  \\
+ \mu_{y}  &\sim  \text{Normal}(0, 1); \\
+ \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
+\end{align}
 

### 2017-06-09 mers-structure.tex
First draft.

--- 2017-03-06 mers-structure.tex
@@
 \usepackage{grffile}
+\usepackage{subfig}
 %\DeclareGraphicsExtensions{.pdf,.png,.jpg}
@@
 \author[2]{Luiz Max Carvalho}
-\author[1]{Allison Black}
+\author[2,3,4]{Andrew Rambaut}
 \author[1]{Trevor Bedford}
-\author[2,4,5]{Andrew Rambaut}
 
@@
 
-Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus causing significant mortality and morbidity largely in the Arabian Peninsula.
-The epidemiology of the virus is relatively poorly understood with most MERS patients being older males with comorbidities, although large nosocomial have occurred, notably in Jeddah, Riyadh and South Korea.
-Our understanding of MERS-CoV epidemiology is further complicated by some patients not recalling contact with camels, the accepted reservoir for MERS-CoV, or any other livestock, suggesting significant cryptic MERS-CoV circulation in communities.
-Seroepidemiology has been employed extensively during the time since the discovery of MERS-CoV to understand the exposure and therefore risk among camels and humans alike, but various sequencing efforts have not been fully leveraged against this zoonotic virus.
-Here we use existing MERS-CoV sequencing data to estimate the number of times the virus has been introduced into humans and use the distribution of sequence clusters to estimate the reproductive number for the virus from sequence data alone.
-We confirm that MERS-CoV has relatively poor capacity to spread in humans and arrive at an estimate of at least 65 zoonotic introductions of the virus into humans.
+Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
+The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
+Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
+Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
+We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
+By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
 
@@
 Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
-The virus, first discovered in 2012 \cite{WHO_2016}, has gone on to cause more than 1800 infections with at least 600 associated deaths.
-Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities.
-Whilst contact with camels is often reported by patients, many do not recall contact with any livestock, suggesting a significant but unobserved community contribution to the outbreak.
-Studies into MERS-CoV epidemiology often rely on serology to identify factors associated with MERS CoV exposure in potential risk groups or hosts.
-The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios.
-Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios, which are a fundamental part of quantifying risk.
-Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir, nosocomial outbreak investigations and routine surveillance.
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 1900 infections and numerous deaths, according to the World Health Organization (WHO).
+Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
+Whilst contact with camels is often reported, patients may not recall contact with livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
+Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
+Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015} compared to the general population.
+Traditional epidemiological approaches have also been used to look at distributions of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.
 
-It is accepted that human MERS-CoV infections are a result of multiple introductions of the virus into humans, but no study has attempted to go beyond educated guesses about this number.
-Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts: humans and camels.
-By explicitly modelling the evolution of MERS-CoV between these two distinct host populations we show that human MERS infections are the result of frequent and unidirectional spillover events from camels into humans.
-Inter-outbreak evolution of the virus occurs exclusively in camels, highlighting the need for genomic surveillance of MERS-CoV in the reservoir as well.
-We also use the sequence clusters recovered from our population structure analyses to estimate the reproductive number for the virus (R$_{0}$), which is in agreement with previous findings that MERS-CoV is, on average, poor at spreading in human populations.
-Our analyses also capture seasonal variation in zoonotic transmission patterns, with indications that particularly large outbreaks are the result of zoonotic transmissions that take place at the beginning of each new year.
+Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much remains opaque to such approaches.
+The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
+Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
+Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
+A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether introductions were of a single origin and homogenous or a result of increased incidence \citep{fagbo_molecular_2015}.
+Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.
 
-%Here we use 307 largely complete MERS-CoV genomes, 91 of which are from camel viruses, to explicitly model the population structure of the virus in humans and camels in order to arrive at an estimate of the number of times the virus has been introduced into human populations.
-%We then explore the resulting sequence clusters that can be thought of as replicates of the transmission process, albeit with different settings, as the data include large nosocomial outbreaks.
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO's guidelines.
+Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
+Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of population structure \citep{vaughan_efficient_2014}.
+Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly estimate distinct branching rates in each deme separately.
+By jointly estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two proposed demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
+Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
+Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
+Using approximate Bayesian computation we show that R$_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that conversely the virus has been introduced into humans hundreds of times.
 
@@
 
-\subsection*{Recombination occurs in the reservoir}
-Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
-Since detectable and problematic recombination in RNA viruses requires co-infection with distinct viral genotypes humans were ruled out as a suitable recombination environment \textit{a priori}.
-Here we confirm this assumption by showing that evolution of MERS-CoV within human clusters is clonal.
-This is not surprising, since zoonotic transmission of MERS-CoV into humans is relatively rare, of limited diversity at the point of introduction and viral diversity in humans is ultimately transient.
-All of these factors act to make co-infection of humans with MERS-CoV exceedingly rare.
-It is worth highlighting that MERS-CoV almost certainly recombines during human infection, but the limited diversity of virus populations in individual humans renders it invisible and of little consequence.
+\subsection*{MERS-CoV is predominantly a camel virus}
+The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (figure \ref{exploded}).
+Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel to human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
+Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (figure \ref{mcc}).
+We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
+It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows the model to consider humans as a potential alternative host.
+The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
+Our finding of negligible human-to-camel transmission is robust to choice of prior (figure \ref{prior}).
 
-Although we do not detect recombination in human clusters, we see evidence for introduction of recombinant viruses into humans which evolve clonally after the transfer.
-This is not unexpected either - there are strong indications that the Korean MERS outbreak was caused by a recombinant virus crossing the species barrier.
-The frequency of MERS-CoV recombination in the reservoir implies that co-infection of camels is quite common, and by extension that the virus is likely to be very prevalent in this host.
-
-\subsection*{Effects of recombination are minimal}
-Most phylogenetic approaches model sequence evolution as a strictly clonal process and when recombination is present traditional phylogenetic approaches can fail to recover correct topologies, branch lengths or demographic history.
-All coronavirus genera are known to be recombinogenic, and MERS-CoV is no exception, but whereas recombination is obligate for some eukaryotes, detectable RNA virus recombination can only occur during co-infection with two or more distinct viral genotypes.
-Previous studies have inferred prevalent recombination in MERS-CoV and found direct evidence for it.
-When not directly looking for recombination, however, most studies carried out their objectives without regard for potential effects of recombination.
-Our independent analyses of the alignment split into two fragments reveals that whilst some effects of recombination are certainly detectable, they do not seem to be systematically biasing phylogenetic inference.
-%So whereas we find some evidence of recombinant viruses crossing the species barrier into humans
-
-\subsection*{Numerous cross-species transmissions of MERS-CoV}
-It is generally accepted that MERS-CoV is a continuously re-emerging human pathogen, however the exact number of cross-species transmission events remains unknown.
-We confirm this by identifying the camel reservoir as the host population where most of MERS-CoV evolution takes place (see figure \ref{mcc}).
-The human side of MERS outbreaks, on the other hand, are caused by repetitive zoonotic transmissions from camels, which we estimated by using an explicit model of population structure.
-We infer that across 216 MERS-CoV sequences collected from humans, we detect a median of 67 individual camel to human cross-species transmissions (95\% highest posterior density: 60--72), with little support for back-transmission scenarios (median 3, 95\% HPDs: 0--13).
-There is also apparent heterogeneity in resulting human cluster sizes, which is not unexpected.
-Large clusters of MERS are apparent and mostly correspond to known large transmission chains, often in a nosocomial setting, such as the outbreak in South Korea and Riyadh.
-We also note a general predisposition for large MERS-CoV clusters to have crossed the species barrier into humans early in each year.
+The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (figure \ref{mcc} and \ref{exploded}).
+We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
+These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
+Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and robust to choice of prior.
+We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.
 
@@
 	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
-	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV sequences from humans and camels.}
-	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered using the structured coalescent.
+	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from humans and camels.}
+	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
 	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
+  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
+  Some of the branches leading up to zoonotic transmission of MERS-CoV into humans are too long to have spent much of their time in humans, given our \textit{a priori} belief that MERS-CoV cannot circulate in humans for long periods of time.
+  Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
 	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
@@
 
-\subsection*{MERS epidemiology}
-Although sequences in an outbreak necessarily represent a subset of all the cases, their clustering and resulting cluster sizes can be highly informative of underlying epidemiological processes that are otherwise inaccesible or unfeasible to study.
-In our case we have used sequence data to disentangle the evolution of MERS-CoV into two discrete host compartments.
-By removing the confounding effects of shared ancestry for MERS-CoV sequences we can treat every viral introduction into humans as an independent replicate of the transmission process in humans.
+\subsection*{Signals of MERS-CoV seasonality}
+Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to recover probable distribution of migration times.
+We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
+We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
+Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
+It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to seasonal variation in force of infection in humans.
+Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology.
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
+	\caption{\textbf{MERS-CoV seasonality.}
+A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
+Posterior means are indicated with circles.
+Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
+Grey area spanning October to February indicates the camel calving season.
+B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
+95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
+The black dotted line indicates the joint probability density for introductions.
+Grey areas spanning October to February indicate the reported camel calving season.
+	}
+	\label{seasonality}
+\end{figure}
+
+%\begin{figure}[h]
+% \centering
+%	\includegraphics[width=0.45\textwidth]{figures/}
+%	\caption{\textbf{}
+%
+%	}
+%	\label{}
+%\end{figure}
+
+\subsection*{MERS-CoV is poorly suited for human transmission}
+Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
+However, we wanted to translate this observation into the basic reproductive number, R$_{0}$, which is more familiar to epidemiologists, as well as extrapolate the migration process observed from sequence data to case data.
+In order to achieve this we perform approximate Bayesian computation (ABC) analyses with two steps.
+We simulate outbreaks using equation \ref{clusterLikelihood} with a fixed dispersion parameter and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing.
+Sequencing simulations take place at different levels of bias, where case clusters likely to be sequenced become even more likely to be sequenced and \textit{vice versa} for cases that would be sequenced rarely.
+This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from index cases will be sampled exceedingly rarely.
+We record the R$_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses.
+
+We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in figure \ref{ABC}).
+For increasing levels of bias mean R$_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
+Whilst the 95\% percentiles for R$_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
+Under unbiased sequencing only 2282 (out of 363 000) simulations fit our observations in structured coalescent analyses of MERS-CoV which increase to 9763 and 9923 for bias levels of 2.0 and 3.0, respectively.
+This is apparent in the heights of the histograms in figure \ref{ABC}.
+Simulations with bias suggest that R$_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
+
+Lower values for R$_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans.
+The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
+These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously seem to produce considerably more MERS-CoV-like data.
+Our results suggest a large number of unobserved MERS index cases.
+Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
+In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (figure \ref{extra_cases}).
+Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated R$_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
+Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices appear sufficient to explain observed epidemiological patterns of MERS-CoV.
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
+	\caption{\textbf{Results of ABC simulations.}
+Each row corresponds to a different bias value used to concentrate the hypergeometric distribution that simulates sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias).
+Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the R$_{0}$ value used for the simulation.
+The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
+The distribution of R$_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
+Bins falling inside the 95\% percentiles are coloured by R$_{0}$, as in the leftmost scatter plot.
+The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
+Darker shade of grey indicates bins falling within the 95\% percentiles.
+ABC indicates R$_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
+	}
+	\label{ABC}
+\end{figure}
+
+%\begin{figure}[h]
+% \centering
+%	\includegraphics[width=0.65\textwidth]{figures/}
+%	\caption{\textbf{}
+%
+%	}
+%	\label{}
+%\end{figure}
+
+\subsection*{Recombination shapes MERS-CoV diversity}
+Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
+In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of data corresponding to inferred MERS-CoV clades.
+Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
+One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, that appear as repeated mutations taking place in the recipient lineage, somewhere else in the tree.
+PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
+Another hallmark of recombination is spatial clustering of alleles in the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
+3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.
+
+Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
+PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (figure \ref{recombination_tree}).
+Neither method can identify where in the tree recombination occurred, but each full asterisk in figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
+This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
+As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
+Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
+Though the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
+This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur.
+
+Conversely, our results argue strongly that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
+We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (figure \ref{recombination}A).
+And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (figure \ref{flower}).
+Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue in MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
+Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
+Conversely, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
+
+% \begin{figure}[h]
+%  \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_chain.png}
+%   \includegraphics[width=0.65\textwidth]{figures/mers_fragments.png}
+% 	\caption{\textbf{Tangled chain of MERS-CoV genome and its parts.}
+% 	Tangled chain of the full genome (a), all nucleotides up to position 21000 (b) and all nucleotides past position 21000 (c) of MERS-CoV.
+% 	The same taxa, if available in neighbouring trees, are connected via coloured lines (human sequences in blue, camel sequences in red).
+% 	Tree branches are coloured by inferred ancestral host state (human in grey, camel in black).
+% 	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in the camel host.
+% 	Human clusters in grey change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
+% 	This is evidence for recombinant viruses generated in the reservoir entering human populations.
+% 	Coding sequences in the MERS-CoV genome are indicated with arrows at the bottom of the plot with position 21000 indicated by the black line.}
+% 	\label{mers_chain}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+%
+% 	\caption{\textbf{Pairwise co-occurence matrix for human clusters.}
+% Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
+% Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
+% Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
+% 	}
+% 	\label{coocurrence}
+% \end{figure}
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{figures/mers_fragments.png} }}%
+    \qquad
+    \subfloat{{\includegraphics[width=5cm]{figures/mers_chain.png} }}%
+    \caption{\textbf{Recombinant features of MERS-CoV phylogenies}.
+    A) Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
+    Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
+    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
+    B) Phylogenies derived from MERS-CoV genome nucleotides up to position 21000 (left) and all nucleotides past position 21000 (right), reduced to just the human tips.
+  	Same taxa are connected via coloured lines indicating their vertical position in the tree on the right.
+  	Tree branches are coloured by inferred ancestral host state (human in blue, camel in orange).
+  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels.
+  	Human clusters in blue change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
+  	This is evidence for recombinant viruses generated in the reservoir entering human populations.
+    }%
+    \label{recombination}%
+\end{figure}
+
+\subsection*{Demographic trajectory of MERS-CoV in camels}
+Since MERS-CoV in humans evolves under a distinct demographic regime, we split our sequence dataset based on whether sequences were collected in humans or camels.
+We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
+This removes conflicting demographic signals sampled during human outbreaks, where densely sampled closely related sequences from humans could be misconstrued as evidence of demographic crash in the viral population.
+
+Our attempt at controlling the confounding effects of different sampling regimes between humans and camels revealed a troubling pattern, however.
+We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions but do not necessarily converge on the same distributions.
+Of the five runs we set up two chains converged to one stationary distribution, two to another and the last one to a third.
+Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
+This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have nudged MCMC towards one of the multiple stationary distributions.
+
+Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (figure \ref{skygrid}).
+However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+This raises the possibility that MERS-CoV populations undergo lineage turnover at a rate not often seen in other livestock viruses (\textit{e.g.} swine influenza A viruses \citep{vijaykrishna_long-term_2011}).
+Part of this could be caused by MERS-CoV ability to re-infect camels previously exposed to the virus \citep{ali_systematic_2017}, which may result in a process akin to antigenic drift, where the depletion of susceptible calves leads to ever increasing selective pressure for the virus to evolve antigenically.
+Whilst this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.
+
+% In order to disentangle distinct demographic modes within the phylogeny of MERS-CoV we split our alignment according to whether the virus was collected from camels or humans.
+% For sequences collected from humans and inferred to be in the same cluster we only keep the earliest sequence, and arrive at a sequence dataset that for the most part is comprised of viral diversity present in camels, and supplemented with human sequences that are treated as sentinel samples.
 
@@
 
-\subsection*{MERS-CoV in camels}
-Our observation of recombinant MERS coronaviruses being transmitted to humans is not cause for concern by itself, however, the mere presence of detectable recombination implies that co-infection of camels with MERS-CoV is common.
-By extension, frequent co-infection implies a high prevalence of the virus in camels, which is the main unsettling conclusion.
-There is ample evidence for this from seroepidemiology, which has shown that XX\% of camels are infected by age XX.
-Our results show that zoonotic transmission of MERS-CoV into humans is the main drivers for the continuation of the outbreak in the Arabian Peninsula. 
-Prevalence of MERS-CoV in camels could easily be translated into force of infection and given the role of zoonotic transmission as the main driver for the outbreak only reiterates the importance of controlling the MERS coronavirus in the reservoir if the human side of the outbreak is to be brought under control.
+\subsection*{Role of population structure in phylodynamics}
+Population structure yields important insight into any biological system.
+Population structure often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
+Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
+In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
+Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.
+
+We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (figures \ref{mcc}, \ref{exploded}).
+This inference succeeds under different choices of priors for unknown demographic parameters (figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (figure \ref{ABC}).
+From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (figure \ref{ABC}).
+We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
+Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (figure \ref{seasonality}).
+As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (figure \ref{ABC}).
+Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
+Perhaps a clearer indicator of this has been the lack of a global MERS pandemic, which would have certainly taken place had R$_{0}$ been above the epidemic threshold.
+
+Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
+As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
+Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
+Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
+Some convergence issues (figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
+As computational infrastructure improves \citep{ayres_beagle:_2012}, it is hopefully only a matter of time before phylogenetic population structure models mature to the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) can be analysed trivially.
+
+% Similar approaches have been used to quantify population structure in American Zika viruses, where repeated introductions into Florida observed in sequence data were used to extrapolate probable numbers of viral introductions
+
+% Population structure is often inferred via phylogeny labelling approaches where discrete states are assigned to taxa, and ancestral states are reconstructed via a continuous time Markov chain the same way nucleotide sites would be.
+% These methods, when used to infer ancestral migration history across a phylogeny, have thus been called ``mugration'' models.
+% When the number of possible discrete states is limited, data are ample and questions are formulated well, phylogeny labelling approaches are computationally efficient, accurate and highly informative.
+% However, ``mugration'' models ignore potentially relevant information, such as markedly different demographic histories in distinct populations.
+%
+% Sequencing of MERS-CoV has been a challenging issue, especially from the reservoir.
+% MERS-CoV seroprevalence in camels, as expected of a reservoir, is very high, yet sequencing efforts have overwhelmingly focused on human infections.
+% At present MERS-CoV sequences collected from humans outnumber those from camels at a ratio of over 2-to-1, and sequences from camels appear late in the MERS epidemic, largely because of the time taken to identify camels as the reservoir.
+% Hence, using MERS-CoV sequence data in ``mugration'' models have led to contradictory findings, such as inference of MERS-CoV as a repeated epizootic in camels of largely human origins \citep{zhang_evolutionary_2016}.
+% Traditional epidemiological approaches fare no better, since links between human cases outside of known family clusters or hospital outbreaks are exceedingly difficult to establish \citep{cauchemez_unraveling_2016}.
+%
+% Our study has shown how significantly different demographic regimes between hosts such as those observed in MERS-CoV can be leveraged against biased sequence data to arrive at conclusions perfectly in line with prior observations.
+% MERS-CoV sequences collected from camels are, on average, relatively distantly related to each other, whilst human outbreaks fall into densely sampled clusters of no apparent relation to each other and of no consequence to further evolution of the virus.
+% These hallmarks of distinct demographic modes were easily picked up by the explicit population structure model we employ, and reveals a cohesive picture of MERS-CoV epidemiology in agreement with previous empirical studies.
+% However, we do note that at present phylogenetic models with population structure are limited in terms of computational efficiency than current models, even when approximations are used.
+
+% \subsection*{Issues with recombination}
+% Recombination in MERS-CoV is now an accepted feature of the system.
+% Prevalent recombination can seriously impede phylogenetic and other sequence analysis techniques.
+% We argue that although MERS-CoV sequences contain overwhelming evidence for recombination, we do not expect humans to play a major role in this process.
+% As human infection with MERS-CoV remains very low, every introduction of the virus into a human population is expected to establish a clonally evolving and short-lived population.
+% However, lineages crossing the species barrier are very likely to have experienced recombination at some point in their past, in some cases occurring right before zoonotic transmission, as was the case in the Korean outbreak of MERS.
+% We thus conclude that whilst sequence analyses of MERS-CoV are perfectly legitimate for well-defined human outbreaks, extreme care must be taken when inferring their origins in the context of the reservoir, or even when comparing related outbreaks in humans.
+% We also strongly advise against drawing hasty conclusions from MERS-CoV phylogenies recovered from human sequences if they span more than four months of evolution or camel sequences under virtually all circumstances, as there is no guarantee that phylogenies will cover the clonal evolutionary history of the virus in humans.
+
+\subsection*{Sequence data as a universal epidemiology tool}
+As the field of epidemiology enters the age of genomics sequence data are likely to be brought to the forefront of epidemiological investigations.
+At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
+With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
+This was used to great effect towards the end of the Ebola virus epidemic in West Africa, where the final transmission chains were tracked and confirmed as such with sequence data \citep{arias_rapid_2016}.
+Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows data from wildly different studies to be combined trivially.
+In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.
 
@@
 Sequences were annotated with available collection dates and hosts, designated as camel or human.
-For partitioned analyses the alignment was split into two fragments, the first containing all nucleotides up to nucleotide 21000, and the rest of the genome being assigned to the second fragment.
-Despite their divergence both Severe Acute Respiratory Syndrome (SARS) and MERS betacoronaviruses appear to recombine across position 21000, which may indicate the existence of a recombinogenic secondary RNA structure that remains to be described in its vicinity.
-%% "Work by Zhang et al.
-(66) suggested that the presence of a stable hairpin between two direct repeats increases the rate of homologous recombination in MLV.
-Their data led to the conclusion that stable secondary structures decrease processive RT synthesis and, further, that a functional NC can increase RT processivity.
-However, in this case, there were no corresponding in vitro data (such as primer extension assays) to correlate with the in vivo observations.
-Here, we have demonstrated for the first time a direct relationship between pausing in vitro and viral recombination in vivo."
 
 \subsection*{Structured coalescent analyses}
-MultiTypeTree module was used in BEAUti v2.4.3 to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
-Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ nucleotide substitution models specified for codon positions 1+2 and 3.
-A relaxed molecular clock with branch rates drawn from a lognormal distribution was used to infer the evolutionary rate from date calibrated tips.
+MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
+Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
+A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
 Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-An identical set up was used for analysing alignments split into two fragments, with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
-Eight independent whole-genome MCMC analyses were run, for 200 million states each subsampling every 20000 states.
-For two-fragment MCMC analyses ten independent runs were set up.
-Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin.
+All analyses were run across ten independent Markov chains.
+To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
+
+We also analysed a dataset where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
+All analyses were set to run for 200 million states, subsampling every 20 000 states.
+Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin for single fragment analyses and 20\% for two-fragment analyses.
+Chains not converging after 200 million states were discarded.
+For single fragment analysis with migration rate prior set to exponential with mean 1.0 there were three chains out of ten that did not converge and two for the analysis where the migration rate prior was an exponential with mean 10.0.
+For the two fragment structured coalescent analyses three chains out of ten failed to converge.
+
+\subsection*{Demographic inference of MERS-CoV in the reservoir}
+In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
+The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
+These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
+Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
+A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
+We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
+This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
 
 \subsection*{Epidemiological analyses}
+We simulate final transmission chain sizes using equation \ref{clusterLikelihood} \citep{lloyd-smith_superspreading_2005,blumberg_inference_2013} until we reach an epidemic comprised of 2000 cases.
+As of 10 May 2017, the World Health Organization has been notified of 1952 cases of MERS-CoV.
+10 000 simulation were run for 121 uniformly spaced values of R$_{0}$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1.
+Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where K is the number of clusters generated.
+We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing.
+The resulting sequence cluster size vector $S$ contains K entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
+Since the sampling scheme operates under equi-probable sequencing we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
+This makes clusters likely to be `sequenced' even more likely to be sequenced and \textit{vice versa}.
+We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to simulate a dramatically underreported epidemic.
 
 \begin{equation}
-r_{j} = \frac{\Gamma(kj+j-1)}{\Gamma(kj)\Gamma(j+1)} \frac{(\frac{R_{0}}{k})^{j-1}}{(1+\frac{R_{0}}{k})^{kj+j-1}}
+p(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}
+\label{clusterLikelihood}
 \end{equation}
 
+Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.
+Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species transmission of the virus from camels into humans.
+The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying distribution of case clusters.
+We employ an approximate Bayesian computation (ABC) approach to identify simulations where the recovered distribution of sequence cluster sizes fall within the 95\% highest posterior density intervals for three summary statistics of MERS-CoV sequence cluster sizes recovered via structured coalescent analyses: mean, median and standard deviation.
+These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
+
+\subsubsection*{Introduction seasonality}
+We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
+This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
+We model the variable $Y_{ij} = \sum_{k = 1}^M Z_{ijk}$ with the hierarchical model:
+\begin{align*}
+  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
+  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
+  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
+  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
+  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
+  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
+  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
+\end{align*}
+
 \subsection*{Data availability}
+Sequence data and all analytical code is publicly available at https://github.com/blab/structured-mers.
 
 \section*{Acknowledgements}
+We would like to thank Allison Black for useful discussion and advice.
+GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
 
@@
 
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_exploded.png}
+	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
+Same tree as shown in figure \ref{mcc}, but all contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top) and humans (bottom).
+This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
+	}
+	\label{exploded}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
+	\caption{\textbf{Result are robust to prior choice.}
+Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.
+Plots on the left show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (left) and \textit{vice versa} (right).
+Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (top) or 10.0 (bottom).
+	}
+	\label{prior}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}
+	\caption{\textbf{Results of ABC simulations with vast underestimation of cases.}
+The plot is identical to figure \ref{ABC}, but instead of 2000 cases, the currently reported number of MERS-CoV cases, simulations were run with 4000 cases.
+With more unobserved cases the R$_{0}$ values matching observed MERS-CoV sequence clusters can only be smaller, with a corresponding increase in numbers of zoonotic transmissions.
+However, the numbers of simulations that match MERS-CoV data go down too.
+	}
+	\label{extra_cases}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_recombination.png}
+	\caption{\textbf{Tests of recombination across MERS-CoV clades.}}
+Maximum clade credibility tree of MERS-CoV genomes annotated with results of two recombination detection tests (PHI and 3Seq) applied to descendent sequences of each clade.
+Both tests identify large portions of existing sequence data as containing signals of recombination.
+Note that markings do not indicate where recombinations have occurred on the tree, merely the minimum distance in sequence/time space between recombining lineages.
+	\label{recombination_tree}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_incompatibilities.png}
+	\caption{\textbf{MERS-CoV genomes exhibit high numbers of non-clonal loci.}
+Ancestral state reconstruction (right) identifies a large number of loci that have occurred more than once in the tree (homoplasies, orange) or are reversions (red) from a state arising in an ancestor.
+Mutations that apparently only occur once in the tree (synapomorphies) are shown in grey.
+The maximum likelihood phylogeny on the left is coloured by whether sequences were sampled in humans (blue) or camels (orange).
+	}
+	\label{incompatibilities}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_flower.png}
+	\caption{\textbf{Human clade sharing between genomic fragments 1 and 2.}
+Central scatter plot shows the posterior probability of human clades shared between genomic fragments 1 and 2, in their respective trees.
+Left and bottom scatter plots track the posterior probability of human clades only observed in fragment 2 (left) or fragment 1 (bottom).
+The cumulative probability of human clades present in either tree are tracked by plots on the right (fragment 2) and top (fragment 1).
+Most of the probability density is concentrated within human clades that are present in trees of both genomic fragment 1 and 2 (0.9701 and 0.9474 of all human clades across posteriors, respectively).
+	}
+	\label{flower}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid.png}
+	\caption{\textbf{Demographic history of MERS-CoV in Arabian peninsula camels.}
+Demographic history of MERS-CoV in camels, as inferred via a skygrid coalescent tree prior \citep{gill_2013}.
+Two skygrid reconstructions are shown, one for each of the stationary distributions reached by MCMC.
+Shaded interval indicates the 95\% highest posterior density interval for the product of generation time and effective population size, $N_{e}\tau$.
+Midline tracks the inferred median of $N_{e}\tau$.
+Maximum clade credibility (MCC) tree from the skygrid was inferred is shown in the background, with camel sequences highlighted in orange and human sequences highlighted in blue.
+	}
+	\label{skygrid}
+\end{figure}
+
 \end{document}

### 2017-06-20 mers-structure.tex
Comments and revisons through Results/Seasonality
This is excellent. Stylistic changes (like "figure 1" to "Figure 1") are attempts to move closer to eLife style.

@@
 \usepackage{float}
+\usepackage{url}
+\urlstyle{same}
 
@@
 The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
-Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
+Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
 Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
 We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
+\tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'.
+I think you want a different word here.
+Maybe `transient host'?}
 By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
+\tbc{Powerful and succient abstract. Great.}
 
@@
 Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
-The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 1900 infections and numerous deaths, according to the World Health Organization (WHO).
-Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
+Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
 Whilst contact with camels is often reported, patients may not recall contact with livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
 Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
-Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015} compared to the general population.
-Traditional epidemiological approaches have also been used to look at distributions of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.
+Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
+Epidemiological modeling approaches have been used to look at traditional data sources of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.
 
-Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much remains opaque to such approaches.
-The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
+Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much, inevitably, remains opaque to such approaches.
+The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
 Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
 Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
-A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether introductions were of a single origin and homogenous or a result of increased incidence \citep{fagbo_molecular_2015}.
+A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
 Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.
 
-It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO's guidelines.
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines.
 Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
-Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of population structure \citep{vaughan_efficient_2014}.
-Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly estimate distinct branching rates in each deme separately.
-By jointly estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two proposed demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
+Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
+Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
+By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
 Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
 Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
-Using approximate Bayesian computation we show that R$_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that conversely the virus has been introduced into humans hundreds of times.
+Using approximate Bayesian computation we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.
 
@@
 \subsection*{MERS-CoV is predominantly a camel virus}
-The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (figure \ref{exploded}).
-Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel to human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
-Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (figure \ref{mcc}).
+The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure \ref{exploded}).
+Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
+Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure \ref{mcc}).
 We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
@@
 The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
-Our finding of negligible human-to-camel transmission is robust to choice of prior (figure \ref{prior}).
+Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure \ref{prior}).
 
-The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (figure \ref{mcc} and \ref{exploded}).
+The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure \ref{mcc} and \ref{exploded}).
 We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
 These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
-Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and robust to choice of prior.
+Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
 We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.
@@
 \subsection*{Signals of MERS-CoV seasonality}
-Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to recover probable distribution of migration times.
-We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
-We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
+Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to estimate migration times.
+We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
+We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
 Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
@@
 B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
-95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
+95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (Figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
 The black dotted line indicates the joint probability density for introductions.
@@
 
-We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in figure \ref{ABC}).
+We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in Figure \ref{ABC}).
 For increasing levels of bias mean R$_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
@@
 Under unbiased sequencing only 2282 (out of 363 000) simulations fit our observations in structured coalescent analyses of MERS-CoV which increase to 9763 and 9923 for bias levels of 2.0 and 3.0, respectively.
-This is apparent in the heights of the histograms in figure \ref{ABC}.
+This is apparent in the heights of the histograms in Figure \ref{ABC}.
 Simulations with bias suggest that R$_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
@@
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
-In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (figure \ref{extra_cases}).
+In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
 Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated R$_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
@@
 Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
-PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (figure \ref{recombination_tree}).
-Neither method can identify where in the tree recombination occurred, but each full asterisk in figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
+PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (Figure \ref{recombination_tree}).
+Neither method can identify where in the tree recombination occurred, but each full asterisk in Figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
 This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
-As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
+As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
 Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
@@
 Conversely, our results argue strongly that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
-We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (figure \ref{recombination}A).
-And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (figure \ref{flower}).
+We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (Figure \ref{recombination}A).
+And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
 Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue in MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
@@
 
-Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (figure \ref{skygrid}).
+Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (Figure \ref{skygrid}).
 However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
@@
 
-We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (figures \ref{mcc}, \ref{exploded}).
-This inference succeeds under different choices of priors for unknown demographic parameters (figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (figure \ref{ABC}).
-From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (figure \ref{ABC}).
+We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
+This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{ABC}).
+From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{ABC}).
 We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
-Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (figure \ref{seasonality}).
-As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (figure \ref{ABC}).
+Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
+As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{ABC}).
 Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
@@
 Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
-Some convergence issues (figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
+Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
 As computational infrastructure improves \citep{ayres_beagle:_2012}, it is hopefully only a matter of time before phylogenetic population structure models mature to the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) can be analysed trivially.
@@
 	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
-Same tree as shown in figure \ref{mcc}, but all contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top) and humans (bottom).
+Same tree as shown in Figure \ref{mcc}, but all contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top) and humans (bottom).
 This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
@@
 Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (top) or 10.0 (bottom).
+\tbc{I think this would be more clear if the left and right sides shared the same x-axis.
+As it stands it looks at first glance like there's different priors for camel-to-human compared to human-to-camel.
+It's okay if the blue human-to-camel rates gets smushed up against the left of the plot.}
 	}
@@
 	\caption{\textbf{Results of ABC simulations with vast underestimation of cases.}
-The plot is identical to figure \ref{ABC}, but instead of 2000 cases, the currently reported number of MERS-CoV cases, simulations were run with 4000 cases.
+The plot is identical to Figure \ref{ABC}, but instead of 2000 cases, the currently reported number of MERS-CoV cases, simulations were run with 4000 cases.
 With more unobserved cases the R$_{0}$ values matching observed MERS-CoV sequence clusters can only be smaller, with a corresponding increase in numbers of zoonotic transmissions.

### 2017-06-21 mers-structure.tex
minor maths typo
Through epi section of results

@@
 Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
-It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to seasonal variation in force of infection in humans.
-Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology.
+It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to increased force of infection sparking human outbreaks.
+Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.
 
@@
 \subsection*{MERS-CoV is poorly suited for human transmission}
-Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
-However, we wanted to translate this observation into the basic reproductive number, R$_{0}$, which is more familiar to epidemiologists, as well as extrapolate the migration process observed from sequence data to case data.
-In order to achieve this we perform approximate Bayesian computation (ABC) analyses with two steps.
-We simulate outbreaks using equation \ref{clusterLikelihood} with a fixed dispersion parameter and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing.
-Sequencing simulations take place at different levels of bias, where case clusters likely to be sequenced become even more likely to be sequenced and \textit{vice versa} for cases that would be sequenced rarely.
+Structured coalescent approaches clearly show humans to be a terminal host \tbc{again, I don't know if this is the right term or not} for MERS-CoV, implying poor transmissibility.
+However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides expectations of epidemic behavior.
+The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation \ref{clusterLikelihood}).
+We estimate $R_{0}$ along with other relevant parameters via approximate Bayesian computation (ABC) in two steps.
+We simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
+Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
 This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from index cases will be sampled exceedingly rarely.
-We record the R$_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses.
+We record the $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses (left-hand panels in Figure \ref{ABC}).
 
-We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in Figure \ref{ABC}).
-For increasing levels of bias mean R$_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
-Whilst the 95\% percentiles for R$_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
-Under unbiased sequencing only 2282 (out of 363 000) simulations fit our observations in structured coalescent analyses of MERS-CoV which increase to 9763 and 9923 for bias levels of 2.0 and 3.0, respectively.
+We find that observed phylogenetic patterns sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
+For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
+Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
+Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
 This is apparent in the heights of the histograms in Figure \ref{ABC}.
-Simulations with bias suggest that R$_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
+This is equivalent to a higher marginal likelihood for bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).
+Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
 
-Lower values for R$_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans.
+Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{ABC}).
 The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
-These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously seem to produce considerably more MERS-CoV-like data.
+These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
 Our results suggest a large number of unobserved MERS index cases.
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
-In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
-Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated R$_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
+In those sets of analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
+
+Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
+\tbc{I didn't understand what you meant by `sequence sampling wholly unrepresentative of viral diversity found in humans'.}
 Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices appear sufficient to explain observed epidemiological patterns of MERS-CoV.
+\tbc{It would be really nice to be able to reference something here on hospital spread of MERS (or SARS) owing to hygiene practices.}
 
@@
 Each row corresponds to a different bias value used to concentrate the hypergeometric distribution that simulates sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias).
-Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the R$_{0}$ value used for the simulation.
+Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
 The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
-The distribution of R$_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
-Bins falling inside the 95\% percentiles are coloured by R$_{0}$, as in the leftmost scatter plot.
+The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
+Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
 The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
 Darker shade of grey indicates bins falling within the 95\% percentiles.
-ABC indicates R$_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
+ABC indicates $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
+\tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to eleviate confusion.
+I'm picturing a set of tables numbered with case counts in a top row and then a bottom row with sequence counts.
+This could also work better as a separate figure referenced from the Methods.}
 	}
@@
 \subsection*{Recombination shapes MERS-CoV diversity}
-Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
-In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of data corresponding to inferred MERS-CoV clades.
+Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \tbc{needs citations}.
+In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
 Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
@@
 This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
-We model the variable $Y_{ij} = \sum_{k = 1}^M Z_{ijk}$ with the hierarchical model:
+We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
 \begin{align*}

### 2017-07-07 mers-structure.tex
Note on dataset size
Implement @trvrb suggestions.
Methods figure + change prior figure.

@@
 \def\tbc#1{\textcolor{purple}{[#1]}}
+\def\gdc#1{\textcolor{blue}{[#1]}}
 
@@
 Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
-We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
+We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
 \tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'.
@@
 Maybe `transient host'?}
+\gdc{Good point about the term. Hope the definition leans towards `one not involved in maintenance' rather than `limited to index case'. Asked a friend working on zoonoses about this.}
 By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
-\tbc{Powerful and succient abstract. Great.}
+% \tbc{Powerful and succinct abstract. Great.}
 
@@
 Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
-Using approximate Bayesian computation we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.
+Using Monte Carlo simulations we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.
 
@@
 \subsection*{MERS-CoV is poorly suited for human transmission}
-Structured coalescent approaches clearly show humans to be a terminal host \tbc{again, I don't know if this is the right term or not} for MERS-CoV, implying poor transmissibility.
+Structured coalescent approaches clearly show humans to be a terminal host %\tbc{again, I don't know if this is the right term or not}
+for MERS-CoV, implying poor transmissibility.
 However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides expectations of epidemic behavior.
 The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation \ref{clusterLikelihood}).
-We estimate $R_{0}$ along with other relevant parameters via approximate Bayesian computation (ABC) in two steps.
+We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
 We simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
@@
 
-Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
-\tbc{I didn't understand what you meant by `sequence sampling wholly unrepresentative of viral diversity found in humans'.}
-Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices appear sufficient to explain observed epidemiological patterns of MERS-CoV.
-\tbc{It would be really nice to be able to reference something here on hospital spread of MERS (or SARS) owing to hygiene practices.}
+Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
+% \tbc{I didn't understand what you meant by `sequence sampling wholly unrepresentative of viral diversity found in humans'.}
+Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
+%\tbc{It would be really nice to be able to reference something here on hospital spread of MERS (or SARS) owing to hygiene practices.}
 
@@
 	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
-	\caption{\textbf{Results of ABC simulations.}
+	\caption{\textbf{Results of Monte Carlo simulations.}
 Each row corresponds to a different bias value used to concentrate the hypergeometric distribution that simulates sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias).
@@
 Darker shade of grey indicates bins falling within the 95\% percentiles.
-ABC indicates $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
-\tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to eleviate confusion.
+Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
+% \tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to alleviate confusion.
 I'm picturing a set of tables numbered with case counts in a top row and then a bottom row with sequence counts.
@@
 \subsection*{Recombination shapes MERS-CoV diversity}
-Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \tbc{needs citations}.
+Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \citep{lai_1985,makino_1986,keck_1988,kottier_1995,herrewegh_1998}. %\tbc{needs citations}.
 In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
@@
 The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying distribution of case clusters.
-We employ an approximate Bayesian computation (ABC) approach to identify simulations where the recovered distribution of sequence cluster sizes fall within the 95\% highest posterior density intervals for three summary statistics of MERS-CoV sequence cluster sizes recovered via structured coalescent analyses: mean, median and standard deviation.
+We employ a Monte Carlo simulation approach to identify simulations where the recovered distribution of sequence cluster sizes fall within the 95\% highest posterior density intervals for three summary statistics of MERS-CoV sequence cluster sizes recovered via structured coalescent analyses: mean, median and standard deviation.
 These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
@@
 Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.
-Plots on the left show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (left) and \textit{vice versa} (right).
-Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (top) or 10.0 (bottom).
-\tbc{I think this would be more clear if the left and right sides shared the same x-axis.
+Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (blue) and \textit{vice versa} (orange).
+Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
+%\tbc{I think this would be more clear if the left and right sides shared the same x-axis.
 As it stands it looks at first glance like there's different priors for camel-to-human compared to human-to-camel.
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_method.pdf}
+	\caption{\textbf{Monte Carlo simulation method.}
+Case clusters are simulated according to \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
+We sample 174 cases from each simulation to represent sequencing of human MERS cases.
+`Sequencing' is carried out by using multivariate hypergeometric sampling, where at each iteration case cluster size $C_{i}$ is the number of successes in the population, $\sum_{i=1}^{K} C_{i} - C_{i}$ is the number of failures and number of trials is equal to leftover sequencing capacity, that is $\sum_{i=1}^{K} S_{i}$ minus whatever has been `sequenced' up to that iteration.
+Sequencing simulations take place at three levels of bias - 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
+Simulated sequence clusters are summarised by their mean and standard deviation, both of which reflect some aspect of the partitioning of sequences into clusters.
+A simulations is considered to match if its sequence cluster size mean and standard deviation fell within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
+$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are summarised in Figure \ref{ABC}.
+	}
+	\label{method}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}
-	\caption{\textbf{Results of ABC simulations with vast underestimation of cases.}
+	\caption{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
 The plot is identical to Figure \ref{ABC}, but instead of 2000 cases, the currently reported number of MERS-CoV cases, simulations were run with 4000 cases.

### 2017-07-08 mers-structure.tex
@maxbiostat's edits/comments

@@
 \def\gdc#1{\textcolor{blue}{[#1]}}
+\def\lmc#1{\textcolor{green}{[#1]}}
 
 % symbols
-\newcommand{\chiSq}{\chi^{2}_{df}}
+\newcommand{\chiSq}{\chi^{2}_{df}} %LM: ancient DNA? =P
 \newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
@@
 Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
-The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 2000  infections and over 700 deaths according to the World Health Organization (WHO) \citep{mersfactsheet}.
 Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
-Whilst contact with camels is often reported, patients may not recall contact with livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
+Whilst contact with camels is often reported, patients may not recall contact with \lmc{other?} livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
 Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
@@
 
-Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much, inevitably, remains opaque to such approaches.
-The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
-Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
+Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches \lmc{why, specifically?}.
+The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
+These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
+Often only sequence data can \lmc{help} pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
 Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
-A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
+A large portion of MERS-CoV sequences come from hospital outbreaks \lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
 Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.
 
-It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines.
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \lmc{how about a citation here?}.
 Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
@@
 \subsection*{MERS-CoV is predominantly a camel virus}
-The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure \ref{exploded}).
+
+The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
 Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
-Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure \ref{mcc}).
+Whilst we identify \lmc{estimate?} a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
 We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
-It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows the model to consider humans as a potential alternative host.
+It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows \lmc{is ``allow'' the best word here?
+I understand you' rather not use ``biases'' or ``misleads'', though...} the model to consider humans as a potential alternative host.
 The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
-Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure \ref{prior}).
+Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).
 
-The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure \ref{mcc} and \ref{exploded}).
+The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
 We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
 These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
-Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
+Our analyses recover these results despite sequence data heavily skewed towards non-uniformly \lmc{maybe ``non-uniformly'' is redundant here} sampled human cases and are robust to choice of prior.
 We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.
@@
 \subsection*{Signals of MERS-CoV seasonality}
-Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to estimate migration times.
+Unlike the most basic \lmc{simpler?} ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, structured population models possess the ability to estimate migration times.
 We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
@@
 Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
-It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to increased force of infection sparking human outbreaks.
-Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.
+It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
+Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.
 
@@
 The black dotted line indicates the joint probability density for introductions.
-Grey areas spanning October to February indicate the reported camel calving season.
+Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
 	}
@@
 \subsection*{MERS-CoV is poorly suited for human transmission}
+
 Structured coalescent approaches clearly show humans to be a terminal host %\tbc{again, I don't know if this is the right term or not}
 for MERS-CoV, implying poor transmissibility.
-However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides expectations of epidemic behavior.
-The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation \ref{clusterLikelihood}).
+However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides expectations\lmc{insights?} of epidemic behavior.
+The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by a dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}).
 We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
-We simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
+First, we simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
 Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
@@
 
-We find that observed phylogenetic patterns sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
+We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
 For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
 Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
-Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
-This is apparent in the heights of the histograms in Figure \ref{ABC}.
-This is equivalent to a higher marginal likelihood for bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).
+Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%\lmc{again?} of simulations fit for bias levels of 2.0 and 3.0, respectively.
+This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate a higher marginal likelihood for bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).\lmc{I don't follow this argument...}
 Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
@@
 These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Our results suggest a large number of unobserved MERS index cases.
+Our results suggest a large number of unobserved MERS index cases.\lmc{does the definition of ``index case'' allow for ``multiple''?}
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
@@
 
-Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
+Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
 % \tbc{I didn't understand what you meant by `sequence sampling wholly unrepresentative of viral diversity found in humans'.}
@@
 	\caption{\textbf{Results of Monte Carlo simulations.}
-Each row corresponds to a different bias value used to concentrate the hypergeometric distribution that simulates sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias).
+Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias \lmc{make sequences from large clusters more likely to be picked} ).
 Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
@@
 Darker shade of grey indicates bins falling within the 95\% percentiles.
-Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
+These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
 % \tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to alleviate confusion.
@@
 This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
-As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
+As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
 Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
-Though the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
-This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur.
+Although the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
+This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur\lmc{stupid biostatistician here: is it?}.
 
-Conversely, our results argue strongly that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
+Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
 We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (Figure \ref{recombination}A).
@@
 Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
-Conversely, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
+On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
 
@@
   	Same taxa are connected via coloured lines indicating their vertical position in the tree on the right.
-  	Tree branches are coloured by inferred ancestral host state (human in blue, camel in orange).
-  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels.
-  	Human clusters in blue change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
-  	This is evidence for recombinant viruses generated in the reservoir entering human populations.
+  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
+  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels.\lmc{I'm sure this is important, but I didn't get why.
+Referees might not either...}
+  	Human clusters in blue change phylogenetic positions between the trees wholesale\lmc{wholesale??}, with minor incongruences within clusters.
+  	This is evidence for recombinant viruses generated in the reservoir\lmc{(camels)} entering human populations.
     }%
@@
 Our attempt at controlling the confounding effects of different sampling regimes between humans and camels revealed a troubling pattern, however.
-We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions but do not necessarily converge on the same distributions.
-Of the five runs we set up two chains converged to one stationary distribution, two to another and the last one to a third.
+We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions which are not necessarily the same.
+Out of the five runs we set up two chains converged to one stationary distribution, two to another and the last one to a third.
 Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
-This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have nudged MCMC towards one of the multiple stationary distributions.
+This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
 
-Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (Figure \ref{skygrid}).
+Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity (N$_{e}\tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
 However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
@@
 \subsection*{Role of population structure in phylodynamics}
-Population structure yields important insight into any biological system.
-Population structure often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
+\lmc{Detecting?} Population structure yields important insight into any biological system.
+It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
 Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
@@
 \subsection*{Sequence data}
-All MERS-CoV sequences were downloaded from GenBank.
+All MERS-CoV sequences were downloaded from GenBank.\lmc{accession numbers are given in Table X}.
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
@@
 Sequences were annotated with available collection dates and hosts, designated as camel or human.
+The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections.
 
 \subsection*{Structured coalescent analyses}
+
 MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
@@
 Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-All analyses were run across ten independent Markov chains.
+All analyses were run across ten independent Markov chains (MCMC runs).
 To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
@@
 We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
-This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
+This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.\lmc{already reported above.
+Why not remove?}
 
 \subsection*{Epidemiological analyses}
-We simulate final transmission chain sizes using equation \ref{clusterLikelihood} \citep{lloyd-smith_superspreading_2005,blumberg_inference_2013} until we reach an epidemic comprised of 2000 cases.
+
 As of 10 May 2017, the World Health Organization has been notified of 1952 cases of MERS-CoV.
-10 000 simulation were run for 121 uniformly spaced values of R$_{0}$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1.
-Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where K is the number of clusters generated.
+We thus simulate final transmission chain sizes using equation \ref{clusterLikelihood} \citep{lloyd-smith_superspreading_2005,blumberg_inference_2013} until we reach an epidemic comprised of 2000 cases.
+10 000 simulations were run for 121 uniformly spaced values of R$_{0}$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1.
+Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where $K$ is the number of clusters generated.
 We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing.
-The resulting sequence cluster size vector $S$ contains K entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
+The resulting sequence cluster size vector $S$ contains $K$ entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
 Since the sampling scheme operates under equi-probable sequencing we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
 This makes clusters likely to be `sequenced' even more likely to be sequenced and \textit{vice versa}.
-We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to simulate a dramatically underreported epidemic.
+We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
 
+Let $R_{0}$ be the basic reproductive number ($ 0 < R_{0} < 1$),  and $\omega > 0$ be the a dispersion parameter, then the probability of observing a stuttering chain of size (cluster) $r$ of size $j$ is~\citep{blumberg_inference_2013}
 \begin{equation}
-p(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}
+Pr(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}.
 \label{clusterLikelihood}
@@
 
-Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.
+Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.\lmc{other than the ``obvious'' limitation that they're \textbf{not} the actual infection process, uh?
+>.>}
 Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species transmission of the virus from camels into humans.
@@
 \subsubsection*{Introduction seasonality}
+
 We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
@@
 \end{align*}
+
+Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
 
@@
 \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_method.pdf}
+	\includegraphics[scale=0.25]{figures/mers_method.pdf}
 	\caption{\textbf{Monte Carlo simulation method.}
 Case clusters are simulated according to \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
-We sample 174 cases from each simulation to represent sequencing of human MERS cases.
+We sample 174 cases from each simulation to represent sequencing of human MERS cases.\lmc{is this implicitly a binomial sampling with $p = 1/2$?} 
 `Sequencing' is carried out by using multivariate hypergeometric sampling, where at each iteration case cluster size $C_{i}$ is the number of successes in the population, $\sum_{i=1}^{K} C_{i} - C_{i}$ is the number of failures and number of trials is equal to leftover sequencing capacity, that is $\sum_{i=1}^{K} S_{i}$ minus whatever has been `sequenced' up to that iteration.
@@
 Simulated sequence clusters are summarised by their mean and standard deviation, both of which reflect some aspect of the partitioning of sequences into clusters.
-A simulations is considered to match if its sequence cluster size mean and standard deviation fell within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
-$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are summarised in Figure \ref{ABC}.
+A simulation is considered to match if its sequence cluster size mean and standard deviation fell within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
+$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are kept as estimates.
+These estimates are summarised in Figure \ref{ABC}.
 	}
@@
 	\caption{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
-The plot is identical to Figure \ref{ABC}, but instead of 2000 cases, the currently reported number of MERS-CoV cases, simulations were run with 4000 cases.
+The plot is identical to Figure \ref{ABC}, but instead of 2000 cases,
+% the currently reported number of MERS-CoV cases,
+simulations were run with 4000 cases.
 With more unobserved cases the R$_{0}$ values matching observed MERS-CoV sequence clusters can only be smaller, with a corresponding increase in numbers of zoonotic transmissions.
-However, the numbers of simulations that match MERS-CoV data go down too.
+However, the numbers of simulations that match MERS-CoV data go down as well.
 	}
@@
 The cumulative probability of human clades present in either tree are tracked by plots on the right (fragment 2) and top (fragment 1).
-Most of the probability density is concentrated within human clades that are present in trees of both genomic fragment 1 and 2 (0.9701 and 0.9474 of all human clades across posteriors, respectively).
+Most of the probability mass is concentrated within human clades that are present in trees of both genomic fragment 1 and 2 (0.9701 and 0.9474 of all human clades across posteriors, respectively).
 	}

### 2017-07-08 mers-structure.tex
forgot an important scolding

@@
 This was used to great effect towards the end of the Ebola virus epidemic in West Africa, where the final transmission chains were tracked and confirmed as such with sequence data \citep{arias_rapid_2016}.
-Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows data from wildly different studies to be combined trivially.
+Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows data from wildly different studies to be combined trivially\lmc{TRIVIALLY???}.
 In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.

### 2017-07-09 mers-structure.tex
A few comments but just checking them in before someone else commits and I have to copy them across again

@@
 Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
-We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
+We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host. 
 \tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'.
@@
 \section*{Introduction}
-Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
-The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 2000  infections and over 700 deaths according to the World Health Organization (WHO) \citep{mersfactsheet}.
+%% Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
+% AR - a virus isn't an infection and not all infections in the Arabian Peninsula
+Middle East Respiratory Syndrome Coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans.
+% ?
+something like this ?
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
 Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
@@
 Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
-Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
+
+%Unlike discrete trait analyses (DTA, or ``mugration'' models) %% AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
+%that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
+Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) [cite here, Philippe's bat rabies and Sam's avian flu?, others?], population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
 By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.

### 2017-07-10 mers-structure.tex
Incorporate @maxbiostat changes/comments.

@@
 \urlstyle{same}
+
+\usepackage{longtable}
 
@@
 % symbols
-\newcommand{\chiSq}{\chi^{2}_{df}} %LM: ancient DNA? =P
-\newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
-\newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
-\newcommand{\dspr}{d_\mathrm{SPR}}
+% \newcommand{\chiSq}{\chi^{2}_{df}} %LM: ancient DNA? =P %GD: quite :)
+% \newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
+% \newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
+% \newcommand{\dspr}{d_\mathrm{SPR}}
 
@@
 
-\date{\today}
+% \date{\today}
 
@@
 Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
-Whilst contact with camels is often reported, patients may not recall contact with \lmc{other?} livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
+Whilst contact with camels is often reported, patients may not recall contact with %\lmc{other?}
+any livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
 Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
@@
 
-Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches \lmc{why, specifically?}.
+Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches %\lmc{why, specifically?}
+due to difficulties in linking cases into clusters in the absence of detailed information.
 The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
 These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
-Often only sequence data can \lmc{help} pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
+Often only sequence data can %\lmc{help}
+help pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
 Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
-A large portion of MERS-CoV sequences come from hospital outbreaks \lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
+A large portion of MERS-CoV sequences come from %\lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}
+outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
 Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.
 
-It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \lmc{how about a citation here?}.
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
+\citep{who_2017}.
 Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
@@
 %that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
-Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) [cite here, Philippe's bat rabies and Sam's avian flu?, others?], population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
+Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) %[cite here, Philippe's bat rabies and Sam's avian flu?, others?]
+\citep{faria_simultaneously_2013,global_consortium_for_h5n8_and_related_influenza_viruses_role_2016}, population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
 By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
@@
 Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
-Whilst we identify \lmc{estimate?} a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
+Whilst we %\lmc{estimate?}
+estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
 We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
-It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows \lmc{is ``allow'' the best word here?
-I understand you' rather not use ``biases'' or ``misleads'', though...} the model to consider humans as a potential alternative host.
+It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels %\lmc{is ``allow'' the best word here?
+I understand you' rather not use ``biases'' or ``misleads'', though...}
+the model cannot dismiss humans as a potential alternative host.
 The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
@@
 These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
-Our analyses recover these results despite sequence data heavily skewed towards non-uniformly \lmc{maybe ``non-uniformly'' is redundant here} sampled human cases and are robust to choice of prior.
+Our analyses recover these results despite sequence data heavily skewed towards non-uniformly %\lmc{maybe ``non-uniformly'' is redundant here} \gdc{disagree, the point I am trying to make is that the data are mostly human and that human data are further skewed in how they are sampled}
+sampled human cases and are robust to choice of prior.
 We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.
@@
 \subsection*{Signals of MERS-CoV seasonality}
-Unlike the most basic \lmc{simpler?} ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, structured population models possess the ability to estimate migration times.
+Unlike the most simpler %\lmc{simpler?}
+``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, structured population models possess the ability to estimate migration times.
 We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
@@
 The black dotted line indicates the joint probability density for introductions.
-Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
+%Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
 	}
@@
 for MERS-CoV, implying poor transmissibility.
-However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides expectations\lmc{insights?} of epidemic behavior.
+However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides %\lmc{insights?}
+insight into epidemic behavior.
 The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by a dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}).
@@
 Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
-Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%\lmc{again?} of simulations fit for bias levels of 2.0 and 3.0, respectively.
-This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate a higher marginal likelihood for bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).\lmc{I don't follow this argument...}
+Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%%\lmc{again?}
+(nearly 10 000 out of 363 000 for both) of simulations fit for bias levels of 2.0 and 3.0, respectively.
+This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate that a higher number of simulations generate MERS-CoV-like data at bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).%\lmc{I don't follow this argument...}
 Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
@@
 These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Our results suggest a large number of unobserved MERS index cases.\lmc{does the definition of ``index case'' allow for ``multiple''?}
+Our results suggest a large number of unobserved MERS index cases.%\lmc{does the definition of ``index case'' allow for ``multiple''?} \gdc{not in epi literature I don't think}
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
@@
 	\caption{\textbf{Results of Monte Carlo simulations.}
-Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias \lmc{make sequences from large clusters more likely to be picked} ).
+Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced%\lmc{make sequences from large clusters more likely to be picked}
+).
 Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
@@
 Although the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
-This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur\lmc{stupid biostatistician here: is it?}.
+This provides ideal conditions for co-infection with distinct genotypes, which is a pre-requisite for detectable RNA virus recombination to occur%\lmc{stupid biostatistician here: is it?} \gdc{fair comment actually, it's a pre-requisite for detectable recombination to occur usually}
+.
 
@@
   	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
-  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels.\lmc{I'm sure this is important, but I didn't get why.
+  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels, adding support to the idea that camels are often co-infected.%\lmc{I'm sure this is important, but I didn't get why.
 Referees might not either...}
-  	Human clusters in blue change phylogenetic positions between the trees wholesale\lmc{wholesale??}, with minor incongruences within clusters.
-  	This is evidence for recombinant viruses generated in the reservoir\lmc{(camels)} entering human populations.
+  	Human clusters in blue change phylogenetic positions between the trees %\lmc{wholesale??}
+    together, with minor incongruences within clusters.
+  	This is evidence for recombinant viruses generated in the camel reservoir%\lmc{(camels)}
+    entering human populations.
     }%
@@
 \subsection*{Role of population structure in phylodynamics}
-\lmc{Detecting?} Population structure yields important insight into any biological system.
+%\lmc{Detecting?}
+Understanding population structure yields important insight into any biological system.
 It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
@@
 This was used to great effect towards the end of the Ebola virus epidemic in West Africa, where the final transmission chains were tracked and confirmed as such with sequence data \citep{arias_rapid_2016}.
-Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows data from wildly different studies to be combined trivially\lmc{TRIVIALLY???}.
+Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows data from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}
+.
 In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.
@@
 \subsection*{Sequence data}
-All MERS-CoV sequences were downloaded from GenBank.\lmc{accession numbers are given in Table X}.
+All MERS-CoV sequences were downloaded from GenBank.%\lmc{accession numbers are given in Table X}.
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
@@
 Sequences were annotated with available collection dates and hosts, designated as camel or human.
-The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections.
+The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).
+
+\begin{longtable}{ | l | l | l | l | l | }
+
+  \caption{Strain names, accessions (where available), identified host and reported collection dates for MERS-CoV genomes used in this study.} \label{sequences} \\
+
+  \hline & strain & accession & host & collection date \\ \hline
+  \endfirsthead
+
+  \multicolumn{5}{l}%
+  {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
+
+  \hline & strain & accession & host & collection date \\ \hline
+  \endhead
+
+  \hline \multicolumn{5}{|r|}{{Continued on next page}} \\ \hline
+  \endfoot
+
+  \hline \hline
+  \endlastfoot
+
+  1 & KSA-378 & KJ713296 & camel & 2013-11 \\
+  2 & KSA-363 & KJ713298 & camel & 2013-11 \\
+  3 & KSA-503 & KJ713297 & camel & 2013-11 \\
+  4 & KSA-376 & KJ713299 & camel & 2013-11 \\
+  5 & KSA-505 & KJ713295 & camel & 2013-11 \\
+  6 & Jeddah-1 & KF917527 & camel & 2013-11-08 \\
+  7 & NRCE-HKU205 & KJ477102 & camel & 2013-11-15 \\
+  8 & KFU-HKU1 & KJ650297 & camel & 2013-11-30 \\
+  9 & KFU-HKU13 & KJ650295 & camel & 2013-12-30 \\
+  10 & Camel\_Egypt\_NRCE-HKU271 &  & camel & 2013-12-30 \\
+  11 & Camel\_Egypt\_NRCE-HKU270 &  & camel & 2013-12-30 \\
+  12 & KFU-HKU19Dam & KJ650296 & camel & 2013-12-30 \\
+  13 & Qatar\_2\_2014 & KJ650098 & camel & 2014-02-16 \\
+  14 & UAE/D469-14 & KU242424 & camel & 2014-03-04 \\
+  15 & UAE/D511-14 & KU242423 & camel & 2014-03-12 \\
+  16 & Jeddah/F13A/2014 & KT368824 & camel & 2014-05 \\
+  17 & UAE/D1164.10/2014 & KP719928 & camel & 2014-06 \\
+  18 & UAE/D1339.2/2014 & KP719931 & camel & 2014-06 \\
+  19 & UAE/D1164.11/2014 & KP719929 & camel & 2014-06 \\
+  20 & UAE/D1164.9/2014 & KP719927 & camel & 2014-06 \\
+  21 & UAE/D1209/2014 & KP719933 & camel & 2014-06 \\
+  22 & UAE/D1164.14/2014 & KP719930 & camel & 2014-06 \\
+  23 & UAE/D1243.12/2014 & KP719932 & camel & 2014-06 \\
+  24 & D1164.1/14 & KX108937 & camel & 2014-06-02 \\
+  25 & Riyadh/Ry23N/2014 & KT368825 & camel & 2014-07 \\
+  26 & Riyadh/Ry84N/2014 & KT368826 & camel & 2014-07 \\
+  27 & Jeddah/S93/2014 & KT368855 & camel & 2014-09 \\
+  28 & Jeddah/401/2014 & KT368827 & camel & 2014-09 \\
+  29 & Jeddah/S100/2014 & KT368853 & camel & 2014-09 \\
+  30 & Jeddah/S99/2014 & KT368857 & camel & 2014-09 \\
+  31 & Jeddah/S94/2014 & KT368856 & camel & 2014-09 \\
+  32 & Jeddah/S73/2014 & KT368854 & camel & 2014-09 \\
+  33 & Jeddah/O47b/2014 & KT368852 & camel & 2014-10 \\
+  34 & Jeddah/O23b/2014 & KT368849 & camel & 2014-10 \\
+  35 & Jeddah/O24/2014 & KT368850 & camel & 2014-10 \\
+  36 & Jeddah/O30/2014 & KT368851 & camel & 2014-10 \\
+  37 & Jeddah/N51/2014 & KT368846 & camel & 2014-11 \\
+  38 & Jeddah/N68b/2014 & KT368848 & camel & 2014-11 \\
+  39 & Jeddah/N62b/2014 & KT368847 & camel & 2014-11 \\
+  40 & Jeddah/D40/2014 & KT368834 & camel & 2014-12 \\
+  41 & Jeddah/D90/2014 & KT368844 & camel & 2014-12 \\
+  42 & Jeddah/D88/2014 & KT368843 & camel & 2014-12 \\
+  43 & Jeddah/D36/2014 & KT368832 & camel & 2014-12 \\
+  44 & Jeddah/D35/2014 & KT368831 & camel & 2014-12 \\
+  45 & Jeddah/D92/2014 & KT368845 & camel & 2014-12 \\
+  46 & Jeddah/D49/2014 & KT368841 & camel & 2014-12 \\
+  47 & Jeddah/D34/2014 & KT368830 & camel & 2014-12 \\
+  48 & Jeddah/D33b/2014 & KT368829 & camel & 2014-12 \\
+  49 & Jeddah/D42/2014 & KT368835 & camel & 2014-12 \\
+  50 & Jeddah/D50b/2014 & KT368842 & camel & 2014-12 \\
+  51 & Jeddah/D45/2014 & KT368837 & camel & 2014-12 \\
+  52 & Jeddah/D46b/2014 & KT368838 & camel & 2014-12 \\
+  53 & Jeddah/D43b/2014 & KT368836 & camel & 2014-12 \\
+  54 & Jeddah/D100/2014 & KT368828 & camel & 2014-12 \\
+  55 & Jeddah/D47/2014 & KT368839 & camel & 2014-12 \\
+  56 & Jeddah/D38b/2014 & KT368833 & camel & 2014-12 \\
+  57 & Jeddah/D48/2014 & KT368840 & camel & 2014-12 \\
+  58 & D2597.2/14 & KX108938 & camel & 2014-12-13 \\
+  59 & Egypt\_NRCE-NC163/2014 & KU740200 & camel & 2014-12-17 \\
+  60 & Jeddah/Jd7/2015 & KT368861 & camel & 2015-01 \\
+  61 & Jeddah/Jd86/2015 & KT368863 & camel & 2015-01 \\
+  62 & Jeddah/Jd90/2015 & KT368865 & camel & 2015-01 \\
+  63 & Jeddah/Jd1b/2015 & KT368858 & camel & 2015-01 \\
+  64 & Jeddah/Jd4/2015 & KT368859 & camel & 2015-01 \\
+  65 & Jeddah/Jd85/2015 & KT368862 & camel & 2015-01 \\
+  66 & Jeddah/Jd6b/2015 & KT368860 & camel & 2015-01 \\
+  67 & Jeddah/Jd87/2015 & KT368864 & camel & 2015-01 \\
+  68 & D252/15 & KX108939 & camel & 2015-01-30 \\
+  69 & Jeddah/Jd199/2015 & KT368867 & camel & 2015-02 \\
+  70 & Jeddah/Jd175/2015 & KT368866 & camel & 2015-02 \\
+  71 & D374/15 & KX108940 & camel & 2015-02-12 \\
+  72 & D383/15 & KX108941 & camel & 2015-02-14 \\
+  73 & D389/15 & KX108942 & camel & 2015-02-15 \\
+  74 & Riyadh/Ry63/2015 & KT368876 & camel & 2015-03 \\
+  75 & Riyadh/Ry136/2015 & KT368868 & camel & 2015-03 \\
+  76 & Riyadh/Ry178/2015 & KT368874 & camel & 2015-03 \\
+  77 & Riyadh/Ry162/2015 & KT368871 & camel & 2015-03 \\
+  78 & Riyadh/Ry86/2015 & KT368879 & camel & 2015-03 \\
+  79 & Taif/T150/2015 & KT368889 & camel & 2015-03 \\
+  80 & Riyadh/Ry137/2015 & KT368869 & camel & 2015-03 \\
+  81 & Riyadh/Ry179/2015 & KT368875 & camel & 2015-03 \\
+  82 & Riyadh/Ry177/2015 & KT368873 & camel & 2015-03 \\
+  83 & Riyadh/Ry79/2015 & KT368878 & camel & 2015-03 \\
+  84 & Riyadh/Ry173/2015 & KT368872 & camel & 2015-03 \\
+  85 & Taif/T157b/2015 & KT368890 & camel & 2015-03 \\
+  86 & Riyadh/Ry159b/2015 & KT368870 & camel & 2015-03 \\
+  87 & Riyadh/Ry64/2015 & KT368877 & camel & 2015-03 \\
+  88 & Taif/T3/2015 & KT368880 & camel & 2015-04 \\
+  89 & Taif/T16/2015 & KT368882 & camel & 2015-04 \\
+  90 & Taif/T22/2015 & KT368883 & camel & 2015-04 \\
+  91 & Taif/T92/2015 & KT368887 & camel & 2015-04 \\
+  92 & Taif/T7/2015 & KT368881 & camel & 2015-04 \\
+  93 & Taif/T91b/2015 & KT368886 & camel & 2015-04 \\
+  94 & Taif/T68/2015 & KT368884 & camel & 2015-04 \\
+  95 & Taif/T89/2015 & KT368885 & camel & 2015-04 \\
+  96 & Taif/T98/2015 & KT368888 & camel & 2015-04 \\
+  97 & D998/15 & KX108943 & camel & 2015-04-23 \\
+  98 & D1157/15 & KX108944 & camel & 2015-05-12 \\
+  99 & D1189.1/15 & KX108946 & camel & 2015-05-18 \\
+  100 & D1271/15 & KX108945 & camel & 2015-05-29 \\
+  101 & Jordan-N3/2012 & KC776174 & human & 2012-04-15 \\
+  102 & EMC/2012 & JX869059 & human & 2012-06-13 \\
+  103 & England/1/2012 & KC164505 & human & 2012-09-11 \\
+  104 & Riyadh\_1\_2012 & KF600612 & human & 2012-10-23 \\
+  105 & Riyadh\_2\_2012 & KF600652 & human & 2012-10-30 \\
+  106 & Riyadh\_3\_2013 & KF600613 & human & 2013-02-05 \\
+  107 & England/3/2013 & KM210278 & human & 2013-02-10 \\
+  108 & England/2/2013 & KM015348 & human & 2013-02-10 \\
+  109 & England/4/2013 & KM210277 & human & 2013-02-13 \\
+  110 & Riyadh\_4\_2013 & KJ156952 & human & 2013-03-01 \\
+  111 & Munich/AbuDhabi/2013 & KF192507 & human & 2013-03-22 \\
+  112 & Al-Hasa\_2\_2013 & KF186566 & human & 2013-04-21 \\
+  113 & Al-Hasa\_3\_2013 & KF186565 & human & 2013-04-22 \\
+  114 & UAE-FRA1\_1627-2013\_BAL & KJ361500 & human & 2013-04-26 \\
+  115 & Al-Hasa\_4\_2013 & KF186564 & human & 2013-05-01 \\
+  116 & Al-Hasa\_7\_2013 & KF600623 & human & 2013-05-01 \\
+  117 & Al-Hasa\_8\_2013 & KF600618 & human & 2013-05-01 \\
+  118 & Al-Hasa\_25\_2013 & KJ156866 & human & 2013-05-02 \\
+  119 & Al-Hasa\_11\_2013 & KF600629 & human & 2013-05-03 \\
+  120 & Al-Hasa\_12\_2013 & KF600627 & human & 2013-05-07 \\
+  121 & Al-Hasa\_14\_2013 & KF600615 & human & 2013-05-08 \\
+  122 & Al-Hasa\_1\_2013 & KF186567 & human & 2013-05-09 \\
+  123 & Al-Hasa\_15\_2013 & KF600645 & human & 2013-05-11 \\
+  124 & Al-Hasa\_16\_2013 & KF600644 & human & 2013-05-12 \\
+  125 & Buraidah\_1\_2013 & KF600630 & human & 2013-05-13 \\
+  126 & Al-Hasa\_23\_2013 & KJ156860 & human & 2013-05-13 \\
+  127 & Al-Hasa\_17\_2013 & KF600647 & human & 2013-05-15 \\
+  128 & Al-Hasa\_19\_2013 & KF600632 & human & 2013-05-23 \\
+  129 & Al-Hasa\_18\_2013 & KF600651 & human & 2013-05-23 \\
+  130 & Al-Hasa\_21\_2013 & KF600634 & human & 2013-05-30 \\
+  131 & Hafr-Al-Batin\_1\_2013 & KF600628 & human & 2013-06-04 \\
+  132 & Wadi-Ad-Dawasir\_1\_2013 & KJ156881 & human & 2013-06-12 \\
+  133 & Taif\_1\_2013 & KJ156949 & human & 2013-06-12 \\
+  134 & Taif\_2\_2013 & KJ156896 & human & 2013-06-12 \\
+  135 & Taif\_3\_2013 & KJ156938 & human & 2013-06-13 \\
+  136 & Al-Hasa\_26\_2013 & KJ156882 & human & 2013-06-18 \\
+  137 & Al-Hasa\_27\_2013 & KJ156943 & human & 2013-06-19 \\
+  138 & Al-Hasa\_28\_2013 & KJ156887 & human & 2013-06-22 \\
+  139 & Riyadh\_6\_2013 & KJ156879 & human & 2013-07-02 \\
+  140 & Riyadh\_5\_2013 & KJ156944 & human & 2013-07-02 \\
+  141 & Riyadh\_7\_2013 & KJ156937 & human & 2013-07-15 \\
+  142 & Riyadh\_8\_2013 & KJ156880 & human & 2013-07-17 \\
+  143 & Riyadh\_9\_2013 & KJ156869 & human & 2013-07-17 \\
+  144 & Hafr-Al-Batin\_2\_2013 & KJ156910 & human & 2013-08-05 \\
+  145 & Asir\_2\_2013 & KJ156863 & human & 2013-08-05 \\
+  146 & Riyadh\_11\_2013 & KJ156946 & human & 2013-08-06 \\
+  147 & Riyadh\_12\_2013 & KJ156926 & human & 2013-08-08 \\
+  148 & Riyadh\_13\_2013 & KJ156888 & human & 2013-08-13 \\
+  149 & Riyadh\_14\_2013 & KJ156934 & human & 2013-08-15 \\
+  150 & Hafr-Al-Batin\_4\_2013 & KJ156931 & human & 2013-08-25 \\
+  151 & Hafr-Al-Batin\_5\_2013 & KJ156951 & human & 2013-08-25 \\
+  152 & Riyadh\_17\_2013 & KJ156918 & human & 2013-08-26 \\
+  153 & Hafr-Al-Batin\_6\_2013 & KJ156874 & human & 2013-08-28 \\
+  154 & Riyadh\_10\_2013 & KJ156891 & human & 2013-09-05 \\
+  155 & Madinah\_3b\_2013 & KJ156916 & human & 2013-09-11 \\
+  156 & Qatar3 & KF961221 & human & 2013-10-13 \\
+  157 & Qatar4 & KF961222 & human & 2013-10-17 \\
+  158 & Oman\_2285\_2013 & KT156560 & human & 2013-10-28 \\
+  159 & Jeddah-1 & KF958702 & human & 2013-11-05 \\
+  160 & AbuDhabi\_UAE\_9\_2013 & KP209312 & human & 2013-11-15 \\
+  161 & Oman\_2874\_2013 & KT156561 & human & 2013-12-28 \\
+  162 & AbuDhabi/Gayathi\_UAE\_2\_2014 & KP209310 & human & 2014-03-07 \\
+  163 & Jeddah\_C7569/KSA & KM027256 & human & 2014-04-03 \\
+  164 & Jeddah\_C7149/KSA & KM027255 & human & 2014-04-05 \\
+  165 & Jeddah\_C7770/KSA & KM027257 & human & 2014-04-07 \\
+  166 & AbuDhabi\_UAE\_8\_2014 & KP209306 & human & 2014-04-07 \\
+  167 & AbuDhabi\_UAE\_16\_2014 & KP209308 & human & 2014-04-10 \\
+  168 & AbuDhabi\_UAE\_18\_2014 & KP209307 & human & 2014-04-10 \\
+  169 & Jeddah\_C8826/KSA & KM027258 & human & 2014-04-12 \\
+  170 & AbuDhabi\_UAE\_26\_2014 & KP209313 & human & 2014-04-13 \\
+  171 & Jeddah\_C9055/KSA & KM027259 & human & 2014-04-14 \\
+  172 & Makkah\_C9355/KSA/Makkah & KM027261 & human & 2014-04-15 \\
+  173 & AbuDhabi\_UAE\_33\_2014 & KP209311 & human & 2014-04-17 \\
+  174 & AbuDhabi\_UAE\_30\_2014 & KP209309 & human & 2014-04-19 \\
+  175 & Jeddah\_C10306/KSA & KM027260 & human & 2014-04-21 \\
+  176 & Riyadh\_2014KSA\_683/KSA/2014 & KM027262 & human & 2014-04-22 \\
+  177 & Riyadh-KKUH-90b &  & human & 2014-04-24 \\
+  178 & Riyadh-KKUH-105 &  & human & 2014-04-25 \\
+  179 & Riyadh-KKUH-104 &  & human & 2014-04-25 \\
+  180 & KFMC-1 & KT121580 & human & 2014-04-28 \\
+  181 & KFMC-8 & KT121579 & human & 2014-04-30 \\
+  182 & Indiana/USA-1\_SaudiArabia\_2014 & KJ813439 & human & 2014-04-30 \\
+  183 & KFMC-10 & KT121578 & human & 2014-05-01 \\
+  184 & KFMC-7 & KT121581 & human & 2014-05-03 \\
+  185 & Riyadh-KKUH-291 &  & human & 2014-05-06 \\
+  186 & KFMC-9 & KT121574 & human & 2014-05-07 \\
+  187 & KFMC-3 & KT121573 & human & 2014-05-09 \\
+  188 & Florida/USA-2\_SaudiArabia\_2014 & KJ829365 & human & 2014-05-10 \\
+  189 & KFMC-2 & KT121577 & human & 2014-05-11 \\
+  190 & KFMC-4 & KT121575 & human & 2014-05-12 \\
+  191 & KFMC-5 & KT121572 & human & 2014-05-12 \\
+  192 & Riyadh-KKUH-368 &  & human & 2014-05-13 \\
+  193 & KFMC-6 & KT121576 & human & 2014-05-18 \\
+  194 & Riyadh\_2014KSA\_158/KSA/2014 & KM027281 & human & 2014-05-20 \\
+  195 & Jeddah-KFH-285TA &  & human & 2014-06-03 \\
+  196 & Jeddah-KFH-605TD &  & human & 2014-06-09 \\
+  197 & Jeddah-KFH-668TD &  & human & 2014-06-09 \\
+  198 & Jeddah-KFH-899NF &  & human & 2014-06-16 \\
+  199 & Jeddah-KFH-949NSG1 &  & human & 2014-06-18 \\
+  200 & Riyadh-KKUH-643 &  & human & 2014-11-02 \\
+  201 & Taif/KSA-7032/2014 & KU710264 & human & 2014-11-04 \\
+  202 & Riyadh-KKUH-665 &  & human & 2014-11-19 \\
+  203 & Riyadh-KSA-2049/2015 & KR011266 & human & 2015-01-06 \\
+  204 & Riyadh-KSA-2343/2015 & KR011264 & human & 2015-01-21 \\
+  205 & Riyadh-KSA-2345/2015 & KR011263 & human & 2015-01-21 \\
+  206 & Riyadh-KSA-2466/2015 & KR011265 & human & 2015-01-26 \\
+  207 & Kharj-KSA-2599/2015 & KT806052 & human & 2015-02-02 \\
+  208 & Kharj-KSA-2598/2015 & KT806053 & human & 2015-02-02 \\
+  209 & Riyadh-KSA-2716/2015 & KT806051 & human & 2015-02-05 \\
+  210 & Khobar-KSA-6736/2015 & KT806048 & human & 2015-02-07 \\
+  211 & Jeddah-KSA-C20843/2015 & KT806044 & human & 2015-02-09 \\
+  212 & Jeddah-KSA-C20860/2015 & KT806055 & human & 2015-02-10 \\
+  213 & Riyadh\_KSA\_2959\_2015 & KT026453 & human & 2015-02-10 \\
+  214 & Riyadh-KSA-3065/2015 & KT806050 & human & 2015-02-12 \\
+  215 & Najran-KSA-C20915/2015 & KT806054 & human & 2015-02-13 \\
+  216 & Riyadh-KSA-3181/2015 & KT806049 & human & 2015-02-15 \\
+  217 & Riyadh\_KKUH\_0734 &  & human & 2015-02-18 \\
+  218 & Jeddah-KSA-C21271/2015 & KT806045 & human & 2015-02-22 \\
+  219 & Riyadh\_KKUH\_0755 &  & human & 2015-02-23 \\
+  220 & Riyadh\_KKUH\_0756 &  & human & 2015-02-23 \\
+  221 & Riyadh\_KKUH\_0780 &  & human & 2015-02-25 \\
+  222 & Riyadh\_KKUH\_0801 &  & human & 2015-02-27 \\
+  223 & Riyadh\_KKUH\_0826 &  & human & 2015-02-28 \\
+  224 & Riyadh\_KKUH\_0818 &  & human & 2015-02-28 \\
+  225 & Riyadh\_KSA\_4050\_2015 & KT026454 & human & 2015-03-01 \\
+  226 & Riyadh\_KKUH\_0944 &  & human & 2015-03-02 \\
+  227 & Riyadh\_KKUH\_0939 &  & human & 2015-03-02 \\
+  228 & Riyadh\_KKUH\_1080 &  & human & 2015-03-03 \\
+  229 & Riyadh\_KKUH\_1066 &  & human & 2015-03-03 \\
+  230 & Riyadh\_KKUH\_1145 &  & human & 2015-03-04 \\
+  231 & Riyadh\_KKUH\_1217 &  & human & 2015-03-04 \\
+  232 & Riyadh\_KKUH\_1461 &  & human & 2015-03-08 \\
+  233 & Riyadh\_KKUH\_1470 &  & human & 2015-03-08 \\
+  234 & Riyadh\_KKUH\_1522 &  & human & 2015-03-09 \\
+  235 & Germany3/UAE-Dubai/Abu-Dhabi &  & human & 2015-03-11 \\
+  236 & Hufuf-KSA-9158/2015 & KT806047 & human & 2015-03-27 \\
+  237 & Hufuf-KSA-11002/2015 & KT806046 & human & 2015-05-10 \\
+  238 & KOR/KNIH/002\_05\_2015 & KT029139 & human & 2015-05-20 \\
+  239 & ChinaGD01 & KT036372 & human & 2015-05-28 \\
+  240 & KOR/Seoul/014-2015 & KX034093 & human & 2015-05-30 \\
+  241 & KOREA/Seoul/014-1-2015 & KT374052 & human & 2015-05-31 \\
+  242 & KOREA/Seoul/035-1-2015 & KT374054 & human & 2015-06-03 \\
+  243 & KOR/Seoul/066-2015 & KX034095 & human & 2015-06-04 \\
+  244 & Korea/Seoul/SNU1-035/2015 & KU308549 & human & 2015-06-08 \\
+  245 & KOR/CNUH\_SNU/030\_06\_2015 & KT868868 & human & 2015-06-08 \\
+  246 & KOR/CNUH\_SNU/024\_06\_2015 & KT868867 & human & 2015-06-08 \\
+  247 & KOR/CNUH\_SNU/054\_06\_2015 & KT868871 & human & 2015-06-09 \\
+  248 & KOR/CNUH\_SNU/038\_06\_2015 & KT868870 & human & 2015-06-10 \\
+  249 & KOR/CNUH\_SNU/148\_06\_2015 & KT868876 & human & 2015-06-10 \\
+  250 & KOR/CNUH\_SNU/122\_06\_2015 & KT868875 & human & 2015-06-10 \\
+  251 & KOR/CNUH\_SNU/082\_06\_2015 & KT868872 & human & 2015-06-10 \\
+  252 & KOR/CNUH\_SNU/085\_06\_2015 & KT868873 & human & 2015-06-10 \\
+  253 & KOR/CNUH\_SNU/016\_06\_2015 & KT868865 & human & 2015-06-11 \\
+  254 & KOR/CNUH\_SNU/023\_06\_2015 & KT868866 & human & 2015-06-11 \\
+  255 & KOR/CNUH\_SNU/031\_06\_2015 & KT868869 & human & 2015-06-11 \\
+  256 & KOR/CNUH\_SNU/110\_06\_2015 & KT868874 & human & 2015-06-11 \\
+  257 & KOR/Seoul/050-1-2015 & KX034094 & human & 2015-06-11 \\
+  258 & THA/CU/17\_06\_2015 & KT225476 & human & 2015-06-17 \\
+  259 & KOR/Seoul/077-2-2015 & KX034096 & human & 2015-06-17 \\
+  260 & KOR/Seoul/080-3-2015 & KX034097 & human & 2015-06-17 \\
+  261 & KOREA/Seoul/163-1-2015 & KT374051 & human & 2015-06-19 \\
+  262 & KOREA/Seoul/168-1-2015 & KT374056 & human & 2015-06-21 \\
+  263 & KOR/Seoul/162-1-2015 & KX034098 & human & 2015-06-22 \\
+  264 & KOR/CNUH\_SNU/172\_06\_2015 & KT868877 & human & 2015-06-22 \\
+  265 & KOR/Seoul/169-2015 & KX034099 & human & 2015-06-26 \\
+  266 & KOR/Seoul/177-3-2015 & KX034100 & human & 2015-07-03 \\
+  267 & Jeddah-KSA-3RS2702/2015 & KU851859 & human & 2015-07-12 \\
+  268 & Riyadh-KSA-16120/2015 & KU851861 & human & 2015-08-24 \\
+  269 & Riyadh-KSA-16117/2015 & KU851862 & human & 2015-08-24 \\
+  270 & Riyadh-KSA-16121/2015 & KU851860 & human & 2015-08-24 \\
+  271 & Riyadh-KSA-16098/2015 & KU851864 & human & 2015-08-24 \\
+  272 & Riyadh-KSA-16077/2015 & KU851863 & human & 2015-08-27 \\
+  273 & Jordan\_1\_2015 & KU233363 & human & 2015-09-17 \\
+  274 & Jordan\_10\_2015 & KU233362 & human & 2015-09-17 \\
+\end{longtable}
 
@@
 We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
-This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.\lmc{already reported above.
-Why not remove?}
+This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above.
+Why not remove?} \gdc{Never hurts to repeat.}
 
@@
 
-Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.\lmc{other than the ``obvious'' limitation that they're \textbf{not} the actual infection process, uh?
->.>}
+Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.%\lmc{other than the ``obvious'' limitation that they're \textbf{not} the actual infection process, uh?
+>.>} \gdc{fair point, but we're talking about cluster inference}
 Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species transmission of the virus from camels into humans.

### 2017-07-10 mers-structure.tex
Some additional comments.

@@
 We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
-\tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'.
+%\tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'.
 I think you want a different word here.
 Maybe `transient host'?}
-\gdc{Good point about the term. Hope the definition leans towards `one not involved in maintenance' rather than `limited to index case'. Asked a friend working on zoonoses about this.}
+%\gdc{Good point about the term. Hope the definition leans towards `one not involved in maintenance' rather than `limited to index case'. Asked a friend working on zoonoses about this.}
 By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
@@
 something like this ?
-The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{mersfactsheet}.
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{who_2017}.
 Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
@@
 It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
-\citep{who_2017}.
+\citep{who_2016_MERS}.
 Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
@@
 
-%Unlike discrete trait analyses (DTA, or ``mugration'' models) %% AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
+%Unlike discrete trait analyses (DTA, or ``mugration'' models) % AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
 %that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
@@
 
+\subsection*{Structured coalescent analyses}
+
+MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
+Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
+A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
+Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
+All analyses were run across ten independent Markov chains (MCMC runs).
+To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
+
+We also analysed a dataset where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
+All analyses were set to run for 200 million states, subsampling every 20 000 states.
+Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin for single fragment analyses and 20\% for two-fragment analyses.
+Chains not converging after 200 million states were discarded.
+For single fragment analysis with migration rate prior set to exponential with mean 1.0 there were three chains out of ten that did not converge and two for the analysis where the migration rate prior was an exponential with mean 10.0.
+For the two fragment structured coalescent analyses three chains out of ten failed to converge.
+
+\subsection*{Demographic inference of MERS-CoV in the reservoir}
+In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
+The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
+These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
+Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
+A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
+We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
+This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above.
+Why not remove?} \gdc{Never hurts to repeat.}
+
+\subsection*{Epidemiological analyses}
+
+As of 10 May 2017, the World Health Organization has been notified of 1952 cases of MERS-CoV.
+We thus simulate final transmission chain sizes using equation \ref{clusterLikelihood} \citep{lloyd-smith_superspreading_2005,blumberg_inference_2013} until we reach an epidemic comprised of 2000 cases.
+10 000 simulations were run for 121 uniformly spaced values of R$_{0}$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1.
+Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where $K$ is the number of clusters generated.
+We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing.
+The resulting sequence cluster size vector $S$ contains $K$ entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
+Since the sampling scheme operates under equi-probable sequencing we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
+This makes clusters likely to be `sequenced' even more likely to be sequenced and \textit{vice versa}.
+We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
+
+Let $R_{0}$ be the basic reproductive number ($ 0 < R_{0} < 1$),  and $\omega > 0$ be the a dispersion parameter, then the probability of observing a stuttering chain of size (cluster) $r$ of size $j$ is~\citep{blumberg_inference_2013}
+\begin{equation}
+Pr(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}.
+\label{clusterLikelihood}
+\end{equation}
+
+Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.%\lmc{other than the ``obvious'' limitation that they're \textbf{not} the actual infection process, uh?
+>.>} \gdc{fair point, but we're talking about cluster inference}
+Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species transmission of the virus from camels into humans.
+The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying distribution of case clusters.
+We employ a Monte Carlo simulation approach to identify simulations where the recovered distribution of sequence cluster sizes fall within the 95\% highest posterior density intervals for three summary statistics of MERS-CoV sequence cluster sizes recovered via structured coalescent analyses: mean, median and standard deviation.
+These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
+
+\subsubsection*{Introduction seasonality}
+
+We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
+This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
+We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
+\begin{align*}
+  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
+  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
+  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
+  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
+  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
+  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
+  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
+\end{align*}
+
+Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
+
+\subsection*{Data availability}
+Sequence data and all analytical code is publicly available at https://github.com/blab/structured-mers.
+
+\section*{Acknowledgements}
+We would like to thank Allison Black for useful discussion and advice.
+GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
+
+\bibliographystyle{mbe.bst}
+\bibliography{mers-structure}
+
+\newpage
+
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
 \begin{longtable}{ | l | l | l | l | l | }
@@
 
-\subsection*{Structured coalescent analyses}
-
-MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
-Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
-A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
-Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-All analyses were run across ten independent Markov chains (MCMC runs).
-To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
-
-We also analysed a dataset where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
-Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
-All analyses were set to run for 200 million states, subsampling every 20 000 states.
-Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin for single fragment analyses and 20\% for two-fragment analyses.
-Chains not converging after 200 million states were discarded.
-For single fragment analysis with migration rate prior set to exponential with mean 1.0 there were three chains out of ten that did not converge and two for the analysis where the migration rate prior was an exponential with mean 10.0.
-For the two fragment structured coalescent analyses three chains out of ten failed to converge.
-
-\subsection*{Demographic inference of MERS-CoV in the reservoir}
-In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
-The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
-These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
-Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
-A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
-We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
-This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above.
-Why not remove?} \gdc{Never hurts to repeat.}
-
-\subsection*{Epidemiological analyses}
-
-As of 10 May 2017, the World Health Organization has been notified of 1952 cases of MERS-CoV.
-We thus simulate final transmission chain sizes using equation \ref{clusterLikelihood} \citep{lloyd-smith_superspreading_2005,blumberg_inference_2013} until we reach an epidemic comprised of 2000 cases.
-10 000 simulations were run for 121 uniformly spaced values of R$_{0}$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1.
-Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where $K$ is the number of clusters generated.
-We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing.
-The resulting sequence cluster size vector $S$ contains $K$ entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
-Since the sampling scheme operates under equi-probable sequencing we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
-This makes clusters likely to be `sequenced' even more likely to be sequenced and \textit{vice versa}.
-We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
-
-Let $R_{0}$ be the basic reproductive number ($ 0 < R_{0} < 1$),  and $\omega > 0$ be the a dispersion parameter, then the probability of observing a stuttering chain of size (cluster) $r$ of size $j$ is~\citep{blumberg_inference_2013}
-\begin{equation}
-Pr(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}.
-\label{clusterLikelihood}
-\end{equation}
-
-Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.%\lmc{other than the ``obvious'' limitation that they're \textbf{not} the actual infection process, uh?
->.>} \gdc{fair point, but we're talking about cluster inference}
-Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species transmission of the virus from camels into humans.
-The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying distribution of case clusters.
-We employ a Monte Carlo simulation approach to identify simulations where the recovered distribution of sequence cluster sizes fall within the 95\% highest posterior density intervals for three summary statistics of MERS-CoV sequence cluster sizes recovered via structured coalescent analyses: mean, median and standard deviation.
-These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
-
-\subsubsection*{Introduction seasonality}
-
-We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
-This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
-We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
-\begin{align*}
-  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
-  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
-  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
-  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
-  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
-  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
-  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
-\end{align*}
-
-Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
-
-\subsection*{Data availability}
-Sequence data and all analytical code is publicly available at https://github.com/blab/structured-mers.
-
-\section*{Acknowledgements}
-We would like to thank Allison Black for useful discussion and advice.
-GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
-
-\bibliographystyle{mbe.bst}
-\bibliography{mers-structure}
-
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-
 \begin{figure}[h]
@@
 Case clusters are simulated according to \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
-We sample 174 cases from each simulation to represent sequencing of human MERS cases.\lmc{is this implicitly a binomial sampling with $p = 1/2$?}
+We sample 174 cases from each simulation to represent sequencing of human MERS cases.%\lmc{is this implicitly a binomial sampling with $p = 1/2$?} \gdc{hypergeometric, since we probably don't resequence people}
 `Sequencing' is carried out by using multivariate hypergeometric sampling, where at each iteration case cluster size $C_{i}$ is the number of successes in the population, $\sum_{i=1}^{K} C_{i} - C_{i}$ is the number of failures and number of trials is equal to leftover sequencing capacity, that is $\sum_{i=1}^{K} S_{i}$ minus whatever has been `sequenced' up to that iteration.

### 2017-07-10 mers-structure.tex
fixing typo

@@
 
-Let $R_{0}$ be the basic reproductive number ($ 0 < R_{0} < 1$),  and $\omega > 0$ be the a dispersion parameter, then the probability of observing a stuttering chain of size (cluster) $r$ of size $j$ is~\citep{blumberg_inference_2013}
+Let $R_{0}$ be the basic reproductive number ($ 0 < R_{0} < 1$),  and $\omega > 0$ be the a dispersion parameter, then the probability of observing a stuttering chain (cluster) $r$ of size $j$ is~\citep{blumberg_inference_2013}
 \begin{equation}

### 2017-07-11 mers-structure.tex
Addressing more of @maxbiostat comments.

@@
 due to difficulties in linking cases into clusters in the absence of detailed information.
-The new era of genomic epidemiology, however, has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
+Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{gire_genomic_2014,quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
 These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
-Often only sequence data can %\lmc{help}
+In the absence of other information often only sequence data can %\lmc{help}
 help pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
@@
 \citep{who_2016_MERS}.
-Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
+Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
 Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
@@
 sampled human cases and are robust to choice of prior.
-We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.
+We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
 
@@
 Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
-This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from index cases will be sampled exceedingly rarely.
+This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
 We record the $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses (left-hand panels in Figure \ref{ABC}).
@@
 These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Our results suggest a large number of unobserved MERS index cases.%\lmc{does the definition of ``index case'' allow for ``multiple''?} \gdc{not in epi literature I don't think}
+Our results suggest a large number of unobserved MERS primary cases.%\lmc{does the definition of ``index case'' allow for ``multiple''?} \gdc{not in epi literature I don't think}
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
@@
 \subsection*{Sequence data as a universal epidemiology tool}
-As the field of epidemiology enters the age of genomics sequence data are likely to be brought to the forefront of epidemiological investigations.
+As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
 At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
 With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
-This was used to great effect towards the end of the Ebola virus epidemic in West Africa, where the final transmission chains were tracked and confirmed as such with sequence data \citep{arias_rapid_2016}.
-Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows data from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}
+This was used to great effect at all stages of the Ebola virus epidemic in West Africa, where sequence data early on were used to show continued circulation of the same strain in the region rather than multiple zoonotic introductions \citep{gire_genomic_2014}, to study viral migration at a broad level \citep{park_ebola_2015,ladner_evolution_2015} as healthcare infrastructure was being overwhelmed during the peak, and towards the end where the final transmission chains were tracked and confirmed as such \citep{arias_rapid_2016}.
+Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}
 .

### 2017-07-12 mers-structure.tex
Add algorithm float.

@@
 \usepackage{longtable}
+\usepackage[]{algorithm2e}
 
@@
 Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where $K$ is the number of clusters generated.
-We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing.
+We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing (algorithm \ref{hypergeometric}).
 The resulting sequence cluster size vector $S$ contains $K$ entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
@@
 These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
+
+\begin{algorithm}[H]
+ \KwData{Array of case cluster sizes in outbreak $[C_{i},C_{i+1},C_{i+2}...C_{K}]$, sequences available $S$, total outbreak size $\sum_{i=1}^{K} C_{i}$}
+ \KwResult{Array of sequence cluster sizes sampled: $[S_{i},S_{i+1},S_{i+2}...S_{K}]$}
+ Draw from $C_{i}$ according to a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $S$ trials\;
+ \While{$i < K$}{
+  $i = i + 1$\;
+  $S = S - S_{i-1}$\;
+  Recover probability mass function $p = [p(0), ..., p(C_{i})]$ for a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $S$ trials\;
+  (Optional) Introduce bias into $p_{bias} = \frac{[p(0)^{bias}, ..., p(C_{i})^{bias}]}{\sum p_{bias}}$\;
+  Draw a sequence cluster size $S_{i}$ from array of potential sequence cluster sizes $[0, ..., C_{i}]$ according to $p$\;
+ }
+ Add remaining sequences to last sequence cluster $C_{K} = S - S_{K-1}$\;
+ \caption{\textbf{Multivariate hypergeometric sampling scheme.}
+ Pseudocode describes the multivariate hypergeometric sampling scheme that simulates sequencing.
+ Probability of sequencing a given number of cases from a case cluster depends on cluster size and sequences left (\textit{i.e.} ``sequencing capacity'').
+ The bias parameter determines how probability mass function of the hypergeometric distribution is concentrated.
+ }
+ \label{hypergeometric}
+\end{algorithm}
 

### 2017-07-14 mers-structure.tex
mathsy changes to algorithm box

--- 2017-07-11 mers-structure.tex
@@
 These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
+
+\begin{algorithm}[H]
+ \KwData{Array of case cluster sizes in outbreak $\boldsymbol C = [C_{1},C_{2}, \ldots, C_{K}]$, sequences available $M$, total outbreak size $\sum_{i=1}^{K} C_{i}$.}
+ \KwResult{Array of sequence cluster sizes sampled: $\boldsymbol S = [S_{1},S_{2}, \ldots, S_{K}]$.}
+ Draw from $S_{i}$ from a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
+ \While{$i < K$}{
+  $i = i + 1$\;
+  $M = M - S_{i-1}$\;
+  Compute the probability mass function (pmf) for all possible values of $S_i$, $\boldsymbol p = [p(0)^{\text{bias}}, p(1), \ldots, p(C_{i})^{\text{bias}}] \times (\sum_i p_i^{\text{bias}}) ^{-1}$, where $p(\cdot)$  is the pmf for a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
+  Draw a sequence cluster size $S_{i}$ from array of potential sequence cluster sizes $[0, 1, \ldots, C_{i}]$ according to $\boldsymbol p$\;
+ }
+ Add remaining sequences to last sequence cluster $C_{K} = M - S_{K-1}$\;
+ \caption{\textbf{Multivariate hypergeometric sampling scheme.}
+ Pseudocode describes the multivariate hypergeometric sampling scheme that simulates sequencing.
+ Probability of sequencing a given number of cases from a case cluster depends on cluster size and sequences left (\textit{i.e.} ``sequencing capacity'').
+ The bias parameter determines how probability mass function of the hypergeometric distribution is concentrated.
+ }
+ \label{hypergeometric}
+\end{algorithm}
 

### 2017-07-14 mers-structure.tex
typo fix

@@
  \KwResult{Array of sequence cluster sizes sampled: $\boldsymbol S = [S_{1},S_{2}, \ldots, S_{K}]$.}
- Draw from $S_{i}$ from a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
+ Draw $S_{i}$ from a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
  \While{$i < K$}{

### 2017-07-14 mers-structure.tex
Minor fix to algorithm.

@@
   $M = M - S_{i-1}$\;
-  Compute the probability mass function (pmf) for all possible values of $S_i$, $\boldsymbol p = [p(0)^{\text{bias}}, p(1), \ldots, p(C_{i})^{\text{bias}}] \times (\sum_i p_i^{\text{bias}}) ^{-1}$, where $p(\cdot)$  is the pmf for a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
+  Compute the probability mass function (pmf) for all possible values of $S_i$, $\boldsymbol p = [p(0)^{\text{bias}}, p(1)^{\text{bias}}, \ldots, p(C_{i})^{\text{bias}}] \times (\sum_i p_i^{\text{bias}}) ^{-1}$, where $p(\cdot)$  is the pmf for a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
   Draw a sequence cluster size $S_{i}$ from array of potential sequence cluster sizes $[0, 1, \ldots, C_{i}]$ according to $\boldsymbol p$\;

### 2017-07-24 coverLetter.tex
Add cover letter.

\documentclass[stdletter,letterpaper,addrfromright,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
\topmarginskip{-0.4in}
\bottommarginskip{-1.5in}
\leftmarginsize{1in}
\rightmarginsize{1.25in}
\sigskipbefore{0.2in}
\sigskipafter{0in}
\noLines
\nolines
\noHeadline
\noheadline
\signature{Gytis Dudas}

\namefrom{}
\addrfrom{Vaccines and Infectious Disease Division \\ Fred Hutchinson Cancer Research Center \\ 1100 Fairview Ave N \\ Seattle \\ USA}
\emailfrom{gdudas@fredhutch.org}

\greetto{Dear Editor,}
\closeline{Sincerely,}

% comments
\usepackage{color}
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}}
\def\tbc#1{\textcolor{purple}{[#1]}}

\begin{document}

\begin{newlfm}
Please find, attached, our manuscript entitled ``Structured population models demystify MERS-CoV epidemiology''.
We would be grateful if you considered it for publication in \textit{eLife}.
Our manuscript uses phylogenetic models that explicitly model population structure applied to Middle East Respiratory Syndrome Coronavirus (MERS-CoV) in order to quantify and partition its evolution between two major known hosts -- humans and camels.
Since its discovery in 2012 the virus has caused over 2000 cases of MERS in humans, largely restricted to the Arabian Peninsula.
Although contact with camels, the presumed reservoir, is a known risk factor in MERS infection, the epidemiology of the virus has remained obscure.
Studies published to date have tried quantifying relative contributions of cross-species and human-to-human transmission in continuation of human MERS outbreaks, but relied heavily on either case data where case clustering is difficult to ascertain or sequence data sampled overwhelmingly from humans.
As a result these studies have reached markedly different conclusions, often at odds with established epidemiological facts.

In our manuscript we employ a structured coalescent model that quantifies patterns of diversity in MERS-CoV sequences (human or camel) by modeling host-specific evolution and cross-species transmission explicitly.
The dynamics of MERS-CoV we infer are characterised by long term viral evolution occurring exclusively in camels and human outbreaks driven largely by cross-species introductions, rather than extensive human-to-human (\textit{i.e.} community) transmission.
This explains why the continued trickle of MERS cases has been limited to the Arabian Peninsula with little evidence of widespread human exposure across the region.
Our findings point towards the central role of the camel reservoir and a minimal, but not negligible, contribution of community transmission to continuation of MERS outbreaks in the region.
We go a step further and use Monte Carlo simulations to integrate MERS case data with structured coalescent inference to quantify the basic reproductive number ($R_{0}$) of MERS-CoV in humans from inferred human sequence clusters.
Our main finding is that MERS-CoV is relatively poor at human-to-human transmission ($R_{0}<1.0$) and that correspondingly, there have been hundreds of zoonotic introductions of the virus into humans, mostly limited to primary cases, and that MERS-CoV sequencing efforts have been uneven owing to the tendency to sequence from large MERS outbreaks in hospitals.
Finally, by leveraging additional features of the structured coalescent we show that cross-species jumps into humans appear to be seasonal, which we attribute to seasonal camel breeding patterns that result in varied force of infection.

We believe our manuscript joins a number of studies bridging phylogenetic and epidemiological fields and shows how methods common to both can be utilised synergistically to understand the drivers of transmission dynamics in rapidly evolving pathogens.
We expect the manuscript to be of great interest to epidemiologists, healthcare professionals, decision makers and colleagues alike.
Our findings clearly summarise MERS-CoV epidemiology from sequence data alone and help contextualise previous studies, rather than add to the perceived mystery of MERS-CoV.
The approaches we employ serve to both highlight the power of population structure modeling, as well as pushing the field towards greater integration of case and sequence data.

\end{newlfm}

\end{document}

### 2017-07-26 coverLetter.tex
comma changes; a lil nagging

@@
 \begin{newlfm}
-Please find, attached, our manuscript entitled ``Structured population models demystify MERS-CoV epidemiology''.
+Please find attached our manuscript entitled ``Structured population models demystify MERS-CoV epidemiology''.
 We would be grateful if you considered it for publication in \textit{eLife}.
-Our manuscript uses phylogenetic models that explicitly model population structure applied to Middle East Respiratory Syndrome Coronavirus (MERS-CoV) in order to quantify and partition its evolution between two major known hosts -- humans and camels.
+Our study uses phylogenetic models that explicitly model population structure applied to Middle East Respiratory Syndrome Coronavirus (MERS-CoV) in order to quantify and partition its evolution between two major known hosts -- humans and camels.
+%LMC: this sentence is oddly located, because the ``solution'' comes before the ``problem'', so the reader has no idea why these fancy phylogenetic models are useful...
 Since its discovery in 2012 the virus has caused over 2000 cases of MERS in humans, largely restricted to the Arabian Peninsula.
@@
 The dynamics of MERS-CoV we infer are characterised by long term viral evolution occurring exclusively in camels and human outbreaks driven largely by cross-species introductions, rather than extensive human-to-human (\textit{i.e.} community) transmission.
-This explains why the continued trickle of MERS cases has been limited to the Arabian Peninsula with little evidence of widespread human exposure across the region.
+This explains why the continued trickle of MERS cases has been limited to the Arabian Peninsula, with little evidence of widespread human exposure across the region.
 Our findings point towards the central role of the camel reservoir and a minimal, but not negligible, contribution of community transmission to continuation of MERS outbreaks in the region.
@@
 
-We believe our manuscript joins a number of studies bridging phylogenetic and epidemiological fields and shows how methods common to both can be utilised synergistically to understand the drivers of transmission dynamics in rapidly evolving pathogens.
+We believe our manuscript joins a number of studies bridging phylogenetics and epidemiology and shows how methods common to both can be utilised synergistically to understand the drivers of transmission dynamics in rapidly evolving pathogens.
 We expect the manuscript to be of great interest to epidemiologists, healthcare professionals, decision makers and colleagues alike.

### 2017-07-26 cover-letter/cover-letter.tex
Revise cover letter

--- 2017-07-26 coverLetter.tex
@@
 \documentclass[stdletter,letterpaper,addrfromright,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
-\topmarginskip{-0.4in}
-\bottommarginskip{-1.5in}
-\leftmarginsize{1in}
-\rightmarginsize{1.25in}
+\topmarginskip{-0.0in}
+\bottommarginskip{-0.0in}
+\leftmarginsize{1.25in}
+\rightmarginsize{1.5in}
 \sigskipbefore{0.2in}
@@
 \namefrom{}
-\addrfrom{Vaccines and Infectious Disease Division \\ Fred Hutchinson Cancer Research Center \\ 1100 Fairview Ave N \\ Seattle \\ USA}
+\addrfrom{
+	\includegraphics[width=6.5cm]{figures/fhcrc_logo} \\
+  Vaccines and Infectious Disease Division \\
+  Fred Hutchinson Cancer Research Center \\
+  1100 Fairview Ave N \\
+  Seattle, WA 98109, USA}
 \emailfrom{gdudas@fredhutch.org}
@@
 %LMC: this sentence is oddly located, because the ``solution'' comes before the ``problem'', so the reader has no idea why these fancy phylogenetic models are useful...
+%TB: I'm okay with the broad sentence on "main point" before getting to the problem
 Since its discovery in 2012 the virus has caused over 2000 cases of MERS in humans, largely restricted to the Arabian Peninsula.
 Although contact with camels, the presumed reservoir, is a known risk factor in MERS infection, the epidemiology of the virus has remained obscure.
-Studies published to date have tried quantifying relative contributions of cross-species and human-to-human transmission in continuation of human MERS outbreaks, but relied heavily on either case data where case clustering is difficult to ascertain or sequence data sampled overwhelmingly from humans.
+Studies published to date have tried to quantify relative contributions of cross-species and human-to-human transmission to MERS-CoV persistence, but have relied heavily on either case data where case clustering is difficult to ascertain or sequence data sampled overwhelmingly from humans.
 As a result these studies have reached markedly different conclusions, often at odds with established epidemiological facts.
 
-In our manuscript we employ a structured coalescent model that quantifies patterns of diversity in MERS-CoV sequences (human or camel) by modeling host-specific evolution and cross-species transmission explicitly.
-The dynamics of MERS-CoV we infer are characterised by long term viral evolution occurring exclusively in camels and human outbreaks driven largely by cross-species introductions, rather than extensive human-to-human (\textit{i.e.} community) transmission.
+Here, we analyze 174 MERS-CoV genomes from human infections and 100 MERS-CoV genomes from camel infections.
+We employ a structured coalescent model that quantifies patterns of diversity in MERS-CoV sequences (human or camel) by modeling host-specific evolution and cross-species transmission explicitly.
+We infer that MERS-CoV is characterised by long-term viral evolution occurring exclusively in camels and human outbreaks driven largely by cross-species introductions, rather than extensive human-to-human (\textit{i.e.}\ community) transmission.
 This explains why the continued trickle of MERS cases has been limited to the Arabian Peninsula, with little evidence of widespread human exposure across the region.
-Our findings point towards the central role of the camel reservoir and a minimal, but not negligible, contribution of community transmission to continuation of MERS outbreaks in the region.
+Our findings point towards the central role of the camel reservoir and a minimal contribution of community transmission to continuation of MERS outbreaks in the region.
 We go a step further and use Monte Carlo simulations to integrate MERS case data with structured coalescent inference to quantify the basic reproductive number ($R_{0}$) of MERS-CoV in humans from inferred human sequence clusters.
@@
 We expect the manuscript to be of great interest to epidemiologists, healthcare professionals, decision makers and colleagues alike.
-Our findings clearly summarise MERS-CoV epidemiology from sequence data alone and help contextualise previous studies, rather than add to the perceived mystery of MERS-CoV.
+Our findings clearly summarise MERS-CoV epidemiology and help contextualise previous studies, rather than add to the perceived mystery of MERS-CoV.
 The approaches we employ serve to both highlight the power of population structure modeling, as well as pushing the field towards greater integration of case and sequence data.

### 2017-07-26 mers-structure.tex
Add line numbers, alter seasonality and method figure.
some late comments and edits

@@
 \usepackage{amsmath}
+
+\usepackage{lineno}
+\linenumbers
 
@@
 Unlike the most simpler %\lmc{simpler?}
-``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, structured population models possess the ability to estimate migration times.
+``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, 
+% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'. 
+structured population models possess the ability to estimate migration times. % So do Markov Jumps. 
 We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
@@
 	\caption{\textbf{MERS-CoV seasonality.}
-A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
+A) %Violin plots% %  (the plot type is irrelevant).
+Posterior density estimates coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
 Posterior means are indicated with circles.
 Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
-Grey area spanning October to February indicates the camel calving season.
+Hatched area spanning October to February indicates the camel calving season.
 B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
@@
 This provides ideal conditions for co-infection with distinct genotypes, which is a pre-requisite for detectable RNA virus recombination to occur%\lmc{stupid biostatistician here: is it?} \gdc{fair comment actually, it's a pre-requisite for detectable recombination to occur usually}
-.
 
+%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
 Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
-We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (Figure \ref{recombination}A).
+We find little evidence that recombination will interfere with the inference of human outbreak clusters %beyond what could be expected of phylogenetic noise %% doesn't make sense to say that recombination is interfering as expected by phylogenetic noise.
+(Figure \ref{recombination}A).
 And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
-Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue in MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
-Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
+Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue for inferring MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
+%Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
+%% a bit overblown 'extreme caution'?
 On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
+%% But why won't recombination be happening in humans?
+Sure it won't be 2 divergent lineages.
 
@@
 \subsection*{Demographic trajectory of MERS-CoV in camels}
-Since MERS-CoV in humans evolves under a distinct demographic regime, we split our sequence dataset based on whether sequences were collected in humans or camels.
+Since MERS-CoV in humans will have a distinct epidemiology
+% evolves under a distinct demographic regime, %% evolves under a demographic regime?
+we split our sequence dataset based on whether sequences were collected in humans or camels.
 We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
@@
 Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
-As computational infrastructure improves \citep{ayres_beagle:_2012}, it is hopefully only a matter of time before phylogenetic population structure models mature to the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) can be analysed trivially.
+%% Don't understand this sentence.
+How is the second clause related to the first?
+As computational infrastructure improves \citep{ayres_beagle:_2012}, %it is hopefully only a matter of time before 
+phylogenetic population structure models will reach the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) will be tractable.
 
@@
 
-\subsection*{Sequence data as a universal epidemiology tool}
+\subsection*{Sequence data as a universal epidemiology tool} %% Drop this paragraph and finish on something better. I think this point has been made already. 
 As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
@@
 This was used to great effect at all stages of the Ebola virus epidemic in West Africa, where sequence data early on were used to show continued circulation of the same strain in the region rather than multiple zoonotic introductions \citep{gire_genomic_2014}, to study viral migration at a broad level \citep{park_ebola_2015,ladner_evolution_2015} as healthcare infrastructure was being overwhelmed during the peak, and towards the end where the final transmission chains were tracked and confirmed as such \citep{arias_rapid_2016}.
-Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}
-.
+%Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}  %% AR - sorry, I hate this sentence - 'fundamental unit of ground truth' WTF?
 In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.
+
+%% AR - this paper is about MERS epidemiology.
+Finish on MERS epidemiology.
+The last 3 paragraphs are discussion about methods.
+For a high impact general journal like eLif, need to talk about the implications of the findings.
+Or send it to MBE.
 

### 2017-07-26 mers-structure.tex
Minor revisions

@@
 We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
-%\tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'.
-I think you want a different word here.
-Maybe `transient host'?}
-%\gdc{Good point about the term. Hope the definition leans towards `one not involved in maintenance' rather than `limited to index case'. Asked a friend working on zoonoses about this.}
 By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
-% \tbc{Powerful and succinct abstract. Great.}
 
@@
 Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
-Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
+Here we use existing MERS-CoV sequence data (174 MERS-CoV genomes from human infections and 100 MERS-CoV genomes from camel infections) to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
 
@@
 We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
-It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels %\lmc{is ``allow'' the best word here?
-I understand you' rather not use ``biases'' or ``misleads'', though...}
-the model cannot dismiss humans as a potential alternative host.
+It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot dismiss humans as a potential alternative host.
 The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
@@
 We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
+\tbc{I think this sentence (``We therefore...'') needs to be spelled out a bit more.
+It feels like the logic is what the Monte Carlo sampling does.
+Either spell out more, or push back to section with figure \ref{ABC}.}
 
@@
 \subsection*{Signals of MERS-CoV seasonality}
-Unlike the most simpler %\lmc{simpler?}
-``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, 
+Unlike simpler ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
 % please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
+% TB: Agree.
 structured population models possess the ability to estimate migration times. % So do Markov Jumps.
@@
 \subsection*{Demographic trajectory of MERS-CoV in camels}
+
+\tbc{Slightly strange to have this at the level of full results subsection and not have a main text figure.}
 Since MERS-CoV in humans will have a distinct epidemiology
@@
 \section*{Acknowledgements}
-We would like to thank Allison Black for useful discussion and advice.
+We would like to thank \tbc{XXX, YYY, ZZZ, etc...} for kindly sharing sequence data, and Allison Black for useful discussion and advice.
 GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
+TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
 

### 2017-07-27 mers-structure.tex
Generalise discussion, AR edits.

@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology}
+\title{\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface}
 
@@
 \subsection*{Signals of MERS-CoV seasonality}
-Unlike simpler ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
+% Unlike the most simpler %\lmc{simpler?}
+% ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
 % please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
-% TB: Agree.
-structured population models possess the ability to estimate migration times. % So do Markov Jumps.
-We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
+% structured population models possess the ability to estimate migration times. % So do Markov Jumps. \gdc{good point, yeah}
+We use the posterior distribution of MERS-CoV introduction times of human clades observed in the MCC tree (Figure \ref{mcc}) recovered from structured coalescent to model seasonal variation in zoonotic transfer of viruses.
 We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
@@
 And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
-Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue for inferring MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
+Observed departures from strictly clonal evolution suggest that whilst recombination is an issue for inferring MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
+MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
 %Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
 %% a bit overblown 'extreme caution'?
-On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
+In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
+% On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
 %% But why won't recombination be happening in humans?
@@
 
-\subsection*{Role of population structure in phylodynamics}
+\subsection*{MERS-CoV epidemiology}
 %\lmc{Detecting?}
@@
 As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{ABC}).
-Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
-Perhaps a clearer indicator of this has been the lack of a global MERS pandemic, which would have certainly taken place had R$_{0}$ been above the epidemic threshold.
 
-Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
-As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
-Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
-Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
-Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
+\subsection*{Transmissibility differences between zoonoses and pandemics}
+Severe Acute Respiratory Syndrome coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
+Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
+Although we describe the epidemiology of MERS-CoV from sequence data, indications that MERS-CoV has poor capacity to spread human-to-human existed prior to any sequence data.
+SARS-CoV swept through the world in a short period of time, but MERS cases trickled slowly and were restricted to the Arabian Peninsula or resulted in self-limiting outbreaks outside of the region, a pattern strongly indicative of repeat zoonotic spillover.
+Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
+
+MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
+Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
+When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
+
+% Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
+% As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
+% Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
+% Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
+% Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods.
 %% Don't understand this sentence.
 How is the second clause related to the first?
-As computational infrastructure improves \citep{ayres_beagle:_2012}, %it is hopefully only a matter of time before
-phylogenetic population structure models will reach the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) will be tractable.
+% As computational infrastructure improves \citep{ayres_beagle:_2012}, %it is hopefully only a matter of time before
+% phylogenetic population structure models will reach the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) will be tractable.
 
@@
 
-\subsection*{Sequence data as a universal epidemiology tool} %% Drop this paragraph and finish on something better. I think this point has been made already.
-As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
-At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
-With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
-This was used to great effect at all stages of the Ebola virus epidemic in West Africa, where sequence data early on were used to show continued circulation of the same strain in the region rather than multiple zoonotic introductions \citep{gire_genomic_2014}, to study viral migration at a broad level \citep{park_ebola_2015,ladner_evolution_2015} as healthcare infrastructure was being overwhelmed during the peak, and towards the end where the final transmission chains were tracked and confirmed as such \citep{arias_rapid_2016}.
-%Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}  %% AR - sorry, I hate this sentence - 'fundamental unit of ground truth' WTF?
-In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.
+% \subsection*{Sequence data as a universal epidemiology tool} %% Drop this paragraph and finish on something better. I think this point has been made already.
+% As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
+% At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
+% With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
+% This was used to great effect at all stages of the Ebola virus epidemic in West Africa, where sequence data early on were used to show continued circulation of the same strain in the region rather than multiple zoonotic introductions \citep{gire_genomic_2014}, to study viral migration at a broad level \citep{park_ebola_2015,ladner_evolution_2015} as healthcare infrastructure was being overwhelmed during the peak, and towards the end where the final transmission chains were tracked and confirmed as such \citep{arias_rapid_2016}.
+% %Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}  %% AR - sorry, I hate this sentence - 'fundamental unit of ground truth' WTF?
+% In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.
 
@@
 GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
-TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
+We gratefully acknowledge the contribution of the following scientists for sharing MERS-CoV sequence data before publication:
+
+Ali M.
+Somily$^{1}$, Mazin Barry$^{1}$, Sarah S.
+Al Subaie$^{1}$, Abdulaziz A.
+BinSaeed$^{1}$, Fahad A.
+Alzamil$^{1}$, Waleed Zaher$^{1}$, Theeb Al Qahtani$^{1}$, Khaldoon Al Jerian$^{1}$, Scott J.N.
+McNabb$^{2}$, Imad A.
+Al-Jahdali$^{3}$, Ahmed M.
+Alotaibi$^{4}$, Nahid A.
+Batarfi$^{5}$, Matthew Cotten$^{6}$, Simon J.
+Watson$^{6}$, Spela Binter$^{6}$, Paul Kellam$^{6}$.
+
+$^{1}$College of Medicine, King Saud University, Riyadh, Kingdom of Saudi Arabia \\
+$^{2}$Rollins School of Public Health, Emory University, Atlanta, GA, USA \\
+$^{3}$Deputy Minister. Ex. General Director King Fahad General Hospital, Jeddah and Occupational and environmental medicine, Um AlQura University, Kingdom of Saudi Arabia \\
+$^{4}$Department of Intensive Care, Prince Mohammed bin Abdulaziz Hospital, Riyadh, Kingdom of Saudi Arabia \\
+$^{5}$Epidemiology section, Command and Control Center (CCC) Ministry of Health, Jeddah \\
+$^{6}$Wellcome Trust Sanger Institute, Hinxton, United Kingdom \\
 

### 2017-07-31 mers-structure.tex
Revision pass

@@
 
-Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
+Middle East respiratory syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
 The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
-Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
+While traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
 Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
 We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
-By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
+By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that human outbreaks in the Arabian peninsula are driven by seasonally varying zoonotic transfer of viruses from camels.
+Without heretofore unseen evolution of host tropism, MERS-CoV is unlikely to become endemic in humans.
 
@@
 \section*{Introduction}
-%% Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
-% AR - a virus isn't an infection and not all infections in the Arabian Peninsula
-Middle East Respiratory Syndrome Coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans.
-% ?
-something like this ?
+Middle East respiratory syndrome coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans.
 The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{who_2017}.
-Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
-Whilst contact with camels is often reported, patients may not recall contact with %\lmc{other?}
-any livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
-Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
-Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
-Epidemiological modeling approaches have been used to look at traditional data sources of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.
+Its epidemiology remains obscure, largely because infections are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
+While contact with camels is often reported, other patients do not recall contact with any livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
+Previous studies on MERS-CoV epidemiology have used serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
+Such data have shown high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
+Separately, epidemiological modelling approaches have been used to look at incidence reports through time, space and across hosts \citep{cauchemez_unraveling_2016}.
 
-Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches %\lmc{why, specifically?}
-due to difficulties in linking cases into clusters in the absence of detailed information.
-Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{gire_genomic_2014,quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
+Although such traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
+Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}, where data are relatively cheap to produce.
 These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
-In the absence of other information often only sequence data can %\lmc{help}
-help pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
+Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
 Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
-A large portion of MERS-CoV sequences come from %\lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}
-outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
+A large portion of MERS-CoV sequences come from outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
 Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.
 
-It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
-\citep{who_2016_MERS}.
-Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
-Here we use existing MERS-CoV sequence data (174 MERS-CoV genomes from human infections and 100 MERS-CoV genomes from camel infections) to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \citep{who_2016_MERS}.
+Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional epidemiological approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
+Here we use a dataset of 274 MERS-CoV genomes to investigate transmission patterns of the virus between humans and camels.
 
-%Unlike discrete trait analyses (DTA, or ``mugration'' models) % AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
-%that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
-Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) %[cite here, Philippe's bat rabies and Sam's avian flu?, others?]
-\citep{faria_simultaneously_2013,global_consortium_for_h5n8_and_related_influenza_viruses_role_2016}, population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
+Here, we use an explicit model of metapopulation structure and migration between discrete subpopulations, referred to here as demes \citep{vaughan_efficient_2014}, derived from the structured coalescent \citep{notohara_structured_coalescent_1990}.
+Unlike approaches that model host species as a discrete phylogenetic trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) \citep{faria_simultaneously_2013,lycett_h5n8_2016}, population structure models explicitly incorporate distinct sampling patterns, population dynamics within demes and migration between demes.
 By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
@@
 
-%\begin{figure}[h]
-% \centering
-%	\includegraphics[width=0.45\textwidth]{figures/}
-%	\caption{\textbf{}
-%
-%	}
-%	\label{}
-%\end{figure}
-
 \section*{Results}
@@
 
-The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
-Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
-Whilst we %\lmc{estimate?}
-estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
-We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
-It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot dismiss humans as a potential alternative host.
+The structured coalescent approach we employ (see Methods) identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), while human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
+Across 174 MERS-CoV genomes collected from humans, we estimate a median of 56 separate camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
+While we estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
+This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
+This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of \tbc{XXX (95\% HPD: XXX--XXX)} and human viruses showing a scaled effective population of \tbc{XXX (95\% HPD: XXX--XXX)}.
+
+We believe that the small number of inferred human-to-camel migrations are induced by the migration rate prior, while some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
+This is most apparent in lineages in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot completely dismiss humans as a potential alternative host.
 The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
@@
 These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
-Our analyses recover these results despite sequence data heavily skewed towards non-uniformly %\lmc{maybe ``non-uniformly'' is redundant here} \gdc{disagree, the point I am trying to make is that the data are mostly human and that human data are further skewed in how they are sampled}
-sampled human cases and are robust to choice of prior.
-We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
-\tbc{I think this sentence (``We therefore...'') needs to be spelled out a bit more.
-It feels like the logic is what the Monte Carlo sampling does.
-Either spell out more, or push back to section with figure \ref{ABC}.}
+Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
+This suggests that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
 
@@
  \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
-	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from humans and camels.}
+	\includegraphics[width=0.75\textwidth]{figures/mers_mcc.png}
+	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from 174 human viruses and 100 camel viruses.}
 	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
-	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
+	The vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
   Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
@@
   Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
-	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
+  \tbc{The previous two sentences (``Some of the...'') seem unnecessary and confusingly written. I would just drop them.}
+	While large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
 	}
@@
 
-\subsection*{Signals of MERS-CoV seasonality}
-% Unlike the most simpler %\lmc{simpler?}
-% ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
-% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
-% structured population models possess the ability to estimate migration times. % So do Markov Jumps. \gdc{good point, yeah}
-We use the posterior distribution of MERS-CoV introduction times of human clades observed in the MCC tree (Figure \ref{mcc}) recovered from structured coalescent to model seasonal variation in zoonotic transfer of viruses.
+\subsection*{MERS-CoV shows seasonal introductions}
+We use the posterior distribution of MERS-CoV introduction events from camels to humans (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
 We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
 Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
-It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
+It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached \citep{martinez_seasonality_2014}, leading to a sharp rise in prevalence of the virus in camels and resulting in increased force of infection into the human population.
 Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.
+
+\tbc{I think this is the best place to discuss the lack of correlation between date and size of human cluster.
+This can be referenced later on in Discussion when talking about pandemic potential.
+Also, seems relevant to look at seasonality in size of cluster here, yes?
+This would get at whether there are seasonal factors on the human side.}
 
@@
 \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
-	\caption{\textbf{MERS-CoV seasonality.}
-A) %Violin plots% %  (the plot type is irrelevant).
-Posterior density estimates coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
+	\includegraphics[width=0.75\textwidth]{figures/mers_seasonality.png}
+	\caption{\textbf{Seasonality of MERS-CoV introduction events.}
+A) Posterior density estimates partitioned by month showing the 95\% highest posterior density interval for relative odds ratios of MERS-CoV introductions into humans.
 Posterior means are indicated with circles.
@@
 Hatched area spanning October to February indicates the camel calving season.
-B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
-95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (Figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
+B) Sequence cluster sizes and inferred dates of introduction events.
+Each introduction event is shown as a vertical line positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendant sequences recovered from human cases.
+95\% highest posterior density intervals for introductions of MERS-CoV into humans are indicated with coloured lines, coloured by median estimated introduction time.
 The black dotted line indicates the joint probability density for introductions.
-%Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
+We find little correlation between date and size of introduction (Spearman $\rho = 0.06$, $p=0.68$).
 	}
@@
 
-%\begin{figure}[h]
-% \centering
-%	\includegraphics[width=0.45\textwidth]{figures/}
-%	\caption{\textbf{}
-%
-%	}
-%	\label{}
-%\end{figure}
-
 \subsection*{MERS-CoV is poorly suited for human transmission}
 
-Structured coalescent approaches clearly show humans to be a terminal host %\tbc{again, I don't know if this is the right term or not}
-for MERS-CoV, implying poor transmissibility.
-However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides %\lmc{insights?}
-insight into epidemic behavior.
-The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by a dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}).
+Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
+However, we wanted to translate this observation into an estimate of the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides insight into epidemic behavior.
+The parameter $R_{0}$ determines expected number of secondary cases in a single infections as well as the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}, Methods).
 We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
-First, we simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
+First, we simulate case counts across multiple outbreaks totaling 2000 cases using Equation \ref{clusterLikelihood} and then we subsample from each case cluster to simulate sequencing of a fraction of cases.
 Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
 This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
-We record the $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses (left-hand panels in Figure \ref{ABC}).
+We record the mean, median and standard deviation of sequence cluster sizes in each simulation (left-hand panels in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
+We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
+A schematic of this Monte Carlo produre is shown in Figure \ref{mc_method}.
+Generally, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters (Figure \ref{mers_epi}).
 
-We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
+We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
 For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
-Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
-Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%%\lmc{again?}
-(nearly 10 000 out of 363 000 for both) of simulations fit for bias levels of 2.0 and 3.0, respectively.
-This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate that a higher number of simulations generate MERS-CoV-like data at bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).%\lmc{I don't follow this argument...}
-Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).
+While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
+Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
+Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
+Model averaging would suggest plausible $R_0$ values between \tbc{XXX} and \tbc{XXX}.
 
-Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{ABC}).
-The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
+Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
+The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
 These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Our results suggest a large number of unobserved MERS primary cases.%\lmc{does the definition of ``index case'' allow for ``multiple''?} \gdc{not in epi literature I don't think}
+Model averaging would suggest plausible numbers of introductions between \tbc{XXX} and \tbc{XXX}.
+Our results suggest a large number of unobserved MERS primary cases.
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
-In those sets of analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
+In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
 
 Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
-% \tbc{I didn't understand what you meant by `sequence sampling wholly unrepresentative of viral diversity found in humans'.}
 Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
-%\tbc{It would be really nice to be able to reference something here on hospital spread of MERS (or SARS) owing to hygiene practices.}
 
@@
 \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
-	\caption{\textbf{Results of Monte Carlo simulations.}
-Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced%\lmc{make sequences from large clusters more likely to be picked}
-).
-Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
-The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
-The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
+	\includegraphics[width=0.75\textwidth]{figures/mers_epi.png}
+	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
+Each row corresponds to a different bias value used to concentrate the subsampling of sequences from cases, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced).
+Leftmost scatter plots show results of individual Monte Carlo simulations, coloured by the $R_{0}$ value used for the simulation.
+The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for empirical MERS-CoV data.
+The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same $y$-axis across all levels of bias.
 Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
-The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
+The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
 Darker shade of grey indicates bins falling within the 95\% percentiles.
 These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
-% \tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to alleviate confusion.
-I'm picturing a set of tables numbered with case counts in a top row and then a bottom row with sequence counts.
-This could also work better as a separate figure referenced from the Methods.}
 	}
-	\label{ABC}
+	\label{mers_epi}
 \end{figure}
-
-%\begin{figure}[h]
-% \centering
-%	\includegraphics[width=0.65\textwidth]{figures/}
-%	\caption{\textbf{}
-%
-%	}
-%	\label{}
-%\end{figure}
 
@@
 Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
-One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, that appear as repeated mutations taking place in the recipient lineage, somewhere else in the tree.
-PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
-Another hallmark of recombination is spatial clustering of alleles in the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
+One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, which appear as repeated mutations taking place in the recipient lineage, somewhere else in the tree.
+The PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
+Another hallmark of recombination is spatial clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
 3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.
@@
 As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
-Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
-Although the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
-This provides ideal conditions for co-infection with distinct genotypes, which is a pre-requisite for detectable RNA virus recombination to occur%\lmc{stupid biostatistician here: is it?} \gdc{fair comment actually, it's a pre-requisite for detectable recombination to occur usually}
+% Homoplasies are present in multiple strains at a time, indicating recombination events that have been successful.
+Although the evolutionary centrality of camel viruses (Figure \ref{mcc}) may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
+This provides ideal conditions for co-infection with distinct genotypes, which is a pre-requisite for detectable RNA virus recombination to occur.
 
@@
 Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
-We find little evidence that recombination will interfere with the inference of human outbreak clusters %beyond what could be expected of phylogenetic noise %% doesn't make sense to say that recombination is interfering as expected by phylogenetic noise.
-(Figure \ref{recombination}A).
-And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
-Observed departures from strictly clonal evolution suggest that whilst recombination is an issue for inferring MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
+We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
+\tbc{Summarize Figure \ref{recombinant_features}A with something like: ``We find that 95\% of human viruses fall within the same introduction event in genomic fragment 1 and in genomic fragment 2.''}
+And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
+Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
 MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
@@
 %% a bit overblown 'extreme caution'?
+%TB: Agree to leave out this sentence.
 In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
@@
 % 	Tree branches are coloured by inferred ancestral host state (human in grey, camel in black).
-% 	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in the camel host.
+% 	While some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in the camel host.
 % 	Human clusters in grey change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
@@
     \centering
-    \subfloat{{\includegraphics[width=7cm]{figures/mers_fragments.png} }}%
-    \qquad
-    \subfloat{{\includegraphics[width=5cm]{figures/mers_chain.png} }}%
-    \caption{\textbf{Recombinant features of MERS-CoV phylogenies}.
-    A) Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
-    Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
+    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
+    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
+    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
+    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to \tbc{XXX}.
+    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
+    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
     Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
-    B) Phylogenies derived from MERS-CoV genome nucleotides up to position 21000 (left) and all nucleotides past position 21000 (right), reduced to just the human tips.
-  	Same taxa are connected via coloured lines indicating their vertical position in the tree on the right.
+    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
+  	Identical taxa are connected via coloured lines with colour indicating vertical position in the tree on the right.
   	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
-  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels, adding support to the idea that camels are often co-infected.%\lmc{I'm sure this is important, but I didn't get why.
+  	While some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels, adding support to the idea that camels are often co-infected.
+%\lmc{I'm sure this is important, but I didn't get why.
 Referees might not either...}
-  	Human clusters in blue change phylogenetic positions between the trees %\lmc{wholesale??}
-    together, with minor incongruences within clusters.
-  	This is evidence for recombinant viruses generated in the camel reservoir%\lmc{(camels)}
-    entering human populations.
-    }%
-    \label{recombination}%
+    \tbc{I would favor just dropping the ``While some of apparent...'' sentence. Confusing and I don't see what it adds.}
+  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
+  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
+    }
+    \label{recombination}
 \end{figure}
@@
 \tbc{Slightly strange to have this at the level of full results subsection and not have a main text figure.}
-Since MERS-CoV in humans will have a distinct epidemiology
+Here we attempt to investigate MERS-CoV demographic patterns in the camel reservoir.
+%Since MERS-CoV in humans has a distinct epidemiology
 % evolves under a distinct demographic regime, %% evolves under a demographic regime?
-we split our sequence dataset based on whether sequences were collected in humans or camels.
+%we split our sequence dataset based on whether sequences were collected in humans or camels.
 We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
@@
 We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions which are not necessarily the same.
-Out of the five runs we set up two chains converged to one stationary distribution, two to another and the last one to a third.
+Out of five runs, two chains converged to one stationary distribution, two to another and the last one to a third.
 Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
 This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
+\tbc{This entire paragraph should be moved to Methods. Too much detail for Results. Below just say that ``Multiple MCMC chains showed similar patternso of $N_e \tau$ through time.''}
 
-Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity (N$_{e}\tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
+Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
 However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
@@
 Part of this could be caused by MERS-CoV ability to re-infect camels previously exposed to the virus \citep{ali_systematic_2017}, which may result in a process akin to antigenic drift, where the depletion of susceptible calves leads to ever increasing selective pressure for the virus to evolve antigenically.
-Whilst this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.
+While this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.
+\tbc{This section seems like its reaching pretty far, but if keeping could mention that human seasonal coronaviruses undergo antigenic drift.}
 
@@
 It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
+\tbc{This seems way overly broad to me.
+I much preferred the Introductions take on genomic epi revealing important population structure.
+Just revise to keep this focused on pathogen transmission.
+Identification of reservoirs, source-sink, all very important.}
 Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
 In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
-Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.
+While there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as $R_0$ or the number of zoonotic introductions of MERS-CoV into humans.
+\tbc{This sentence (``While there...'') is overly like Intro.}
 
-We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
-This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{ABC}).
-From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{ABC}).
-We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
+We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV genomic data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
+This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{mers_epi}).
+From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{mers_epi}).
+We also looked at potential seasonality in MERS-CoV spillover into humans.
 Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
-As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{ABC}).
+As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{mers_epi}).
 
 \subsection*{Transmissibility differences between zoonoses and pandemics}
-Severe Acute Respiratory Syndrome coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
+Severe acute respiratory syndrome (SARS) coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
 Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
@@
 Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
+\tbc{I like the basic approach, this last segue could be cleaner however.}
 
@@
 When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
+
+\tbc{Perhaps close with discussion of evolution of host tropism?
+So far outbreaks have been self-limiting, but continued circulation in camels and continued spill-over into humans provides continued chances for evolution of a more human transmissible virus.}
 
@@
 % Our study has shown how significantly different demographic regimes between hosts such as those observed in MERS-CoV can be leveraged against biased sequence data to arrive at conclusions perfectly in line with prior observations.
-% MERS-CoV sequences collected from camels are, on average, relatively distantly related to each other, whilst human outbreaks fall into densely sampled clusters of no apparent relation to each other and of no consequence to further evolution of the virus.
+% MERS-CoV sequences collected from camels are, on average, relatively distantly related to each other, while human outbreaks fall into densely sampled clusters of no apparent relation to each other and of no consequence to further evolution of the virus.
 % These hallmarks of distinct demographic modes were easily picked up by the explicit population structure model we employ, and reveals a cohesive picture of MERS-CoV epidemiology in agreement with previous empirical studies.
@@
 % However, lineages crossing the species barrier are very likely to have experienced recombination at some point in their past, in some cases occurring right before zoonotic transmission, as was the case in the Korean outbreak of MERS.
-% We thus conclude that whilst sequence analyses of MERS-CoV are perfectly legitimate for well-defined human outbreaks, extreme care must be taken when inferring their origins in the context of the reservoir, or even when comparing related outbreaks in humans.
+% We thus conclude that while sequence analyses of MERS-CoV are perfectly legitimate for well-defined human outbreaks, extreme care must be taken when inferring their origins in the context of the reservoir, or even when comparing related outbreaks in humans.
 % We also strongly advise against drawing hasty conclusions from MERS-CoV phylogenies recovered from human sequences if they span more than four months of evolution or camel sequences under virtually all circumstances, as there is no guarantee that phylogenies will cover the clonal evolutionary history of the virus in humans.
@@
 % As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
-% At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
+% At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, while also offering insight into a pathogen's evolutionary history.
 % With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
@@
 All analyses were set to run for 200 million states, subsampling every 20 000 states.
-Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burnin for single fragment analyses and 20\% for two-fragment analyses.
+Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in for single fragment analyses and 20\% for two-fragment analyses.
 Chains not converging after 200 million states were discarded.
@@
 \section*{Acknowledgements}
-We would like to thank \tbc{XXX, YYY, ZZZ, etc...} for kindly sharing sequence data, and Allison Black for useful discussion and advice.
+We would like to thank Allison Black for useful discussion and advice.
 GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
+TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
 We gratefully acknowledge the contribution of the following scientists for sharing MERS-CoV sequence data before publication:
@@
 
-\bibliographystyle{mbe.bst}
+\bibliographystyle{mbe}
 \bibliography{mers-structure}
@@
 	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
-Same tree as shown in Figure \ref{mcc}, but all contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top) and humans (bottom).
+This is the same tree as shown in Figure \ref{mcc}, but with contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top in orange) and humans (bottom in blue).
 This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
@@
 \centering
-	\includegraphics[scale=0.25]{figures/mers_method.pdf}
-	\caption{\textbf{Monte Carlo simulation method.}
-Case clusters are simulated according to \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
-We sample 174 cases from each simulation to represent sequencing of human MERS cases.%\lmc{is this implicitly a binomial sampling with $p = 1/2$?} \gdc{hypergeometric, since we probably don't resequence people}
-`Sequencing' is carried out by using multivariate hypergeometric sampling, where at each iteration case cluster size $C_{i}$ is the number of successes in the population, $\sum_{i=1}^{K} C_{i} - C_{i}$ is the number of failures and number of trials is equal to leftover sequencing capacity, that is $\sum_{i=1}^{K} S_{i}$ minus whatever has been `sequenced' up to that iteration.
-Sequencing simulations take place at three levels of bias - 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
-Simulated sequence clusters are summarised by their mean and standard deviation, both of which reflect some aspect of the partitioning of sequences into clusters.
-A simulation is considered to match if its sequence cluster size mean and standard deviation fell within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
-$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are kept as estimates.
-These estimates are summarised in Figure \ref{ABC}.
+	\includegraphics[width=0.75\textwidth]{figures/mers_mc_method.pdf}
+	\caption{\textbf{Monte Carlo simulation schematic.}
+Case clusters are simulated according to Equation \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
+We sample 174 cases from each simulation to represent sequencing of human MERS cases.
+`Sequencing' is carried out by using multivariate hypergeometric sampling, representing sampling cases without replacement to be sequenced.
+Sequencing simulations take place at three levels of bias: 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
+The distribution of simulated sequence clusters is summarised by its mean, median and standard deviation.
+A simulation is considered to match if the mean, median and standard deviation of its sequence cluster sizes falls within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
+$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are retained as estimates.
+These estimates are summarised in Figure \ref{mers_epi}.
 	}
-	\label{method}
+	\label{mc_method}
 \end{figure}
@@
 	\caption{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
-The plot is identical to Figure \ref{ABC}, but instead of 2000 cases,
+The plot is identical to Figure \ref{mers_epi}, but instead of 2000 cases,
 % the currently reported number of MERS-CoV cases,
@@
 \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_recombination.png}
-	\caption{\textbf{Tests of recombination across MERS-CoV clades.}}
+	\includegraphics[width=0.65\textwidth]{figures/mers_recombination_tree.png}
+	\caption{\textbf{Tests of recombination across MERS-CoV clades.}
 Maximum clade credibility tree of MERS-CoV genomes annotated with results of two recombination detection tests (PHI and 3Seq) applied to descendent sequences of each clade.
 Both tests identify large portions of existing sequence data as containing signals of recombination.
-Note that markings do not indicate where recombinations have occurred on the tree, merely the minimum distance in sequence/time space between recombining lineages.
+Note that markings do not indicate where recombinations have occurred on the tree, merely the minimum distance in sequence/time space between recombining lineages.}
 	\label{recombination_tree}
@@
 \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_incompatibilities.png}
+	\includegraphics[width=0.75\textwidth]{figures/mers_incompatibilities.png}
 	\caption{\textbf{MERS-CoV genomes exhibit high numbers of non-clonal loci.}
-Ancestral state reconstruction (right) identifies a large number of loci that have occurred more than once in the tree (homoplasies, orange) or are reversions (red) from a state arising in an ancestor.
+Ancestral state reconstruction (right) identifies a large number of sites in which mutations have occurred more than once in the tree (homoplasies, orange) or are reversions (red) from a state arising in an ancestor.
 Mutations that apparently only occur once in the tree (synapomorphies) are shown in grey.

### 2017-08-01 mers-structure.tex
Start on Methods edits

@@
 \subsection*{Sequence data}
-All MERS-CoV sequences were downloaded from GenBank.%\lmc{accession numbers are given in Table X}.
+All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
+\tbc{Mention what's going on with accession-less sequences.}
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
+\tbc{This should be obvious in Table \ref{sequences}. Some strains should be multiple accessions. However, I don't see any.}
+\tbc{What we the criteria for inclusion? How many bases were necessary to be considered a `genome'?}
+Sequences were annotated with available collection dates and hosts, designated as camel or human.
 Protein coding sequences were extracted and concatenated.
-Sequences were annotated with available collection dates and hosts, designated as camel or human.
+\tbc{Leaving how many bases?}
+\tbc{How were they aligned?}
 The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).
@@
 
-MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes - humans and camels - based on GenBank records.
+MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes -- humans and camels.
+This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
 Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
@@
 All analyses were run across ten independent Markov chains (MCMC runs).
+\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
 To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
 
-We also analysed a dataset where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+As a secondary analysis, we also considered a scenario where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
 Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.

### 2017-08-01 mers-structure.tex
High resolution figures, some text changes.

--- 2017-07-31 mers-structure.tex
@@
 This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
-This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of \tbc{XXX (95\% HPD: XXX--XXX)} and human viruses showing a scaled effective population of \tbc{XXX (95\% HPD: XXX--XXX)}.
+This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of 3.49 (95\% HPD: 2.71--4.38) and human viruses showing a scaled effective population of 0.24 (95\% HPD: 0.14--0.34).
 
@@
   Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
-  Some of the branches leading up to zoonotic transmission of MERS-CoV into humans are too long to have spent much of their time in humans, given our \textit{a priori} belief that MERS-CoV cannot circulate in humans for long periods of time.
-  Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
-  \tbc{The previous two sentences (``Some of the...'') seem unnecessary and confusingly written. I would just drop them.}
 	While large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
@@
   	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
-  	While some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels, adding support to the idea that camels are often co-infected.
-%\lmc{I'm sure this is important, but I didn't get why.
-Referees might not either...}
-    \tbc{I would favor just dropping the ``While some of apparent...'' sentence. Confusing and I don't see what it adds.}
   	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
@@
 Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
-In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
-While there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as $R_0$ or the number of zoonotic introductions of MERS-CoV into humans.
-\tbc{This sentence (``While there...'') is overly like Intro.}
+In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula by using sequence data collected from both hosts (174 from humans and 100 from camels).
 
@@
 When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
-
-\tbc{Perhaps close with discussion of evolution of host tropism?
-So far outbreaks have been self-limiting, but continued circulation in camels and continued spill-over into humans provides continued chances for evolution of a more human transmissible virus.}
+Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans does not eliminate the possibility that a more transmissible variant of MERS-CoV will emerge.
+We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.
 

### 2017-08-01 mers-structure.tex
Add scripts, some data, last hi-res figures, some text changes.

@@
 
+Although we find evidence of seasonality in zoonotic spill over timing, no such relationship exists for sizes of human sequence clusters (\textit{i.e. post-spillover clades})(?).
+This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
+Similarly, we do not observe any trend in human sequence cluster sizes over time, suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
+This is not surprising either, since MERS-CoV is a camel virus that experiences little to no selective pressure to improve transmissibility between humans.
+
 \tbc{I think this is the best place to discuss the lack of correlation between date and size of human cluster.
@@
 Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
-Model averaging would suggest plausible $R_0$ values between \tbc{XXX} and \tbc{XXX}.
+Model averaging would suggest plausible $R_0$ values between 0.57 and 0.91.
 
@@
 These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Model averaging would suggest plausible numbers of introductions between \tbc{XXX} and \tbc{XXX}.
+Model averaging would suggest plausible numbers of introductions between 299 and 818.
 Our results suggest a large number of unobserved MERS primary cases.
@@
     A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
-    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to \tbc{XXX}.
+    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to 29364.
     The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
@@
 
-Our attempt at controlling the confounding effects of different sampling regimes between humans and camels revealed a troubling pattern, however.
-We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions which are not necessarily the same.
-Out of five runs, two chains converged to one stationary distribution, two to another and the last one to a third.
-Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
-This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
-\tbc{This entire paragraph should be moved to Methods. Too much detail for Results. Below just say that ``Multiple MCMC chains showed similar patternso of $N_e \tau$ through time.''}
-
 Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
@@
 %\lmc{Detecting?}
-Understanding population structure yields important insight into any biological system.
-It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
-\tbc{This seems way overly broad to me.
-I much preferred the Introductions take on genomic epi revealing important population structure.
-Just revise to keep this focused on pathogen transmission.
-Identification of reservoirs, source-sink, all very important.}
-Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
 In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula by using sequence data collected from both hosts (174 from humans and 100 from camels).
-
 We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV genomic data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
@@
 
+Strong population structure in viruses often arises through limited gene flow, either due to geography \citep{dudas_virus_2017}, ecology \citep{smith_dating_2009} or evolutionary forces \citep{turner_genomic_2005,dudas_reassortment_2015}.
+On a smaller scale population structure can unveil important details about transmission patterns, such as identifying reservoirs and understanding spill over trends and risk, much as we have done here.
+When properly understood naturally arising barriers to gene flow can be exploited for more efficient disease control and prevention, as well as risk management.
+%\tbc{This seems way overly broad to me.
+I much preferred the Introductions take on genomic epi revealing important population structure.
+Just revise to keep this focused on pathogen transmission.
+Identification of reservoirs, source-sink, all very important.}
+
 \subsection*{Transmissibility differences between zoonoses and pandemics}
@@
 Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
-\tbc{I like the basic approach, this last segue could be cleaner however.}
+Even though our results show that MERS-CoV does not appear to present a serious global threat, we would like to highlight that its epidemiology is nonetheless concerning.
+%\tbc{I like the basic approach, this last segue could be cleaner however.}
 
@@
 All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
-\tbc{Mention what's going on with accession-less sequences.}
+Sequences without accessions were kindly shared by Ali M.
+Somily, Mazin Barry, Sarah S.
+Al Subaie, Abdulaziz A.
+BinSaeed, Fahad A.
+Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N.
+McNabb, Imad A.
+Al-Jahdali, Ahmed M.
+Alotaibi, Nahid A.
+Batarfi, Matthew Cotten, Simon J.
+Watson, Spela Binter, and Paul Kellam prior to publication.
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
-\tbc{This should be obvious in Table \ref{sequences}. Some strains should be multiple accessions. However, I don't see any.}
-\tbc{What we the criteria for inclusion? How many bases were necessary to be considered a `genome'?}
-Sequences were annotated with available collection dates and hosts, designated as camel or human.
-Protein coding sequences were extracted and concatenated.
-\tbc{Leaving how many bases?}
-\tbc{How were they aligned?}
+Only sequences covering at least 50\% of MERS-CoV genome were kept.
+Sequences were annotated with available collection dates and hosts, designated as camel or human and aligned with MAFFT \citep{katoh_mafft_2013} and edited manually.
+Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364.
 The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).
@@
 Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-All analyses were run across ten independent Markov chains (MCMC runs).
-\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
+All analyses were run across for 200 million states across ten independent Markov chains (MCMC runs).
+%\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
+Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
+Three chains out of ten did not converge and were discarded altogether.
 To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
+For these analyses two chains did not converge.
 
@@
 All analyses were set to run for 200 million states, subsampling every 20 000 states.
-Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in for single fragment analyses and 20\% for two-fragment analyses.
 Chains not converging after 200 million states were discarded.
-For single fragment analysis with migration rate prior set to exponential with mean 1.0 there were three chains out of ten that did not converge and two for the analysis where the migration rate prior was an exponential with mean 10.0.
-For the two fragment structured coalescent analyses three chains out of ten failed to converge.
+20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
+Three chains out of ten did not converge.
 
@@
 Why not remove?} \gdc{Never hurts to repeat.}
+Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
+This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
 
@@
   115 & Al-Hasa\_4\_2013 & KF186564 & human & 2013-05-01 \\
-  116 & Al-Hasa\_7\_2013 & KF600623 & human & 2013-05-01 \\
-  117 & Al-Hasa\_8\_2013 & KF600618 & human & 2013-05-01 \\
+  116 & Al-Hasa\_7\_2013 & KF600623, KF600655 & human & 2013-05-01 \\
+  117 & Al-Hasa\_8\_2013 & KF600618, KF600626, KF600635, KF600638 & human & 2013-05-01 \\
   118 & Al-Hasa\_25\_2013 & KJ156866 & human & 2013-05-02 \\
-  119 & Al-Hasa\_11\_2013 & KF600629 & human & 2013-05-03 \\
+  119 & Al-Hasa\_11\_2013 & KF600629, KF600636, KF600646 & human & 2013-05-03 \\
   120 & Al-Hasa\_12\_2013 & KF600627 & human & 2013-05-07 \\
-  121 & Al-Hasa\_14\_2013 & KF600615 & human & 2013-05-08 \\
+  121 & Al-Hasa\_14\_2013 & KF600615, KF600643 & human & 2013-05-08 \\
   122 & Al-Hasa\_1\_2013 & KF186567 & human & 2013-05-09 \\
@@
   125 & Buraidah\_1\_2013 & KF600630 & human & 2013-05-13 \\
-  126 & Al-Hasa\_23\_2013 & KJ156860 & human & 2013-05-13 \\
+  126 & Al-Hasa\_23\_2013 & KJ156860, KJ156894, KJ156929, KJ156923, KJ156862 & human & 2013-05-13 \\
   127 & Al-Hasa\_17\_2013 & KF600647 & human & 2013-05-15 \\
@@
   133 & Taif\_1\_2013 & KJ156949 & human & 2013-06-12 \\
-  134 & Taif\_2\_2013 & KJ156896 & human & 2013-06-12 \\
-  135 & Taif\_3\_2013 & KJ156938 & human & 2013-06-13 \\
-  136 & Al-Hasa\_26\_2013 & KJ156882 & human & 2013-06-18 \\
-  137 & Al-Hasa\_27\_2013 & KJ156943 & human & 2013-06-19 \\
-  138 & Al-Hasa\_28\_2013 & KJ156887 & human & 2013-06-22 \\
-  139 & Riyadh\_6\_2013 & KJ156879 & human & 2013-07-02 \\
+  134 & Taif\_2\_2013 & KJ156896, KJ156876 & human & 2013-06-12 \\
+  135 & Taif\_3\_2013 & KJ156938, KJ156897, KJ156922, KJ156868, KJ156921, KJ156915,
+KJ156906 & human & 2013-06-13 \\
+  136 & Al-Hasa\_26\_2013 & KJ156882, KJ156941, KJ156872 & human & 2013-06-18 \\
+  137 & Al-Hasa\_27\_2013 & KJ156943, KJ156939 & human & 2013-06-19 \\
+  138 & Al-Hasa\_28\_2013 & KJ156887, KJ156940, KJ156889, KJ156893, KJ156884, KJ156930,
+KJ156928, KJ156909 & human & 2013-06-22 \\
+  139 & Riyadh\_6\_2013 & KJ156879, KJ156947, KJ156890, KJ156908, KJ156927 & human & 2013-07-02 \\
   140 & Riyadh\_5\_2013 & KJ156944 & human & 2013-07-02 \\
-  141 & Riyadh\_7\_2013 & KJ156937 & human & 2013-07-15 \\
-  142 & Riyadh\_8\_2013 & KJ156880 & human & 2013-07-17 \\
+  141 & Riyadh\_7\_2013 & KJ156937, KJ156905 & human & 2013-07-15 \\
+  142 & Riyadh\_8\_2013 & KJ156880, KJ156942 & human & 2013-07-17 \\
   143 & Riyadh\_9\_2013 & KJ156869 & human & 2013-07-17 \\
   144 & Hafr-Al-Batin\_2\_2013 & KJ156910 & human & 2013-08-05 \\
-  145 & Asir\_2\_2013 & KJ156863 & human & 2013-08-05 \\
-  146 & Riyadh\_11\_2013 & KJ156946 & human & 2013-08-06 \\
-  147 & Riyadh\_12\_2013 & KJ156926 & human & 2013-08-08 \\
-  148 & Riyadh\_13\_2013 & KJ156888 & human & 2013-08-13 \\
+  145 & Asir\_2\_2013 & KJ156863, KJ156899, KJ156912, KJ156900, KJ156898, KJ156945,
+KJ156932 & human & 2013-08-05 \\
+  146 & Riyadh\_11\_2013 & KJ156946, KJ156911 & human & 2013-08-06 \\
+  147 & Riyadh\_12\_2013 & KJ156926, KJ156901 & human & 2013-08-08 \\
+  148 & Riyadh\_13\_2013 & KJ156888, KJ156873 & human & 2013-08-13 \\
   149 & Riyadh\_14\_2013 & KJ156934 & human & 2013-08-15 \\
-  150 & Hafr-Al-Batin\_4\_2013 & KJ156931 & human & 2013-08-25 \\
-  151 & Hafr-Al-Batin\_5\_2013 & KJ156951 & human & 2013-08-25 \\
-  152 & Riyadh\_17\_2013 & KJ156918 & human & 2013-08-26 \\
+  150 & Hafr-Al-Batin\_4\_2013 & KJ156931, KJ156895, KJ156864, KJ156861 & human & 2013-08-25 \\
+  151 & Hafr-Al-Batin\_5\_2013 & KJ156951, KJ156924, KJ156954, KJ156913 & human & 2013-08-25 \\
+  152 & Riyadh\_17\_2013 & KJ156918, KJ156920, KJ156865 & human & 2013-08-26 \\
   153 & Hafr-Al-Batin\_6\_2013 & KJ156874 & human & 2013-08-28 \\
-  154 & Riyadh\_10\_2013 & KJ156891 & human & 2013-09-05 \\
-  155 & Madinah\_3b\_2013 & KJ156916 & human & 2013-09-11 \\
+  154 & Riyadh\_10\_2013 & KJ156891, KJ156936, KJ156907 & human & 2013-09-05 \\
+  155 & Madinah\_3b\_2013 & KJ156950, KJ156916 & human & 2013-09-11 \\
   156 & Qatar3 & KF961221 & human & 2013-10-13 \\
@@
 Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
-%\tbc{I think this would be more clear if the left and right sides shared the same x-axis.
-As it stands it looks at first glance like there's different priors for camel-to-human compared to human-to-camel.
-It's okay if the blue human-to-camel rates gets smushed up against the left of the plot.}
 	}

### 2017-08-02 mers-structure.tex

@@
 \usepackage{float}
+\usepackage{microtype}
 \usepackage{url}
 \urlstyle{same}
+
+\usepackage[hidelinks]{hyperref}
+\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}
 
@@
 While we estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
+Correspondingly, we observe substantially higher camel-to-human migration rate estimates than human-to-camel migration rate estimates (Figure~\ref{prior}).
 This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
-This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of 3.49 (95\% HPD: 2.71--4.38) and human viruses showing a scaled effective population of 0.24 (95\% HPD: 0.14--0.34).
+This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of 3.49 years (95\% HPD: 2.71--4.38) and human viruses showing a scaled effective population of 0.24 years (95\% HPD: 0.14--0.34).
 
@@
 
-Although we find evidence of seasonality in zoonotic spill over timing, no such relationship exists for sizes of human sequence clusters (\textit{i.e. post-spillover clades})(?).
+Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters \tbc{reference a figure}.
 This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
-Similarly, we do not observe any trend in human sequence cluster sizes over time, suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
-This is not surprising either, since MERS-CoV is a camel virus that experiences little to no selective pressure to improve transmissibility between humans.
-
-\tbc{I think this is the best place to discuss the lack of correlation between date and size of human cluster.
-This can be referenced later on in Discussion when talking about pandemic potential.
-Also, seems relevant to look at seasonality in size of cluster here, yes?
-This would get at whether there are seasonal factors on the human side.}
+Similarly, we do not observe any trend in human sequence cluster sizes over time (Figure \ref{seasonality}B, Spearman $\rho = 0.06$, $p=0.68$), suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
+This is not surprising either, since MERS-CoV is a camel virus that has to date, experienced little-to-no selective pressure to improve transmissibility between humans.
 
@@
 
-Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
+Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.91$ and sequence sampling likely to be biased towards large hospital outbreaks.
 Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
@@
 And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
+Critically, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
 Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
 MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
-%Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
-%% a bit overblown 'extreme caution'?
-%TB: Agree to leave out this sentence.
 In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
-% On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.
-%% But why won't recombination be happening in humans?
-Sure it won't be 2 divergent lineages.
 
@@
     }
-    \label{recombination}
+    \label{recombinant_features}
 \end{figure}
 
-\subsection*{Demographic trajectory of MERS-CoV in camels}
+\subsection*{MERS-CoV shows population turnover in camels}
 
-\tbc{Slightly strange to have this at the level of full results subsection and not have a main text figure.}
 Here we attempt to investigate MERS-CoV demographic patterns in the camel reservoir.
@@
 While this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.
-\tbc{This section seems like its reaching pretty far, but if keeping could mention that human seasonal coronaviruses undergo antigenic drift.}
+\tbc{Include results of N and S to indicate that surface protein is evolving quickly supporting antigenic evolution. Or not, if the data says otherwise...}
+
+\tbc{Okay for me to leave Figure \ref{skygrid} as supplementary or it can be main text.
+I'm happy either way.
+Suggest second panel showing correspondence between fragmented genome and full genome trajectories.}
 
@@
 Strong population structure in viruses often arises through limited gene flow, either due to geography \citep{dudas_virus_2017}, ecology \citep{smith_dating_2009} or evolutionary forces \citep{turner_genomic_2005,dudas_reassortment_2015}.
-On a smaller scale population structure can unveil important details about transmission patterns, such as identifying reservoirs and understanding spill over trends and risk, much as we have done here.
+On a smaller scale population structure can unveil important details about transmission patterns, such as identifying reservoirs and understanding spillover trends and risk, much as we have done here.
 When properly understood naturally arising barriers to gene flow can be exploited for more efficient disease control and prevention, as well as risk management.
-%\tbc{This seems way overly broad to me.
-I much preferred the Introductions take on genomic epi revealing important population structure.
-Just revise to keep this focused on pathogen transmission.
-Identification of reservoirs, source-sink, all very important.}
 
@@
 Even though our results show that MERS-CoV does not appear to present a serious global threat, we would like to highlight that its epidemiology is nonetheless concerning.
-%\tbc{I like the basic approach, this last segue could be cleaner however.}
 
@@
 When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
-Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans does not eliminate the possibility that a more transmissible variant of MERS-CoV will emerge.
+Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
 We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.
@@
 \subsection*{Data availability}
-Sequence data and all analytical code is publicly available at https://github.com/blab/structured-mers.
+Sequence data and all analytical code is publicly available at \href{https://github.com/blab/structured-mers}{github.com/blab/structured-mers}.
 
@@
 
-\begin{longtable}{ | l | l | l | l | l | }
+\begin{longtable}{ | r | l | p{2cm} | l | l | }
 
@@
 	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
-	\caption{\textbf{Result are robust to prior choice.}
+	\caption{\textbf{Posterior migration rate estimates for two choices of prior.}
 Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.

### 2017-08-02 mers-structure.tex
Revise Methods.
Resolves #7.

@@
 
-MultiTypeTree module \citep{vaughan_efficient_2014} was used in BEAUti v2.4.3 \citep{bouckaert_beast_2014} to specify a structured coalescent model with two demes -- humans and camels.
+For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
 This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
@@
 Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-All analyses were run across for 200 million states across ten independent Markov chains (MCMC runs).
-%\tbc{What was burn-in? How many steps in the MCMC? How was convergence assessed? How were chains combined?}
-Due to the increased complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
+All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every \tbc{XXX} steps.
+Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
 Three chains out of ten did not converge and were discarded altogether.
-To confirm that demographic inference was derived from the data rather than the prior we set up an additional ten independent analyses where the mean of the exponential distribution prior for migration rate was set to 10.0.
-For these analyses two chains did not converge.
+This left \tbc{XXX} states on which to base posterior inference.
 
-As a secondary analysis, we also considered a scenario where alignments split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+As a secondary analysis to test robustness to choice of prior, we set up an analysis where we increased the mean of the exponential distribution prior for migration rate to 10.0.
+All other parameters were identical to the primary analysis and as before 10 independent MCMC chains were run.
+In this case, two chains did not converge.
+This left \tbc{XXX} states on which to base posterior inference.
+
+As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
 Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
@@
 Three chains out of ten did not converge.
+This left \tbc{XXX} states on which to base posterior inference.
 
-\subsection*{Demographic inference of MERS-CoV in the reservoir}
-In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
-The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
-These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
-Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
-A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity (N$_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
-We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
-This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.%\lmc{already reported above.
-Why not remove?} \gdc{Never hurts to repeat.}
-Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
-This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
+\subsubsection*{Introduction seasonality}
+
+We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
+This distribution of introduction times was then discretised as follows: for sample  $k = 1, 2, \ldots, L$ from the posterior,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
+We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
+\begin{align*}
+  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
+  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
+  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
+  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
+  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
+  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
+  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
+\end{align*}
+
+Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
+\tbc{Please spell out what parameters are referring to. In English, what is $\alpha_j$ and what is $\beta_i$? I'm having a hard time figuring out what's going on here.}
+\tbc{Can you move `\textit{logit}' to the right hand side of the equation (if not too messy)?}
 
@@
 
-\subsubsection*{Introduction seasonality}
-
-We extracted the times of camel-to-human introductions from the posterior distribution of multitype trees.
-This distribution of introduction times was then discretised as follows: for state  $k = 1, 2, \ldots, L$ in the chain,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
-We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
-\begin{align*}
-  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
-  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
-  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
-  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
-  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
-  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
-  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
-\end{align*}
-
-Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
+\subsection*{Demographic inference of MERS-CoV in the reservoir}
+In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
+The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
+These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
+Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
+A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
+We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
+This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
+Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
+\tbc{This statement seemd a bit strong, as it looked like the $N_{e}\tau$ trajectory was highly similar between these different stationary distributions.}
+This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
 

### 2017-08-02 mers-structure.tex
Text changes, add skygrid as main figure.

@@
 
-Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters \tbc{reference a figure}.
+Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters (figure \ref{seasonality}C?
+\gdc{note - lettering might change}).
 This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
@@
 \subsection*{Recombination shapes MERS-CoV diversity}
-Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \citep{lai_1985,makino_1986,keck_1988,kottier_1995,herrewegh_1998}. %\tbc{needs citations}.
+Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \citep{lai_1985,makino_1986,keck_1988,kottier_1995,herrewegh_1998}.
 In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
@@
 We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
-\tbc{Summarize Figure \ref{recombinant_features}A with something like: ``We find that 95\% of human viruses fall within the same introduction event in genomic fragment 1 and in genomic fragment 2.''}
+Between the two trees in figure \ref{recombinant_features}B four `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
+For the reverse comparison there are five `human' clades in fragment 2 that are expanded.
+All such clades have low divergence (figure \ref{recombinant_features}B) and thus incongruences in human clades are more likely to be caused by differences in deme assignment rather than actual recombination.
 And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
@@
 % For sequences collected from humans and inferred to be in the same cluster we only keep the earliest sequence, and arrive at a sequence dataset that for the most part is comprised of viral diversity present in camels, and supplemented with human sequences that are treated as sentinel samples.
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid.png}
+	\caption{\textbf{Demographic history of MERS-CoV in Arabian peninsula camels.}
+Demographic history of MERS-CoV in camels, as inferred via a skygrid coalescent tree prior \citep{gill_2013}.
+Three skygrid reconstructions are shown, red and orange for each of the stationary distributions reached by MCMC with the whole genome and a black one where the genome was split into ten partitions.
+Shaded interval indicates the 95\% highest posterior density interval for the product of generation time and effective population size, $N_{e}\tau$.
+Midline tracks the inferred median of $N_{e}\tau$.
+%Maximum clade credibility (MCC) tree from the skygrid was inferred is shown in the background, with camel sequences highlighted in orange and human sequences highlighted in blue.
+	}
+	\label{skygrid}
+\end{figure}
 
@@
 Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every \tbc{XXX} steps.
+All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every 20 000 steps.
 Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
 Three chains out of ten did not converge and were discarded altogether.
-This left \tbc{XXX} states on which to base posterior inference.
+This left 70 000 states on which to base posterior inference.
 
@@
 In this case, two chains did not converge.
-This left \tbc{XXX} states on which to base posterior inference.
+This left 80 000 states on which to base posterior inference.
 
-As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-29364), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
 Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
@@
 Three chains out of ten did not converge.
-This left \tbc{XXX} states on which to base posterior inference.
+This left 70 000 states on which to base posterior inference.
 
@@
 This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
-Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
-\tbc{This statement seemd a bit strong, as it looked like the $N_{e}\tau$ trajectory was highly similar between these different stationary distributions.}
+Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto subtly different tree topologies.
 This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
@@
 
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid.png}
-	\caption{\textbf{Demographic history of MERS-CoV in Arabian peninsula camels.}
-Demographic history of MERS-CoV in camels, as inferred via a skygrid coalescent tree prior \citep{gill_2013}.
-Two skygrid reconstructions are shown, one for each of the stationary distributions reached by MCMC.
-Shaded interval indicates the 95\% highest posterior density interval for the product of generation time and effective population size, $N_{e}\tau$.
-Midline tracks the inferred median of $N_{e}\tau$.
-Maximum clade credibility (MCC) tree from the skygrid was inferred is shown in the background, with camel sequences highlighted in orange and human sequences highlighted in blue.
-	}
-	\label{skygrid}
-\end{figure}
-
 \end{document}

### 2017-08-03 mers-structure.tex
Incorporate skygrid comparison figure.

@@
 Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
-A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
+A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}, under a relaxed molecular clock with rates drawn from a lognormal distribution \citep{drummond_2006} and codon position partitioned (positions $1+2$ and $3$) HKY$+\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models.
 We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
@@
 This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.
+
+To ensure that recombination was not interfering with the skygrid reconstruction we also split our MERS-CoV alignment into ten parts 2937 nucleotides long.
+These were then used as separate partitions with independent trees and clock rates in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
+Nucleotide substitution and relaxed clock models were set up identically to the whole genome skygrid analysis described above \citep{drummond_2006,hky_1985,yang_1994}.
+Skygrid coalescent tree prior \citep{gill_2013} was used jointly across all ten partitions for demographic inference.
+Five MCMC chains were set up, each running for 200 million states, sampling every 20 000 states.
 
@@
 
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid_comparison.png}
+	\caption{\textbf{Skygrid comparison between whole and fragmented genomes.}
+Inferred median $N_{e}\tau$ recovered using a skygrid tree prior on whole genome (bottom) and ten genomic fragments with independent trees (left), coloured by time.
+Dotted line indicates the one-to-one line.
+	}
+	\label{skygrid_comparison}
+\end{figure}
+
 \end{document}

### 2017-08-03 mers-structure.tex
Minor text changes.

@@
 These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
-Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
+Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014,quick_rapid_2015}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
 Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
@@
 We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
-Between the two trees in figure \ref{recombinant_features}B four `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
-For the reverse comparison there are five `human' clades in fragment 2 that are expanded.
+Between the two trees in figure \ref{recombinant_features}B four (out of 54) `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
+For the reverse comparison there are five `human' clades (out of 53) in fragment 2 that are expanded.
 All such clades have low divergence (figure \ref{recombinant_features}B) and thus incongruences in human clades are more likely to be caused by differences in deme assignment rather than actual recombination.
@@
 For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
-This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
+This model estimates four parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
 Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.

### 2017-08-05 mers-structure.tex
Add N vs Ne text

@@
 Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
-However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+However, we do note that estimates of relative genetic diversity are relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years, and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+This estimate of $N_e\tau$ can be translated into an estimate of effective prevalence $N_e$ given an estimate of generation time or serial intervals.
+Given a 20 day duration of infection in camels \tbc{please add citation}, we expect a 10 day serial interval or $\tau$ of 0.027 years per generation.
+Dividing by $\tau$ yields an estimate of $N_e$ of 129.3 individuals.
+
+Here, we compare this estimate of effective prevalence to a rough estimate of census prevalence.
+Given extremely high seroprevalence estimates within camels in Saudi Arabia \tbc{please add citation}, we expect camels to usually be infected within their first year of life.
+Therefore we can estimate the rough number of camel infections per year as the number of calves produced each year.
+We find there are 270 000 camels in Saudi Arabia and that in camel farming practices a camel female can be bred for about 22 years and about 10 calves can be produced within the 22 year period \tbc{cite Elmi 1998}.
+This gives an expectation of each female camel producing, on average, 0.45 calves per year.
+Thus, we expect $270000 \times 0.5 \times 0.45 = 60750$ new calves produced yearly and correspondingly 60750 new infections every year, which spread over 10 day serial intervals gives an average of $N = 1660$ infections.
+
 This raises the possibility that MERS-CoV populations undergo lineage turnover at a rate not often seen in other livestock viruses (\textit{e.g.} swine influenza A viruses \citep{vijaykrishna_long-term_2011}).

### 2017-08-05 mers-structure.tex
Finish paragraph on N vs Ne

@@
 Here we attempt to investigate MERS-CoV demographic patterns in the camel reservoir.
-%Since MERS-CoV in humans has a distinct epidemiology
-% evolves under a distinct demographic regime, %% evolves under a demographic regime?
-%we split our sequence dataset based on whether sequences were collected in humans or camels.
 We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
@@
 Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
-However, we do note that estimates of relative genetic diversity are relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years, and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+However, we do note that estimates of relative genetic diversity are relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
 This estimate of $N_e\tau$ can be translated into an estimate of effective prevalence $N_e$ given an estimate of generation time or serial intervals.
 Given a 20 day duration of infection in camels \tbc{please add citation}, we expect a 10 day serial interval or $\tau$ of 0.027 years per generation.
-Dividing by $\tau$ yields an estimate of $N_e$ of 129.3 individuals.
+Dividing by $\tau$ yields an estimate of $N_e$ of 129 (100--162) individuals.
 
@@
 Therefore we can estimate the rough number of camel infections per year as the number of calves produced each year.
-We find there are 270 000 camels in Saudi Arabia and that in camel farming practices a camel female can be bred for about 22 years and about 10 calves can be produced within the 22 year period \tbc{cite Elmi 1998}.
-This gives an expectation of each female camel producing, on average, 0.45 calves per year.
-Thus, we expect $270000 \times 0.5 \times 0.45 = 60750$ new calves produced yearly and correspondingly 60750 new infections every year, which spread over 10 day serial intervals gives an average of $N = 1660$ infections.
-
-This raises the possibility that MERS-CoV populations undergo lineage turnover at a rate not often seen in other livestock viruses (\textit{e.g.} swine influenza A viruses \citep{vijaykrishna_long-term_2011}).
-Part of this could be caused by MERS-CoV ability to re-infect camels previously exposed to the virus \citep{ali_systematic_2017}, which may result in a process akin to antigenic drift, where the depletion of susceptible calves leads to ever increasing selective pressure for the virus to evolve antigenically.
-While this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.
-\tbc{Include results of N and S to indicate that surface protein is evolving quickly supporting antigenic evolution. Or not, if the data says otherwise...}
-
-\tbc{Okay for me to leave Figure \ref{skygrid} as supplementary or it can be main text.
-I'm happy either way.
-Suggest second panel showing correspondence between fragmented genome and full genome trajectories.}
-
-% In order to disentangle distinct demographic modes within the phylogeny of MERS-CoV we split our alignment according to whether the virus was collected from camels or humans.
-% For sequences collected from humans and inferred to be in the same cluster we only keep the earliest sequence, and arrive at a sequence dataset that for the most part is comprised of viral diversity present in camels, and supplemented with human sequences that are treated as sentinel samples.
+We find there are $830\,000$ camels in Saudi Arabia \citep{abdallah_camel_farming_2013} and that female camels in Saudi Arabia have an average fecundity of 45\% \citep{abdallah_camel_farming_2013}.
+Thus, we expect $830\,000 \times 0.50 \times 0.45 = 186\,750$ new calves produced yearly and correspondingly $186\,750$ new infections every year, which spread over 10 day serial intervals gives an average of $N = 5042$ infections.
+
+We thus arrive a ratio of $N/N_e = 39.1$ (31.1--50.4).
+This is less than the equivalent ratio for human measles virus \citep{bedford_strength_2011} and is in line with the expectation from neutral evolutionary dynamics plus some degree of transmission heterogeneity \citep{volz_phylodynamics_2013} and seasonal troughs in prevalence.
+Thus, we believe that the ladder-like appearance of the MERS-CoV tree can likely be explained by entirely demographic factors.
+
+\tbc{Say something like: ``We find that this result is robust to recombination by...'' and reference Figure \ref{skygrid_comparison}.}
 

### 2017-08-06 mers-structure.tex
Move figures within .tex, add citations, combine figs.

@@
 
-\usepackage{lineno}
-\linenumbers
+%\usepackage{lineno}
+%\linenumbers
 
@@
 % comments
-\usepackage{ulem}
+%\usepackage{ulem}
 \definecolor{purple}{rgb}{0.459,0.109,0.538}
@@
 
-We believe that the small number of inferred human-to-camel migrations are induced by the migration rate prior, while some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
-This is most apparent in lineages in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot completely dismiss humans as a potential alternative host.
-The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
-Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).
-
-The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
-We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
-These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
-Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
-This suggests that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
-
 \begin{figure}[h]
@@
 
+We believe that the small number of inferred human-to-camel migrations are induced by the migration rate prior, while some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
+This is most apparent in lineages in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot completely dismiss humans as a potential alternative host.
+The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
+Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).
+
+The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
+We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
+These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
+Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
+This suggests that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
+
 \subsection*{MERS-CoV shows seasonal introductions}
@@
 Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.
-
-Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters (figure \ref{seasonality}C?
-\gdc{note - lettering might change}).
-This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
-Similarly, we do not observe any trend in human sequence cluster sizes over time (Figure \ref{seasonality}B, Spearman $\rho = 0.06$, $p=0.68$), suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
-This is not surprising either, since MERS-CoV is a camel virus that has to date, experienced little-to-no selective pressure to improve transmissibility between humans.
 
@@
 
+Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters (figure \ref{seasonality}B).
+This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
+Similarly, we do not observe any trend in human sequence cluster sizes over time (Figure \ref{seasonality}B, Spearman $\rho = 0.06$, $p=0.68$), suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
+This is not surprising either, since MERS-CoV is a camel virus that has to date, experienced little-to-no selective pressure to improve transmissibility between humans.
+
 \subsection*{MERS-CoV is poorly suited for human transmission}
@@
 
-We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
-For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
-While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
-Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
-Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
-Model averaging would suggest plausible $R_0$ values between 0.57 and 0.91.
-
-Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
-The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
-These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Model averaging would suggest plausible numbers of introductions between 299 and 818.
-Our results suggest a large number of unobserved MERS primary cases.
-Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
-In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
-
-Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.91$ and sequence sampling likely to be biased towards large hospital outbreaks.
-Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
-
 \begin{figure}[h]
@@
 
+We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
+For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
+While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
+Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
+Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
+Model averaging would suggest plausible $R_0$ values between 0.57 and 0.91.
+
+Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
+The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
+These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
+Model averaging would suggest plausible numbers of introductions between 299 and 818.
+Our results suggest a large number of unobserved MERS primary cases.
+Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
+In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
+
+Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.91$ and sequence sampling likely to be biased towards large hospital outbreaks.
+Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
+
 \subsection*{Recombination shapes MERS-CoV diversity}
@@
 3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.
+
+\begin{figure}%
+    \centering
+    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
+    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
+    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
+    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to 29364.
+    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
+    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
+    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
+    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
+  	Taxa descended from the same introduction in fragment 1 are connected to their counterparts in fragment 2 with combinations of line dotting and colour unique to each introduction clade.
+  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
+  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
+  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
+    }
+    \label{recombinant_features}
+\end{figure}
 
@@
 % \end{figure}
-\begin{figure}%
-    \centering
-    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
-    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
-    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
-    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to 29364.
-    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
-    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
-    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
-    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
-  	Identical taxa are connected via coloured lines with colour indicating vertical position in the tree on the right.
-  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
-  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
-  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
-    }
-    \label{recombinant_features}
-\end{figure}
 
@@
 This removes conflicting demographic signals sampled during human outbreaks, where densely sampled closely related sequences from humans could be misconstrued as evidence of demographic crash in the viral population.
-
-Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
-However, we do note that estimates of relative genetic diversity are relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
-This estimate of $N_e\tau$ can be translated into an estimate of effective prevalence $N_e$ given an estimate of generation time or serial intervals.
-Given a 20 day duration of infection in camels \tbc{please add citation}, we expect a 10 day serial interval or $\tau$ of 0.027 years per generation.
-Dividing by $\tau$ yields an estimate of $N_e$ of 129 (100--162) individuals.
-
-Here, we compare this estimate of effective prevalence to a rough estimate of census prevalence.
-Given extremely high seroprevalence estimates within camels in Saudi Arabia \tbc{please add citation}, we expect camels to usually be infected within their first year of life.
-Therefore we can estimate the rough number of camel infections per year as the number of calves produced each year.
-We find there are $830\,000$ camels in Saudi Arabia \citep{abdallah_camel_farming_2013} and that female camels in Saudi Arabia have an average fecundity of 45\% \citep{abdallah_camel_farming_2013}.
-Thus, we expect $830\,000 \times 0.50 \times 0.45 = 186\,750$ new calves produced yearly and correspondingly $186\,750$ new infections every year, which spread over 10 day serial intervals gives an average of $N = 5042$ infections.
-
-We thus arrive a ratio of $N/N_e = 39.1$ (31.1--50.4).
-This is less than the equivalent ratio for human measles virus \citep{bedford_strength_2011} and is in line with the expectation from neutral evolutionary dynamics plus some degree of transmission heterogeneity \citep{volz_phylodynamics_2013} and seasonal troughs in prevalence.
-Thus, we believe that the ladder-like appearance of the MERS-CoV tree can likely be explained by entirely demographic factors.
-
-\tbc{Say something like: ``We find that this result is robust to recombination by...'' and reference Figure \ref{skygrid_comparison}.}
 
@@
 \end{figure}
+
+Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
+We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid_comparison}).
+However, we do note that estimates of relative genetic diversity are relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+This estimate of $N_e\tau$ can be translated into an estimate of effective prevalence $N_e$ given an estimate of generation time or serial intervals.
+Given a 20 day duration of infection in camels \citep{adney_replication_2014}, we expect a 10 day serial interval or $\tau$ of 0.027 years per generation.
+Dividing by $\tau$ yields an estimate of $N_e$ of 129 (100--162) individuals.
+
+Here, we compare this estimate of effective prevalence to a rough estimate of census prevalence.
+Given extremely high seroprevalence estimates within camels in Saudi Arabia \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}, we expect camels to usually be infected within their first year of life.
+Therefore we can estimate the rough number of camel infections per year as the number of calves produced each year.
+We find there are $830\,000$ camels in Saudi Arabia \citep{abdallah_camel_farming_2013} and that female camels in Saudi Arabia have an average fecundity of 45\% \citep{abdallah_camel_farming_2013}.
+Thus, we expect $830\,000 \times 0.50 \times 0.45 = 186\,750$ new calves produced yearly and correspondingly $186\,750$ new infections every year, which spread over 10 day serial intervals gives an average of $N = 5042$ infections.
+
+We thus arrive a ratio of $N/N_e = 39.1$ (31.1--50.4).
+This is less than the equivalent ratio for human measles virus \citep{bedford_strength_2011} and is in line with the expectation from neutral evolutionary dynamics plus some degree of transmission heterogeneity \citep{volz_phylodynamics_2013} and seasonal troughs in prevalence.
+Thus, we believe that the ladder-like appearance of the MERS-CoV tree can likely be explained by entirely demographic factors.
 
@@
 Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
-\tbc{Please spell out what parameters are referring to. In English, what is $\alpha_j$ and what is $\beta_i$? I'm having a hard time figuring out what's going on here.}
-\tbc{Can you move `\textit{logit}' to the right hand side of the equation (if not too messy)?}
+%\tbc{Please spell out what parameters are referring to. In English, what is $\alpha_j$ and what is $\beta_i$? I'm having a hard time figuring out what's going on here.}
+%\tbc{Can you move `\textit{logit}' to the right hand side of the equation (if not too messy)?}
 

### 2017-08-06 mers-structure.tex
Revise Seasonality Methods section
Fixes #12

@@
 This distribution of introduction times was then discretised as follows: for sample  $k = 1, 2, \ldots, L$ from the posterior,  $Z_{ijk}$ was $1$ if there as an introduction in month $i$ and year $j$ and $0$ otherwise.
-We model the variable $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
+We model the sum of introductions at month $i$ and year $j$ across the posterior sample $Y_{ij} = \sum_{k = 1}^L Z_{ijk}$ with the hierarchical model:
 \begin{align*}
-  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}); \\
-  logit(\theta_{ij}) &= \alpha_j + \beta_i; \\
-  \beta_i &\sim \text{Normal}(0, \sigma_{m}); \\
-  \sigma_{m} &\sim \text{Cauchy}(0, 2.5); \\
-  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}); \\
-  \mu_{y}  &\sim  \text{Normal}(0, 1); \\
-  \sigma_{y} &\sim \text{Cauchy}(0, 2.5).
+  Y_{ij} &\sim \text{Binomial}(L, \theta_{ij}) \\
+  \theta_{ij} &= \mathrm{logistic}(\alpha_j + \beta_i) \\
+  \alpha_j &\sim \text{Normal}(\mu_{y}, \sigma_{y}) \\
+  \mu_{y}  &\sim  \text{Normal}(0, 1) \\
+  \sigma_{y} &\sim \text{Cauchy}(0, 2.5) \\
+  \beta_i &\sim \text{Normal}(0, \sigma_{m}) \\
+  \sigma_{m} &\sim \text{Cauchy}(0, 2.5), \\
 \end{align*}
-
-Odds ratios of introductions can then be computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
+where $\alpha_j$ represents the contribution of year to expected introduction count and $\beta_i$ represents the contribution of month to expected introduction count.
+Here, $\mathrm{logistic}(\alpha_j + \beta_i) =  \frac{\mathrm{exp}(\alpha_j + \beta_i)}{\mathrm{exp}(\alpha_j + \beta_i) + 1}$.
+We sampled posterior values from this model via the Markov chain Monte Carlo methods implemented in Stan \citep{carpenter_stan_2016}.
+Odds ratios of introductions were computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
 %\tbc{Please spell out what parameters are referring to. In English, what is $\alpha_j$ and what is $\beta_i$? I'm having a hard time figuring out what's going on here.}

### 2017-08-07 mers-structure.tex
Revamp of Monte Carlo Methods section

@@
 Middle East respiratory syndrome coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans.
-The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{who_2017}.
+The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{who_mers_summary_2017}.
 Its epidemiology remains obscure, largely because infections are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
@@
 
-It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \citep{who_2016_MERS}.
+It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \citep{who_mers_guidelines_2016}.
 Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional epidemiological approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
@@
 
-Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters (figure \ref{seasonality}B).
+Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters (Figure \ref{seasonality}B).
 This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
@@
 Odds ratios of introductions were computed for each month $i$ as $\text{OR}_i = \exp(\beta_i)$.
-%\tbc{Please spell out what parameters are referring to. In English, what is $\alpha_j$ and what is $\beta_i$? I'm having a hard time figuring out what's going on here.}
-%\tbc{Can you move `\textit{logit}' to the right hand side of the equation (if not too messy)?}
 
@@
 
-As of 10 May 2017, the World Health Organization has been notified of 1952 cases of MERS-CoV.
-We thus simulate final transmission chain sizes using equation \ref{clusterLikelihood} \citep{lloyd-smith_superspreading_2005,blumberg_inference_2013} until we reach an epidemic comprised of 2000 cases.
-10 000 simulations were run for 121 uniformly spaced values of R$_{0}$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1.
-Each simulation results in a vector of outbreak sizes $C$, where $C_{i}$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} C_{i} = 2000$, where $K$ is the number of clusters generated.
-We sample from the case cluster size vector according to a multivariate hypergeometric distribution to simulate sequencing (algorithm \ref{hypergeometric}).
-The resulting sequence cluster size vector $S$ contains $K$ entries, some of which are zero (\textit{i.e.} case clusters not sequenced), but $\sum_{i=1}^{K} S_{i} = 174$ which reflects the number of MERS-CoV sequences used in this study.
-Since the sampling scheme operates under equi-probable sequencing we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
-This makes clusters likely to be `sequenced' even more likely to be sequenced and \textit{vice versa}.
-We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
-
-Let $R_{0}$ be the basic reproductive number ($ 0 < R_{0} < 1$),  and $\omega > 0$ be the a dispersion parameter, then the probability of observing a stuttering chain (cluster) $r$ of size $j$ is~\citep{blumberg_inference_2013}
+Here, we emply a Monte Carlo simulation approach to identify parameters consistent with observed patterns of sequence clustering (Figure \ref{mc_method}).
+Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species spillover of the virus from camels into humans.
+The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying tranmission process.
+At heart, we expect fewer larger clusters if fundamental reproductive number $R_0$ is large and more smaller clusters if $R_0$ is small.
+A similar approach was used in \citet{grubaugh_multiple_2017} to estimate $R_0$ for Zika introductions into Florida.
+
+Branching process theory provides an analytical distribution for the number of eventual cases $j$ in a transmission chain resulting from a single introduction event with $R_0$ and dispersion parameter $\omega$ \citep{blumberg_inference_2013}.
+This distribution follows
 \begin{equation}
-Pr(r = j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j)\Gamma(j+1)} \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}.
+\mathrm{Pr}(j | R_{0}, \omega) = \frac{\Gamma(\omega j+j-1)}{\Gamma(\omega j) \, \Gamma(j+1)} \; \frac{(\frac{R_{0}}{\omega})^{j-1}}{(1+\frac{R_{0}}{\omega})^{\omega j+j-1}}.
 \label{clusterLikelihood}
 \end{equation}
-
-Although case clusters and their sizes are difficult to infer directly and require detailed epidemiological follow up, sequence data have fewer limitations.%\lmc{other than the ``obvious'' limitation that they're \textbf{not} the actual infection process, uh?
->.>} \gdc{fair point, but we're talking about cluster inference}
-Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species transmission of the virus from camels into humans.
-The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying distribution of case clusters.
-We employ a Monte Carlo simulation approach to identify simulations where the recovered distribution of sequence cluster sizes fall within the 95\% highest posterior density intervals for three summary statistics of MERS-CoV sequence cluster sizes recovered via structured coalescent analyses: mean, median and standard deviation.
-These values are 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
+Here, $R_0$ represents the expected number of secondary cases following a single infection and $\omega$ represents the dispersion parameter assuming secondary cases follow a negative binomial distribution \citep{lloyd-smith_superspreading_2005}, so that smaller values represent larger degrees of heterogeneity in the transmission process.
+
+As of 10 May 2017, the World Health Organization has been notified of 1952 cases of MERS-CoV \citep{who_mers_summary_2017}.
+We thus simulated final transmission chain sizes using Equation \ref{clusterLikelihood} until we reached an epidemic comprised of $N=2000$ cases.
+$10\,000$ simulations were run for 121 uniformly spaced values of $R_0$ across the range [0.5--1.1] with dispersion parameter $\omega$ fixed to 0.1 following expectations from previous studies of coronavirus behavior \citep{lloyd-smith_superspreading_2005}.
+Each simulation results in a vector of outbreak sizes $\mathbf{c}$, where $c_i$ is the size of the \textit{i}th transmission cluster and $\sum_{i=1}^{K} c_i = 2000$ and $K$ is the number of clusters generated.
+
+Following the underlying transmission process generating case clusters $\mathbf{c}$ we simulate a secondary process of sampling some fraction of cases and sequencing them to generate data analogous to what we empirically observe.
+We sample from the case cluster size vector $\mathbf{c}$ without replacement according to a multivariate hypergeometric distribution (Algorithm \ref{hypergeometric}).
+The resulting sequence cluster size vector $\mathbf{s}$ contains $K$ entries, some of which are zero (\textit{i.e.}\ case clusters not sequenced), but $\sum_{i=1}^{K} s_i = 174$ which reflects the number of human MERS-CoV sequences used in this study.
+As the default sampling scheme operates under equiprobable sequencing, we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
+Here, bias enriches the hypergeometric distribution so that sequences are sampled with weights proportional to $(c_1^\mathrm{bias}, c_2^\mathrm{bias}, \ldots, c_k^\mathrm{bias})$.
+Thus, bias makes larger clusters more likely to be `sequenced'.
+
+After simulations were completed, we identified simulations in which the recovered distribution of sequence cluster sizes $\mathbf{s}$ fell within the 95\% highest posterior density intervals for three summary statistics of empirical MERS-CoV sequence cluster sizes recovered via structured coalescent analysis: mean, median and standard deviation.
+These values were 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
+
+Additionally, we performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
 
 \begin{algorithm}[H]
- \KwData{Array of case cluster sizes in outbreak $\boldsymbol C = [C_{1},C_{2}, \ldots, C_{K}]$, sequences available $M$, total outbreak size $\sum_{i=1}^{K} C_{i}$.}
- \KwResult{Array of sequence cluster sizes sampled: $\boldsymbol S = [S_{1},S_{2}, \ldots, S_{K}]$.}
- Draw $S_{i}$ from a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
+ \KwData{Array of case cluster sizes in outbreak $\mathbf{c} = (c_1, c_2, \ldots, c_K)$, sequences available $M$, total outbreak size $N$, where $N = \sum_{i=1}^{K} c_{i}$.}
+ \KwResult{Array of sequence cluster sizes sampled: $\mathbf{s} = (s_1, s_2, \ldots, s_K)$.}
+ Draw $s_i$ from a hypergeometric distribution with $c_i$ successes, $N-c_i$ failures after $M$ trials\;
  \While{$i < K$}{
   $i = i + 1$\;
-  $M = M - S_{i-1}$\;
-  Compute the probability mass function (pmf) for all possible values of $S_i$, $\boldsymbol p = [p(0)^{\text{bias}}, p(1)^{\text{bias}}, \ldots, p(C_{i})^{\text{bias}}] \times (\sum_i p_i^{\text{bias}}) ^{-1}$, where $p(\cdot)$  is the pmf for a hypergeometric distribution with $C_{i}$ successes, $\sum_{i=1}^{K} C_{i}-C_{i}$ failures after $M$ trials\;
-  Draw a sequence cluster size $S_{i}$ from array of potential sequence cluster sizes $[0, 1, \ldots, C_{i}]$ according to $\boldsymbol p$\;
- }
- Add remaining sequences to last sequence cluster $C_{K} = M - S_{K-1}$\;
+  $M = M - s_{i-1}$\;
+  Compute the probability mass function (pmf) for all possible values of $s_i$, $\mathbf{p} = (p(0)^{\mathrm{bias}}, p(1)^{\mathrm{bias}}, \ldots, p(c_i)^{\mathrm{bias}}) \times (\sum_i p_i^{\text{bias}}) ^{-1}$, where $p(\cdot)$ is the pmf for a hypergeometric distribution with $c_i$ successes, $N-c_i$ failures after $M$ trials\;
+  Draw a sequence cluster size $s_i$ from array of potential sequence cluster sizes $(0, 1, \ldots, c_i)$ according to $\boldsymbol p$\;
+ }
+ Add remaining sequences to last sequence cluster $c_K = M - s_{K-1}$\;
  \caption{\textbf{Multivariate hypergeometric sampling scheme.}
  Pseudocode describes the multivariate hypergeometric sampling scheme that simulates sequencing.
- Probability of sequencing a given number of cases from a case cluster depends on cluster size and sequences left (\textit{i.e.} ``sequencing capacity'').
+ Probability of sequencing a given number of cases from a case cluster depends on cluster size and sequences left (\textit{i.e.}\ ``sequencing capacity'').
  The bias parameter determines how probability mass function of the hypergeometric distribution is concentrated.

### 2017-08-07 mers-structure.tex
Minor fixes to spelling, numbering and spacing

--- 2017-08-06 mers-structure.tex
@@
 Correspondingly, we observe substantially higher camel-to-human migration rate estimates than human-to-camel migration rate estimates (Figure~\ref{prior}).
-This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
+This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show significantly higher diversity and less clustering.
 This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of 3.49 years (95\% HPD: 2.71--4.38) and human viruses showing a scaled effective population of 0.24 years (95\% HPD: 0.14--0.34).
@@
 We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
-A schematic of this Monte Carlo produre is shown in Figure \ref{mc_method}.
+A schematic of this Monte Carlo procedure is shown in Figure \ref{mc_method}.
 Generally, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters (Figure \ref{mers_epi}).
@@
 All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
-Sequences without accessions were kindly shared by Ali M.
-Somily, Mazin Barry, Sarah S.
-Al Subaie, Abdulaziz A.
-BinSaeed, Fahad A.
-Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N.
-McNabb, Imad A.
-Al-Jahdali, Ahmed M.
-Alotaibi, Nahid A.
-Batarfi, Matthew Cotten, Simon J.
-Watson, Spela Binter, and Paul Kellam prior to publication.
+Sequences without accessions were kindly shared by Ali M.\ Somily, Mazin Barry, Sarah S.\ Al Subaie, Abdulaziz A.\ BinSaeed, Fahad A.\ Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N.\ McNabb, Imad A.\ Al-Jahdali, Ahmed M.\ Alotaibi, Nahid A.\ Batarfi, Matthew Cotten, Simon J.\ Watson, Spela Binter, and Paul Kellam prior to publication.
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
@@
 Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
-All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every 20 000 steps.
+All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every $20\,000$ steps.
 Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
 Three chains out of ten did not converge and were discarded altogether.
-This left 70 000 states on which to base posterior inference.
+This left $70\,000$ states on which to base posterior inference.
 
@@
 In this case, two chains did not converge.
-This left 80 000 states on which to base posterior inference.
+This left $80\,000$ states on which to base posterior inference.
 
@@
 Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
-All analyses were set to run for 200 million states, subsampling every 20 000 states.
+All analyses were set to run for 200 million states, subsampling every $20\,000$ states.
 Chains not converging after 200 million states were discarded.
@@
 Three chains out of ten did not converge.
-This left 70 000 states on which to base posterior inference.
+This left $70\,000$ states on which to base posterior inference.
 
@@
 A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}, under a relaxed molecular clock with rates drawn from a lognormal distribution \citep{drummond_2006} and codon position partitioned (positions $1+2$ and $3$) HKY$+\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models.
-We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
+We set up five independent MCMC chains to run for 500 million states, sampling every $50\,000$ states.
 This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
@@
 Skygrid coalescent tree prior \citep{gill_2013} was used jointly across all ten partitions for demographic inference.
-Five MCMC chains were set up, each running for 200 million states, sampling every 20 000 states.
+Five MCMC chains were set up, each running for 200 million states, sampling every $20\,000$ states.
 
@@
 
-Ali M.
-Somily$^{1}$, Mazin Barry$^{1}$, Sarah S.
-Al Subaie$^{1}$, Abdulaziz A.
-BinSaeed$^{1}$, Fahad A.
-Alzamil$^{1}$, Waleed Zaher$^{1}$, Theeb Al Qahtani$^{1}$, Khaldoon Al Jerian$^{1}$, Scott J.N.
-McNabb$^{2}$, Imad A.
-Al-Jahdali$^{3}$, Ahmed M.
-Alotaibi$^{4}$, Nahid A.
-Batarfi$^{5}$, Matthew Cotten$^{6}$, Simon J.
-Watson$^{6}$, Spela Binter$^{6}$, Paul Kellam$^{6}$.
+Ali M.\ Somily$^{1}$, Mazin Barry$^{1}$, Sarah S.\ Al Subaie$^{1}$, Abdulaziz A.\ BinSaeed$^{1}$, Fahad A.\ Alzamil$^{1}$, Waleed Zaher$^{1}$, Theeb Al Qahtani$^{1}$, Khaldoon Al Jerian$^{1}$, Scott J.N.\ McNabb$^{2}$, Imad A.\ Al-Jahdali$^{3}$, Ahmed M.\ Alotaibi$^{4}$, Nahid A.\ Batarfi$^{5}$, Matthew Cotten$^{6}$, Simon J.\ Watson$^{6}$, Spela Binter$^{6}$, Paul Kellam$^{6}$.
 
@@
 $^{2}$Rollins School of Public Health, Emory University, Atlanta, GA, USA \\
-$^{3}$Deputy Minister. Ex. General Director King Fahad General Hospital, Jeddah and Occupational and environmental medicine, Um AlQura University, Kingdom of Saudi Arabia \\
+$^{3}$Deputy Minister.\ Ex.\ General Director King Fahad General Hospital, Jeddah and Occupational and environmental medicine, Um AlQura University, Kingdom of Saudi Arabia \\
 $^{4}$Department of Intensive Care, Prince Mohammed bin Abdulaziz Hospital, Riyadh, Kingdom of Saudi Arabia \\

### 2017-08-07 mers-structure.tex
Updating AR's affiliations and grants

@@
 \author[2]{Luiz Max Carvalho}
-\author[2,3,4]{Andrew Rambaut}
+\author[2,3]{Andrew Rambaut}
 \author[1]{Trevor Bedford}
@@
 \affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
-\affil[4]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}
 
@@
 TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
+AR was supported in part by the European Union Seventh Framework Programme for research, technological development and demonstration under Grant Agreement no.
+278433-PREDEMICS and no.
+725422-RESERVOIRDOCS, and the Wellcome Trust through project 206298/Z/17/Z.
+MAS is partly supported by NSF grant DMS 1264153 and NIH grants R01 HG006139 and R01 AI107034.
+We gratefully acknowledge support from NVIDIA Corporation with the donation of parallel computing resources used for this research.
+
 We gratefully acknowledge the contribution of the following scientists for sharing MERS-CoV sequence data before publication:

### 2017-10-19 mers-structure.tex
Updating AR's affiliations and grants
Fix issue #24

@@
 Here, we use an explicit model of metapopulation structure and migration between discrete subpopulations, referred to here as demes \citep{vaughan_efficient_2014}, derived from the structured coalescent \citep{notohara_structured_coalescent_1990}.
-Unlike approaches that model host species as a discrete phylogenetic trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) \citep{faria_simultaneously_2013,lycett_h5n8_2016}, population structure models explicitly incorporate distinct sampling patterns, population dynamics within demes and migration between demes.
+Unlike approaches that model host species as a discrete phylogenetic trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) \citep{faria_simultaneously_2013,lycett_h5n8_2016}, population structure models explicitly incorporate contrasts in deme effective population sizes and migration between demes.
 By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
@@
 725422-RESERVOIRDOCS, and the Wellcome Trust through project 206298/Z/17/Z.
-MAS is partly supported by NSF grant DMS 1264153 and NIH grants R01 HG006139 and R01 AI107034.
-We gratefully acknowledge support from NVIDIA Corporation with the donation of parallel computing resources used for this research.
 

### 2017-10-23 mers-structure.tex
Working on issue #23
Justifying some choices a bit better in text. Need another sentence to explicitly say why BASTA, MASCOT and PhyDyn were not used.

@@
 Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
-Only sequences covering at least 50\% of MERS-CoV genome were kept.
-Sequences were annotated with available collection dates and hosts, designated as camel or human and aligned with MAFFT \citep{katoh_mafft_2013} and edited manually.
-Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364.
+Only sequences covering at least 50\% of MERS-CoV genome were kept, to facilitate later analyses where the alignment is split into two parts in order to account for effects of recombination \citep{dudas_mers-cov_2016}.
+Sequences were annotated with available collection dates and hosts, designated as camel or human, aligned with MAFFT \citep{katoh_mafft_2013}, and edited manually.
+Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364, which allowed for codon-partitioned substitution models to be used.
 The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).
@@
 For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
-This model estimates four parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
+At time of writing structured population models are available in BEAST v2 \citep{bouckaert_beast_2014} but not in BEAST v1 \citep{drummond_bayesian_2012}.
+We use the more computationally intensive MultiTypeTree module \citep{vaughan_efficient_2014} over approximate methods also available in BEAST v2, such as BASTA \citep{maio_new_2015}, MASCOT \citep{mueller_mascot:_2017}, and PhyDyn \citep{volz_complex_2011}.
+Structured coalescent model implemented in the MultiTypeTree module \citep{vaughan_efficient_2014} estimates four parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
 Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
@@
 
-As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-29364), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
+Due to concerns that recombination might affect our conclusions \citep{dudas_mers-cov_2016}, as an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-29364), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
 Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
@@
 
-Here, we emply a Monte Carlo simulation approach to identify parameters consistent with observed patterns of sequence clustering (Figure \ref{mc_method}).
+Here, we employ a Monte Carlo simulation approach to identify parameters consistent with observed patterns of sequence clustering (Figure \ref{mc_method}).
 Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species spillover of the virus from camels into humans.
-The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying tranmission process.
+The distribution of the number of these cross-species transmissions and their sizes thus contain information about the underlying transmission process.
 At heart, we expect fewer larger clusters if fundamental reproductive number $R_0$ is large and more smaller clusters if $R_0$ is small.
@@
 In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
-The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
-These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
+The oldest sequence from each cluster introduced into humans was kept for further analysis.
+This procedure removes lineages coalescing rapidly in humans, which would otherwise introduce a strong signal of low effective population size.
+These subsampled MERS-CoV sequences from humans were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
 Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
 A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}, under a relaxed molecular clock with rates drawn from a lognormal distribution \citep{drummond_2006} and codon position partitioned (positions $1+2$ and $3$) HKY$+\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models.
+At time of writing advanced flexible coalescent tree priors from the skyline family, such as skygrid \citep{gill_2013} are available in BEAST v1 \citep{drummond_bayesian_2012} but not in BEAST v2 \citep{bouckaert_beast_2014}.
 We set up five independent MCMC chains to run for 500 million states, sampling every $50\,000$ states.

### 2017-10-23 mers-structure.tex
Address #22 somewhat.
Add discussion paragraph that's 50% "we will all die" and 50% "eh, it'll be fine". Enough to close issue?

@@
 Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
-Even though our results show that MERS-CoV does not appear to present a serious global threat, we would like to highlight that its epidemiology is nonetheless concerning.
+Even though our results show that MERS-CoV does not appear to present an imminent global threat, we would like to highlight that its epidemiology is nonetheless concerning.
 
@@
 Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
+Previous research \citep{antia_role_2003} suggests that zoonotic pathogens become disproportionately more likely to evolve towards $R_{0}>1.0$ if $R_{0}$ is close to $1.0$ in the new host to begin with, which leaves MERS-CoV in a precarious position: on one hand our estimate of $R_{0}$ at $\approx 0.7$ makes MERS-CoV a potential pandemic hazard, on the other nothing is known of fitness landscapes or adaptive potential of MERS-CoV.
 We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.

### 2017-10-23 mers-structure.tex
Fix #32

@@
 The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
-While traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
+While case-based and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
 Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
@@
 
-Although such traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
+Although such epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
 Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}, where data are relatively cheap to produce.
@@
 It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \citep{who_mers_guidelines_2016}.
-Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional epidemiological approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
+Previous studies attempting to quantify the actual number of spillover infections have either relied on case-based epidemiological approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
 Here we use a dataset of 274 MERS-CoV genomes to investigate transmission patterns of the virus between humans and camels.

### 2017-10-23 mers-structure.tex
Fix #33
Rewrite offending sentence.

@@
 Although such epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
-Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}, where data are relatively cheap to produce.
+Where sequence data are relatively cheap to produce, genomic epidemiological approaches can fill this critical gap in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}.
 These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.

### 2017-10-23 mers-structure.tex
Fix #35
R0 is not only for epidemiologists.

@@
 Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
-However, we wanted to translate this observation into an estimate of the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides insight into epidemic behavior.
+However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide a more intuitive epidemic behaviour metric.
 The parameter $R_{0}$ determines expected number of secondary cases in a single infections as well as the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}, Methods).

### 2017-10-23 mers-structure.tex
Fix #34
Don't do sequencing as diagnostics.

@@
 Where sequence data are relatively cheap to produce, genomic epidemiological approaches can fill this critical gap in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}.
-These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
+These data often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
 Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014,quick_rapid_2015}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.

### 2017-10-23 mers-structure.tex
Address #37
Might be fixed already.

@@
 The PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
-Another hallmark of recombination is spatial clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
-3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.
+Another hallmark of recombination is clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
+3Seq relies on the clustering of nucleotide similarities along the genome between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.
 

### 2017-10-25 mers-structure.tex
Addressing #29 and #25
Add k-justifying simulations and their summary figure. Switch mers_epi figure to show mean x skewness. Add Jupyter notebook with code to do revision sims + plot results, adjust estimates with what's in the notebook, update epi notebook itself (#29). Add more understandable detail on Monte Carlo (#25).

@@
 First, we simulate case counts across multiple outbreaks totaling 2000 cases using Equation \ref{clusterLikelihood} and then we subsample from each case cluster to simulate sequencing of a fraction of cases.
-Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
+Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
+In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones.
 This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
-We record the mean, median and standard deviation of sequence cluster sizes in each simulation (left-hand panels in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
+We record the number, mean, standard deviation and skewness (third standard moment) of sequence cluster sizes in each simulation (left-hand panels in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
 We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
@@
 	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
-Each row corresponds to a different bias value used to concentrate the subsampling of sequences from cases, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced).
-Leftmost scatter plots show results of individual Monte Carlo simulations, coloured by the $R_{0}$ value used for the simulation.
-The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for empirical MERS-CoV data.
-The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same $y$-axis across all levels of bias.
+From top to bottom each row corresponds to departures from completely random sequencing efforts with respect to case cluster size (bias parameter=1) to sequencing increasingly biased towards capturing large case clusters (bias=2, bias=3).
+Leftmost scatter plots show the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) that fell within the 95\% HPDs for number of sequence clusters, mean sequence cluster size, sequence cluster size standard deviation, and sequence cluster size skewness recovered for empirical data, coloured by the $R_{0}$ value used for the simulation.
+The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and skewness observed for empirical MERS-CoV data.
+The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same $y$-axis across all levels of the bias parameter.
 Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
@@
 We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
-For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
-While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
-Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
-Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
-Model averaging would suggest plausible $R_0$ values between 0.57 and 0.91.
+For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.84, 0.73, and 0.68, respectively.
+While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.70--0.97) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
+Under unbiased sequencing only 0.29\% of simulations fit our phylogenetic observations, while 1.2\% and 1.1\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
+Correspondingly, we estimate 11\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 44\% support for a model with bias level 3.0.
+Model averaging would suggest plausible $R_0$ values between 0.57 and 0.90.
 
 Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
-The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
-These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Model averaging would suggest plausible numbers of introductions between 299 and 818.
+The median number of cross-species introductions observed in simulations matching empirical data without bias are 344 (95\% percentiles 264--448).
+These numbers jump up to 567 (95\% percentiles 432--730) for bias = 2 and 654 (95\% percentiles 491--847) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
+Model averaging would suggest plausible numbers of introductions between 311 and 816.
 Our results suggest a large number of unobserved MERS primary cases.
@@
 
-After simulations were completed, we identified simulations in which the recovered distribution of sequence cluster sizes $\mathbf{s}$ fell within the 95\% highest posterior density intervals for three summary statistics of empirical MERS-CoV sequence cluster sizes recovered via structured coalescent analysis: mean, median and standard deviation.
-These values were 2.90--3.70 for mean sequence cluster size, 4.87--6.07 for standard deviation of sequence cluster sizes and a median sequence cluster size of 1.
-
-Additionally, we performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
+After simulations were completed, we identified simulations in which the recovered distribution of sequence cluster sizes $\mathbf{s}$ fell within the 95\% highest posterior density intervals for four summary statistics of empirical MERS-CoV sequence cluster sizes recovered via structured coalescent analysis: number of sequence clusters, mean, standard deviation and skewness (third central moment).
+These values were 48-61 for number of sequence clusters, 2.87--3.65 for mean sequence cluster size, 4.84--6.02 for standard deviation of sequence cluster sizes, and 415.40--621.06 for skewness of sequence cluster sizes.
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
+	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
+Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
+More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
+	}
+	\label{marginals}
+\end{figure}
+
+We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
+Additionally, we performed additional smaller set of simulations on a rougher grid of $R_{0}$ values (23 values, 0.50--1.05), with 5 values of dispersion parameter $\omega$ (0.002, 0.04, 0.1, 0.5, 1.0) and 3 levels of bias (1, 2, 3) to justify our choice of dispersion parameter $\omega$ that was fixed to 0.1 in the main analyses.
 

### 2017-10-30 mers-structure.tex
Address #20
Add DTA results and associated text. Might be it for this issue.

@@
 These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
-Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
-This suggests that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
+Our findings suggest that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
+Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
+Comparing these results with a currently standard discrete trait analysis \citep{lemey_bayesian_2009} approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure \ref{dta}).
+Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
+A similar approach has been attempted previously \citep{zhang_evolutionary_2016}, but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
+We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
 
 \subsection*{MERS-CoV shows seasonal introductions}
+
 We use the posterior distribution of MERS-CoV introduction events from camels to humans (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
@@
 \subsection*{Sequence data}
+
 All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
@@
 This left $70\,000$ states on which to base posterior inference.
+Posterior sets of typed (where migrating branches from structured coalescent are collapsed, and migration information is left as a switch in state between parent and descendant nodes) trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
 
@@
 This left $80\,000$ states on which to base posterior inference.
+Posterior sets of typed trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
 
@@
 This left $70\,000$ states on which to base posterior inference.
+Posterior sets of typed trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
+
+\subsubsection*{Discrete trait analysis}
+
+A currently widely used approach to infer ancestral states in phylogenies relies on treating traits of interest (such as geography, host, \textit{etc.}) as features evolving along a phylogeny as continuous time Markov chains with an arbitrary number of states \citep{lemey_bayesian_2009}.
+Unlike structured coalescent methods, such discrete trait approaches are independent from the tree (\textit{i.e.} demographic) prior and thus unable to influence coalescence rates under different trait states.
+Such models have been used in the past to infer the number of MERS-CoV host jumps \citep{zhang_evolutionary_2016} with results contradicting other sources of information.
+In order to test how a discrete trait approach compares to the structured coalescent we used our 274 MERS-CoV genome data set in BEAST v2.4.3 \citep{bouckaert_beast_2014} and specified identical codon-partitioned nucleotide substitution and molecular clock models to our structured coalescent analysis.
+To give the most comparable results we used a constant population size coalescent model, as this is the demographic function for each deme in the structured coalescent model.
+Five replicate MCMC analyses were run for 200 million states, three of which converged onto the same posterior distribution.
+The converging chains were combined after removing 20 million states as burn-in to give a total of $27\,000$ trees drawn from the posterior distribution.
+These trees were then summarised using TreeAnnotator v2.4.5 with the common ancestor heights option \citep{heled_looking_2013}.
 
@@
 These values were 48-61 for number of sequence clusters, 2.87--3.65 for mean sequence cluster size, 4.84--6.02 for standard deviation of sequence cluster sizes, and 415.40--621.06 for skewness of sequence cluster sizes.
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
-	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
-Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
-More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
-	}
-	\label{marginals}
-\end{figure}
 
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_dta_mcc.png}
+	\caption{\textbf{Maximum clade credibility (MCC) tree with ancestral state reconstruction according to a discrete trait model.}
+MCC tree is presented the same as Figure \ref{mcc}, with colours indicating the most probable state reconstruction at internal nodes.
+Unlike the structured coalescent summary shown in Figure \ref{mcc} where camels are reconstructed as the main host where MERS-CoV persists, the discrete trait approach identifies both camels and humans as major hosts with humans being the source of MERS-CoV infection in camels.
+	}
+	\label{dta}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
+	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
+Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
+More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
+	}
+	\label{marginals}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}

### 2017-10-31 mers-structure.tex
Fix #26

@@
 This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
-As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
+As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of genomic clustering of such alleles.
 % Homoplasies are present in multiple strains at a time, indicating recombination events that have been successful.
@@
 In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
+Overall, recombination presents a serious hurdle for detailed phylodynamic analyses in MERS-CoV, especially where lineages evolving in camels are concerned.
+Amongst other parameters of interest, recombination is expected to interfere with molecular clocks, where transferred genomic regions can give the impression of branches undergoing rapid evolution, or branches where recombination results in reversions appearing to evolve slow.
+In addition to its potential to influence tree topology, recombination in molecular sequence data is an erratic force with unpredictable effects.
+We suspect that the effects of recombination in MERS-CoV data are reigned in by a relatively small effective population size of the virus in Saudi Arabia (see next section) where haplotypes are fixed or nearly fixed, thus preventing an accumulation of genetic diversity that would then be reshuffled via recombination.
+Nevertheless, we choose not to report on any particular estimates for times of common ancestors (tMRCAs), even though these are expected to be somewhat robust for dating human clusters, and we do not report on the evolutionary rate of the virus, even though it appears to fall firmly within the expected range for RNA viruses (data not shown).
 

### 2017-10-31 mers-structure.tex
Fix #27
Rephrase sentence about low Ne.

@@
 We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid_comparison}).
-However, we do note that estimates of relative genetic diversity are relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
 This estimate of $N_e\tau$ can be translated into an estimate of effective prevalence $N_e$ given an estimate of generation time or serial intervals.

### 2017-10-31 mers-structure.tex
Address #29
Add sentence pointing out that sequencing capacity does not vary over time.
Fix #31

@@
 The resulting sequence cluster size vector $\mathbf{s}$ contains $K$ entries, some of which are zero (\textit{i.e.}\ case clusters not sequenced), but $\sum_{i=1}^{K} s_i = 174$ which reflects the number of human MERS-CoV sequences used in this study.
+Note that this ``sequencing capacity'' parameter does not vary over time, even though MERS-CoV sequencing efforts have varied in intensity, starting out slow due to lack of awareness, methods and materials and increasing in response to hospital outbreaks later.
 As the default sampling scheme operates under equiprobable sequencing, we also simulated biased sequencing by using concentrated hypergeometric distributions where the probability mass function is squared (bias = 2) or cubed (bias = 3) and then normalized.
@@
 These subsampled MERS-CoV sequences from humans were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
-Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
+Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in Saudi Arabia and Egypt are distinct epidemics with relatively poor sampling in the latter.
+Were more sequences of MERS-CoV available from other parts of Africa we speculate they would fall outside of the diversity that has been sampled in Saudi Arabia and cluster with early MERS-CoV sequences from Jordan and sequences from Egyptian camels.
+However, currently there are no indications of what MERS-CoV diversity looks like in camels east of Saudi Arabia.
 A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}, under a relaxed molecular clock with rates drawn from a lognormal distribution \citep{drummond_2006} and codon position partitioned (positions $1+2$ and $3$) HKY$+\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models.

### 2017-11-07 mers-structure.tex
Adressing #19 and #21

@@
 We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
+We can replicate these results by either applying the same model to current data (Figure \ref{dta}) or by enforcing equal coalescence rates within each deme in the structured coalescent model (Figure \ref{}).
 
@@
 This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{mers_epi}).
+When rapid coalescence in the human deme is not allowed (Figure \ref{equal_sizes}) structured coalescent inference loses power and ancestral state reconstruction is nearly identical to that of discrete trait analysis (Figure \ref{dta}).
+
 From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{mers_epi}).
+Although we recover migration rates from our model (Figure \ref{prior}), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
 We also looked at potential seasonality in MERS-CoV spillover into humans.
@@
 
-\subsection*{Structured coalescent analyses}
+\subsection*{Phylogenetic analyses}
+
+\subsubsection*{Primary analysis, structured coalescent}
 
@@
 Posterior sets of typed (where migrating branches from structured coalescent are collapsed, and migration information is left as a switch in state between parent and descendant nodes) trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
+A maximum likelihood phylogeny showing just the genetic relationships between MERS-CoV genomes from camels and humans was recovered using PhyML \citep{guindon_simple_2003} under a HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution model and is shown in Figure \ref{ml}.
+
+\subsubsection*{Control, structured coalescent with different prior}
 
@@
 Posterior sets of typed trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
+
+\subsubsection*{Control, structured coalescent with equal deme sizes}
+
+To better understand where statistical power of the structured coalescent model lies we set up a tertiary analysis where a model was set up identically to the first structured coalescent analysis, but deme population sizes were enforced to have the same size.
+This analysis allowed us to differentiate whether statistical power in our analysis is coming from effective population size contrasts between demes or the backwards-in-time migration rate estimation.
+Five replicate chains were set up, two of which failed to converge after 200 million states.
+Combining the three converging runs left us with $15\,000$ trees sampled from the posterior distribution, which were summarised in TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
+
+\subsubsection*{Control, structured coalescent with more than one tree per genome}
 
@@
 
-\subsubsection*{Discrete trait analysis}
+\subsubsection*{Control, discrete trait analysis}
 
@@
 	\caption{\textbf{Maximum clade credibility (MCC) tree with ancestral state reconstruction according to a discrete trait model.}
-MCC tree is presented the same as Figure \ref{mcc}, with colours indicating the most probable state reconstruction at internal nodes.
+MCC tree is presented the same as Figure \ref{mcc} and Figure \ref{equal_sizes}, with colours indicating the most probable state reconstruction at internal nodes.
 Unlike the structured coalescent summary shown in Figure \ref{mcc} where camels are reconstructed as the main host where MERS-CoV persists, the discrete trait approach identifies both camels and humans as major hosts with humans being the source of MERS-CoV infection in camels.
@@
 	\label{dta}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_es_mcc.png}
+	\caption{\textbf{Maximum clade credibility (MCC) tree of structured coalescent model with enforced equal coalescence rates.}
+MCC tree is presented the same as Figures \ref{mcc} and \ref{dta}, with colours indicating the most probable state reconstruction at internal nodes.
+Similar to Figure \ref{dta} enforcing equal coalescence rates between demes in a structured coalescent model identifies humans as a major MERS-CoV host and the source of viruses in camels.
+	}
+	\label{equal_sizes}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_ML.png}
+	\caption{\textbf{Maximum likelihood (ML) tree of MERS-CoV genomes coloured by origin of sequence.}
+Maximum likelihood tree shows genetic divergence between MERS-CoV genomes collected from camels (orange tips) and humans (blue tips).
+	}
+	\label{ml}
 \end{figure}

### 2017-11-08 mers-structure.tex
Fix typos in epi figures, address #39

@@
 Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.91$ and sequence sampling likely to be biased towards large hospital outbreaks.
+All matching simulations exhibit highly skewed distributions of case cluster sizes with long tails (Figure \ref{cluster_distributions}), which is qualitatively similar to the results of \citep{cauchemez_unraveling_2016}.
 Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_distributions.png}
+	\caption{\textbf{Boxplots of matching simulated case and sequence cluster distributions.}
+Boxplots indicate frequency of case (blue, top) and sequence (red, bottom) cluster sizes across simulations at different bias levels, marginalised across $R_{0}$ values.
+Outliers are shown with transparency, medians are indicated with thick black lines.
+Case clusters exhibit a strong skew with large numbers of singleton introductions and a substantial tail at higher levels of bias.
+	}
+	\label{cluster_distributions}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}

### 2017-11-09 mers-structure.tex
Address #38, add forgotten fig for #39

@@
 We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
-We can replicate these results by either applying the same model to current data (Figure \ref{dta}) or by enforcing equal coalescence rates within each deme in the structured coalescent model (Figure \ref{}).
+We can replicate these results by either applying the same model to current data (Figure \ref{dta}) or by enforcing equal coalescence rates within each deme in the structured coalescent model (Figure \ref{equal_sizes}).
 
@@
     \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
-    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
-    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to 29364.
-    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
-    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
+    A) Marginal posterior probabilities of taxa collected from humans belonging to the same clade in phylogenies derived from different parts of the genome.
+    Taxa are ordered according to phylogeny of fragment 2 (genome positions 21001 to 29364) reduced to just the human tips and is displayed on the left.
     Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
-    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
-  	Taxa descended from the same introduction in fragment 1 are connected to their counterparts in fragment 2 with combinations of line dotting and colour unique to each introduction clade.
-  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
-  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
-  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
+    B) Tanglegram connecting the same taxa between a phylogeny derived from fragment 1 (left, genome positions 1 to 21000) and fragment 2 (right, genome positions 21001 to 29364), reduced to just the human tips and branches with posterior probability $<0.1$ collapsed.
+    Human clusters exhibit limited diversity and corresponding low levels of incongruence within an introduction cluster.
     }

### 2017-11-09 mers-structure.tex
Address #19

@@
 As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{mers_epi}).
+Having said that, structured population models explicitly relating epidemiological parameters to the branching process observed in sequence data \citep{kuhnert_phylodynamics_2016} should ideally be used in cases like MERS-CoV, but in our case lack of good prior information and MCMC convergence issues prevented us from employing such models here.
 
@@
 On a smaller scale population structure can unveil important details about transmission patterns, such as identifying reservoirs and understanding spillover trends and risk, much as we have done here.
+There is much room for improvement, however.
+The population structure model applied in this study does not have an implementation allowing for changes in effective population size through time.
 When properly understood naturally arising barriers to gene flow can be exploited for more efficient disease control and prevention, as well as risk management.

### 2017-11-17 mers-structure.tex
Textual changes to address #38

@@
 Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \citep{lai_1985,makino_1986,keck_1988,kottier_1995,herrewegh_1998}.
-In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
+In order to quantify the degree to recombination has shaped MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
 Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
@@
 % Homoplasies are present in multiple strains at a time, indicating recombination events that have been successful.
-Although the evolutionary centrality of camel viruses (Figure \ref{mcc}) may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
-This provides ideal conditions for co-infection with distinct genotypes, which is a pre-requisite for detectable RNA virus recombination to occur.
+These results are consistent with high incidence of MERS-CoV in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}, allowing for co-infection with distinct genotypes and thus recombination to occur.
 
 %%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
-Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
-We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
+Owing to these findings, we performed a sensitivity analysis in which we partitioned the MERS-CoV genome into two fragments and identified human outbreak clusters within each fragment.
+We find strong similarity in the grouping of human cases into outbreak clusters between fragments (Figure \ref{recombinant_features}A).
 Between the two trees in figure \ref{recombinant_features}B four (out of 54) `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
@@
 And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
-Critically, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
-Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
-MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
-In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
-Overall, recombination presents a serious hurdle for detailed phylodynamic analyses in MERS-CoV, especially where lineages evolving in camels are concerned.
+Additionally, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
+
+Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected if humans represent a transient host with little opportunity for co-infection.
+MERS-CoV evolution on the reservoir side is complicated by recombination, though is nonetheless still largely amenable to phylogenetic methods.
+% In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
+% Overall, recombination presents a serious hurdle for detailed phylodynamic analyses in MERS-CoV, especially where lineages evolving in camels are concerned.
 Amongst other parameters of interest, recombination is expected to interfere with molecular clocks, where transferred genomic regions can give the impression of branches undergoing rapid evolution, or branches where recombination results in reversions appearing to evolve slow.

### 2017-11-20 mers-structure.tex
State findings differently. Fixes #30.
Address #29, #19 and reformat for eLife

@@
-\documentclass[11pt,oneside,letterpaper]{article}
+% \documentclass[11pt,oneside,letterpaper]{article}
+\documentclass[9pt,lineno]{elife}
 
@@
 
-\usepackage[hidelinks]{hyperref}
-\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}
+% \usepackage[hidelinks]{hyperref}
+% \hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}
 
@@
 
-\author[1]{Gytis Dudas}
+\author[1*]{Gytis Dudas}
 \author[2]{Luiz Max Carvalho}
@@
 
-% \date{\today}
+\corr{gdudas@fredhutch.org}{GD}
 
@@
 Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
-Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
+However, we find that MERS-CoV, once introduced into humans, follows transient transmission chains that soon abate.
 Using Monte Carlo simulations we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.
@@
 
-\bibliographystyle{mbe}
+% \bibliographystyle{mbe}
 \bibliography{mers-structure}

### 2017-11-21 mers-structure.tex
Explain SC logic, per #21

@@
 When rapid coalescence in the human deme is not allowed (Figure \ref{equal_sizes}) structured coalescent inference loses power and ancestral state reconstruction is nearly identical to that of discrete trait analysis (Figure \ref{dta}).
+When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme.
+This in turn, informs the inferred state of long ancestral branches in the phylogeny, i.e.\ because these long branches are not immediately coalescing, they are most likely in camels.
 

### 2017-11-21 mers-structure.tex
Address #25.

--- 2017-11-20 mers-structure.tex
@@
 Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
-In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones.
+In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones (for its effects see Figure \ref{mers_epi_grid}).
 This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
-We record the number, mean, standard deviation and skewness (third standard moment) of sequence cluster sizes in each simulation (left-hand panels in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
+We record the number, mean, standard deviation and skewness (third standard moment) of sequence cluster sizes in each simulation (left-hand panel in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
 We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
 A schematic of this Monte Carlo procedure is shown in Figure \ref{mc_method}.
-Generally, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters (Figure \ref{mers_epi}).
+Since the total number of cases is fixed at 2000, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters.
 
@@
 	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
-From top to bottom each row corresponds to departures from completely random sequencing efforts with respect to case cluster size (bias parameter=1) to sequencing increasingly biased towards capturing large case clusters (bias=2, bias=3).
-Leftmost scatter plots show the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) that fell within the 95\% HPDs for number of sequence clusters, mean sequence cluster size, sequence cluster size standard deviation, and sequence cluster size skewness recovered for empirical data, coloured by the $R_{0}$ value used for the simulation.
+Leftmost scatter plot shows the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) coloured by the $R_{0}$ value used for the simulation.
 The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and skewness observed for empirical MERS-CoV data.
-The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same $y$-axis across all levels of the bias parameter.
+The distribution of $R_{0}$ values that fall within 95\% HPDs for sequence cluster size mean, standard deviation, skewness and number of introductions, is shown in the middle, on the same $y$-axis.
 Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
-The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
+The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown on the right.
 Darker shade of grey indicates bins falling within the 95\% percentiles.
-These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
+Monte Carlo simulations indicate $R_{0}$ for MERS-CoV in humans is likely to be below 1.0, with numbers of zoonotic transmissions numbering in the hundreds.
 	}
@@
 
-We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
-For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.84, 0.73, and 0.68, respectively.
-While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.70--0.97) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
-Under unbiased sequencing only 0.29\% of simulations fit our phylogenetic observations, while 1.2\% and 1.1\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
-Correspondingly, we estimate 11\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 44\% support for a model with bias level 3.0.
-Model averaging would suggest plausible $R_0$ values between 0.57 and 0.90.
-
-Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
-The median number of cross-species introductions observed in simulations matching empirical data without bias are 344 (95\% percentiles 264--448).
-These numbers jump up to 567 (95\% percentiles 432--730) for bias = 2 and 654 (95\% percentiles 491--847) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
-Model averaging would suggest plausible numbers of introductions between 311 and 816.
+We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panel in Figure \ref{mers_epi}).
+Mean $R_{0}$ value observed in matching simulations is 0.72 (95\% percentiles 0.57--0.90), suggesting the inability of the virus to sustain transmission in humans.
+Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panel in Figure \ref{mers_epi}).
+Median number of cross-species introductions observed in matching simulations is 592 (95\% percentiles 311-811).
 Our results suggest a large number of unobserved MERS primary cases.
@@
 
-Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.91$ and sequence sampling likely to be biased towards large hospital outbreaks.
+Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.90$ and sequence sampling likely to be biased towards large hospital outbreaks (Figure \ref{mers_epi_grid}).
 All matching simulations exhibit highly skewed distributions of case cluster sizes with long tails (Figure \ref{cluster_distributions}), which is qualitatively similar to the results of \citep{cauchemez_unraveling_2016}.
@@
 \centering
+	\includegraphics[width=0.75\textwidth]{figures/mers_epi_grid.png}
+	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
+From top to bottom each row corresponds to departures from completely random sequencing efforts with respect to case cluster size (bias parameter=1.0) to sequencing increasingly biased towards capturing large case clusters (bias=2.0, bias=3.0).
+Leftmost scatter plots show the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) coloured by the $R_{0}$ value used for the simulation.
+The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and skewness observed for empirical MERS-CoV data.
+The distribution of $R_{0}$ values matching empirical data are shown in the middle, on the same $y$-axis across all levels of the bias parameter.
+Under unbiased sequencing (bias=1.0) only 0.45\% of simulations fit our phylogenetic observations, while 1.79\% and 1.67\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
+Correspondingly, we estimate 11.6\% support for a model with bias level 1.0, 45.7\% support for a model with bias level 2.0, and 42.7\% support for a model with bias level 3.0.
+Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
+While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.71--0.98) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters.
+
+The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
+Darker shade of grey indicates bins falling within the 95\% percentiles.
+The median number of cross-species introductions observed in simulations matching empirical data without bias are 346 (95\% percentiles 262--439).
+These numbers jump up to 568 (95\% percentiles 430--727) for bias = 2.0 and 656 (95\% percentiles 488--853) for bias = 3.0 simulations.
+Model averaging would suggest plausible numbers of introductions between 311 and 811.
+	}
+	\label{mers_epi_grid}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}
 	\caption{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
-The plot is identical to Figure \ref{mers_epi}, but instead of 2000 cases,
+The plot is identical to Figure \ref{mers_epi_grid}, but instead of 2000 cases,
 % the currently reported number of MERS-CoV cases,

### 2017-11-21 mers-structure.tex
Address #29

@@
 All matching simulations exhibit highly skewed distributions of case cluster sizes with long tails (Figure \ref{cluster_distributions}), which is qualitatively similar to the results of \citep{cauchemez_unraveling_2016}.
+We find that simulated sequence cluster sizes resemble observed sequence cluster sizes in the posterior distribution, with a slight reduction in mid-sized clusters in simulated data (Figure \ref{qq}).
 Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_qq.png}
+	\caption{\textbf{Quantile-quantile (Q-Q) plot of empirical and simulated sequence cluster sizes.}
+Density of sequence cluster size percentiles (1st--99th, calculated across a grid of 50 values) calculated for random states from the posterior distribution ($x$-axis) and matching simulations ($y$-axis).
+Most values fall on the one-to-one line, with a heavier tail in mid-sized sequence clusters in empirical data, manifesting as a greater density of points below the one-to-one line in the middle.
+	}
+	\label{qq}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.75\textwidth]{figures/mers_epi_grid.png}
@@
 While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.71--0.98) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters.
-
 The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.

### 2017-11-21 mers-structure.tex

@@
 
-MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
-Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
-When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
-Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
-Previous research \citep{antia_role_2003} suggests that zoonotic pathogens become disproportionately more likely to evolve towards $R_{0}>1.0$ if $R_{0}$ is close to $1.0$ in the new host to begin with, which leaves MERS-CoV in a precarious position: on one hand our estimate of $R_{0}$ at $\approx 0.7$ makes MERS-CoV a potential pandemic hazard, on the other nothing is known of fitness landscapes or adaptive potential of MERS-CoV.
-We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.
+Pathogens \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004} and influenza A are able to jump species barriers but only influenza A viruses have historically resulted in pandemics \citep{lipsitch_viral_2016}.
+MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host.
+Since its emergence in 2012, MERS-CoV has caused self-limiting outbreaks in humans with no evidence of worsening outbreaks over time.
+However, sustained evolution of the virus in the reservoir and continued flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
+Previous modeling studies \citep{antia_role_2003, park_multiple_2013} suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
+This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
+With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
+In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
 

### 2017-11-22 mers-structure.tex
Reformulate Ne calculations per #28

@@
 However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
-This estimate of $N_e\tau$ can be translated into an estimate of effective prevalence $N_e$ given an estimate of generation time or serial intervals.
-Given a 20 day duration of infection in camels \citep{adney_replication_2014}, we expect a 10 day serial interval or $\tau$ of 0.027 years per generation.
-Dividing by $\tau$ yields an estimate of $N_e$ of 129 (100--162) individuals.
-
-Here, we compare this estimate of effective prevalence to a rough estimate of census prevalence.
+
+This empirically estimated effectived population can be compared to the expected effective population size in a simple epidemiological model.
+At endemic equilibrium, we expect scaled effective population size $N_{e}\tau$ to follow $I \, / \, 2 \beta$, where $\beta$ is the equilibrium rate of transmission and $I$ is the equilibrium number of infecteds \citep{frost_viral_2010}.
+We assume that $\beta$ is constant and is equal to the rate of recovery.
+Given a 20 day duration of infection in camels \citep{adney_replication_2014}, we arrive at $\beta = 365/20 = 18.25$ infections per year.
 Given extremely high seroprevalence estimates within camels in Saudi Arabia \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}, we expect camels to usually be infected within their first year of life.
@@
 We find there are $830\,000$ camels in Saudi Arabia \citep{abdallah_camel_farming_2013} and that female camels in Saudi Arabia have an average fecundity of 45\% \citep{abdallah_camel_farming_2013}.
-Thus, we expect $830\,000 \times 0.50 \times 0.45 = 186\,750$ new calves produced yearly and correspondingly $186\,750$ new infections every year, which spread over 10 day serial intervals gives an average of $N = 5042$ infections.
-
-We thus arrive a ratio of $N/N_e = 39.1$ (31.1--50.4).
+Thus, we expect $830\,000 \times 0.50 \times 0.45 = 186\,750$ new calves produced yearly and correspondingly $186\,750$ new infections every year, which spread over 20 day intervals gives an average prevalence of $I = 10\,233$ infections.
+This results in an expected scaled effective population size $N_{e}\tau=280.4$ years.
+
+Comparing expected $N_{e}\tau$ to empirical $N_{e}\tau$ we arrive at a ratio of 80.3 (64.0--103.5).
 This is less than the equivalent ratio for human measles virus \citep{bedford_strength_2011} and is in line with the expectation from neutral evolutionary dynamics plus some degree of transmission heterogeneity \citep{volz_phylodynamics_2013} and seasonal troughs in prevalence.

### 2017-11-22 mers-structure.tex
Address #19

@@
 As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{mers_epi}).
-Having said that, structured population models explicitly relating epidemiological parameters to the branching process observed in sequence data \citep{kuhnert_phylodynamics_2016} should ideally be used in cases like MERS-CoV, but in our case lack of good prior information and MCMC convergence issues prevented us from employing such models here.
+Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models \citep{rasmussen_phylodynamic_2014} which would be more appropriate for MERS-CoV.
+Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.
+Outside of coalescent-based models there are population structure models that explicitly relate epidemiological parameters to the branching process observed in sequence data \citep{kuhnert_phylodynamics_2016}, but often rely on specifying numerous informative priors and can suffer from MCMC convergence issues.
 
@@
 On a smaller scale population structure can unveil important details about transmission patterns, such as identifying reservoirs and understanding spillover trends and risk, much as we have done here.
-There is much room for improvement, however.
-The population structure model applied in this study does not have an implementation allowing for changes in effective population size through time.
 When properly understood naturally arising barriers to gene flow can be exploited for more efficient disease control and prevention, as well as risk management.

### 2017-11-27 mers-structure.tex
Address #26 and #40.

@@
 We suspect that the effects of recombination in MERS-CoV data are reigned in by a relatively small effective population size of the virus in Saudi Arabia (see next section) where haplotypes are fixed or nearly fixed, thus preventing an accumulation of genetic diversity that would then be reshuffled via recombination.
-Nevertheless, we choose not to report on any particular estimates for times of common ancestors (tMRCAs), even though these are expected to be somewhat robust for dating human clusters, and we do not report on the evolutionary rate of the virus, even though it appears to fall firmly within the expected range for RNA viruses (data not shown).
+Nevertheless, we choose not to report on any particular estimates for times of common ancestors (tMRCAs), even though these are expected to be somewhat robust for dating human clusters, and we do not report on the evolutionary rate of the virus, even though it appears to fall firmly within the expected range for RNA viruses: regression of nucleotide differences to Jordan-N3/2012 genome against sequence collection dates yields a rate of $4.59 \times 10^{-4}$ subs/site/year, Bayesian structured coalescent estimate from primary analysis $9.57 \times 10^{-4}$ (95\% HPDs: $8.28-10.9 \times 10^{-4}$) subs/site/year.
 

### 2017-11-27 mers-structure.tex
Better way of addressing #35.

@@
 Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
-However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide a more intuitive epidemic behaviour metric.
+However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide an intuitive epidemic behaviour metric that does not require expertise in reading phylogenies.
 The parameter $R_{0}$ determines expected number of secondary cases in a single infections as well as the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}, Methods).

### 2017-11-29 mers-structure.tex
Reorder some supplementary figures.

@@
 We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
-Additionally, we performed additional smaller set of simulations on a rougher grid of $R_{0}$ values (23 values, 0.50--1.05), with 5 values of dispersion parameter $\omega$ (0.002, 0.04, 0.1, 0.5, 1.0) and 3 levels of bias (1, 2, 3) to justify our choice of dispersion parameter $\omega$ that was fixed to 0.1 in the main analyses.
+Additionally, we performed additional smaller set of simulations on a rougher grid of $R_{0}$ values (23 values, 0.50--1.05), with 5 values of dispersion parameter $\omega$ (0.002, 0.04, 0.1, 0.5, 1.0) and 3 levels of bias (1, 2, 3) to justify our choice of dispersion parameter $\omega$ that was fixed to 0.1 in the main analyses (Figure \ref{marginals}).
 
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
+	\caption{\textbf{Posterior backwards migration rate estimates for two choices of prior.}
+Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice (note that rates are backwards in time).
+Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme looking backwards in time (orange) and \textit{vice versa} (blue).
+Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
+	}
+	\label{prior}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_dta_mcc.png}
@@
 	\label{equal_sizes}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_ML.png}
-	\caption{\textbf{Maximum likelihood (ML) tree of MERS-CoV genomes coloured by origin of sequence.}
-Maximum likelihood tree shows genetic divergence between MERS-CoV genomes collected from camels (orange tips) and humans (blue tips).
-	}
-	\label{ml}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
-	\caption{\textbf{Posterior migration rate estimates for two choices of prior.}
-Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice.
-Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme (blue) and \textit{vice versa} (orange).
-Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
-	}
-	\label{prior}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_mc_method.pdf}
-	\caption{\textbf{Monte Carlo simulation schematic.}
-Case clusters are simulated according to Equation \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
-We sample 174 cases from each simulation to represent sequencing of human MERS cases.
-`Sequencing' is carried out by using multivariate hypergeometric sampling, representing sampling cases without replacement to be sequenced.
-Sequencing simulations take place at three levels of bias: 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
-The distribution of simulated sequence clusters is summarised by its mean, median and standard deviation.
-A simulation is considered to match if the mean, median and standard deviation of its sequence cluster sizes falls within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
-$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are retained as estimates.
-These estimates are summarised in Figure \ref{mers_epi}.
-	}
-	\label{mc_method}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
-	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
-Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
-More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
-	}
-	\label{marginals}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_distributions.png}
-	\caption{\textbf{Boxplots of matching simulated case and sequence cluster distributions.}
-Boxplots indicate frequency of case (blue, top) and sequence (red, bottom) cluster sizes across simulations at different bias levels, marginalised across $R_{0}$ values.
-Outliers are shown with transparency, medians are indicated with thick black lines.
-Case clusters exhibit a strong skew with large numbers of singleton introductions and a substantial tail at higher levels of bias.
-	}
-	\label{cluster_distributions}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_qq.png}
-	\caption{\textbf{Quantile-quantile (Q-Q) plot of empirical and simulated sequence cluster sizes.}
-Density of sequence cluster size percentiles (1st--99th, calculated across a grid of 50 values) calculated for random states from the posterior distribution ($x$-axis) and matching simulations ($y$-axis).
-Most values fall on the one-to-one line, with a heavier tail in mid-sized sequence clusters in empirical data, manifesting as a greater density of points below the one-to-one line in the middle.
-	}
-	\label{qq}
 \end{figure}
@@
 \centering
+	\includegraphics[width=0.75\textwidth]{figures/mers_mc_method.pdf}
+	\caption{\textbf{Monte Carlo simulation schematic.}
+Case clusters are simulated according to Equation \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
+We sample 174 cases from each simulation to represent sequencing of human MERS cases.
+`Sequencing' is carried out by using multivariate hypergeometric sampling, representing sampling cases without replacement to be sequenced.
+Sequencing simulations take place at three levels of bias: 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
+The distribution of simulated sequence clusters is summarised by its mean, median and standard deviation.
+A simulation is considered to match if the mean, median and standard deviation of its sequence cluster sizes falls within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
+$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are retained as estimates.
+These estimates are summarised in Figure \ref{mers_epi}.
+	}
+	\label{mc_method}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}
@@
 \centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_distributions.png}
+	\caption{\textbf{Boxplots of matching simulated case and sequence cluster distributions.}
+Boxplots indicate frequency of case (blue, top) and sequence (red, bottom) cluster sizes across simulations at different bias levels, marginalised across $R_{0}$ values.
+Outliers are shown with transparency, medians are indicated with thick black lines.
+Case clusters exhibit a strong skew with large numbers of singleton introductions and a substantial tail at higher levels of bias.
+	}
+	\label{cluster_distributions}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_qq.png}
+	\caption{\textbf{Quantile-quantile (Q-Q) plot of empirical and simulated sequence cluster sizes.}
+Density of sequence cluster size percentiles (1st--99th, calculated across a grid of 50 values) calculated for random states from the posterior distribution ($x$-axis) and matching simulations ($y$-axis).
+Most values fall on the one-to-one line, with a heavier tail in mid-sized sequence clusters in empirical data, manifesting as a greater density of points below the one-to-one line in the middle.
+	}
+	\label{qq}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 	\includegraphics[width=0.65\textwidth]{figures/mers_recombination_tree.png}
@@
 
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_ML.png}
+	\caption{\textbf{Maximum likelihood (ML) tree of MERS-CoV genomes coloured by origin of sequence.}
+Maximum likelihood tree shows genetic divergence between MERS-CoV genomes collected from camels (orange tips) and humans (blue tips).
+	}
+	\label{ml}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
+	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
+Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
+More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
+	}
+	\label{marginals}
+\end{figure}
+
 \end{document}

### 2017-11-29 response_1/response_1.tex
Add response text.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{hyperref}

% text layout
\usepackage{geometry}
\geometry{textwidth=15.25cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace}
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

% notation
\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}

\newgeometry{top=4cm}

Dear eLife editorial board,

Thank you for your thorough and insightful comments. Please find attached our revised manuscript entitled ``MERS-CoV spillover at the camel-human interface''.  This is an update to submission 14-08-2017-RA-eLife-31257. The editorial assessment identified five essential revisions, in addition to minor points raised by individual reviewers.

Point-by-point review responses are attached.

Sincerely,\\
Gytis Dudas

\newpage

\section*{Reviewer responses}

Original comments are in plain text.  Our responses follow in \textbf{bold}.

\section*{Essential revisions}
1. The population genetic model (the particular form of structured coalescent) is highly idealised and this may influence the quantitative conclusions, although we suspect the conclusions are quite robust qualitatively. This model specifically estimates the rate of a lineage moving between demes going backwards in time; the numbers cited for the camel->human rate is really the rate that a lineage in humans goes to a camel going down the tree. The relationship between these migration rates and the epidemiologically meaningful transmission rate is complex and depends among other things on the ratio of population sizes in both demes. Per-capita transmission rates could be estimated using an epidemiologically structured coalescent model (see e.g. papers by Volz \& Rasmussen), which would ideally be stochastic due to bursty dynamics in humans. But this would be a large undertaking and so we suggest that for now the distinction is clarified. Overall, a little more discussion of the complexity and pitfalls when relating idealised population genetic models (like the island model used here) to a noisy nonlinear epidemic like this one might be merited.

\textbf{Yes, we agree that the structured coalescent approach is idealised and does not reflect a meaningful rate of zoonotic transfer of lineages, which is the reason we restricted any mention of rates to supplementary figures and do not attach numbers whenever rates are mentioned, but still report the number of introductions observed in the sequence data. We have altered Figure S2 in the reviewed manuscript to reflect that the rates shown are backwards in time. We have added the following sentences to the discussion to highlight the fact that the coalescent model employed is not ideal:}

\begin{quotation}
Although we recover migration rates from our model (Figure S2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
\end{quotation}

\begin{quotation}
Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models (Rasmussen \textit{et al}., 2014) which would be more appropriate for MERS-CoV.
Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.
Outside of coalescent-based models there are population structure models that explicitly relate epidemiological parameters to the branching process observed in sequence data (K\"{u}hnert \textit{et al}., 2016), but often rely on specifying numerous informative priors and can suffer from MCMC convergence issues.
\end{quotation}

% \begin{quotation}
%
% \end{quotation}
2. p4, `Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior'. This is a quite nice result and raises the question if skewed sampling would bias estimates if using a substitution model approach (`discrete trait analysis', DTA). It would strengthen the paper to include a comparison of the structured coalescent estimates to another method for ancestral states; the most popular approach in beast has been substitution models (DTA). These may give divergent results because of skewed sampling. It would be rather easy for the authors to run a DTA and if biased, this would serve as a good cautionary example when sampling is highly skewed towards one deme.

\textbf{An excellent suggestion, thank you. We have run this analysis and include the results as a new supplementary figure S3. As expected, the skewed sampling results in a reconstruction of ancestral states that puts humans as the source of most MERS-CoV lineages in camels. We have added the appropriate description of methods as well as the following paragraph in results:}

\begin{quotation}
Our findings suggest that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al.}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
A similar approach has been attempted previously (Zhang \textit{et al.}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
\end{quotation}

3. A comparison to ML tree reconstruction could potentially be illuminating. We think you could be clearer about what drives the results in your paper. It is unusual for a phylogenetic ancestral reconstruction, that the results seem to be determined as much by the coalescent assumptions as by the tree topology. The two-patch model had a much higher coalescent rate in the human deme than in the camel deme - so long branches are only really possible in the camel deme. This may be why for example, staring at the top clade of figure 1, one can see camel ancestry to a whole bunch of human sequences that are not topologically separated by camel sequences. If this is correct, these results may not necessarily be wrong, but it made us slightly uncomfortable that the results are driven by the coalescent model, not the tree topology. Please elaborate, either correcting us, or explaining better. A simple test of this hypothesis would be that an ML ancestral reconstruction on the ML tree would not give the same clusters. I don't think that would make the ML result correct, but it might be an enlightening comparison. Or you may prefer another way to address this.

\textbf{A very good point. We share the suspicion that the results are largely driven by contrasts in effective population sizes between demes. In addition to the requested maximum likelihood phylogeny (supplementary Figure S14) we also ran a structured coalescent analysis where deme sizes are enforced to be the same (supplementary Figure S4). This model fails in a similar way to a DTA reconstruction shown in the new Figure S3. We now explain how the structured coalescent arrives at the tree shown in Figure 1 in the discussion:}

\begin{quotation}
When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme. 
This in turn, informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
\end{quotation}

4. Easily addressed, but important. The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger. Antia et al Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0. So R0~0.8 is much worse than R0~0.3. More sophisticated models have been done since, especially by Llyod-smith's group, but the basic result is sound. In the light of this theoretical work, your findings are not at all reassuring.

\textbf{We agree that this is an important point, but also feel that it is difficult to formulate warnings about pandemic potential without overstating the case. We also believe that adaptive landscapes play a considerable role in emerging pandemics. We added an additional sentence to the discussion and refer to the Antia et al study:}

\begin{quotation}
Previous modeling studies (Antia \textit{et al}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
\end{quotation}

5. More generally, the model choices need better explaining. Why delve into a structured coalescent in BEAST2 for the ancestral reconstruction, but go back to the Skygrid in BEAST1 for computations of Ne? We assume this is a pragmatic choice, and for the latter you carefully reduced the human clusters to reduce bias, but we think the rationale for your choices need laying out more clearly. Even if pragmatic rather than principled, (e.g. there are no structure coalescent options in BEAST1), we think it still needs to be stated why you made the choices you did. Especially since there are other recently-developed BEAST2 packages that could be used to fit the same structured coalescent model: BASTA \& MASCOT, as well as the very flexible PhyDyn package (which might offer improvements in computation time).

\textbf{In hindsight we see how our choices may have seemed arbitrary. Indeed all cases of using BEAST 1 vs BEAST 2 came down to what models were implemented in which package. We have tried clarified our choices throughout the manuscript.}

% \begin{quotation}
%
% \end{quotation}

\section*{Reviewer \#1}

1. p3, `...population dynamics within demes...'
The models used here assumed constant effective size in each deme.

\textbf{The reviewer is absolutely correct here. This has been fixed in the text.}

2. p6-7, I found the description of the simulation study and simulation parameters difficult to follow. For example, greater clarity would be appreciated for terms like `bias level' in Figure 3.
These terms are clarified in the supporting methods, but I found the main text incomprehensible without seeing that section first. I would make an effort to transfer more detailed methods related to the branching process model here.

\textbf{We agree that the simulation study could have been explained better. We elaborated the description of simulations in results to say:}

\begin{quotation}
Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones.
\end{quotation}

3. p 8-11, What are the implications of recombinant MERS CoV in camels for estimation of evolutionary rates and TMRCAs?

\textbf{An excellent point. Concerns over artefacts of recombination is what prevented us from reporting on evolutionary rates and tMRCAs of interest, such as our postulated introduction of the virus into the Arabian peninsula. We added a discussion about the potential effects of recombination in the results section and report evolutionary rates recovered by two methods:}

\begin{quotation}
Overall, recombination presents a serious hurdle for detailed phylodynamic analyses in MERS-CoV, especially where lineages evolving in camels are concerned.
Amongst other parameters of interest, recombination is expected to interfere with molecular clocks, where transferred genomic regions can give the impression of branches undergoing rapid evolution, or branches where recombination results in reversions appearing to evolve slow.
In addition to its potential to influence tree topology, recombination in molecular sequence data is an erratic force with unpredictable effects.
We suspect that the effects of recombination in MERS-CoV data are reigned in by a relatively small effective population size of the virus in Saudi Arabia (see next section) where haplotypes are fixed or nearly fixed, thus preventing an accumulation of genetic diversity that would then be reshuffled via recombination.
Nevertheless, we choose not to report on any particular estimates for times of common ancestors (tMRCAs), even though these are expected to be somewhat robust for dating human clusters, and we do not report on the evolutionary rate of the virus, even though it appears to fall firmly within the expected range for RNA viruses: regression of nucleotide differences to Jordan-N3/2012 genome against sequence collection dates yields a rate of $4.59 \times 10^{-4}$ subs/site/year, Bayesian structured coalescent estimate from primary analysis $9.57 \times 10^{-4}$ (95\% HPDs: $8.28-10.9 \times 10^{-4}$) subs/site/year.
\end{quotation}

4. p 10, `...relatively low overall with a mean estimate of N of 3.49 years (95\% HPD: 2.71-4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies...' I would not say that low Ne directly implies a ladder-like tree or vice versa, although it may produce such a tree in combination with highly dispersed sample times.

\textbf{We agree that casually calling the effective population size `low' does not explain the shape of the MERS-CoV phylogeny. We have rephrased this sentence to emphasise that the rate of coalescence observed in that particular part of the MERS-CoV phylogeny is high compared to the period over which sampling took place:}

\begin{quotation}
However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies.
\end{quotation}

5. p 11, I think these back-of-the-envelope calculations to relate Ne to epidemic size are illuminating, however I would do some things differently:

5.1. A 10 day generation interval is assumed based on a 20 day infectious period, however this is strongly dependent on assumptions about how infectiousness varies over the course of infection. Note that I prefer the term `generation interval' for this particular situation, see eg https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2365921/. In a simple SIR model, the mean generation interval will equal the mean infectious period, and the choice of 10 days seems a little strange. There is certainly considerable uncertainty about parameters here, so at least the authors could compute this over a range of values.

\textbf{}

5.2. Other theoretical work suggests a slightly different relationship between Ne tau and I(t), specifically that it will depend more on transmission rate than generation time. However if making the approximation that I(t) is constant (which the authors are) , the formulas will be the same up to a factor of 2x. See Koelle, 2011, DOI: 10.1098/rsif.2011.0495 \& Volz and Frost 2013, DOI: 10.1098/rstb.2010.0060. Specifically, I would use (Ne tau ) = I(t) / 2 * [per capita transmission rate], but some of the formulas in Koelle's paper may be more realistic for MERS.
and equating beta = 1 / [generation time] gives 2x the authors' estimate if solving for I.
Note that if using a 20 day generation interval and the above formula, you would have I = 127, so this doesn't make any practical difference, but is more elegant in my opinion, since we are exact about the epidemic model that the estimate is based on. If some estimate of the dispersion parameter for R0 could be obtained from the cluster size distribution, a better estimate for I could be obtained using equations in Koelle 2011.

\textbf{}

6. P 15+, I identify the following weaknesses in the simulation study, and I would candidly discuss these in the Discussion section:

6.1. Dispersion parameter is set at 0.1 based on a previous branching process study, however there's not great support for re-using the value for a different epidemic. Was an effort made to estimate this parameter?

\textbf{The dispersion parameter was indeed chosen from a previous study. We have run small scale analyses over a grid of three bias levels (same as the main analyses), 23 levels of $R_{0}$ (0.50--1.05), and five levels of dispersion (0.004, 0.02, 0.10, 0.50 and 1.0) with 2000 replicate simulations for each combination of variables. Numbers of simulations matching empirical data are shown in Figure S15.}

6.2. The sampling model accounts for `sequencing capacity', which is good. But is it the case that sequencing probability varied over the course of the epidemic?

\textbf{A very good point. We do not expect `sequencing capacity' to have been good in the initial stages of the outbreak and perhaps even to have peaked and declined at various stages, \textit{e.g.} during the Korean MERS outbreak. We think that addressing this analytically is perhaps a dimension too far, especially given the number of possible arrangements of sequencing intensity. We have added the following sentence to the results section:}

\begin{quotation}
Note that this ``sequencing capacity'' parameter does not vary over time, even though MERS-CoV sequencing efforts have varied in intensity, starting out slow due to lack of awareness, methods and materials and increasing in response to hospital outbreaks later.
\end{quotation}

6.3. The `model fitting' methods are ad-hoc. This is a situation where approximate Bayes computation would be appropriate. But I would try GEE/method of moments. I suspect the first three standardized moments of cluster sizes would be more useful than using mean/median/sd. - The moments could also be computed analytically if you dispense with the `sequencing capacity' aspect of the sampling model, so would be very fast.

\textbf{Our approach is indeed somewhat ad-hoc, though closely related to ABC. We have replaced the tolerance parameter of ABC with sequence data uncertainty and otherwise treat parameters of interest as being under uniform priors. We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical. We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.}

6.4. Can you show a comparison of the empirical distribution of cluster sizes and the fitted distribution? A QQ plot? Are there any aspects of the empirical distribution that are not well characterised by the branching process model, eg at the tails?

\textbf{We have generated QQ plots of sequence cluster size distribution observed across the posterior distribution of structured coalescent trees (picked at random) against matching sequence cluster size distributions from the primary Monte Carlo simulation with bias parameter set to 2.0. Simulated sequence clusters tend to exhibit a left skew near the middle of the distribution, but not so much near the tails. We have added figure S9 to show this.}

\section*{Reviewer \#2}

1. Page 3 - where the authors state that ``Once human MERS-CoV infections are established, hoever, we find that MERS0CoV is poor at transmitted between humans" some references should be added.

\textbf{The sentence was meant to summarise our findings. We have changed the sentence to clarify that the statement is based on our findings:}

\begin{quotation}
However, we find that MERS-CoV, once introduced into humans, follows transient transmission chains that soon abate.
\end{quotation}

2. In discussion: Could the author speculate how more sequences from humans and animals, particularly recent ones, or sequences from other geographic areas (east africa, north africa, pakistan) would potentially impace their results?

\textbf{We agree that understanding the distribution of MERS zoonotic risk to humans is of paramout importance, but feel that such speculation is outside the remit of the current study.
During the time the study was performed (first half of 2017) human MERS-CoV sequences comprised a large majority of available MERS-CoV sequences.
A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al.}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
We are also aware of another study into camels across much of Africa.
These are all consistent with general statements we have made in the manuscript in the methods section so far, which we have expanded further in response to this request:}

\begin{quotation}
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in Saudi Arabia and Egypt are distinct epidemics with relatively poor sampling in the latter.

Were more sequences of MERS-CoV available from other parts of Africa we speculate they would fall outside of the diversity that has been sampled in Saudi Arabia and cluster with early MERS-CoV sequences from Jordan and sequences from Egyptian camels.
However, currently there are no indications of what MERS-CoV diversity looks like in camels east of Saudi Arabia.
\end{quotation}

\textbf{We also intend to eventually update the manuscript with a structured coalescent analysis that includes all the sequence data that were made available recently.}

\section*{Reviewer \#3}

1. Page 2, introduction. I am not sure I would refer to the Cauchemez et al PNAS paper as traditional epidemiology. It's sophisticated statistical modelling of spatio-temporal data and the partial observation process. I think you are using different tools in epidemiology, obtaining complimentary insights, neither particularly traditional. You could just delete `traditional' or use `case-based'.

\textbf{Agreed. We have either removed or changed the three instances of `traditional' as requested.}

2. In the same paragraph the sentence starting `Genetic epidemiology' uses poor language (who is it who has shown?)

\textbf{We apologise for the poor choice of words. We have changed the sentence to say:}

\begin{quotation}
Where sequence data are relatively cheap to produce, genomic epidemiological approaches can fill this critical gap in outbreak scenarios.
\end{quotation}

3. Same page, why make a controversial statement about sequencing standing in for diagnostics? Really?

\textbf{Apologies for that sentence as well. We have removed that part of the sentence altogether.}

4. Page 5 `R0 more familiar to epidemiologists' More familiar than what? There are no previous quantitative estimates presented. Rephrase.

\textbf{Apologies, we meant to say that $R_{0}$ is a more intuitive measure of epidemic behaviour than extrapolating risk from subtrees of a phylogeny. We have changed the sentence to say:}

\begin{quotation}
However, we wanted to translate this observation into an estimate of the basic reproductive number $R_{0}$ to provide an intuitive epidemic behaviour metric that does not require expertise in reading phylogenies.
\end{quotation}

5. Page 5. Your summary statistics are mean, median and standard deviation of cluster size. It seems odd not to have a statistic measuring skewness, given how skew the cluster sizes seem to be. Why did you not use one? Do you think you would have got similar results with one?

\textbf{Thank you for the comment. It was our thinking that mean, median and standard deviation of cluster sizes would be able to capture both the overall magnitude and skew of the empirical distribution. We also accidentally omitted the number of clusters as a fourth parameter by which we filter simulated data. In retrospect we should have been using skewness from the beginning. We have since fixed both these issues and now use mean, standard deviation, skewness and number of clusters as filters for simulated data throughout the manuscript. This does not seem affect any of our results.}

6. Page 8. You refer repeatedly to spatial statistics to refer to location on genome. I think you should refer to clustering on the location on the genome.

\textbf{Agree. We have changed references to `spatial' when referring to genome positions to `clustering':}

\begin{quotation}
Another hallmark of recombination is clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
3Seq relies on the clustering of nucleotide similarities along the genome between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences.
\end{quotation}

7. Page 9. I think the section on recombination is verbose, and potentially misleading. You didn't perform an ARG, you performed a perfectly reasonable sensitivity analysis to different partitions of the genome. I would be more upfront about this, and try to explain it more simply and clearly.

\textbf{We have tried our best to clean up the section on recombination.}

8. Discussion. I think the results of this analysis on R0 and cluster size distribution could be compared with Cauchemez et al PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters. An important result of that paper was that the cluster size distributions were very skew - I think it would be good to look at skewness here.

\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al}. study did not share any data beyond the raw input file they used. Since the actual numbers in Fig 1C in the manuscript are not reported and a publicly available implementation of the MCMC algorithm do not exist we have included an additional figure that visualises the distribution of cases recovered by simulation. These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al}.}

% \bibliography{./../mers-structure}

\end{document}

### 2017-11-30 response_1/response_1.tex
don't agree with this comma =D

@@
 When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme.
-This in turn, informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
+This in turn informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
 \end{quotation}

### 2017-11-30 mers-structure.tex
don't agree with this comma =D

@@
 When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme.
-This in turn, informs the inferred state of long ancestral branches in the phylogeny, i.e.\ because these long branches are not immediately coalescing, they are most likely in camels.
+This in turn informs the inferred state of long ancestral branches in the phylogeny, i.e.\ because these long branches are not immediately coalescing, they are most likely in camels.
 

### 2017-11-30 response_1/response_1.tex
there is a period, you know? :x

@@
 \begin{quotation}
-Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models (Rasmussen \textit{et al}., 2014) which would be more appropriate for MERS-CoV.
+Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models (Rasmussen \textit{et al.}., 2014) which would be more appropriate for MERS-CoV.
 Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.
-Outside of coalescent-based models there are population structure models that explicitly relate epidemiological parameters to the branching process observed in sequence data (K\"{u}hnert \textit{et al}., 2016), but often rely on specifying numerous informative priors and can suffer from MCMC convergence issues.
+Outside of coalescent-based models there are population structure models that explicitly relate epidemiological parameters to the branching process observed in sequence data (K\"{u}hnert \textit{et al.}., 2016), but often rely on specifying numerous informative priors and can suffer from MCMC convergence issues.
 \end{quotation}
@@
 Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
-Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al.}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
+Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al..}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
 Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
-A similar approach has been attempted previously (Zhang \textit{et al.}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
+A similar approach has been attempted previously (Zhang \textit{et al..}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
 We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
@@
 The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger.
-Antia et al Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0.
+Antia et al.
+Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0.
 So R0~0.8 is much worse than R0~0.3.
@@
 We also believe that adaptive landscapes play a considerable role in emerging pandemics.
-We added an additional sentence to the discussion and refer to the Antia et al study:}
+We added an additional sentence to the discussion and refer to the Antia et al.
+study:}
 
 \begin{quotation}
-Previous modeling studies (Antia \textit{et al}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
+Previous modeling studies (Antia \textit{et al.}., 2003; Part \textit{et al..}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
 This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
@@
 During the time the study was performed (first half of 2017) human MERS-CoV sequences comprised a large majority of available MERS-CoV sequences.
-A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al.}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
+A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al..}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
 We are also aware of another study into camels across much of Africa.
@@
 Page 2, introduction.
-I am not sure I would refer to the Cauchemez et al PNAS paper as traditional epidemiology.
+I am not sure I would refer to the Cauchemez et al.
+PNAS paper as traditional epidemiology.
 It's sophisticated statistical modelling of spatio-temporal data and the partial observation process.
@@
 Discussion.
-I think the results of this analysis on R0 and cluster size distribution could be compared with Cauchemez et al PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters.
+I think the results of this analysis on R0 and cluster size distribution could be compared with Cauchemez et al.
+PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters.
 An important result of that paper was that the cluster size distributions were very skew - I think it would be good to look at skewness here.
 
-\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al}.
+\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al.}.
 study did not share any data beyond the raw input file they used.
 Since the actual numbers in Fig 1C in the manuscript are not reported and a publicly available implementation of the MCMC algorithm do not exist we have included an additional figure that visualises the distribution of cases recovered by simulation.
-These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al}.}
+These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al.}.}
 

### 2017-11-30 response_1/response_1.tex
oops

@@
 Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
-Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al..}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
+Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al.}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
 Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
-A similar approach has been attempted previously (Zhang \textit{et al..}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
+A similar approach has been attempted previously (Zhang \textit{et al.}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
 We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
@@
 \begin{quotation}
-Previous modeling studies (Antia \textit{et al.}., 2003; Part \textit{et al..}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
+Previous modeling studies (Antia \textit{et al.}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
 This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
@@
 During the time the study was performed (first half of 2017) human MERS-CoV sequences comprised a large majority of available MERS-CoV sequences.
-A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al..}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
+A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al.}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
 We are also aware of another study into camels across much of Africa.

### 2017-11-30 response_1/response_1.tex
amending what I had 'fixed' before; 'concerns' is plural

@@
 \begin{quotation}
-Previous modeling studies (Antia \textit{et al.}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
+Previous modeling studies (Antia \textit{et al}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
 This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
@@
 \textbf{An excellent point.
-Concerns over artefacts of recombination is what prevented us from reporting on evolutionary rates and tMRCAs of interest, such as our postulated introduction of the virus into the Arabian peninsula.
+Concerns over artefacts of recombination are what prevented us from reporting on evolutionary rates and tMRCAs of interest, such as our postulated introduction of the virus into the Arabian peninsula.
 We added a discussion about the potential effects of recombination in the results section and report evolutionary rates recovered by two methods:}

### 2017-12-01 mers-structure.tex
not sure what changed, sorry...

@@
 We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid_comparison}).
-However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
+However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder, as often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
 

### 2017-12-01 response_1/response_1.tex
waffling on about why we don't do ABC

@@
 \begin{quotation}
-However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies.
+However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder, as often seen in human influenza A virus phylogenies.
 \end{quotation}
@@
 \textbf{Our approach is indeed somewhat ad-hoc, though closely related to ABC.
-We have replaced the tolerance parameter of ABC with sequence data uncertainty and otherwise treat parameters of interest as being under uniform priors.
+We have essentially replaced the loss function in ABC with a slightly more complicated criterion based on  sequence data uncertainty.
+Our approach is also equivalent to a fully Bayesian setting where the parameters of interest are assigned uniform priors.
+Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.
 We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
+This is to be expected, since the sample mean cluster size is the sufficient statistic for $R_0$ and the MLE for $k$ is proportional to the second (non-standardised) moment of the cluster sizes.
 We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.}

### 2017-12-01 response_1/response_1.tex
'finally' better be in the end

@@
 Our approach is also equivalent to a fully Bayesian setting where the parameters of interest are assigned uniform priors.
-Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.
 We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
@@
 We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.}
+Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.
 

### 2017-12-03 response_1/response_1.tex
small fix

--- 2017-12-01 response_1/response_1.tex
@@
 Our approach is also equivalent to a fully Bayesian setting where the parameters of interest are assigned uniform priors.
-Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.
 We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
 This is to be expected, since the sample mean cluster size is the sufficient statistic for $R_0$ and the MLE for $k$ is proportional to the second (non-standardised) moment of the cluster sizes.
-We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.}
+We do see somewhat better separation of simulations when plotting mean sequence cluster size versus their skewness in all figures summarising the results of Monte Carlo simulations.
+Finally, while we completely agree with the reviewer that ABC would lend itself  very well to this estimation problem, it would also entail a good deal of calibration and experimentation, which we believe would be outside the scope of this paper.
+%LM: might want to re-order...
+}
 
@@
 \textbf{Thank you for the comment.
-It was our thinking that mean, median and standard deviation of cluster sizes would be able to capture both the overall magnitude and skew of the empirical distribution.
+% It was our thinking that mean, median and standard deviation of cluster sizes would be able to capture both the overall magnitude and skew of the empirical distribution.
+The reason for choosing the sample mean, median 
 We also accidentally omitted the number of clusters as a fourth parameter by which we filter simulated data.

### 2017-12-05 response_1/response_1.tex

@@
 
-1. Page 3 - where the authors state that ``Once human MERS-CoV infections are established, hoever, we find that MERS0CoV is poor at transmitted between humans" some references should be added.
+1. Page 3 - where the authors state that ``Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitted between humans" some references should be added.
 
@@
 % It was our thinking that mean, median and standard deviation of cluster sizes would be able to capture both the overall magnitude and skew of the empirical distribution.
-The reason for choosing the sample mean, median 
+The reason for choosing the sample mean, median and standard deviation was connected to the likelihood function we employed to model the distribution of cluster sizes.
+As stated in the reply to Reviewer \#1's comment 6.3, the sample mean is the sufficient statistic for $R0$ and the MLE for the dispersion parameter depends on the sample mean and standard deviation only.
 We also accidentally omitted the number of clusters as a fourth parameter by which we filter simulated data.
-In retrospect we should have been using skewness from the beginning. 
-We have since fixed both these issues and now use mean, standard deviation, skewness and number of clusters as filters for simulated data throughout the manuscript.
-This does not seem affect any of our results.}
+% In retrospect we should have been using skewness from the beginning. %LM: nah, I don't quite agree... Good idea in general, irrelevant here...
+In the interest of investigating robustness, we now use mean, standard deviation, skewness and number of clusters as filters for simulated data throughout the manuscript.
+This does not seem to affect any of our results.}
 

### 2017-12-05 mers-structure.tex

@@
 This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
-We record the number, mean, standard deviation and skewness (third standard moment) of sequence cluster sizes in each simulation (left-hand panel in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
+We record the number, mean, standard deviation and skewness (third standardised moment) of sequence cluster sizes in each simulation (left-hand panel in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
 We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.

### 2017-12-08 response_1/response_1.tex
Revision pass to response letter

@@
 The population genetic model (the particular form of structured coalescent) is highly idealised and this may influence the quantitative conclusions, although we suspect the conclusions are quite robust qualitatively.
-This model specifically estimates the rate of a lineage moving between demes going backwards in time; the numbers cited for the camel->human rate is really the rate that a lineage in humans goes to a camel going down the tree.
+This model specifically estimates the rate of a lineage moving between demes going backwards in time; the numbers cited for the camel$\rightarrow$human rate is really the rate that a lineage in humans goes to a camel going down the tree.
 The relationship between these migration rates and the epidemiologically meaningful transmission rate is complex and depends among other things on the ratio of population sizes in both demes.
-Per-capita transmission rates could be estimated using an epidemiologically structured coalescent model (see e.g.
-papers by Volz \& Rasmussen), which would ideally be stochastic due to bursty dynamics in humans.
+Per-capita transmission rates could be estimated using an epidemiologically structured coalescent model (see e.g.\ papers by Volz \& Rasmussen), which would ideally be stochastic due to bursty dynamics in humans.
 But this would be a large undertaking and so we suggest that for now the distinction is clarified.
@@
 \begin{quotation}
-Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models (Rasmussen \textit{et al.}., 2014) which would be more appropriate for MERS-CoV.
+Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate time varying migration rates and population sizes, integrate alternative sources of information and fit to stochastic nonlinear models (Rasmussen \textit{et al.}, 2014), which would be more appropriate for MERS-CoV.
 Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.
@@
 A similar approach has been attempted previously (Zhang \textit{et al.}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
-We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
+We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans ($n=174$) than from camels ($n=100$) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
 \end{quotation}
@@
 It is unusual for a phylogenetic ancestral reconstruction, that the results seem to be determined as much by the coalescent assumptions as by the tree topology.
-The two-patch model had a much higher coalescent rate in the human deme than in the camel deme - so long branches are only really possible in the camel deme.
+The two-patch model had a much higher coalescent rate in the human deme than in the camel deme -- so long branches are only really possible in the camel deme.
 This may be why for example, staring at the top clade of figure 1, one can see camel ancestry to a whole bunch of human sequences that are not topologically separated by camel sequences.
@@
 The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger.
-Antia et al.
-Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0.
-So R0~0.8 is much worse than R0~0.3.
-More sophisticated models have been done since, especially by Llyod-smith's group, but the basic result is sound.
+Antia et al.\ Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with $R_0>1$ increases dramatically as a function of the wild-type $R_0$.
+So $R_0\sim0.8$ is much worse than $R_0\sim0.3$.
+More sophisticated models have been done since, especially by Llyod-Smith's group, but the basic result is sound.
 In the light of this theoretical work, your findings are not at all reassuring.
@@
 We also believe that adaptive landscapes play a considerable role in emerging pandemics.
-We added an additional sentence to the discussion and refer to the Antia et al.
-study:}
+We added an additional sentence to the discussion and refer to the Antia et al.\ study:}
 
@@
 4.
-p 10, `...relatively low overall with a mean estimate of N of 3.49 years (95\% HPD: 2.71-4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies...' I would not say that low Ne directly implies a ladder-like tree or vice versa, although it may produce such a tree in combination with highly dispersed sample times.
+p 10, `...relatively low overall with a mean estimate of $N_e\tau$ of 3.49 years (95\% HPD: 2.71-4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies...' I would not say that low $N_e$ directly implies a ladder-like tree or vice versa, although it may produce such a tree in combination with highly dispersed sample times.
 
@@
 
-5. p 11, I think these back-of-the-envelope calculations to relate Ne to epidemic size are illuminating, however I would do some things differently:
+5. p 11, I think these back-of-the-envelope calculations to relate $N_e$ to epidemic size are illuminating, however I would do some things differently:
 
@@
 
-\textbf{}
+\textbf{Thank you for pointing this out.
+This was an error on our part.
+We've switched from a 10 day generation interval to a 20 day generation interval consistent with existing data and a simple SIR model.}
 
@@
 Other theoretical work suggests a slightly different relationship between Ne tau and I(t), specifically that it will depend more on transmission rate than generation time.
-However if making the approximation that I(t) is constant (which the authors are) , the formulas will be the same up to a factor of 2x.
+However if making the approximation that $I(t)$ is constant (which the authors are) , the formulas will be the same up to a factor of 2$\times$.
 See Koelle, 2011, DOI: 10.1098/rsif.2011.0495 \& Volz and Frost 2013, DOI: 10.1098/rstb.2010.0060.
-Specifically, I would use (Ne tau ) = I(t) / 2 * [per capita transmission rate], but some of the formulas in Koelle's paper may be more realistic for MERS.
-and equating beta = 1 / [generation time] gives 2x the authors' estimate if solving for I.
-Note that if using a 20 day generation interval and the above formula, you would have I = 127, so this doesn't make any practical difference, but is more elegant in my opinion, since we are exact about the epidemic model that the estimate is based on.
-If some estimate of the dispersion parameter for R0 could be obtained from the cluster size distribution, a better estimate for I could be obtained using equations in Koelle 2011.
+Specifically, I would use $N_e\tau = I(t) \, / \, 2 \, \times $ [per capita transmission rate], but some of the formulas in Koelle's paper may be more realistic for MERS.
+and equating $\beta = 1 \, / $ [generation time] gives 2$\times$ the authors' estimate if solving for $I$.
+Note that if using a 20 day generation interval and the above formula, you would have $I = 127$, so this doesn't make any practical difference, but is more elegant in my opinion, since we are exact about the epidemic model that the estimate is based on.
+If some estimate of the dispersion parameter for $R_0$ could be obtained from the cluster size distribution, a better estimate for $I$ could be obtained using equations in Koelle 2011.
 
-\textbf{}
+\textbf{This was helpful.
+We've reformulated this calculation to follow Frost and Volz (2010).
+We believe this reformulation streamlined the description.
+However, the switch from 10 day generation interval to 20 day interval resulted in a 2$\times$ difference to our final estimate.
+However, the final result of XXX is still consistent with broad conclusions of a congruence between camel incidence rates and observed $N_e$.
+Relevant text in Results has been reworked:}
+
+\begin{quotation}
+This empirically estimated effectived population can be compared to the expected effective population size in a simple epidemiological model.
+At endemic equilibrium, we expect scaled effective population size $N_{e}\tau$ to follow $I \, / \, 2 \beta$, where $\beta$ is the equilibrium rate of transmission and $I$ is the equilibrium number of infecteds (Frost and Volz, 2010).
+We assume that $\beta$ is constant and is equal to the rate of recovery.
+Given a 20 day duration of infection in camels (Adnet \textit{et al.}, 2014), we arrive at $\beta = 365/20 = 18.25$ infections per year.
+Given extremely high seroprevalence estimates within camels in Saudi Arabia (M\"uller \textit{et al.}, 2014; Corman \textit{et al.}, 2014; \textit{et al.}, 2014; Reusken \textit{et al.}, 2013, 2014), we expect camels to usually be infected within their first year of life.
+Therefore we can estimate the rough number of camel infections per year as the number of calves produced each year.
+We find there are $830\,000$ camels in Saudi Arabia (Abdallah and Faye, 2013) and that female camels in Saudi Arabia have an average fecundity of 45\% (Abdallah and Faye, 2013).
+Thus, we expect $830\,000 \times 0.50 \times 0.45 = 186\,750$ new calves produced yearly and correspondingly $186\,750$ new infections every year, which spread over 20 day intervals gives an average prevalence of $I = 10\,233$ infections.
+This results in an expected scaled effective population size $N_{e}\tau=280.4$ years.
+\end{quotation}
 
@@
 We have run small scale analyses over a grid of three bias levels (same as the main analyses), 23 levels of $R_{0}$ (0.50--1.05), and five levels of dispersion (0.004, 0.02, 0.10, 0.50 and 1.0) with 2000 replicate simulations for each combination of variables.
-Numbers of simulations matching empirical data are shown in Figure S15.}
+Numbers of simulations matching empirical data are shown in Figure S15.
+In this analysis, we find strongest support for dispersion $k$ = 0.1.}
 
@@
 I suspect the first three standardized moments of cluster sizes would be more useful than using mean/median/sd.
-- The moments could also be computed analytically if you dispense with the `sequencing capacity' aspect of the sampling model, so would be very fast.
+-- The moments could also be computed analytically if you dispense with the `sequencing capacity' aspect of the sampling model, so would be very fast.
 
@@
 We have followed the suggestion here to switch from mean/median/stdev to the first three standardised moments of sequence cluster size and found the results to be virtually identical.
+We have additionally included a 4th statistic of the number of observed sequence clusters.
 This is to be expected, since the sample mean cluster size is the sufficient statistic for $R_0$ and the MLE for $k$ is proportional to the second (non-standardised) moment of the cluster sizes.
@@
 2.
-In discussion: Could the author speculate how more sequences from humans and animals, particularly recent ones, or sequences from other geographic areas (east africa, north africa, pakistan) would potentially impace their results?
+In discussion: Could the author speculate how more sequences from humans and animals, particularly recent ones, or sequences from other geographic areas (East Africa, North Africa, Pakistan) would potentially impace their results?
 
@@
 Page 2, introduction.
-I am not sure I would refer to the Cauchemez et al.
-PNAS paper as traditional epidemiology.
+I am not sure I would refer to the Cauchemez et al.\ PNAS paper as traditional epidemiology.
 It's sophisticated statistical modelling of spatio-temporal data and the partial observation process.
@@
 
-4. Page 5 `R0 more familiar to epidemiologists' More familiar than what? There are no previous quantitative estimates presented. Rephrase.
+4. Page 5 `$R_0$ more familiar to epidemiologists' More familiar than what? There are no previous quantitative estimates presented. Rephrase.
 
@@
 The reason for choosing the sample mean, median and standard deviation was connected to the likelihood function we employed to model the distribution of cluster sizes.
-As stated in the reply to Reviewer \#1's comment 6.3, the sample mean is the sufficient statistic for $R0$ and the MLE for the dispersion parameter depends on the sample mean and standard deviation only.
+As stated in the reply to Reviewer \#1's comment 6.3, the sample mean is the sufficient statistic for $R_0$ and the MLE for the dispersion parameter depends on the sample mean and standard deviation only.
 We also accidentally omitted the number of clusters as a fourth parameter by which we filter simulated data.
@@
 In the interest of investigating robustness, we now use mean, standard deviation, skewness and number of clusters as filters for simulated data throughout the manuscript.
-This does not seem to affect any of our results.}
+This does not seem to affect our results.}
 
@@
 Discussion.
-I think the results of this analysis on R0 and cluster size distribution could be compared with Cauchemez et al.
-PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters.
+I think the results of this analysis on $R_0$ and cluster size distribution could be compared with Cauchemez et al.\ PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters.
 An important result of that paper was that the cluster size distributions were very skew - I think it would be good to look at skewness here.
@@
 study did not share any data beyond the raw input file they used.
-Since the actual numbers in Fig 1C in the manuscript are not reported and a publicly available implementation of the MCMC algorithm do not exist we have included an additional figure that visualises the distribution of cases recovered by simulation.
+Since the actual numbers in Fig 1C in the manuscript are not reported and a publicly available implementation of the MCMC algorithm do not exist we have included an additional figure (supplementary Figure S8) that visualises the distribution of cases recovered by simulation.
 These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al.}.}

### 2017-12-08 mers-structure.tex
Revision pass to response letter

@@
 A similar approach has been attempted previously \citep{zhang_evolutionary_2016}, but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
-We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
+We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans ($n=174$) than from camels ($n=100$) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
 We can replicate these results by either applying the same model to current data (Figure \ref{dta}) or by enforcing equal coalescence rates within each deme in the structured coalescent model (Figure \ref{equal_sizes}).
@@
 When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme.
-This in turn informs the inferred state of long ancestral branches in the phylogeny, i.e.\ because these long branches are not immediately coalescing, they are most likely in camels.
+This in turn informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
 
@@
 As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{mers_epi}).
-Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models \citep{rasmussen_phylodynamic_2014} which would be more appropriate for MERS-CoV.
+Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate time varying migration rates and population sizes, integrate alternative sources of information and fit to stochastic nonlinear models \citep{rasmussen_phylodynamic_2014}, which would be more appropriate for MERS-CoV.
 Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.

### 2017-12-08 response_1/response_1.tex
Fix missing statistic

@@
 However, the switch from 10 day generation interval to 20 day interval resulted in a 2$\times$ difference to our final estimate.
-However, the final result of XXX is still consistent with broad conclusions of a congruence between camel incidence rates and observed $N_e$.
+However, the final result of $N_{e}\tau=280.4$ years is still consistent with broad conclusions of a congruence (or at least no worse a congruence than measles in humans) between camel incidence rates and observed $N_e\tau$.
 Relevant text in Results has been reworked:}

### 2017-12-11 response_1/response_1.tex
Last changes to response.

@@
 Although we recover migration rates from our model (Figure S2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
-\end{quotation}
+%\end{quotation}
 
-\begin{quotation}
+[...]
+
+%\begin{quotation}
 Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate time varying migration rates and population sizes, integrate alternative sources of information and fit to stochastic nonlinear models (Rasmussen \textit{et al.}, 2014), which would be more appropriate for MERS-CoV.
@@
 \begin{quotation}
-Previous modeling studies (Antia \textit{et al}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
+Previous modeling studies (Antia \textit{et al}., 2003; Park \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
 This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
@@
 Indeed all cases of using BEAST 1 vs BEAST 2 came down to what models were implemented in which package.
-We have tried clarified our choices throughout the manuscript.}
+We have clarified our choices throughout the manuscript.}
 
@@
 
-\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al.}.
+\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al}.
 study did not share any data beyond the raw input file they used.
-Since the actual numbers in Fig 1C in the manuscript are not reported and a publicly available implementation of the MCMC algorithm do not exist we have included an additional figure (supplementary Figure S8) that visualises the distribution of cases recovered by simulation.
-These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al.}.}
+Since neither the actual numbers in Fig 1C of Cauchemez \textit{et al}., nor an implementation of the MCMC algorithm exist in the public domain we have included an additional figure (supplementary Figure S8) that visualises the distribution of cases recovered by simulation.
+These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al}.}
 

### 2017-12-12 mers-structure.tex
Reduce abstract to ~150 words.

@@
 
-Middle East respiratory syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
-The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
-While case-based and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
-Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
+Middle East respiratory syndrome coronavirus (MERS-CoV) is a zoonotic virus from camels causing significant mortality and morbidity in humans in the Arabian Peninsula.
+The epidemiology of the virus remains poorly understood, and while case-based and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential.
+Here we use existing MERS-CoV sequence data to explore its phylodynamics in two of its known major hosts, humans and camels.
 We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.

### 2017-12-12 mers-structure.tex
Reformat from supplementary to supplements.

@@
 \usepackage{xr}
-\externaldocument{mers-structure_supp}
+%\externaldocument{mers-structure_supp}
 
@@
 
-The structured coalescent approach we employ (see Methods) identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), while human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
+The structured coalescent approach we employ (see Methods) identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), while human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{mcc}-Figure supplement 1).
 Across 174 MERS-CoV genomes collected from humans, we estimate a median of 56 separate camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
 While we estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
-Correspondingly, we observe substantially higher camel-to-human migration rate estimates than human-to-camel migration rate estimates (Figure~\ref{prior}).
+Correspondingly, we observe substantially higher camel-to-human migration rate estimates than human-to-camel migration rate estimates (Figure~\ref{mcc}-Figure supplement 2).
 This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show significantly higher diversity and less clustering.
@@
  \centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_mcc.png}
+	\includegraphics[width=0.75\textwidth]{figures/figure 1.png}
 	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from 174 human viruses and 100 camel viruses.}
@@
 	\label{mcc}
+
+  \figsupp{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
+  This is the same tree as shown in Figure \ref{mcc}, but with contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top in orange) and humans (bottom in blue).
+  This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.}{\includegraphics[width=0.65\textwidth]{figures/figure 1-supplement 1.png}}
+
+  \figsupp{\textbf{Posterior backwards migration rate estimates for two choices of prior.}
+  Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice (note that rates are backwards in time).
+  Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme looking backwards in time (orange) and \textit{vice versa} (blue).
+  Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).}{\includegraphics[width=0.65\textwidth]{figures/figure 1-supplement 2.png}}
+
+  \figsupp{\textbf{Maximum clade credibility (MCC) tree with ancestral state reconstruction according to a discrete trait model.}
+  MCC tree is presented the same as Figure \ref{mcc} and Figure \ref{mcc}-Figure supplement 4, with colours indicating the most probable state reconstruction at internal nodes.
+  Unlike the structured coalescent summary shown in Figure \ref{mcc} where camels are reconstructed as the main host where MERS-CoV persists, the discrete trait approach identifies both camels and humans as major hosts with humans being the source of MERS-CoV infection in camels.}{\includegraphics[width=0.65\textwidth]{figures/figure 1-supplement 3.png}}
+
+  \figsupp{\textbf{Maximum clade credibility (MCC) tree of structured coalescent model with enforced equal coalescence rates.}
+  MCC tree is presented the same as Figures \ref{mcc} and \ref{mcc}-Figure supplement 3, with colours indicating the most probable state reconstruction at internal nodes.
+  Similar to Figure \ref{mcc}-Figure supplement 3 enforcing equal coalescence rates between demes in a structured coalescent model identifies humans as a major MERS-CoV host and the source of viruses in camels.}{\includegraphics[width=0.65\textwidth]{figures/figure 1-supplement 4.png}}
+
+  \figsupp{\textbf{Maximum likelihood (ML) tree of MERS-CoV genomes coloured by origin of sequence.}
+  Maximum likelihood tree shows genetic divergence between MERS-CoV genomes collected from camels (orange tips) and humans (blue tips).}{\includegraphics[width=0.65\textwidth]{figures/figure 1-supplement 5.png}}
+
 \end{figure}
@@
 The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
-Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).
-
-The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
+Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{mcc}-Figure supplement 2).
+
+The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{mcc}-Figure supplement 1).
 We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
@@
 Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
-Comparing these results with a currently standard discrete trait analysis \citep{lemey_bayesian_2009} approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure \ref{dta}).
+Comparing these results with a currently standard discrete trait analysis \citep{lemey_bayesian_2009} approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure \ref{mcc}-Figure supplement 3).
 Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
@@
 We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans ($n=174$) than from camels ($n=100$) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
-We can replicate these results by either applying the same model to current data (Figure \ref{dta}) or by enforcing equal coalescence rates within each deme in the structured coalescent model (Figure \ref{equal_sizes}).
+We can replicate these results by either applying the same model to current data (Figure \ref{mcc}-Figure supplement 3) or by enforcing equal coalescence rates within each deme in the structured coalescent model (Figure \ref{mcc}-Figure supplement 4).
 
@@
 \centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_seasonality.png}
+	\includegraphics[width=0.75\textwidth]{figures/figure 2.png}
 	\caption{\textbf{Seasonality of MERS-CoV introduction events.}
@@
 Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
-In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones (for its effects see Figure \ref{mers_epi_grid}).
+In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones (for its effects see Figure \ref{mers_epi}-Figure supplement 1).
 This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
@@
 We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
-A schematic of this Monte Carlo procedure is shown in Figure \ref{mc_method}.
+A schematic of this Monte Carlo procedure is shown in Figure \ref{mers_epi}-Figure supplement 2.
 Since the total number of cases is fixed at 2000, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters.
@@
 \centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_epi.png}
+	\includegraphics[width=0.75\textwidth]{figures/figure 3.png}
 	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
@@
 	\label{mers_epi}
+
+  \figsupp{\textbf{Monte Carlo simulations of human transmission clusters.}
+  From top to bottom each row corresponds to departures from completely random sequencing efforts with respect to case cluster size (bias parameter=1.0) to sequencing increasingly biased towards capturing large case clusters (bias=2.0, bias=3.0).
+  Leftmost scatter plots show the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) coloured by the $R_{0}$ value used for the simulation.
+  The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and skewness observed for empirical MERS-CoV data.
+  The distribution of $R_{0}$ values matching empirical data are shown in the middle, on the same $y$-axis across all levels of the bias parameter.
+  Under unbiased sequencing (bias=1.0) only 0.45\% of simulations fit our phylogenetic observations, while 1.79\% and 1.67\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
+  Correspondingly, we estimate 11.6\% support for a model with bias level 1.0, 45.7\% support for a model with bias level 2.0, and 42.7\% support for a model with bias level 3.0.
+  Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
+  While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.71--0.98) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters.
+  The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
+  Darker shade of grey indicates bins falling within the 95\% percentiles.
+  The median number of cross-species introductions observed in simulations matching empirical data without bias are 346 (95\% percentiles 262--439).
+  These numbers jump up to 568 (95\% percentiles 430--727) for bias = 2.0 and 656 (95\% percentiles 488--853) for bias = 3.0 simulations.
+  Model averaging would suggest plausible numbers of introductions between 311 and 811.}{\includegraphics[width=0.75\textwidth]{figures/figure 3-supplement 1.png}}
+
+  \figsupp{\textbf{Monte Carlo simulation schematic.}
+  Case clusters are simulated according to Equation \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
+  We sample 174 cases from each simulation to represent sequencing of human MERS cases.
+  `Sequencing' is carried out by using multivariate hypergeometric sampling, representing sampling cases without replacement to be sequenced.
+  Sequencing simulations take place at three levels of bias: 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
+  The distribution of simulated sequence clusters is summarised by its mean, median and standard deviation.
+  A simulation is considered to match if the mean, median and standard deviation of its sequence cluster sizes falls within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
+  $R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are retained as estimates.
+  These estimates are summarised in Figure \ref{mers_epi}.}{\includegraphics[width=0.75\textwidth]{figures/figure 3-supplement 2.pdf}}
+
+  \figsupp{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
+  The plot is identical to Figure \ref{mers_epi}-Figure supplement 1, but instead of 2000 cases,
+  % the currently reported number of MERS-CoV cases,
+  simulations were run with 4000 cases.
+  With more unobserved cases the R$_{0}$ values matching observed MERS-CoV sequence clusters can only be smaller, with a corresponding increase in numbers of zoonotic transmissions.
+  However, the numbers of simulations that match MERS-CoV data go down as well.}{\includegraphics[width=0.65\textwidth]{figures/figure 3-supplement 3.png}}
+
+  \figsupp{\textbf{Boxplots of matching simulated case and sequence cluster distributions.}
+  Boxplots indicate frequency of case (blue, top) and sequence (red, bottom) cluster sizes across simulations at different bias levels, marginalised across $R_{0}$ values.
+  Outliers are shown with transparency, medians are indicated with thick black lines.
+  Case clusters exhibit a strong skew with large numbers of singleton introductions and a substantial tail at higher levels of bias.}{\includegraphics[width=0.65\textwidth]{figures/figure 3-supplement 4.png}}
+
+  \figsupp{\textbf{Quantile-quantile (Q-Q) plot of empirical and simulated sequence cluster sizes.}
+  Density of sequence cluster size percentiles (1st--99th, calculated across a grid of 50 values) calculated for random states from the posterior distribution ($x$-axis) and matching simulations ($y$-axis).
+  Most values fall on the one-to-one line, with a heavier tail in mid-sized sequence clusters in empirical data, manifesting as a greater density of points below the one-to-one line in the middle.}{\includegraphics[width=0.65\textwidth]{figures/figure 3-supplement 5.png}}
+
+  \figsupp{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
+  Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
+  More simulations match observations with bias$>1$ and $\omega \approx 0.1$.}{\includegraphics[width=0.65\textwidth]{figures/figure 3-supplement 6.png}}
+
 \end{figure}
@@
 Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
-In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).
-
-Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.90$ and sequence sampling likely to be biased towards large hospital outbreaks (Figure \ref{mers_epi_grid}).
-All matching simulations exhibit highly skewed distributions of case cluster sizes with long tails (Figure \ref{cluster_distributions}), which is qualitatively similar to the results of \citep{cauchemez_unraveling_2016}.
-We find that simulated sequence cluster sizes resemble observed sequence cluster sizes in the posterior distribution, with a slight reduction in mid-sized clusters in simulated data (Figure \ref{qq}).
+In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{mers_epi}-Figure supplement 3).
+
+Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<0.90$ and sequence sampling likely to be biased towards large hospital outbreaks (Figure \ref{mers_epi}-Figure supplement 1).
+All matching simulations exhibit highly skewed distributions of case cluster sizes with long tails (Figure \ref{mers_epi}-Figure supplement 4), which is qualitatively similar to the results of \citep{cauchemez_unraveling_2016}.
+We find that simulated sequence cluster sizes resemble observed sequence cluster sizes in the posterior distribution, with a slight reduction in mid-sized clusters in simulated data (Figure \ref{mers_epi}-Figure supplement 5).
 Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
+    \includegraphics[width=0.95\textwidth]{figures/figure 4.png}
     \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
@@
     \label{recombinant_features}
+
+    \figsupp{\textbf{Tests of recombination across MERS-CoV clades.}
+    Maximum clade credibility tree of MERS-CoV genomes annotated with results of two recombination detection tests (PHI and 3Seq) applied to descendent sequences of each clade.
+    Both tests identify large portions of existing sequence data as containing signals of recombination.
+    Note that markings do not indicate where recombinations have occurred on the tree, merely the minimum distance in sequence/time space between recombining lineages.}{\includegraphics[width=0.65\textwidth]{figures/figure 4-supplement 1.png}}
+
+    \figsupp{\textbf{MERS-CoV genomes exhibit high numbers of non-clonal loci.}
+    Ancestral state reconstruction (right) identifies a large number of sites in which mutations have occurred more than once in the tree (homoplasies, orange) or are reversions (red) from a state arising in an ancestor.
+    Mutations that apparently only occur once in the tree (synapomorphies) are shown in grey.
+    The maximum likelihood phylogeny on the left is coloured by whether sequences were sampled in humans (blue) or camels (orange).}{\includegraphics[width=0.75\textwidth]{figures/figure 4-supplement 2.png}}
+
+    \figsupp{\textbf{Human clade sharing between genomic fragments 1 and 2.}
+    Central scatter plot shows the posterior probability of human clades shared between genomic fragments 1 and 2, in their respective trees.
+    Left and bottom scatter plots track the posterior probability of human clades only observed in fragment 2 (left) or fragment 1 (bottom).
+    The cumulative probability of human clades present in either tree are tracked by plots on the right (fragment 2) and top (fragment 1).
+    Most of the probability mass is concentrated within human clades that are present in trees of both genomic fragment 1 and 2 (0.9701 and 0.9474 of all human clades across posteriors, respectively).}{\includegraphics[width=0.65\textwidth]{figures/figure 4-supplement 3.png}}
+
 \end{figure}
@@
 Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
-PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (Figure \ref{recombination_tree}).
-Neither method can identify where in the tree recombination occurred, but each full asterisk in Figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
+PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (Figure \ref{recombinant_features}-Figure supplement 1).
+Neither method can identify where in the tree recombination occurred, but each full asterisk in Figure \ref{recombinant_features}-Figure supplement 1 should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
 This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
-As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{incompatibilities}) with some evidence of genomic clustering of such alleles.
+As done previously \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (Figure \ref{recombinant_features}-Figure supplement 2) with some evidence of genomic clustering of such alleles.
 % Homoplasies are present in multiple strains at a time, indicating recombination events that have been successful.
@@
 All such clades have low divergence (figure \ref{recombinant_features}B) and thus incongruences in human clades are more likely to be caused by differences in deme assignment rather than actual recombination.
-And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
+And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{recombinant_features}-Figure supplement 3).
 Additionally, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
@@
 \centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid.png}
+	\includegraphics[width=0.65\textwidth]{figures/figure 5.png}
 	\caption{\textbf{Demographic history of MERS-CoV in Arabian peninsula camels.}
@@
 	\label{skygrid}
+
+  \figsupp{\textbf{Skygrid comparison between whole and fragmented genomes.}
+  Inferred median $N_{e}\tau$ recovered using a skygrid tree prior on whole genome (bottom) and ten genomic fragments with independent trees (left), coloured by time.
+  Dotted line indicates the one-to-one line.}{\includegraphics[width=0.65\textwidth]{figures/figure 5-supplement 1.png}}
 \end{figure}
@@
 Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
-We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid_comparison}).
+We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid}-Figure supplement 1).
 However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder, as often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
@@
 In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula by using sequence data collected from both hosts (174 from humans and 100 from camels).
-We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV genomic data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
-This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{mers_epi}).
-When rapid coalescence in the human deme is not allowed (Figure \ref{equal_sizes}) structured coalescent inference loses power and ancestral state reconstruction is nearly identical to that of discrete trait analysis (Figure \ref{dta}).
+We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV genomic data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{mcc}-Figure supplement 1).
+This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{mcc}-Figure supplement 2) and in the presence of strong biases in sequence sampling schemes (Figure \ref{mers_epi}).
+When rapid coalescence in the human deme is not allowed (Figure \ref{mcc}-Figure supplement 4) structured coalescent inference loses power and ancestral state reconstruction is nearly identical to that of discrete trait analysis (Figure \ref{mcc}-Figure supplement 3).
 When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme.
@@
 From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{mers_epi}).
-Although we recover migration rates from our model (Figure \ref{prior}), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
+Although we recover migration rates from our model (Figure \ref{mcc}-Figure supplement 2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
 We also looked at potential seasonality in MERS-CoV spillover into humans.
@@
 
-All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
+All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Supplementary File 1.% Table \ref{sequences}.
 Sequences without accessions were kindly shared by Ali M.\ Somily, Mazin Barry, Sarah S.\ Al Subaie, Abdulaziz A.\ BinSaeed, Fahad A.\ Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N.\ McNabb, Imad A.\ Al-Jahdali, Ahmed M.\ Alotaibi, Nahid A.\ Batarfi, Matthew Cotten, Simon J.\ Watson, Spela Binter, and Paul Kellam prior to publication.
@@
 Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364, which allowed for codon-partitioned substitution models to be used.
-The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).
+The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Supplementary File 1). %(Table \ref{sequences}).
 
@@
 Posterior sets of typed (where migrating branches from structured coalescent are collapsed, and migration information is left as a switch in state between parent and descendant nodes) trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.
-A maximum likelihood phylogeny showing just the genetic relationships between MERS-CoV genomes from camels and humans was recovered using PhyML \citep{guindon_simple_2003} under a HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution model and is shown in Figure \ref{ml}.
+A maximum likelihood phylogeny showing just the genetic relationships between MERS-CoV genomes from camels and humans was recovered using PhyML \citep{guindon_simple_2003} under a HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution model and is shown in Figure \ref{mcc}-Figure supplement 5.
 
@@
 
-Here, we employ a Monte Carlo simulation approach to identify parameters consistent with observed patterns of sequence clustering (Figure \ref{mc_method}).
+Here, we employ a Monte Carlo simulation approach to identify parameters consistent with observed patterns of sequence clustering (Figure \ref{mers_epi}-Figure supplement 2).
 Our structured coalescent analyses indicate that every MERS outbreak is a contained cross-species spillover of the virus from camels into humans.
@@
 We performed a smaller set of simulations with 2500 replicates and twice the number of cases, \textit{i.e.} $\sum_{i=1}^{K} C_{i} = 4000$, to explore a dramatically underreported epidemic.
-Additionally, we performed additional smaller set of simulations on a rougher grid of $R_{0}$ values (23 values, 0.50--1.05), with 5 values of dispersion parameter $\omega$ (0.002, 0.04, 0.1, 0.5, 1.0) and 3 levels of bias (1, 2, 3) to justify our choice of dispersion parameter $\omega$ that was fixed to 0.1 in the main analyses (Figure \ref{marginals}).
+Additionally, we performed additional smaller set of simulations on a rougher grid of $R_{0}$ values (23 values, 0.50--1.05), with 5 values of dispersion parameter $\omega$ (0.002, 0.04, 0.1, 0.5, 1.0) and 3 levels of bias (1, 2, 3) to justify our choice of dispersion parameter $\omega$ that was fixed to 0.1 in the main analyses (Figure \ref{mers_epi}-Figure supplement 6).
 
@@
 
-\newpage
-
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-
-\begin{longtable}{ | r | l | p{2cm} | l | l | }
-
-  \caption{Strain names, accessions (where available), identified host and reported collection dates for MERS-CoV genomes used in this study.} \label{sequences} \\
-
-  \hline & strain & accession & host & collection date \\ \hline
-  \endfirsthead
-
-  \multicolumn{5}{l}%
-  {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
-
-  \hline & strain & accession & host & collection date \\ \hline
-  \endhead
-
-  \hline \multicolumn{5}{|r|}{{Continued on next page}} \\ \hline
-  \endfoot
-
-  \hline \hline
-  \endlastfoot
-
-  1 & KSA-378 & KJ713296 & camel & 2013-11 \\
-  2 & KSA-363 & KJ713298 & camel & 2013-11 \\
-  3 & KSA-503 & KJ713297 & camel & 2013-11 \\
-  4 & KSA-376 & KJ713299 & camel & 2013-11 \\
-  5 & KSA-505 & KJ713295 & camel & 2013-11 \\
-  6 & Jeddah-1 & KF917527 & camel & 2013-11-08 \\
-  7 & NRCE-HKU205 & KJ477102 & camel & 2013-11-15 \\
-  8 & KFU-HKU1 & KJ650297 & camel & 2013-11-30 \\
-  9 & KFU-HKU13 & KJ650295 & camel & 2013-12-30 \\
-  10 & Camel\_Egypt\_NRCE-HKU271 &  & camel & 2013-12-30 \\
-  11 & Camel\_Egypt\_NRCE-HKU270 &  & camel & 2013-12-30 \\
-  12 & KFU-HKU19Dam & KJ650296 & camel & 2013-12-30 \\
-  13 & Qatar\_2\_2014 & KJ650098 & camel & 2014-02-16 \\
-  14 & UAE/D469-14 & KU242424 & camel & 2014-03-04 \\
-  15 & UAE/D511-14 & KU242423 & camel & 2014-03-12 \\
-  16 & Jeddah/F13A/2014 & KT368824 & camel & 2014-05 \\
-  17 & UAE/D1164.10/2014 & KP719928 & camel & 2014-06 \\
-  18 & UAE/D1339.2/2014 & KP719931 & camel & 2014-06 \\
-  19 & UAE/D1164.11/2014 & KP719929 & camel & 2014-06 \\
-  20 & UAE/D1164.9/2014 & KP719927 & camel & 2014-06 \\
-  21 & UAE/D1209/2014 & KP719933 & camel & 2014-06 \\
-  22 & UAE/D1164.14/2014 & KP719930 & camel & 2014-06 \\
-  23 & UAE/D1243.12/2014 & KP719932 & camel & 2014-06 \\
-  24 & D1164.1/14 & KX108937 & camel & 2014-06-02 \\
-  25 & Riyadh/Ry23N/2014 & KT368825 & camel & 2014-07 \\
-  26 & Riyadh/Ry84N/2014 & KT368826 & camel & 2014-07 \\
-  27 & Jeddah/S93/2014 & KT368855 & camel & 2014-09 \\
-  28 & Jeddah/401/2014 & KT368827 & camel & 2014-09 \\
-  29 & Jeddah/S100/2014 & KT368853 & camel & 2014-09 \\
-  30 & Jeddah/S99/2014 & KT368857 & camel & 2014-09 \\
-  31 & Jeddah/S94/2014 & KT368856 & camel & 2014-09 \\
-  32 & Jeddah/S73/2014 & KT368854 & camel & 2014-09 \\
-  33 & Jeddah/O47b/2014 & KT368852 & camel & 2014-10 \\
-  34 & Jeddah/O23b/2014 & KT368849 & camel & 2014-10 \\
-  35 & Jeddah/O24/2014 & KT368850 & camel & 2014-10 \\
-  36 & Jeddah/O30/2014 & KT368851 & camel & 2014-10 \\
-  37 & Jeddah/N51/2014 & KT368846 & camel & 2014-11 \\
-  38 & Jeddah/N68b/2014 & KT368848 & camel & 2014-11 \\
-  39 & Jeddah/N62b/2014 & KT368847 & camel & 2014-11 \\
-  40 & Jeddah/D40/2014 & KT368834 & camel & 2014-12 \\
-  41 & Jeddah/D90/2014 & KT368844 & camel & 2014-12 \\
-  42 & Jeddah/D88/2014 & KT368843 & camel & 2014-12 \\
-  43 & Jeddah/D36/2014 & KT368832 & camel & 2014-12 \\
-  44 & Jeddah/D35/2014 & KT368831 & camel & 2014-12 \\
-  45 & Jeddah/D92/2014 & KT368845 & camel & 2014-12 \\
-  46 & Jeddah/D49/2014 & KT368841 & camel & 2014-12 \\
-  47 & Jeddah/D34/2014 & KT368830 & camel & 2014-12 \\
-  48 & Jeddah/D33b/2014 & KT368829 & camel & 2014-12 \\
-  49 & Jeddah/D42/2014 & KT368835 & camel & 2014-12 \\
-  50 & Jeddah/D50b/2014 & KT368842 & camel & 2014-12 \\
-  51 & Jeddah/D45/2014 & KT368837 & camel & 2014-12 \\
-  52 & Jeddah/D46b/2014 & KT368838 & camel & 2014-12 \\
-  53 & Jeddah/D43b/2014 & KT368836 & camel & 2014-12 \\
-  54 & Jeddah/D100/2014 & KT368828 & camel & 2014-12 \\
-  55 & Jeddah/D47/2014 & KT368839 & camel & 2014-12 \\
-  56 & Jeddah/D38b/2014 & KT368833 & camel & 2014-12 \\
-  57 & Jeddah/D48/2014 & KT368840 & camel & 2014-12 \\
-  58 & D2597.2/14 & KX108938 & camel & 2014-12-13 \\
-  59 & Egypt\_NRCE-NC163/2014 & KU740200 & camel & 2014-12-17 \\
-  60 & Jeddah/Jd7/2015 & KT368861 & camel & 2015-01 \\
-  61 & Jeddah/Jd86/2015 & KT368863 & camel & 2015-01 \\
-  62 & Jeddah/Jd90/2015 & KT368865 & camel & 2015-01 \\
-  63 & Jeddah/Jd1b/2015 & KT368858 & camel & 2015-01 \\
-  64 & Jeddah/Jd4/2015 & KT368859 & camel & 2015-01 \\
-  65 & Jeddah/Jd85/2015 & KT368862 & camel & 2015-01 \\
-  66 & Jeddah/Jd6b/2015 & KT368860 & camel & 2015-01 \\
-  67 & Jeddah/Jd87/2015 & KT368864 & camel & 2015-01 \\
-  68 & D252/15 & KX108939 & camel & 2015-01-30 \\
-  69 & Jeddah/Jd199/2015 & KT368867 & camel & 2015-02 \\
-  70 & Jeddah/Jd175/2015 & KT368866 & camel & 2015-02 \\
-  71 & D374/15 & KX108940 & camel & 2015-02-12 \\
-  72 & D383/15 & KX108941 & camel & 2015-02-14 \\
-  73 & D389/15 & KX108942 & camel & 2015-02-15 \\
-  74 & Riyadh/Ry63/2015 & KT368876 & camel & 2015-03 \\
-  75 & Riyadh/Ry136/2015 & KT368868 & camel & 2015-03 \\
-  76 & Riyadh/Ry178/2015 & KT368874 & camel & 2015-03 \\
-  77 & Riyadh/Ry162/2015 & KT368871 & camel & 2015-03 \\
-  78 & Riyadh/Ry86/2015 & KT368879 & camel & 2015-03 \\
-  79 & Taif/T150/2015 & KT368889 & camel & 2015-03 \\
-  80 & Riyadh/Ry137/2015 & KT368869 & camel & 2015-03 \\
-  81 & Riyadh/Ry179/2015 & KT368875 & camel & 2015-03 \\
-  82 & Riyadh/Ry177/2015 & KT368873 & camel & 2015-03 \\
-  83 & Riyadh/Ry79/2015 & KT368878 & camel & 2015-03 \\
-  84 & Riyadh/Ry173/2015 & KT368872 & camel & 2015-03 \\
-  85 & Taif/T157b/2015 & KT368890 & camel & 2015-03 \\
-  86 & Riyadh/Ry159b/2015 & KT368870 & camel & 2015-03 \\
-  87 & Riyadh/Ry64/2015 & KT368877 & camel & 2015-03 \\
-  88 & Taif/T3/2015 & KT368880 & camel & 2015-04 \\
-  89 & Taif/T16/2015 & KT368882 & camel & 2015-04 \\
-  90 & Taif/T22/2015 & KT368883 & camel & 2015-04 \\
-  91 & Taif/T92/2015 & KT368887 & camel & 2015-04 \\
-  92 & Taif/T7/2015 & KT368881 & camel & 2015-04 \\
-  93 & Taif/T91b/2015 & KT368886 & camel & 2015-04 \\
-  94 & Taif/T68/2015 & KT368884 & camel & 2015-04 \\
-  95 & Taif/T89/2015 & KT368885 & camel & 2015-04 \\
-  96 & Taif/T98/2015 & KT368888 & camel & 2015-04 \\
-  97 & D998/15 & KX108943 & camel & 2015-04-23 \\
-  98 & D1157/15 & KX108944 & camel & 2015-05-12 \\
-  99 & D1189.1/15 & KX108946 & camel & 2015-05-18 \\
-  100 & D1271/15 & KX108945 & camel & 2015-05-29 \\
-  101 & Jordan-N3/2012 & KC776174 & human & 2012-04-15 \\
-  102 & EMC/2012 & JX869059 & human & 2012-06-13 \\
-  103 & England/1/2012 & KC164505 & human & 2012-09-11 \\
-  104 & Riyadh\_1\_2012 & KF600612 & human & 2012-10-23 \\
-  105 & Riyadh\_2\_2012 & KF600652 & human & 2012-10-30 \\
-  106 & Riyadh\_3\_2013 & KF600613 & human & 2013-02-05 \\
-  107 & England/3/2013 & KM210278 & human & 2013-02-10 \\
-  108 & England/2/2013 & KM015348 & human & 2013-02-10 \\
-  109 & England/4/2013 & KM210277 & human & 2013-02-13 \\
-  110 & Riyadh\_4\_2013 & KJ156952 & human & 2013-03-01 \\
-  111 & Munich/AbuDhabi/2013 & KF192507 & human & 2013-03-22 \\
-  112 & Al-Hasa\_2\_2013 & KF186566 & human & 2013-04-21 \\
-  113 & Al-Hasa\_3\_2013 & KF186565 & human & 2013-04-22 \\
-  114 & UAE-FRA1\_1627-2013\_BAL & KJ361500 & human & 2013-04-26 \\
-  115 & Al-Hasa\_4\_2013 & KF186564 & human & 2013-05-01 \\
-  116 & Al-Hasa\_7\_2013 & KF600623, KF600655 & human & 2013-05-01 \\
-  117 & Al-Hasa\_8\_2013 & KF600618, KF600626, KF600635, KF600638 & human & 2013-05-01 \\
-  118 & Al-Hasa\_25\_2013 & KJ156866 & human & 2013-05-02 \\
-  119 & Al-Hasa\_11\_2013 & KF600629, KF600636, KF600646 & human & 2013-05-03 \\
-  120 & Al-Hasa\_12\_2013 & KF600627 & human & 2013-05-07 \\
-  121 & Al-Hasa\_14\_2013 & KF600615, KF600643 & human & 2013-05-08 \\
-  122 & Al-Hasa\_1\_2013 & KF186567 & human & 2013-05-09 \\
-  123 & Al-Hasa\_15\_2013 & KF600645 & human & 2013-05-11 \\
-  124 & Al-Hasa\_16\_2013 & KF600644 & human & 2013-05-12 \\
-  125 & Buraidah\_1\_2013 & KF600630 & human & 2013-05-13 \\
-  126 & Al-Hasa\_23\_2013 & KJ156860, KJ156894, KJ156929, KJ156923, KJ156862 & human & 2013-05-13 \\
-  127 & Al-Hasa\_17\_2013 & KF600647 & human & 2013-05-15 \\
-  128 & Al-Hasa\_19\_2013 & KF600632 & human & 2013-05-23 \\
-  129 & Al-Hasa\_18\_2013 & KF600651 & human & 2013-05-23 \\
-  130 & Al-Hasa\_21\_2013 & KF600634 & human & 2013-05-30 \\
-  131 & Hafr-Al-Batin\_1\_2013 & KF600628 & human & 2013-06-04 \\
-  132 & Wadi-Ad-Dawasir\_1\_2013 & KJ156881 & human & 2013-06-12 \\
-  133 & Taif\_1\_2013 & KJ156949 & human & 2013-06-12 \\
-  134 & Taif\_2\_2013 & KJ156896, KJ156876 & human & 2013-06-12 \\
-  135 & Taif\_3\_2013 & KJ156938, KJ156897, KJ156922, KJ156868, KJ156921, KJ156915,
-KJ156906 & human & 2013-06-13 \\
-  136 & Al-Hasa\_26\_2013 & KJ156882, KJ156941, KJ156872 & human & 2013-06-18 \\
-  137 & Al-Hasa\_27\_2013 & KJ156943, KJ156939 & human & 2013-06-19 \\
-  138 & Al-Hasa\_28\_2013 & KJ156887, KJ156940, KJ156889, KJ156893, KJ156884, KJ156930,
-KJ156928, KJ156909 & human & 2013-06-22 \\
-  139 & Riyadh\_6\_2013 & KJ156879, KJ156947, KJ156890, KJ156908, KJ156927 & human & 2013-07-02 \\
-  140 & Riyadh\_5\_2013 & KJ156944 & human & 2013-07-02 \\
-  141 & Riyadh\_7\_2013 & KJ156937, KJ156905 & human & 2013-07-15 \\
-  142 & Riyadh\_8\_2013 & KJ156880, KJ156942 & human & 2013-07-17 \\
-  143 & Riyadh\_9\_2013 & KJ156869 & human & 2013-07-17 \\
-  144 & Hafr-Al-Batin\_2\_2013 & KJ156910 & human & 2013-08-05 \\
-  145 & Asir\_2\_2013 & KJ156863, KJ156899, KJ156912, KJ156900, KJ156898, KJ156945,
-KJ156932 & human & 2013-08-05 \\
-  146 & Riyadh\_11\_2013 & KJ156946, KJ156911 & human & 2013-08-06 \\
-  147 & Riyadh\_12\_2013 & KJ156926, KJ156901 & human & 2013-08-08 \\
-  148 & Riyadh\_13\_2013 & KJ156888, KJ156873 & human & 2013-08-13 \\
-  149 & Riyadh\_14\_2013 & KJ156934 & human & 2013-08-15 \\
-  150 & Hafr-Al-Batin\_4\_2013 & KJ156931, KJ156895, KJ156864, KJ156861 & human & 2013-08-25 \\
-  151 & Hafr-Al-Batin\_5\_2013 & KJ156951, KJ156924, KJ156954, KJ156913 & human & 2013-08-25 \\
-  152 & Riyadh\_17\_2013 & KJ156918, KJ156920, KJ156865 & human & 2013-08-26 \\
-  153 & Hafr-Al-Batin\_6\_2013 & KJ156874 & human & 2013-08-28 \\
-  154 & Riyadh\_10\_2013 & KJ156891, KJ156936, KJ156907 & human & 2013-09-05 \\
-  155 & Madinah\_3b\_2013 & KJ156950, KJ156916 & human & 2013-09-11 \\
-  156 & Qatar3 & KF961221 & human & 2013-10-13 \\
-  157 & Qatar4 & KF961222 & human & 2013-10-17 \\
-  158 & Oman\_2285\_2013 & KT156560 & human & 2013-10-28 \\
-  159 & Jeddah-1 & KF958702 & human & 2013-11-05 \\
-  160 & AbuDhabi\_UAE\_9\_2013 & KP209312 & human & 2013-11-15 \\
-  161 & Oman\_2874\_2013 & KT156561 & human & 2013-12-28 \\
-  162 & AbuDhabi/Gayathi\_UAE\_2\_2014 & KP209310 & human & 2014-03-07 \\
-  163 & Jeddah\_C7569/KSA & KM027256 & human & 2014-04-03 \\
-  164 & Jeddah\_C7149/KSA & KM027255 & human & 2014-04-05 \\
-  165 & Jeddah\_C7770/KSA & KM027257 & human & 2014-04-07 \\
-  166 & AbuDhabi\_UAE\_8\_2014 & KP209306 & human & 2014-04-07 \\
-  167 & AbuDhabi\_UAE\_16\_2014 & KP209308 & human & 2014-04-10 \\
-  168 & AbuDhabi\_UAE\_18\_2014 & KP209307 & human & 2014-04-10 \\
-  169 & Jeddah\_C8826/KSA & KM027258 & human & 2014-04-12 \\
-  170 & AbuDhabi\_UAE\_26\_2014 & KP209313 & human & 2014-04-13 \\
-  171 & Jeddah\_C9055/KSA & KM027259 & human & 2014-04-14 \\
-  172 & Makkah\_C9355/KSA/Makkah & KM027261 & human & 2014-04-15 \\
-  173 & AbuDhabi\_UAE\_33\_2014 & KP209311 & human & 2014-04-17 \\
-  174 & AbuDhabi\_UAE\_30\_2014 & KP209309 & human & 2014-04-19 \\
-  175 & Jeddah\_C10306/KSA & KM027260 & human & 2014-04-21 \\
-  176 & Riyadh\_2014KSA\_683/KSA/2014 & KM027262 & human & 2014-04-22 \\
-  177 & Riyadh-KKUH-90b &  & human & 2014-04-24 \\
-  178 & Riyadh-KKUH-105 &  & human & 2014-04-25 \\
-  179 & Riyadh-KKUH-104 &  & human & 2014-04-25 \\
-  180 & KFMC-1 & KT121580 & human & 2014-04-28 \\
-  181 & KFMC-8 & KT121579 & human & 2014-04-30 \\
-  182 & Indiana/USA-1\_SaudiArabia\_2014 & KJ813439 & human & 2014-04-30 \\
-  183 & KFMC-10 & KT121578 & human & 2014-05-01 \\
-  184 & KFMC-7 & KT121581 & human & 2014-05-03 \\
-  185 & Riyadh-KKUH-291 &  & human & 2014-05-06 \\
-  186 & KFMC-9 & KT121574 & human & 2014-05-07 \\
-  187 & KFMC-3 & KT121573 & human & 2014-05-09 \\
-  188 & Florida/USA-2\_SaudiArabia\_2014 & KJ829365 & human & 2014-05-10 \\
-  189 & KFMC-2 & KT121577 & human & 2014-05-11 \\
-  190 & KFMC-4 & KT121575 & human & 2014-05-12 \\
-  191 & KFMC-5 & KT121572 & human & 2014-05-12 \\
-  192 & Riyadh-KKUH-368 &  & human & 2014-05-13 \\
-  193 & KFMC-6 & KT121576 & human & 2014-05-18 \\
-  194 & Riyadh\_2014KSA\_158/KSA/2014 & KM027281 & human & 2014-05-20 \\
-  195 & Jeddah-KFH-285TA &  & human & 2014-06-03 \\
-  196 & Jeddah-KFH-605TD &  & human & 2014-06-09 \\
-  197 & Jeddah-KFH-668TD &  & human & 2014-06-09 \\
-  198 & Jeddah-KFH-899NF &  & human & 2014-06-16 \\
-  199 & Jeddah-KFH-949NSG1 &  & human & 2014-06-18 \\
-  200 & Riyadh-KKUH-643 &  & human & 2014-11-02 \\
-  201 & Taif/KSA-7032/2014 & KU710264 & human & 2014-11-04 \\
-  202 & Riyadh-KKUH-665 &  & human & 2014-11-19 \\
-  203 & Riyadh-KSA-2049/2015 & KR011266 & human & 2015-01-06 \\
-  204 & Riyadh-KSA-2343/2015 & KR011264 & human & 2015-01-21 \\
-  205 & Riyadh-KSA-2345/2015 & KR011263 & human & 2015-01-21 \\
-  206 & Riyadh-KSA-2466/2015 & KR011265 & human & 2015-01-26 \\
-  207 & Kharj-KSA-2599/2015 & KT806052 & human & 2015-02-02 \\
-  208 & Kharj-KSA-2598/2015 & KT806053 & human & 2015-02-02 \\
-  209 & Riyadh-KSA-2716/2015 & KT806051 & human & 2015-02-05 \\
-  210 & Khobar-KSA-6736/2015 & KT806048 & human & 2015-02-07 \\
-  211 & Jeddah-KSA-C20843/2015 & KT806044 & human & 2015-02-09 \\
-  212 & Jeddah-KSA-C20860/2015 & KT806055 & human & 2015-02-10 \\
-  213 & Riyadh\_KSA\_2959\_2015 & KT026453 & human & 2015-02-10 \\
-  214 & Riyadh-KSA-3065/2015 & KT806050 & human & 2015-02-12 \\
-  215 & Najran-KSA-C20915/2015 & KT806054 & human & 2015-02-13 \\
-  216 & Riyadh-KSA-3181/2015 & KT806049 & human & 2015-02-15 \\
-  217 & Riyadh\_KKUH\_0734 &  & human & 2015-02-18 \\
-  218 & Jeddah-KSA-C21271/2015 & KT806045 & human & 2015-02-22 \\
-  219 & Riyadh\_KKUH\_0755 &  & human & 2015-02-23 \\
-  220 & Riyadh\_KKUH\_0756 &  & human & 2015-02-23 \\
-  221 & Riyadh\_KKUH\_0780 &  & human & 2015-02-25 \\
-  222 & Riyadh\_KKUH\_0801 &  & human & 2015-02-27 \\
-  223 & Riyadh\_KKUH\_0826 &  & human & 2015-02-28 \\
-  224 & Riyadh\_KKUH\_0818 &  & human & 2015-02-28 \\
-  225 & Riyadh\_KSA\_4050\_2015 & KT026454 & human & 2015-03-01 \\
-  226 & Riyadh\_KKUH\_0944 &  & human & 2015-03-02 \\
-  227 & Riyadh\_KKUH\_0939 &  & human & 2015-03-02 \\
-  228 & Riyadh\_KKUH\_1080 &  & human & 2015-03-03 \\
-  229 & Riyadh\_KKUH\_1066 &  & human & 2015-03-03 \\
-  230 & Riyadh\_KKUH\_1145 &  & human & 2015-03-04 \\
-  231 & Riyadh\_KKUH\_1217 &  & human & 2015-03-04 \\
-  232 & Riyadh\_KKUH\_1461 &  & human & 2015-03-08 \\
-  233 & Riyadh\_KKUH\_1470 &  & human & 2015-03-08 \\
-  234 & Riyadh\_KKUH\_1522 &  & human & 2015-03-09 \\
-  235 & Germany3/UAE-Dubai/Abu-Dhabi &  & human & 2015-03-11 \\
-  236 & Hufuf-KSA-9158/2015 & KT806047 & human & 2015-03-27 \\
-  237 & Hufuf-KSA-11002/2015 & KT806046 & human & 2015-05-10 \\
-  238 & KOR/KNIH/002\_05\_2015 & KT029139 & human & 2015-05-20 \\
-  239 & ChinaGD01 & KT036372 & human & 2015-05-28 \\
-  240 & KOR/Seoul/014-2015 & KX034093 & human & 2015-05-30 \\
-  241 & KOREA/Seoul/014-1-2015 & KT374052 & human & 2015-05-31 \\
-  242 & KOREA/Seoul/035-1-2015 & KT374054 & human & 2015-06-03 \\
-  243 & KOR/Seoul/066-2015 & KX034095 & human & 2015-06-04 \\
-  244 & Korea/Seoul/SNU1-035/2015 & KU308549 & human & 2015-06-08 \\
-  245 & KOR/CNUH\_SNU/030\_06\_2015 & KT868868 & human & 2015-06-08 \\
-  246 & KOR/CNUH\_SNU/024\_06\_2015 & KT868867 & human & 2015-06-08 \\
-  247 & KOR/CNUH\_SNU/054\_06\_2015 & KT868871 & human & 2015-06-09 \\
-  248 & KOR/CNUH\_SNU/038\_06\_2015 & KT868870 & human & 2015-06-10 \\
-  249 & KOR/CNUH\_SNU/148\_06\_2015 & KT868876 & human & 2015-06-10 \\
-  250 & KOR/CNUH\_SNU/122\_06\_2015 & KT868875 & human & 2015-06-10 \\
-  251 & KOR/CNUH\_SNU/082\_06\_2015 & KT868872 & human & 2015-06-10 \\
-  252 & KOR/CNUH\_SNU/085\_06\_2015 & KT868873 & human & 2015-06-10 \\
-  253 & KOR/CNUH\_SNU/016\_06\_2015 & KT868865 & human & 2015-06-11 \\
-  254 & KOR/CNUH\_SNU/023\_06\_2015 & KT868866 & human & 2015-06-11 \\
-  255 & KOR/CNUH\_SNU/031\_06\_2015 & KT868869 & human & 2015-06-11 \\
-  256 & KOR/CNUH\_SNU/110\_06\_2015 & KT868874 & human & 2015-06-11 \\
-  257 & KOR/Seoul/050-1-2015 & KX034094 & human & 2015-06-11 \\
-  258 & THA/CU/17\_06\_2015 & KT225476 & human & 2015-06-17 \\
-  259 & KOR/Seoul/077-2-2015 & KX034096 & human & 2015-06-17 \\
-  260 & KOR/Seoul/080-3-2015 & KX034097 & human & 2015-06-17 \\
-  261 & KOREA/Seoul/163-1-2015 & KT374051 & human & 2015-06-19 \\
-  262 & KOREA/Seoul/168-1-2015 & KT374056 & human & 2015-06-21 \\
-  263 & KOR/Seoul/162-1-2015 & KX034098 & human & 2015-06-22 \\
-  264 & KOR/CNUH\_SNU/172\_06\_2015 & KT868877 & human & 2015-06-22 \\
-  265 & KOR/Seoul/169-2015 & KX034099 & human & 2015-06-26 \\
-  266 & KOR/Seoul/177-3-2015 & KX034100 & human & 2015-07-03 \\
-  267 & Jeddah-KSA-3RS2702/2015 & KU851859 & human & 2015-07-12 \\
-  268 & Riyadh-KSA-16120/2015 & KU851861 & human & 2015-08-24 \\
-  269 & Riyadh-KSA-16117/2015 & KU851862 & human & 2015-08-24 \\
-  270 & Riyadh-KSA-16121/2015 & KU851860 & human & 2015-08-24 \\
-  271 & Riyadh-KSA-16098/2015 & KU851864 & human & 2015-08-24 \\
-  272 & Riyadh-KSA-16077/2015 & KU851863 & human & 2015-08-27 \\
-  273 & Jordan\_1\_2015 & KU233363 & human & 2015-09-17 \\
-  274 & Jordan\_10\_2015 & KU233362 & human & 2015-09-17 \\
-\end{longtable}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_exploded.png}
-	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
-This is the same tree as shown in Figure \ref{mcc}, but with contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top in orange) and humans (bottom in blue).
-This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
-	}
-	\label{exploded}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
-	\caption{\textbf{Posterior backwards migration rate estimates for two choices of prior.}
-Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice (note that rates are backwards in time).
-Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme looking backwards in time (orange) and \textit{vice versa} (blue).
-Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
-	}
-	\label{prior}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_dta_mcc.png}
-	\caption{\textbf{Maximum clade credibility (MCC) tree with ancestral state reconstruction according to a discrete trait model.}
-MCC tree is presented the same as Figure \ref{mcc} and Figure \ref{equal_sizes}, with colours indicating the most probable state reconstruction at internal nodes.
-Unlike the structured coalescent summary shown in Figure \ref{mcc} where camels are reconstructed as the main host where MERS-CoV persists, the discrete trait approach identifies both camels and humans as major hosts with humans being the source of MERS-CoV infection in camels.
-	}
-	\label{dta}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_es_mcc.png}
-	\caption{\textbf{Maximum clade credibility (MCC) tree of structured coalescent model with enforced equal coalescence rates.}
-MCC tree is presented the same as Figures \ref{mcc} and \ref{dta}, with colours indicating the most probable state reconstruction at internal nodes.
-Similar to Figure \ref{dta} enforcing equal coalescence rates between demes in a structured coalescent model identifies humans as a major MERS-CoV host and the source of viruses in camels.
-	}
-	\label{equal_sizes}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_epi_grid.png}
-	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
-From top to bottom each row corresponds to departures from completely random sequencing efforts with respect to case cluster size (bias parameter=1.0) to sequencing increasingly biased towards capturing large case clusters (bias=2.0, bias=3.0).
-Leftmost scatter plots show the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) coloured by the $R_{0}$ value used for the simulation.
-The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and skewness observed for empirical MERS-CoV data.
-The distribution of $R_{0}$ values matching empirical data are shown in the middle, on the same $y$-axis across all levels of the bias parameter.
-Under unbiased sequencing (bias=1.0) only 0.45\% of simulations fit our phylogenetic observations, while 1.79\% and 1.67\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
-Correspondingly, we estimate 11.6\% support for a model with bias level 1.0, 45.7\% support for a model with bias level 2.0, and 42.7\% support for a model with bias level 3.0.
-Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
-While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.71--0.98) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters.
-The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
-Darker shade of grey indicates bins falling within the 95\% percentiles.
-The median number of cross-species introductions observed in simulations matching empirical data without bias are 346 (95\% percentiles 262--439).
-These numbers jump up to 568 (95\% percentiles 430--727) for bias = 2.0 and 656 (95\% percentiles 488--853) for bias = 3.0 simulations.
-Model averaging would suggest plausible numbers of introductions between 311 and 811.
-  }
-	\label{mers_epi_grid}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_mc_method.pdf}
-	\caption{\textbf{Monte Carlo simulation schematic.}
-Case clusters are simulated according to Equation \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
-We sample 174 cases from each simulation to represent sequencing of human MERS cases.
-`Sequencing' is carried out by using multivariate hypergeometric sampling, representing sampling cases without replacement to be sequenced.
-Sequencing simulations take place at three levels of bias: 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
-The distribution of simulated sequence clusters is summarised by its mean, median and standard deviation.
-A simulation is considered to match if the mean, median and standard deviation of its sequence cluster sizes falls within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
-$R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are retained as estimates.
-These estimates are summarised in Figure \ref{mers_epi}.
-	}
-	\label{mc_method}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}
-	\caption{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
-The plot is identical to Figure \ref{mers_epi_grid}, but instead of 2000 cases,
-% the currently reported number of MERS-CoV cases,
-simulations were run with 4000 cases.
-With more unobserved cases the R$_{0}$ values matching observed MERS-CoV sequence clusters can only be smaller, with a corresponding increase in numbers of zoonotic transmissions.
-However, the numbers of simulations that match MERS-CoV data go down as well.
-	}
-	\label{extra_cases}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_distributions.png}
-	\caption{\textbf{Boxplots of matching simulated case and sequence cluster distributions.}
-Boxplots indicate frequency of case (blue, top) and sequence (red, bottom) cluster sizes across simulations at different bias levels, marginalised across $R_{0}$ values.
-Outliers are shown with transparency, medians are indicated with thick black lines.
-Case clusters exhibit a strong skew with large numbers of singleton introductions and a substantial tail at higher levels of bias.
-	}
-	\label{cluster_distributions}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_qq.png}
-	\caption{\textbf{Quantile-quantile (Q-Q) plot of empirical and simulated sequence cluster sizes.}
-Density of sequence cluster size percentiles (1st--99th, calculated across a grid of 50 values) calculated for random states from the posterior distribution ($x$-axis) and matching simulations ($y$-axis).
-Most values fall on the one-to-one line, with a heavier tail in mid-sized sequence clusters in empirical data, manifesting as a greater density of points below the one-to-one line in the middle.
-	}
-	\label{qq}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_recombination_tree.png}
-	\caption{\textbf{Tests of recombination across MERS-CoV clades.}
-Maximum clade credibility tree of MERS-CoV genomes annotated with results of two recombination detection tests (PHI and 3Seq) applied to descendent sequences of each clade.
-Both tests identify large portions of existing sequence data as containing signals of recombination.
-Note that markings do not indicate where recombinations have occurred on the tree, merely the minimum distance in sequence/time space between recombining lineages.}
-	\label{recombination_tree}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.75\textwidth]{figures/mers_incompatibilities.png}
-	\caption{\textbf{MERS-CoV genomes exhibit high numbers of non-clonal loci.}
-Ancestral state reconstruction (right) identifies a large number of sites in which mutations have occurred more than once in the tree (homoplasies, orange) or are reversions (red) from a state arising in an ancestor.
-Mutations that apparently only occur once in the tree (synapomorphies) are shown in grey.
-The maximum likelihood phylogeny on the left is coloured by whether sequences were sampled in humans (blue) or camels (orange).
-	}
-	\label{incompatibilities}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_flower.png}
-	\caption{\textbf{Human clade sharing between genomic fragments 1 and 2.}
-Central scatter plot shows the posterior probability of human clades shared between genomic fragments 1 and 2, in their respective trees.
-Left and bottom scatter plots track the posterior probability of human clades only observed in fragment 2 (left) or fragment 1 (bottom).
-The cumulative probability of human clades present in either tree are tracked by plots on the right (fragment 2) and top (fragment 1).
-Most of the probability mass is concentrated within human clades that are present in trees of both genomic fragment 1 and 2 (0.9701 and 0.9474 of all human clades across posteriors, respectively).
-	}
-	\label{flower}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid_comparison.png}
-	\caption{\textbf{Skygrid comparison between whole and fragmented genomes.}
-Inferred median $N_{e}\tau$ recovered using a skygrid tree prior on whole genome (bottom) and ten genomic fragments with independent trees (left), coloured by time.
-Dotted line indicates the one-to-one line.
-	}
-	\label{skygrid_comparison}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_ML.png}
-	\caption{\textbf{Maximum likelihood (ML) tree of MERS-CoV genomes coloured by origin of sequence.}
-Maximum likelihood tree shows genetic divergence between MERS-CoV genomes collected from camels (orange tips) and humans (blue tips).
-	}
-	\label{ml}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
-	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
-Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
-More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
-	}
-	\label{marginals}
-\end{figure}
+% \newpage
+
+% \setcounter{figure}{0}
+% \setcounter{table}{0}
+% \renewcommand{\thefigure}{S\arabic{figure}}
+% \renewcommand{\thetable}{S\arabic{table}}
+
+% \begin{longtable}{ | r | l | p{2cm} | l | l | }
+%
+%   \caption{Strain names, accessions (where available), identified host and reported collection dates for MERS-CoV genomes used in this study.} \label{sequences} \\
+%
+%
+%   \hline & strain & accession & host & collection date \\ \hline
+%   \endfirsthead
+%
+%   \multicolumn{5}{l}%
+%   {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
+%
+%   \hline & strain & accession & host & collection date \\ \hline
+%   \endhead
+%
+%   \hline \multicolumn{5}{|r|}{{Continued on next page}} \\ \hline
+%   \endfoot
+%
+%   \hline \hline
+%   \endlastfoot
+%
+%   1 & KSA-378 & KJ713296 & camel & 2013-11 \\
+%   2 & KSA-363 & KJ713298 & camel & 2013-11 \\
+%   3 & KSA-503 & KJ713297 & camel & 2013-11 \\
+%   4 & KSA-376 & KJ713299 & camel & 2013-11 \\
+%   5 & KSA-505 & KJ713295 & camel & 2013-11 \\
+%   6 & Jeddah-1 & KF917527 & camel & 2013-11-08 \\
+%   7 & NRCE-HKU205 & KJ477102 & camel & 2013-11-15 \\
+%   8 & KFU-HKU1 & KJ650297 & camel & 2013-11-30 \\
+%   9 & KFU-HKU13 & KJ650295 & camel & 2013-12-30 \\
+%   10 & Camel\_Egypt\_NRCE-HKU271 &  & camel & 2013-12-30 \\
+%   11 & Camel\_Egypt\_NRCE-HKU270 &  & camel & 2013-12-30 \\
+%   12 & KFU-HKU19Dam & KJ650296 & camel & 2013-12-30 \\
+%   13 & Qatar\_2\_2014 & KJ650098 & camel & 2014-02-16 \\
+%   14 & UAE/D469-14 & KU242424 & camel & 2014-03-04 \\
+%   15 & UAE/D511-14 & KU242423 & camel & 2014-03-12 \\
+%   16 & Jeddah/F13A/2014 & KT368824 & camel & 2014-05 \\
+%   17 & UAE/D1164.10/2014 & KP719928 & camel & 2014-06 \\
+%   18 & UAE/D1339.2/2014 & KP719931 & camel & 2014-06 \\
+%   19 & UAE/D1164.11/2014 & KP719929 & camel & 2014-06 \\
+%   20 & UAE/D1164.9/2014 & KP719927 & camel & 2014-06 \\
+%   21 & UAE/D1209/2014 & KP719933 & camel & 2014-06 \\
+%   22 & UAE/D1164.14/2014 & KP719930 & camel & 2014-06 \\
+%   23 & UAE/D1243.12/2014 & KP719932 & camel & 2014-06 \\
+%   24 & D1164.1/14 & KX108937 & camel & 2014-06-02 \\
+%   25 & Riyadh/Ry23N/2014 & KT368825 & camel & 2014-07 \\
+%   26 & Riyadh/Ry84N/2014 & KT368826 & camel & 2014-07 \\
+%   27 & Jeddah/S93/2014 & KT368855 & camel & 2014-09 \\
+%   28 & Jeddah/401/2014 & KT368827 & camel & 2014-09 \\
+%   29 & Jeddah/S100/2014 & KT368853 & camel & 2014-09 \\
+%   30 & Jeddah/S99/2014 & KT368857 & camel & 2014-09 \\
+%   31 & Jeddah/S94/2014 & KT368856 & camel & 2014-09 \\
+%   32 & Jeddah/S73/2014 & KT368854 & camel & 2014-09 \\
+%   33 & Jeddah/O47b/2014 & KT368852 & camel & 2014-10 \\
+%   34 & Jeddah/O23b/2014 & KT368849 & camel & 2014-10 \\
+%   35 & Jeddah/O24/2014 & KT368850 & camel & 2014-10 \\
+%   36 & Jeddah/O30/2014 & KT368851 & camel & 2014-10 \\
+%   37 & Jeddah/N51/2014 & KT368846 & camel & 2014-11 \\
+%   38 & Jeddah/N68b/2014 & KT368848 & camel & 2014-11 \\
+%   39 & Jeddah/N62b/2014 & KT368847 & camel & 2014-11 \\
+%   40 & Jeddah/D40/2014 & KT368834 & camel & 2014-12 \\
+%   41 & Jeddah/D90/2014 & KT368844 & camel & 2014-12 \\
+%   42 & Jeddah/D88/2014 & KT368843 & camel & 2014-12 \\
+%   43 & Jeddah/D36/2014 & KT368832 & camel & 2014-12 \\
+%   44 & Jeddah/D35/2014 & KT368831 & camel & 2014-12 \\
+%   45 & Jeddah/D92/2014 & KT368845 & camel & 2014-12 \\
+%   46 & Jeddah/D49/2014 & KT368841 & camel & 2014-12 \\
+%   47 & Jeddah/D34/2014 & KT368830 & camel & 2014-12 \\
+%   48 & Jeddah/D33b/2014 & KT368829 & camel & 2014-12 \\
+%   49 & Jeddah/D42/2014 & KT368835 & camel & 2014-12 \\
+%   50 & Jeddah/D50b/2014 & KT368842 & camel & 2014-12 \\
+%   51 & Jeddah/D45/2014 & KT368837 & camel & 2014-12 \\
+%   52 & Jeddah/D46b/2014 & KT368838 & camel & 2014-12 \\
+%   53 & Jeddah/D43b/2014 & KT368836 & camel & 2014-12 \\
+%   54 & Jeddah/D100/2014 & KT368828 & camel & 2014-12 \\
+%   55 & Jeddah/D47/2014 & KT368839 & camel & 2014-12 \\
+%   56 & Jeddah/D38b/2014 & KT368833 & camel & 2014-12 \\
+%   57 & Jeddah/D48/2014 & KT368840 & camel & 2014-12 \\
+%   58 & D2597.2/14 & KX108938 & camel & 2014-12-13 \\
+%   59 & Egypt\_NRCE-NC163/2014 & KU740200 & camel & 2014-12-17 \\
+%   60 & Jeddah/Jd7/2015 & KT368861 & camel & 2015-01 \\
+%   61 & Jeddah/Jd86/2015 & KT368863 & camel & 2015-01 \\
+%   62 & Jeddah/Jd90/2015 & KT368865 & camel & 2015-01 \\
+%   63 & Jeddah/Jd1b/2015 & KT368858 & camel & 2015-01 \\
+%   64 & Jeddah/Jd4/2015 & KT368859 & camel & 2015-01 \\
+%   65 & Jeddah/Jd85/2015 & KT368862 & camel & 2015-01 \\
+%   66 & Jeddah/Jd6b/2015 & KT368860 & camel & 2015-01 \\
+%   67 & Jeddah/Jd87/2015 & KT368864 & camel & 2015-01 \\
+%   68 & D252/15 & KX108939 & camel & 2015-01-30 \\
+%   69 & Jeddah/Jd199/2015 & KT368867 & camel & 2015-02 \\
+%   70 & Jeddah/Jd175/2015 & KT368866 & camel & 2015-02 \\
+%   71 & D374/15 & KX108940 & camel & 2015-02-12 \\
+%   72 & D383/15 & KX108941 & camel & 2015-02-14 \\
+%   73 & D389/15 & KX108942 & camel & 2015-02-15 \\
+%   74 & Riyadh/Ry63/2015 & KT368876 & camel & 2015-03 \\
+%   75 & Riyadh/Ry136/2015 & KT368868 & camel & 2015-03 \\
+%   76 & Riyadh/Ry178/2015 & KT368874 & camel & 2015-03 \\
+%   77 & Riyadh/Ry162/2015 & KT368871 & camel & 2015-03 \\
+%   78 & Riyadh/Ry86/2015 & KT368879 & camel & 2015-03 \\
+%   79 & Taif/T150/2015 & KT368889 & camel & 2015-03 \\
+%   80 & Riyadh/Ry137/2015 & KT368869 & camel & 2015-03 \\
+%   81 & Riyadh/Ry179/2015 & KT368875 & camel & 2015-03 \\
+%   82 & Riyadh/Ry177/2015 & KT368873 & camel & 2015-03 \\
+%   83 & Riyadh/Ry79/2015 & KT368878 & camel & 2015-03 \\
+%   84 & Riyadh/Ry173/2015 & KT368872 & camel & 2015-03 \\
+%   85 & Taif/T157b/2015 & KT368890 & camel & 2015-03 \\
+%   86 & Riyadh/Ry159b/2015 & KT368870 & camel & 2015-03 \\
+%   87 & Riyadh/Ry64/2015 & KT368877 & camel & 2015-03 \\
+%   88 & Taif/T3/2015 & KT368880 & camel & 2015-04 \\
+%   89 & Taif/T16/2015 & KT368882 & camel & 2015-04 \\
+%   90 & Taif/T22/2015 & KT368883 & camel & 2015-04 \\
+%   91 & Taif/T92/2015 & KT368887 & camel & 2015-04 \\
+%   92 & Taif/T7/2015 & KT368881 & camel & 2015-04 \\
+%   93 & Taif/T91b/2015 & KT368886 & camel & 2015-04 \\
+%   94 & Taif/T68/2015 & KT368884 & camel & 2015-04 \\
+%   95 & Taif/T89/2015 & KT368885 & camel & 2015-04 \\
+%   96 & Taif/T98/2015 & KT368888 & camel & 2015-04 \\
+%   97 & D998/15 & KX108943 & camel & 2015-04-23 \\
+%   98 & D1157/15 & KX108944 & camel & 2015-05-12 \\
+%   99 & D1189.1/15 & KX108946 & camel & 2015-05-18 \\
+%   100 & D1271/15 & KX108945 & camel & 2015-05-29 \\
+%   101 & Jordan-N3/2012 & KC776174 & human & 2012-04-15 \\
+%   102 & EMC/2012 & JX869059 & human & 2012-06-13 \\
+%   103 & England/1/2012 & KC164505 & human & 2012-09-11 \\
+%   104 & Riyadh\_1\_2012 & KF600612 & human & 2012-10-23 \\
+%   105 & Riyadh\_2\_2012 & KF600652 & human & 2012-10-30 \\
+%   106 & Riyadh\_3\_2013 & KF600613 & human & 2013-02-05 \\
+%   107 & England/3/2013 & KM210278 & human & 2013-02-10 \\
+%   108 & England/2/2013 & KM015348 & human & 2013-02-10 \\
+%   109 & England/4/2013 & KM210277 & human & 2013-02-13 \\
+%   110 & Riyadh\_4\_2013 & KJ156952 & human & 2013-03-01 \\
+%   111 & Munich/AbuDhabi/2013 & KF192507 & human & 2013-03-22 \\
+%   112 & Al-Hasa\_2\_2013 & KF186566 & human & 2013-04-21 \\
+%   113 & Al-Hasa\_3\_2013 & KF186565 & human & 2013-04-22 \\
+%   114 & UAE-FRA1\_1627-2013\_BAL & KJ361500 & human & 2013-04-26 \\
+%   115 & Al-Hasa\_4\_2013 & KF186564 & human & 2013-05-01 \\
+%   116 & Al-Hasa\_7\_2013 & KF600623, KF600655 & human & 2013-05-01 \\
+%   117 & Al-Hasa\_8\_2013 & KF600618, KF600626, KF600635, KF600638 & human & 2013-05-01 \\
+%   118 & Al-Hasa\_25\_2013 & KJ156866 & human & 2013-05-02 \\
+%   119 & Al-Hasa\_11\_2013 & KF600629, KF600636, KF600646 & human & 2013-05-03 \\
+%   120 & Al-Hasa\_12\_2013 & KF600627 & human & 2013-05-07 \\
+%   121 & Al-Hasa\_14\_2013 & KF600615, KF600643 & human & 2013-05-08 \\
+%   122 & Al-Hasa\_1\_2013 & KF186567 & human & 2013-05-09 \\
+%   123 & Al-Hasa\_15\_2013 & KF600645 & human & 2013-05-11 \\
+%   124 & Al-Hasa\_16\_2013 & KF600644 & human & 2013-05-12 \\
+%   125 & Buraidah\_1\_2013 & KF600630 & human & 2013-05-13 \\
+%   126 & Al-Hasa\_23\_2013 & KJ156860, KJ156894, KJ156929, KJ156923, KJ156862 & human & 2013-05-13 \\
+%   127 & Al-Hasa\_17\_2013 & KF600647 & human & 2013-05-15 \\
+%   128 & Al-Hasa\_19\_2013 & KF600632 & human & 2013-05-23 \\
+%   129 & Al-Hasa\_18\_2013 & KF600651 & human & 2013-05-23 \\
+%   130 & Al-Hasa\_21\_2013 & KF600634 & human & 2013-05-30 \\
+%   131 & Hafr-Al-Batin\_1\_2013 & KF600628 & human & 2013-06-04 \\
+%   132 & Wadi-Ad-Dawasir\_1\_2013 & KJ156881 & human & 2013-06-12 \\
+%   133 & Taif\_1\_2013 & KJ156949 & human & 2013-06-12 \\
+%   134 & Taif\_2\_2013 & KJ156896, KJ156876 & human & 2013-06-12 \\
+%   135 & Taif\_3\_2013 & KJ156938, KJ156897, KJ156922, KJ156868, KJ156921, KJ156915,
+% KJ156906 & human & 2013-06-13 \\
+%   136 & Al-Hasa\_26\_2013 & KJ156882, KJ156941, KJ156872 & human & 2013-06-18 \\
+%   137 & Al-Hasa\_27\_2013 & KJ156943, KJ156939 & human & 2013-06-19 \\
+%   138 & Al-Hasa\_28\_2013 & KJ156887, KJ156940, KJ156889, KJ156893, KJ156884, KJ156930,
+% KJ156928, KJ156909 & human & 2013-06-22 \\
+%   139 & Riyadh\_6\_2013 & KJ156879, KJ156947, KJ156890, KJ156908, KJ156927 & human & 2013-07-02 \\
+%   140 & Riyadh\_5\_2013 & KJ156944 & human & 2013-07-02 \\
+%   141 & Riyadh\_7\_2013 & KJ156937, KJ156905 & human & 2013-07-15 \\
+%   142 & Riyadh\_8\_2013 & KJ156880, KJ156942 & human & 2013-07-17 \\
+%   143 & Riyadh\_9\_2013 & KJ156869 & human & 2013-07-17 \\
+%   144 & Hafr-Al-Batin\_2\_2013 & KJ156910 & human & 2013-08-05 \\
+%   145 & Asir\_2\_2013 & KJ156863, KJ156899, KJ156912, KJ156900, KJ156898, KJ156945,
+% KJ156932 & human & 2013-08-05 \\
+%   146 & Riyadh\_11\_2013 & KJ156946, KJ156911 & human & 2013-08-06 \\
+%   147 & Riyadh\_12\_2013 & KJ156926, KJ156901 & human & 2013-08-08 \\
+%   148 & Riyadh\_13\_2013 & KJ156888, KJ156873 & human & 2013-08-13 \\
+%   149 & Riyadh\_14\_2013 & KJ156934 & human & 2013-08-15 \\
+%   150 & Hafr-Al-Batin\_4\_2013 & KJ156931, KJ156895, KJ156864, KJ156861 & human & 2013-08-25 \\
+%   151 & Hafr-Al-Batin\_5\_2013 & KJ156951, KJ156924, KJ156954, KJ156913 & human & 2013-08-25 \\
+%   152 & Riyadh\_17\_2013 & KJ156918, KJ156920, KJ156865 & human & 2013-08-26 \\
+%   153 & Hafr-Al-Batin\_6\_2013 & KJ156874 & human & 2013-08-28 \\
+%   154 & Riyadh\_10\_2013 & KJ156891, KJ156936, KJ156907 & human & 2013-09-05 \\
+%   155 & Madinah\_3b\_2013 & KJ156950, KJ156916 & human & 2013-09-11 \\
+%   156 & Qatar3 & KF961221 & human & 2013-10-13 \\
+%   157 & Qatar4 & KF961222 & human & 2013-10-17 \\
+%   158 & Oman\_2285\_2013 & KT156560 & human & 2013-10-28 \\
+%   159 & Jeddah-1 & KF958702 & human & 2013-11-05 \\
+%   160 & AbuDhabi\_UAE\_9\_2013 & KP209312 & human & 2013-11-15 \\
+%   161 & Oman\_2874\_2013 & KT156561 & human & 2013-12-28 \\
+%   162 & AbuDhabi/Gayathi\_UAE\_2\_2014 & KP209310 & human & 2014-03-07 \\
+%   163 & Jeddah\_C7569/KSA & KM027256 & human & 2014-04-03 \\
+%   164 & Jeddah\_C7149/KSA & KM027255 & human & 2014-04-05 \\
+%   165 & Jeddah\_C7770/KSA & KM027257 & human & 2014-04-07 \\
+%   166 & AbuDhabi\_UAE\_8\_2014 & KP209306 & human & 2014-04-07 \\
+%   167 & AbuDhabi\_UAE\_16\_2014 & KP209308 & human & 2014-04-10 \\
+%   168 & AbuDhabi\_UAE\_18\_2014 & KP209307 & human & 2014-04-10 \\
+%   169 & Jeddah\_C8826/KSA & KM027258 & human & 2014-04-12 \\
+%   170 & AbuDhabi\_UAE\_26\_2014 & KP209313 & human & 2014-04-13 \\
+%   171 & Jeddah\_C9055/KSA & KM027259 & human & 2014-04-14 \\
+%   172 & Makkah\_C9355/KSA/Makkah & KM027261 & human & 2014-04-15 \\
+%   173 & AbuDhabi\_UAE\_33\_2014 & KP209311 & human & 2014-04-17 \\
+%   174 & AbuDhabi\_UAE\_30\_2014 & KP209309 & human & 2014-04-19 \\
+%   175 & Jeddah\_C10306/KSA & KM027260 & human & 2014-04-21 \\
+%   176 & Riyadh\_2014KSA\_683/KSA/2014 & KM027262 & human & 2014-04-22 \\
+%   177 & Riyadh-KKUH-90b &  & human & 2014-04-24 \\
+%   178 & Riyadh-KKUH-105 &  & human & 2014-04-25 \\
+%   179 & Riyadh-KKUH-104 &  & human & 2014-04-25 \\
+%   180 & KFMC-1 & KT121580 & human & 2014-04-28 \\
+%   181 & KFMC-8 & KT121579 & human & 2014-04-30 \\
+%   182 & Indiana/USA-1\_SaudiArabia\_2014 & KJ813439 & human & 2014-04-30 \\
+%   183 & KFMC-10 & KT121578 & human & 2014-05-01 \\
+%   184 & KFMC-7 & KT121581 & human & 2014-05-03 \\
+%   185 & Riyadh-KKUH-291 &  & human & 2014-05-06 \\
+%   186 & KFMC-9 & KT121574 & human & 2014-05-07 \\
+%   187 & KFMC-3 & KT121573 & human & 2014-05-09 \\
+%   188 & Florida/USA-2\_SaudiArabia\_2014 & KJ829365 & human & 2014-05-10 \\
+%   189 & KFMC-2 & KT121577 & human & 2014-05-11 \\
+%   190 & KFMC-4 & KT121575 & human & 2014-05-12 \\
+%   191 & KFMC-5 & KT121572 & human & 2014-05-12 \\
+%   192 & Riyadh-KKUH-368 &  & human & 2014-05-13 \\
+%   193 & KFMC-6 & KT121576 & human & 2014-05-18 \\
+%   194 & Riyadh\_2014KSA\_158/KSA/2014 & KM027281 & human & 2014-05-20 \\
+%   195 & Jeddah-KFH-285TA &  & human & 2014-06-03 \\
+%   196 & Jeddah-KFH-605TD &  & human & 2014-06-09 \\
+%   197 & Jeddah-KFH-668TD &  & human & 2014-06-09 \\
+%   198 & Jeddah-KFH-899NF &  & human & 2014-06-16 \\
+%   199 & Jeddah-KFH-949NSG1 &  & human & 2014-06-18 \\
+%   200 & Riyadh-KKUH-643 &  & human & 2014-11-02 \\
+%   201 & Taif/KSA-7032/2014 & KU710264 & human & 2014-11-04 \\
+%   202 & Riyadh-KKUH-665 &  & human & 2014-11-19 \\
+%   203 & Riyadh-KSA-2049/2015 & KR011266 & human & 2015-01-06 \\
+%   204 & Riyadh-KSA-2343/2015 & KR011264 & human & 2015-01-21 \\
+%   205 & Riyadh-KSA-2345/2015 & KR011263 & human & 2015-01-21 \\
+%   206 & Riyadh-KSA-2466/2015 & KR011265 & human & 2015-01-26 \\
+%   207 & Kharj-KSA-2599/2015 & KT806052 & human & 2015-02-02 \\
+%   208 & Kharj-KSA-2598/2015 & KT806053 & human & 2015-02-02 \\
+%   209 & Riyadh-KSA-2716/2015 & KT806051 & human & 2015-02-05 \\
+%   210 & Khobar-KSA-6736/2015 & KT806048 & human & 2015-02-07 \\
+%   211 & Jeddah-KSA-C20843/2015 & KT806044 & human & 2015-02-09 \\
+%   212 & Jeddah-KSA-C20860/2015 & KT806055 & human & 2015-02-10 \\
+%   213 & Riyadh\_KSA\_2959\_2015 & KT026453 & human & 2015-02-10 \\
+%   214 & Riyadh-KSA-3065/2015 & KT806050 & human & 2015-02-12 \\
+%   215 & Najran-KSA-C20915/2015 & KT806054 & human & 2015-02-13 \\
+%   216 & Riyadh-KSA-3181/2015 & KT806049 & human & 2015-02-15 \\
+%   217 & Riyadh\_KKUH\_0734 &  & human & 2015-02-18 \\
+%   218 & Jeddah-KSA-C21271/2015 & KT806045 & human & 2015-02-22 \\
+%   219 & Riyadh\_KKUH\_0755 &  & human & 2015-02-23 \\
+%   220 & Riyadh\_KKUH\_0756 &  & human & 2015-02-23 \\
+%   221 & Riyadh\_KKUH\_0780 &  & human & 2015-02-25 \\
+%   222 & Riyadh\_KKUH\_0801 &  & human & 2015-02-27 \\
+%   223 & Riyadh\_KKUH\_0826 &  & human & 2015-02-28 \\
+%   224 & Riyadh\_KKUH\_0818 &  & human & 2015-02-28 \\
+%   225 & Riyadh\_KSA\_4050\_2015 & KT026454 & human & 2015-03-01 \\
+%   226 & Riyadh\_KKUH\_0944 &  & human & 2015-03-02 \\
+%   227 & Riyadh\_KKUH\_0939 &  & human & 2015-03-02 \\
+%   228 & Riyadh\_KKUH\_1080 &  & human & 2015-03-03 \\
+%   229 & Riyadh\_KKUH\_1066 &  & human & 2015-03-03 \\
+%   230 & Riyadh\_KKUH\_1145 &  & human & 2015-03-04 \\
+%   231 & Riyadh\_KKUH\_1217 &  & human & 2015-03-04 \\
+%   232 & Riyadh\_KKUH\_1461 &  & human & 2015-03-08 \\
+%   233 & Riyadh\_KKUH\_1470 &  & human & 2015-03-08 \\
+%   234 & Riyadh\_KKUH\_1522 &  & human & 2015-03-09 \\
+%   235 & Germany3/UAE-Dubai/Abu-Dhabi &  & human & 2015-03-11 \\
+%   236 & Hufuf-KSA-9158/2015 & KT806047 & human & 2015-03-27 \\
+%   237 & Hufuf-KSA-11002/2015 & KT806046 & human & 2015-05-10 \\
+%   238 & KOR/KNIH/002\_05\_2015 & KT029139 & human & 2015-05-20 \\
+%   239 & ChinaGD01 & KT036372 & human & 2015-05-28 \\
+%   240 & KOR/Seoul/014-2015 & KX034093 & human & 2015-05-30 \\
+%   241 & KOREA/Seoul/014-1-2015 & KT374052 & human & 2015-05-31 \\
+%   242 & KOREA/Seoul/035-1-2015 & KT374054 & human & 2015-06-03 \\
+%   243 & KOR/Seoul/066-2015 & KX034095 & human & 2015-06-04 \\
+%   244 & Korea/Seoul/SNU1-035/2015 & KU308549 & human & 2015-06-08 \\
+%   245 & KOR/CNUH\_SNU/030\_06\_2015 & KT868868 & human & 2015-06-08 \\
+%   246 & KOR/CNUH\_SNU/024\_06\_2015 & KT868867 & human & 2015-06-08 \\
+%   247 & KOR/CNUH\_SNU/054\_06\_2015 & KT868871 & human & 2015-06-09 \\
+%   248 & KOR/CNUH\_SNU/038\_06\_2015 & KT868870 & human & 2015-06-10 \\
+%   249 & KOR/CNUH\_SNU/148\_06\_2015 & KT868876 & human & 2015-06-10 \\
+%   250 & KOR/CNUH\_SNU/122\_06\_2015 & KT868875 & human & 2015-06-10 \\
+%   251 & KOR/CNUH\_SNU/082\_06\_2015 & KT868872 & human & 2015-06-10 \\
+%   252 & KOR/CNUH\_SNU/085\_06\_2015 & KT868873 & human & 2015-06-10 \\
+%   253 & KOR/CNUH\_SNU/016\_06\_2015 & KT868865 & human & 2015-06-11 \\
+%   254 & KOR/CNUH\_SNU/023\_06\_2015 & KT868866 & human & 2015-06-11 \\
+%   255 & KOR/CNUH\_SNU/031\_06\_2015 & KT868869 & human & 2015-06-11 \\
+%   256 & KOR/CNUH\_SNU/110\_06\_2015 & KT868874 & human & 2015-06-11 \\
+%   257 & KOR/Seoul/050-1-2015 & KX034094 & human & 2015-06-11 \\
+%   258 & THA/CU/17\_06\_2015 & KT225476 & human & 2015-06-17 \\
+%   259 & KOR/Seoul/077-2-2015 & KX034096 & human & 2015-06-17 \\
+%   260 & KOR/Seoul/080-3-2015 & KX034097 & human & 2015-06-17 \\
+%   261 & KOREA/Seoul/163-1-2015 & KT374051 & human & 2015-06-19 \\
+%   262 & KOREA/Seoul/168-1-2015 & KT374056 & human & 2015-06-21 \\
+%   263 & KOR/Seoul/162-1-2015 & KX034098 & human & 2015-06-22 \\
+%   264 & KOR/CNUH\_SNU/172\_06\_2015 & KT868877 & human & 2015-06-22 \\
+%   265 & KOR/Seoul/169-2015 & KX034099 & human & 2015-06-26 \\
+%   266 & KOR/Seoul/177-3-2015 & KX034100 & human & 2015-07-03 \\
+%   267 & Jeddah-KSA-3RS2702/2015 & KU851859 & human & 2015-07-12 \\
+%   268 & Riyadh-KSA-16120/2015 & KU851861 & human & 2015-08-24 \\
+%   269 & Riyadh-KSA-16117/2015 & KU851862 & human & 2015-08-24 \\
+%   270 & Riyadh-KSA-16121/2015 & KU851860 & human & 2015-08-24 \\
+%   271 & Riyadh-KSA-16098/2015 & KU851864 & human & 2015-08-24 \\
+%   272 & Riyadh-KSA-16077/2015 & KU851863 & human & 2015-08-27 \\
+%   273 & Jordan\_1\_2015 & KU233363 & human & 2015-09-17 \\
+%   274 & Jordan\_10\_2015 & KU233362 & human & 2015-09-17 \\
+% \end{longtable}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_exploded.png}
+% 	\caption{\textbf{Evolutionary history of MERS-CoV partitioned between camels and humans.}
+% This is the same tree as shown in Figure \ref{mcc}, but with contiguous stretches of MERS-CoV evolutionary history split by inferred host: camels (top in orange) and humans (bottom in blue).
+% This visualisation highlights the ephemeral nature of MERS-CoV outbreaks in humans, compared to continuous circulation of the virus in camels.
+% 	}
+% 	\label{exploded}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_prior.png}
+% 	\caption{\textbf{Posterior backwards migration rate estimates for two choices of prior.}
+% Negligible flow of MERS-CoV lineages from humans into camels is recovered regardless of prior choice (note that rates are backwards in time).
+% Plots show the 95\% highest posterior density for the estimated migration rate from the human deme into the camel deme looking backwards in time (orange) and \textit{vice versa} (blue).
+% Dotted lines indicate exponential priors specified for migration rates, with mean 1.0 (bottom) or 10.0 (top).
+% 	}
+% 	\label{prior}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_dta_mcc.png}
+% 	\caption{\textbf{Maximum clade credibility (MCC) tree with ancestral state reconstruction according to a discrete trait model.}
+% MCC tree is presented the same as Figure \ref{mcc} and Figure \ref{mcc}-Figure supplement 4, with colours indicating the most probable state reconstruction at internal nodes.
+% Unlike the structured coalescent summary shown in Figure \ref{mcc} where camels are reconstructed as the main host where MERS-CoV persists, the discrete trait approach identifies both camels and humans as major hosts with humans being the source of MERS-CoV infection in camels.
+% 	}
+% 	\label{dta}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_es_mcc.png}
+% 	\caption{\textbf{Maximum clade credibility (MCC) tree of structured coalescent model with enforced equal coalescence rates.}
+% MCC tree is presented the same as Figures \ref{mcc} and \ref{mcc}-Figure supplement 3, with colours indicating the most probable state reconstruction at internal nodes.
+% Similar to Figure \ref{mcc}-Figure supplement 3 enforcing equal coalescence rates between demes in a structured coalescent model identifies humans as a major MERS-CoV host and the source of viruses in camels.
+% 	}
+% 	\label{equal_sizes}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.75\textwidth]{figures/mers_epi_grid.png}
+% 	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
+% From top to bottom each row corresponds to departures from completely random sequencing efforts with respect to case cluster size (bias parameter=1.0) to sequencing increasingly biased towards capturing large case clusters (bias=2.0, bias=3.0).
+% Leftmost scatter plots show the distribution of individual Monte Carlo simulation sequence cluster size statistics (mean and skewness) coloured by the $R_{0}$ value used for the simulation.
+% The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and skewness observed for empirical MERS-CoV data.
+% The distribution of $R_{0}$ values matching empirical data are shown in the middle, on the same $y$-axis across all levels of the bias parameter.
+% Under unbiased sequencing (bias=1.0) only 0.45\% of simulations fit our phylogenetic observations, while 1.79\% and 1.67\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
+% Correspondingly, we estimate 11.6\% support for a model with bias level 1.0, 45.7\% support for a model with bias level 2.0, and 42.7\% support for a model with bias level 3.0.
+% Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
+% While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.71--0.98) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters.
+% The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
+% Darker shade of grey indicates bins falling within the 95\% percentiles.
+% The median number of cross-species introductions observed in simulations matching empirical data without bias are 346 (95\% percentiles 262--439).
+% These numbers jump up to 568 (95\% percentiles 430--727) for bias = 2.0 and 656 (95\% percentiles 488--853) for bias = 3.0 simulations.
+% Model averaging would suggest plausible numbers of introductions between 311 and 811.
+%   }
+% 	\label{mers_epi_grid}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.75\textwidth]{figures/mers_mc_method.pdf}
+% 	\caption{\textbf{Monte Carlo simulation schematic.}
+% Case clusters are simulated according to Equation \ref{clusterLikelihood} until an outbreak size of 2000 cases is reached.
+% We sample 174 cases from each simulation to represent sequencing of human MERS cases.
+% `Sequencing' is carried out by using multivariate hypergeometric sampling, representing sampling cases without replacement to be sequenced.
+% Sequencing simulations take place at three levels of bias: 1.0, where every case is equally likely to be sequenced, and 2.0 and 3.0, where cases from larger clusters are increasingly more likely to be sequenced.
+% The distribution of simulated sequence clusters is summarised by its mean, median and standard deviation.
+% A simulation is considered to match if the mean, median and standard deviation of its sequence cluster sizes falls within the 95\% highest posterior density interval of observed MERS-CoV sequence clusters.
+% $R_{0}$ values that ultimately generate data matching empirical observations, as well as associated numbers of `introductions' are retained as estimates.
+% These estimates are summarised in Figure \ref{mers_epi}.
+% 	}
+% 	\label{mc_method}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_epi_4000c.png}
+% 	\caption{\textbf{Results of Monte Carlo simulations with vast underestimation of cases.}
+% The plot is identical to Figure \ref{mers_epi}-Figure supplement 1, but instead of 2000 cases,
+% % the currently reported number of MERS-CoV cases,
+% simulations were run with 4000 cases.
+% With more unobserved cases the R$_{0}$ values matching observed MERS-CoV sequence clusters can only be smaller, with a corresponding increase in numbers of zoonotic transmissions.
+% However, the numbers of simulations that match MERS-CoV data go down as well.
+% 	}
+% 	\label{extra_cases}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_distributions.png}
+% 	\caption{\textbf{Boxplots of matching simulated case and sequence cluster distributions.}
+% Boxplots indicate frequency of case (blue, top) and sequence (red, bottom) cluster sizes across simulations at different bias levels, marginalised across $R_{0}$ values.
+% Outliers are shown with transparency, medians are indicated with thick black lines.
+% Case clusters exhibit a strong skew with large numbers of singleton introductions and a substantial tail at higher levels of bias.
+% 	}
+% 	\label{cluster_distributions}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_qq.png}
+% 	\caption{\textbf{Quantile-quantile (Q-Q) plot of empirical and simulated sequence cluster sizes.}
+% Density of sequence cluster size percentiles (1st--99th, calculated across a grid of 50 values) calculated for random states from the posterior distribution ($x$-axis) and matching simulations ($y$-axis).
+% Most values fall on the one-to-one line, with a heavier tail in mid-sized sequence clusters in empirical data, manifesting as a greater density of points below the one-to-one line in the middle.
+% 	}
+% 	\label{qq}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_recombination_tree.png}
+% 	\caption{\textbf{Tests of recombination across MERS-CoV clades.}
+% Maximum clade credibility tree of MERS-CoV genomes annotated with results of two recombination detection tests (PHI and 3Seq) applied to descendent sequences of each clade.
+% Both tests identify large portions of existing sequence data as containing signals of recombination.
+% Note that markings do not indicate where recombinations have occurred on the tree, merely the minimum distance in sequence/time space between recombining lineages.}
+% 	\label{recombination_tree}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.75\textwidth]{figures/mers_incompatibilities.png}
+% 	\caption{\textbf{MERS-CoV genomes exhibit high numbers of non-clonal loci.}
+% Ancestral state reconstruction (right) identifies a large number of sites in which mutations have occurred more than once in the tree (homoplasies, orange) or are reversions (red) from a state arising in an ancestor.
+% Mutations that apparently only occur once in the tree (synapomorphies) are shown in grey.
+% The maximum likelihood phylogeny on the left is coloured by whether sequences were sampled in humans (blue) or camels (orange).
+% 	}
+% 	\label{incompatibilities}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_flower.png}
+% 	\caption{\textbf{Human clade sharing between genomic fragments 1 and 2.}
+% Central scatter plot shows the posterior probability of human clades shared between genomic fragments 1 and 2, in their respective trees.
+% Left and bottom scatter plots track the posterior probability of human clades only observed in fragment 2 (left) or fragment 1 (bottom).
+% The cumulative probability of human clades present in either tree are tracked by plots on the right (fragment 2) and top (fragment 1).
+% Most of the probability mass is concentrated within human clades that are present in trees of both genomic fragment 1 and 2 (0.9701 and 0.9474 of all human clades across posteriors, respectively).
+% 	}
+% 	\label{flower}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid_comparison.png}
+% 	\caption{\textbf{Skygrid comparison between whole and fragmented genomes.}
+% Inferred median $N_{e}\tau$ recovered using a skygrid tree prior on whole genome (bottom) and ten genomic fragments with independent trees (left), coloured by time.
+% Dotted line indicates the one-to-one line.
+% 	}
+% 	\label{skygrid_comparison}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_ML.png}
+% 	\caption{\textbf{Maximum likelihood (ML) tree of MERS-CoV genomes coloured by origin of sequence.}
+% Maximum likelihood tree shows genetic divergence between MERS-CoV genomes collected from camels (orange tips) and humans (blue tips).
+% 	}
+% 	\label{ml}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% 	\includegraphics[width=0.65\textwidth]{figures/mers_simMarginals.png}
+% 	\caption{\textbf{Numbers of epidemiological simulations conforming to empirical observations.}
+% Numbers indicate the total number of epidemiological simulations under each combination of bias and dispersion parameter $\omega$ that result in MERS-CoV-like sequence cluster sizes.
+% More simulations match observations with bias$>1$ and $\omega \approx 0.1$.
+% 	}
+% 	\label{marginals}
+% \end{figure}
+
+\textbf{Supplementary File 1}. Strain names, accessions (where available), identified host and reported collection dates for MERS-CoV genomes used in this study.
 

### 2017-12-12 response_1/response_1.tex
Change references to new figure names in response.

@@
 \textbf{Yes, we agree that the structured coalescent approach is idealised and does not reflect a meaningful rate of zoonotic transfer of lineages, which is the reason we restricted any mention of rates to supplementary figures and do not attach numbers whenever rates are mentioned, but still report the number of introductions observed in the sequence data.
-We have altered Figure S2 in the reviewed manuscript to reflect that the rates shown are backwards in time.
+We have altered Figure S2 (now Figure 1-supplement 2) in the reviewed manuscript to reflect that the rates shown are backwards in time.
 We have added the following sentences to the discussion to highlight the fact that the coalescent model employed is not ideal:}
@@
 \begin{quotation}
-Although we recover migration rates from our model (Figure S2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
+Although we recover migration rates from our model (Figure 1-supplement 2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
 %\end{quotation}
@@
 \textbf{An excellent suggestion, thank you.
-We have run this analysis and include the results as a new supplementary figure S3.
+We have run this analysis and include the results as a new figure supplement (Figure 1-supplement 3).
 As expected, the skewed sampling results in a reconstruction of ancestral states that puts humans as the source of most MERS-CoV lineages in camels.
@@
 Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
-Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al.}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
+Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al.}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure 1-supplement 3).
 Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
@@
 We share the suspicion that the results are largely driven by contrasts in effective population sizes between demes.
-In addition to the requested maximum likelihood phylogeny (supplementary Figure S14) we also ran a structured coalescent analysis where deme sizes are enforced to be the same (supplementary Figure S4).
-This model fails in a similar way to a DTA reconstruction shown in the new Figure S3.
+In addition to the requested maximum likelihood phylogeny (now Figure 1-supplement 5) we also ran a structured coalescent analysis where deme sizes are enforced to be the same (now Figure 1-supplement 4).
+This model fails in a similar way to a DTA reconstruction shown in the Figure 1-supplement 3.
 We now explain how the structured coalescent arrives at the tree shown in Figure 1 in the discussion:}
@@
 We have run small scale analyses over a grid of three bias levels (same as the main analyses), 23 levels of $R_{0}$ (0.50--1.05), and five levels of dispersion (0.004, 0.02, 0.10, 0.50 and 1.0) with 2000 replicate simulations for each combination of variables.
-Numbers of simulations matching empirical data are shown in Figure S15.
+Numbers of simulations matching empirical data are shown in Figure 3-supplement 6.
 In this analysis, we find strongest support for dispersion $k$ = 0.1.}
@@
 Simulated sequence clusters tend to exhibit a left skew near the middle of the distribution, but not so much near the tails.
-We have added figure S9 to show this.}
+We have added Figure 3-supplement 5 to show this.}
 
@@
 study did not share any data beyond the raw input file they used.
-Since neither the actual numbers in Fig 1C of Cauchemez \textit{et al}., nor an implementation of the MCMC algorithm exist in the public domain we have included an additional figure (supplementary Figure S8) that visualises the distribution of cases recovered by simulation.
+Since neither the actual numbers in Fig 1C of Cauchemez \textit{et al}., nor an implementation of the MCMC algorithm exist in the public domain we have included an additional figure (Figure 3-supplement 4) that visualises the distribution of cases recovered by simulation.
 These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al}.}

### 2017-12-12 mers-structure.tex
Add repo citation.
But how does one get the SHA for the commit right if I'm pushing it just now?

@@
 \subsection*{Data availability}
-Sequence data and all analytical code is publicly available at \href{https://github.com/blab/structured-mers}{github.com/blab/structured-mers}.
+Sequence data and all analytical code is publicly available at \href{https://github.com/blab/structured-mers}{github.com/blab/structured-mers} \citep{mers-structure}.
 
